### 2016-01-11 documentation/bias/bias.tex
Edited check-biases and duplication calculator, added a documentation for biases

\documentclass[a4paper,10pt, twocolumn]{article}

\usepackage{float}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\graphicspath{{../../results/}}

\begin{document}
\section{Are EnTrI results biased?}
As transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for two type of biases: the distance from the origin and GC content. As Figure~\ref{fig:distance_bias} shows, the insertion index decreases with the increase of the distance from the origin. Therefore, there is a bias towards the distance from the origin. However, the threshold that has been selected for essential genes (0.2) separates the two groups well.
\begin{figure}[H]
\includegraphics[scale=0.3, page=1]{biases.pdf}
\caption{The insertion index value depends on the distance from the origin}
\label{fig:distance_bias}
\end{figure}

As depicted in Figure~\ref{fig:GC_bias} there is no bias towards the GC content of the genes and the regression line is straight. The nucleotides around the insertion sites (the insertion site and 10 nucleotides on each side) are stacked on top of each other and a sequence logo is generated from these sequences to see if there is any positional bias for any nucleotide. It can be inferred from Figure~\ref{fig:logo} that there is no bias in any position except a negligible bias in the insertion site towards GC pairs. 
\begin{figure}[H]
\includegraphics[scale=0.3, page=2]{biases.pdf}
\caption{The insertion index value is not biased to GC content}
\label{fig:GC_bias}
\end{figure}

\begin{figure*}
\includegraphics[scale=1.2]{logo.pdf}
\caption{There is no bias in the nucleotides around the insertion site}
\label{fig:logo}
\end{figure*}

\end{document}

### 2016-01-11 codes/check-biases.R

@@ -4,13 +4,19 @@ colnames(biasestable) <- c("name", "ii", "dist", "gc")
 
 pdf("../results/biases.pdf")
 
-plot(biasestable$dist, biasestable$ii, pch = '.', ylim=c(0,3))
+mar.default <- c(5,4,4,2) + 0.1
+par(mar = mar.default + c(0, 1, 0, 0))
+
+plot(biasestable$dist, biasestable$ii, pch = '.', ylim=c(0,3), xlab = "Distance from the origin", ylab = "insertion index",
+     main = "Distance bias", cex.lab = 2, cex.axis = 2, cex.main =2)
 lines(lowess(biasestable$dist, biasestable$ii, f = 0.2), col = 2, lwd=5)
+legend(0.26,2.8,c("regression line", "lowess"), col = c(3,2), lty=1, lwd=5, cex = 1.5)
 regln <- lm(biasestable$dist ~ biasestable$ii)
 abline(regln, col=3, lwd=5)
 
-plot(biasestable$gc, biasestable$ii, pch = '.', ylim=c(0,3))
-lines(lowess(biasestable$gc, biasestable$ii, f = 0.2), col = 2, lwd=5)
+plot(biasestable$gc, biasestable$ii, pch = '.', ylim=c(0,3), xlab = "GC content", ylab = "insertion index", main = "GC bias",
+     cex.lab = 2, cex.axis = 2, cex.main =2)
+# lines(lowess(biasestable$gc, biasestable$ii, f = 0.2), col = 2, lwd=5)
 regln <- lm(biasestable$gc ~ biasestable$ii)
 abline(regln, col=3, lwd=5)
 

### 2016-01-11 documentation/bias/bias.tex
Edited bias documentation

@@
 \section{Are EnTrI results biased?}
-As transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for two type of biases: the distance from the origin and GC content.
-As Figure~\ref{fig:distance_bias} shows, the insertion index decreases with the increase of the distance from the origin.
+Since transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for two types of biases: the distance from the origin and GC content.
+As Figure~\ref{fig:distance_bias} shows, although the regression line is flat, the insertion index decreases while increasing the distance from the origin.
 Therefore, there is a bias towards the distance from the origin.
@@
 \begin{figure}[H]
-\includegraphics[scale=0.3, page=1]{biases.pdf}
+\includegraphics[scale=0.28, page=1]{biases.pdf}
 \caption{The insertion index value depends on the distance from the origin}
@@
 
-As depicted in Figure~\ref{fig:GC_bias} there is no bias towards the GC content of the genes and the regression line is straight.
+As depicted in Figure~\ref{fig:GC_bias}, there is no bias towards the GC content of the genes and the regression line is flat.
 The nucleotides around the insertion sites (the insertion site and 10 nucleotides on each side) are stacked on top of each other and a sequence logo is generated from these sequences to see if there is any positional bias for any nucleotide.
 It can be inferred from Figure~\ref{fig:logo} that there is no bias in any position except a negligible bias in the insertion site towards GC pairs.
-\begin{figure}[H]
-\includegraphics[scale=0.3, page=2]{biases.pdf}
+\begin{figure}
+\includegraphics[scale=0.28, page=2]{biases.pdf}
 \caption{The insertion index value is not biased to GC content}
@@
 
-\begin{figure*}
-\includegraphics[scale=1.2]{logo.pdf}
+\begin{figure}
+\includegraphics[scale=.98, angle = -90]{logo.pdf}
 \caption{There is no bias in the nucleotides around the insertion site}
 \label{fig:logo}
-\end{figure*}
+\end{figure}
 

### 2016-01-26 documentation/EnTrI/EnTrI.tex
Documented the results of essentiality trees and heatmaps

\documentclass[a4paper,10pt, twocolumn]{article}

\usepackage{float}
\usepackage{graphicx}
\usepackage{caption}
%\usepackage{subcaption}
%\usepackage{pdfpages}
\graphicspath{{../../results/}}

\begin{document}
\section{Are EnTrI results biased?}
Since transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for two types of biases: the distance from the origin and GC content. As Figure~\ref{fig:distance_bias} shows, the insertion index decreases while increasing the distance from the origin. Therefore, there is a bias towards the distance from the origin. The distance bias in every individual strain is depicted in Figure~\ref{fig:distance_bias_individual}. These plots indicate that the bias is negligible in some strains like Salmonella typhi, while the insertion indices of the genes in other strains need to be normalised by their distances from the origin.

\begin{figure}[H]
\includegraphics[scale=0.28, page=1]{biases.pdf}
\caption{The bias towards the distance from origin (DnaA gene) for all of the species under study. The red curve shows the fitted LOWESS regression curve.}
\label{fig:distance_bias}
\end{figure}

\begin{figure*}
\centering
\begin{tabular}{c c c}
\includegraphics[page=2, scale=0.22]{biases.pdf} & 
\includegraphics[page=3, scale=0.22]{biases.pdf} &
\includegraphics[page=4, scale=0.22]{biases.pdf} \\
\includegraphics[page=5, scale=0.22]{biases.pdf} & 
\includegraphics[page=6, scale=0.22]{biases.pdf} &
\includegraphics[page=7, scale=0.22]{biases.pdf} \\
\includegraphics[page=8, scale=0.22]{biases.pdf} & 
\includegraphics[page=9, scale=0.22]{biases.pdf} &
\includegraphics[page=10, scale=0.22]{biases.pdf} \\
\includegraphics[page=11, scale=0.22]{biases.pdf} & 
\includegraphics[page=12, scale=0.22]{biases.pdf} &
\includegraphics[page=13, scale=0.22]{biases.pdf} \\
& 
\includegraphics[page=14, scale=0.22]{biases.pdf} &
\\
\end{tabular}
\caption{The bias towards the distance from origin for every individual strain. The red curves show the fitted LOWESS curves.}
\label{fig:distance_bias_individual}
\end{figure*}

As depicted in Figure~\ref{fig:GC_bias}, there is no bias towards the GC content of the genes and the regression line is flat. The nucleotides around the insertion sites (the insertion site and 10 nucleotides on each side) are stacked on top of each other and a sequence logo is generated from these sequences to see if there is any positional bias for any nucleotide. It can be inferred from Figure~\ref{fig:logo} that there is no bias in any position.

\begin{figure}
\includegraphics[scale=0.28, page=15]{biases.pdf}
\caption{The bias towards GC content. The regression line is depicted by green.}
\label{fig:GC_bias}
\end{figure}

\begin{figure}
\includegraphics[scale=.98, angle = -90]{logo.pdf}
\caption{The nucleotides around the insertion sites (the insertion site and 10 nucleotides on each side) are stacked on top of each other and a sequence logo is generated from these sequences using webLogo stand-alone package. The height of the letter stack in each position shows how conserved the bases in that position are.}
\label{fig:logo}
\end{figure}

\section{Can we recover phylogenetic information from the essential genes?}
To get the evolutionary relationship among all our strains, we collected all clusters with one and only one gene per genome and concatenated all the genes corresponding to every strain. Then aligned them using mafft and generated a phylogenetic tree using fasttree software. The resulted phylogenetic tree is depicted in Figure~\ref{fig:species-tree}.

\begin{figure}
\includegraphics[scale=0.14]{../speciestree/speciestree.pdf}
\caption{The tree is generated using fasttree software}
\label{fig:species-tree}
\end{figure}

To test if the same tree can be obtained from the essentiality of genes, we have selected all clusters that contain no more than one gene from each strain and made a binary matrix from the essentiality of the genes in these clusters. If a gene is essential in a strain, the corresponding value in the matrix is 1 and if the gene does not exist in the strain or is not essential, the value is 0. Then, we have generated a distance matrix from these values and plotted a phylogenetic tree.  Figure~\ref{fig:essentiality-tree} indicates that the resulting tree does not maintain the phylogenetic information of the species under study.

\begin{figure}
\includegraphics[scale=0.24]{make-essentiality-tree/essentiality-tree.pdf}
\caption{The phylogenetic tree is generated using ``phylip neighbor'' software.}
\label{fig:essentiality-tree}
\end{figure}

We have also compared every pair of strains and calculated the number of genes that are essential in one strain and absent in the other, the number of genes that are essential in both strains, the number of genes that are essential in one strain and present but not essential in the other strain, the number of genes that are present in one strain and absent in the other, and the number of genes that are shared between the two strains. The resulted heatmaps can be seen in Figure~\ref{fig:pairwise-venn}. The dendograms obtained from heatmaps in this figure are consistent with the species tree to some extent.

\begin{figure*}
\centering
\begin{tabular}{c c}
\includegraphics[page=1, scale=0.35]{essentiality-heatmap.pdf} & 
\includegraphics[page=2, scale=0.35]{essentiality-heatmap.pdf} \\
\includegraphics[page=3, scale=0.35]{essentiality-heatmap.pdf} & 
\includegraphics[page=4, scale=0.35]{essentiality-heatmap.pdf} \\
\includegraphics[page=5, scale=0.35]{essentiality-heatmap.pdf} & 
\\
\end{tabular}
\caption{The number of genes that are essential in one strain and absent in the other (upper left), the number of genes that are essential in both strains (upper right), the number of genes that are essential in one strain and present but not essential in the other strain (middle left), the number of genes that are present in one strain and absent in the other (middle right), and the number of genes that are shared between the two strains (lower left).}
\label{fig:pairwise-venn}
\end{figure*}

\end{document}

### 2016-02-09 documentation/EnTrI/EnTrI.tex
Reran codes after removing Ecoli9000 from all codes, added species tree

@@
 Since transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for two types of biases: the distance from the origin and GC content.
-As Figure~\ref{fig:distance_bias} shows, the insertion index decreases while increasing the distance from the origin.
-Therefore, there is a bias towards the distance from the origin.
+% As Figure~\ref{fig:distance_bias} shows, there is an insertion index bias towards the position of the genes for different strains.
 The distance bias in every individual strain is depicted in Figure~\ref{fig:distance_bias_individual}.
 These plots indicate that the bias is negligible in some strains like Salmonella typhi, while the insertion indices of the genes in other strains need to be normalised by their distances from the origin.
+We have used the LOESS curve and for each strain, divided the insertion indices by the predicted LOESS value to normalise the insertion indices.
+As the distribution of new insertion indices will be around 1, we have multiplied the resulted values by the mean of the initial insertion indices to have a distribution around the mean.
+The results are shown in Figure~\ref{fig:normalised_distance_bias}.
 
-\begin{figure}[H]
-\includegraphics[scale=0.28, page=1]{biases.pdf}
-\caption{The bias towards the distance from origin (DnaA gene) for all of the species under study. The red curve shows the fitted LOWESS regression curve.}
-\label{fig:distance_bias}
-\end{figure}
+%\begin{figure}[H]
+%\includegraphics[scale=0.28, page=1]{biases.pdf}
+%\caption{The bias towards the distance from origin (DnaA gene) for all of the species under study. The red curve shows the fitted LOWESS regression curve.}
+%\label{fig:distance_bias}
+%\end{figure}
 
@@
 \includegraphics[page=2, scale=0.22]{biases.pdf} & 
-\includegraphics[page=3, scale=0.22]{biases.pdf} &
-\includegraphics[page=4, scale=0.22]{biases.pdf} \\
-\includegraphics[page=5, scale=0.22]{biases.pdf} & 
-\includegraphics[page=6, scale=0.22]{biases.pdf} &
-\includegraphics[page=7, scale=0.22]{biases.pdf} \\
+\includegraphics[page=4, scale=0.22]{biases.pdf} &
+\includegraphics[page=6, scale=0.22]{biases.pdf} \\
 \includegraphics[page=8, scale=0.22]{biases.pdf} & 
-\includegraphics[page=9, scale=0.22]{biases.pdf} &
-\includegraphics[page=10, scale=0.22]{biases.pdf} \\
-\includegraphics[page=11, scale=0.22]{biases.pdf} & 
-\includegraphics[page=12, scale=0.22]{biases.pdf} &
-\includegraphics[page=13, scale=0.22]{biases.pdf} \\
-& 
+\includegraphics[page=10, scale=0.22]{biases.pdf} &
+\includegraphics[page=12, scale=0.22]{biases.pdf} \\
 \includegraphics[page=14, scale=0.22]{biases.pdf} & 
-\\
+\includegraphics[page=16, scale=0.22]{biases.pdf} &
+\includegraphics[page=18, scale=0.22]{biases.pdf} \\
+\includegraphics[page=20, scale=0.22]{biases.pdf} & 
+\includegraphics[page=22, scale=0.22]{biases.pdf} &
+\includegraphics[page=24, scale=0.22]{biases.pdf}
 \end{tabular}
-\caption{The bias towards the distance from origin for every individual strain. The red curves show the fitted LOWESS curves.}
+\caption{The bias towards the position of the gene for every individual strain. The red curves show the fitted LOESS curves.}
 \label{fig:distance_bias_individual}
@@
 
-As depicted in Figure~\ref{fig:GC_bias}, there is no bias towards the GC content of the genes and the regression line is flat.
-The nucleotides around the insertion sites (the insertion site and 10 nucleotides on each side) are stacked on top of each other and a sequence logo is generated from these sequences to see if there is any positional bias for any nucleotide.
-It can be inferred from Figure~\ref{fig:logo} that there is no bias in any position.
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=3, scale=0.22]{biases.pdf} & 
+\includegraphics[page=5, scale=0.22]{biases.pdf} &
+\includegraphics[page=7, scale=0.22]{biases.pdf} \\
+\includegraphics[page=9, scale=0.22]{biases.pdf} & 
+\includegraphics[page=11, scale=0.22]{biases.pdf} &
+\includegraphics[page=13, scale=0.22]{biases.pdf} \\
+\includegraphics[page=15, scale=0.22]{biases.pdf} & 
+\includegraphics[page=17, scale=0.22]{biases.pdf} &
+\includegraphics[page=19, scale=0.22]{biases.pdf} \\
+\includegraphics[page=21, scale=0.22]{biases.pdf} & 
+\includegraphics[page=23, scale=0.22]{biases.pdf} &
+\includegraphics[page=25, scale=0.22]{biases.pdf}
+\end{tabular}
+\caption{The insertion indices are normalised by the predicted LOESS value and the mean of the insertion indices. The red curves show the fitted LOESS curves.}
+\label{fig:normalised_distance_bias}
+\end{figure*}
 
-\begin{figure}
-\includegraphics[scale=0.28, page=15]{biases.pdf}
-\caption{The bias towards GC content. The regression line is depicted by green.}
+We have also checked for GC bias.
+There was a bias towards the GC content of the genes and we normalised the insertion indices for GC content in the same way as we did for gene position.
+The results, before and after the normalisation, are plotted in Figure~\ref{fig:GC_bias}.
+To see if there is any positional bias for any nucleotide, the nucleotides around the insertion sites (the insertion site and 10 nucleotides on each side) are stacked on top of each other and a sequence logo is generated from these sequences.
+It can be inferred from Figure~\ref{fig:logo} that there is no significant bias in any position.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[scale=0.35, page=26]{biases.pdf} &
+\includegraphics[scale=0.35, page=27]{biases.pdf}
+\end{tabular}
+\caption{The bias towards GC content. The LOESS curve is shown in red.}
 \label{fig:GC_bias}
-\end{figure}
+\end{figure*}
 
@@
 \includegraphics[scale=0.14]{../speciestree/speciestree.pdf}
-\caption{The tree is generated using fasttree software}
+\caption{The tree is generated from all the clusters with one and only one gene per strain.}
 \label{fig:species-tree}
@@
 
-To test if the same tree can be obtained from the essentiality of genes, we have selected all clusters that contain no more than one gene from each strain and made a binary matrix from the essentiality of the genes in these clusters.
-If a gene is essential in a strain, the corresponding value in the matrix is 1 and if the gene does not exist in the strain or is not essential, the value is 0.
-Then, we have generated a distance matrix from these values and plotted a phylogenetic tree.
+To test if the same tree can be obtained from the essentiality of genes, we have selected all clusters that contain exactly one gene from each strain (82 clusters) and made a binary matrix from the essentiality of the genes in these clusters.
+If a gene is essential in a strain, the corresponding value in the matrix is 1 and if the gene  is not essential, the value is 0.
+Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree.
 Figure~\ref{fig:essentiality-tree} indicates that the resulting tree does not maintain the phylogenetic information of the species under study.
@@
 \begin{figure}
-\includegraphics[scale=0.24]{make-essentiality-tree/essentiality-tree.pdf}
+\includegraphics[scale=0.14]{make-essentiality-tree/essentiality-tree.pdf}
 \caption{The phylogenetic tree is generated using ``phylip neighbor'' software.}

### 2016-02-10 documentation/EnTrI/EnTrI.tex
Added the number of core and accessore genes to the document, changed the sequence logo to show the probability of the occurrence of each base

@@
 
+\section{Are conserved genes more likely to be essential?}
+To find out whether conserved genes are more likely to be essential or not we have counted the number of essential, sometimes essential and never essential genes in both core and accessory genomes.
+The results are summarised in Table~\ref{table:core-accessory}.
+Core genes are the ones that have at least one copy per genome, otherwise, the genes are called accessory.
+
+\begin{table*}[t]
+\centering
+\begin{tabular}{| l || c | c | c || c | c | c |}
+\hline
+Species&\multicolumn{3}{c ||}{Core} & \multicolumn{3}{c |}{Accessory}\\
+\cline{2-7}
+Name&Ess&SEss&NEss&Ess&SEss&NEss\\
+\hline
+Salmonella typhimurium A130&266&665&1621&53&327&1528\\
+Salmonella typhimurium D23580&267&670&1624&54&339&1550\\
+Salmonella typhimurium SL1344&266&674&1630&63&350&1546\\
+Salmonella enteritidis P125109&266&679&1615&61&311&1015\\
+Salmonella typhi Ty2&265&628&1564&56&294&1505\\
+Escherichia coli UPEC ST131&299&685&1635&62&323&1767\\
+Escherichia coli ETEC CS17&280&656&1603&107&356&1586\\
+Escherichia coli ETEC H10407&284&656&1623&89&329&1743\\
+Escherichia coli K-12&279&657&1612&174&278&1311\\
+Citrobacter rodentium ICC168&304&685&1630&96&356&1868\\
+Enterobacter cloacae NCTC 9394&280&623&1526&34&129&1128\\
+Klebsiella penumoniae RH201207&357&864&1780&69&279&2314\\
+Klebsiella pneumoniae Ecl8&357&840&1770&61&225&1751\\
+\hline
+\end{tabular}
+\caption{The number of essential (Ess), sometimes essential (SEss) and never essential (NEss) genes in both core and accessory genomes.
+Core genes are the ones that have at least one copy per genome, otherwise, the genes are called accessory.}
+\label{table:core-accessory}
+\end{table*}
+
 \end{document}

### 2016-02-18 documentation/EnTrI/EnTrI.tex
Annotated species tree
Calculated essential, non-essential and sometimes essential genes for each genome, edited essentiality-plot and cluster-size-dist

--- 2016-02-09 documentation/EnTrI/EnTrI.tex
@@
 \begin{figure}
-\includegraphics[scale=0.14]{../speciestree/speciestree.pdf}
+\includegraphics[scale=0.14]{../speciestree/speciestree-annotated.png}
 \caption{The tree is generated from all the clusters with one and only one gene per strain.}
@@
 
+\section{Are conserved genes more likely to be essential?}
+To find out whether conserved genes are more likely to be essential or not we have counted the number of essential, sometimes essential and never essential genes in both core and accessory genomes.
+The results are summarised in Table~\ref{table:core-accessory}.
+Core genes are the ones that have at least one copy per genome, otherwise, the genes are called accessory.
+
+\begin{table*}
+\centering
+\begin{tabular}{| l || c | c | c || c || c | c | c || c |}
+\hline
+Species&\multicolumn{4}{c ||}{Core} & \multicolumn{4}{c |}{Accessory}\\
+\cline{2-9}
+Name&Ess&SEss&NEss&Sum&Ess&SEss&NEss&Sum\\
+\hline
+Salmonella typhimurium A130&266&665&1621&2552&53&327&1528&1908\\
+Salmonella typhimurium D23580&267&670&1624&2561&54&339&1550&1943\\
+Salmonella typhimurium SL1344&266&674&1630&2570&63&350&1546&1959\\
+Salmonella enteritidis P125109&266&679&1615&2560&61&311&1015&1387\\
+Salmonella typhi Ty2&265&628&1564&2457&56&294&1505&1855\\
+Escherichia coli UPEC ST131&299&685&1635&2619&62&323&1767&2152\\
+Escherichia coli ETEC CS17&280&656&1603&2539&107&356&1586&2049\\
+Escherichia coli ETEC H10407&284&656&1623&2563&89&329&1743&2161\\
+Escherichia coli K-12&279&657&1612&2548&174&278&1311&1763\\
+Citrobacter rodentium ICC168&304&685&1630&2619&96&356&1868&2320\\
+Enterobacter cloacae NCTC 9394&280&623&1526&2429&34&129&1128&1291\\
+Klebsiella penumoniae RH201207&357&864&1780&3001&69&279&2314&2662\\
+Klebsiella pneumoniae Ecl8&357&840&1770&2967&61&225&1751&2037\\
+\hline
+\end{tabular}
+\caption{The number of essential (Ess), sometimes essential (SEss) and never essential (NEss) genes in both core and accessory genomes.
+Core genes are the ones that have at least one copy per genome, otherwise, the genes are called accessory.}
+\label{table:core-accessory}
+\end{table*}
+
 \end{document}

### 2016-02-22 codes/make-logos.py
Made define-core-accessory.py lesss stupid!

@@ -8,7 +8,7 @@ from os import listdir, path, remove
 genomespath = '../sequences/fasta-genome/chromosome'
 plotspath = '../sequences/plot-files/chromosome'
 genomextension = 'fa'
-logospath = '../results/logos.txt'
+logospath = '../results/logos/logos.txt'
 complement = {'a': 't', 'c': 'g', 'g': 'c', 't': 'a', 'u': 'a', 'm': 'k', 'k': 'm', 'r': 'y',
               'y': 'r', 'w': 'w', 's': 's', 'v': 'b', 'b': 'v', 'h': 'd', 'd': 'h', 'n': 'n'}
 

### 2016-02-22 codes/test100logos.py

@@ -9,7 +9,7 @@ import numpy as np
 genomespath = '../sequences/fasta-genome/chromosome'
 plotspath = '../sequences/plot-files/chromosome'
 genomextension = 'fa'
-logospath = '../results/100logos.txt'
+logospath = '../results/logos/100logos.txt'
 complement = {'a': 't', 'c': 'g', 'g': 'c', 't': 'a', 'u': 'a', 'm': 'k', 'k': 'm', 'r': 'y',
               'y': 'r', 'w': 'w', 's': 's', 'v': 'b', 'b': 'v', 'h': 'd', 'd': 'h', 'n': 'n'}
 

### 2016-02-23 documentation/EnTrI/EnTrI.tex
Made define-core-accessory.py lesss stupid!
Added insertion index distribution and cluster size distribution to the TeX document

--- 2016-02-09 documentation/EnTrI/EnTrI.tex
@@
 \begin{document}
+\section{The homologue clustering program}
+We have developed a program which clusters homologous genes.
+This program uses jackhmmer from HMMER package for clustering.
+Figure~\ref{fig:cluster-size-dist} shows the results of this clustering on all the genes in the 13 species under study.
+To divide our clusters into ORFan, single copy, and multiple copy groups, we have used the procedure represented in Figure~\ref{fig:orfan-singlecopy-multiplecopy}.
+
+\begin{figure}[H]
+\includegraphics[scale=0.40]{cluster-size-dist.pdf}
+\caption{The distribution of the size of the clusters resulted from clustering all homologous genes in the 13 species under study.}
+\label{fig:cluster-size-dist}
+\end{figure}
+
+\begin{figure}[H]
+\includegraphics[scale=0.25, angle=90]{flowchart_orfan_sing_mult/flowchart-orfansingmult.pdf}
+\caption{The procedure used to divide clusters into three groups: ORFan, single copy, and multiple copy.}
+\label{fig:orfan-singlecopy-multiplecopy}
+\end{figure}
+
 \section{Are EnTrI results biased?}
 Since transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for two types of biases: the distance from the origin and GC content.
-% As Figure~\ref{fig:distance_bias} shows, there is an insertion index bias towards the position of the genes for different strains.
 The distance bias in every individual strain is depicted in Figure~\ref{fig:distance_bias_individual}.
@@
 The results are shown in Figure~\ref{fig:normalised_distance_bias}.
-
-%\begin{figure}[H]
-%\includegraphics[scale=0.28, page=1]{biases.pdf}
-%\caption{The bias towards the distance from origin (DnaA gene) for all of the species under study. The red curve shows the fitted LOWESS regression curve.}
-%\label{fig:distance_bias}
-%\end{figure}
 
@@
 \end{tabular}
-\caption{The bias towards GC content. The LOESS curve is shown in red.}
+\caption{The bias towards GC content (left) and the data after normalising GC bias using LOESS curve and mean (right). The LOESS curve is shown in red.}
 \label{fig:GC_bias}
@@
 
-\begin{figure}
-\includegraphics[scale=.98, angle = -90]{logo.pdf}
+\begin{figure}[H]
+\centering
+\begin{tabular}{c c}
+\includegraphics[scale=.98, angle = -90]{logos/logo-prob.pdf} &
+\includegraphics[scale=.98, angle = -90]{logos/logo-bits.pdf}
+\end{tabular}
 \caption{The nucleotides around the insertion sites (the insertion site and 10 nucleotides on each side) are stacked on top of each other and a sequence logo is generated from these sequences using webLogo stand-alone package.
@@
 
+\section{Gene essentiality}
+After normalising the insertion indices, we have used these values to determine the essentiality of genes.
+The insertion index distribution for each single strain is pictured in Figure~\ref{fig:ii-singlestrain}.
+If the insertion index is less than 0.2, the corresponding gene is regarded as essential, if it is between 0.2 and 2, the gene is non-essential, and otherwise, it is considered as a beneficial loss.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=1, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=2, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=3, scale=0.28]{per-species-insertion-index.pdf} \\ 
+\includegraphics[page=4, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=5, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=6, scale=0.28]{per-species-insertion-index.pdf} \\
+\includegraphics[page=7, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=8, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=9, scale=0.28]{per-species-insertion-index.pdf} \\
+\includegraphics[page=10, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=11, scale=0.28]{per-species-insertion-index.pdf} &
+\includegraphics[page=12, scale=0.28]{per-species-insertion-index.pdf} \\
+\end{tabular}
+\caption{Insertion index distribution for every single strain.}
+\label{fig:ii-singlestrain}
+\end{figure*}
+
+Figure~\ref{fig:ii-clusters} shows the insertion index distribution for all clusters.
+We have divided the clusters into ORFan, single copy, and multiple copy groups and studied the essentiality in these groups.
+
+\begin{figure}[H]
+\includegraphics[scale=0.4]{cluster-essentiality.pdf}
+\caption{Essentiality of clusters in all clusters, ORFan, single copy clusters, and multiple copy clusters.}
+\label{fig:ii-clusters}
+\end{figure}
+
 \section{Can we recover phylogenetic information from the essential genes?}
 To get the evolutionary relationship among all our strains, we collected all clusters with one and only one gene per genome and concatenated all the genes corresponding to every strain.
-Then aligned them using mafft and generated a phylogenetic tree using fasttree software.
+Then, aligned them using mafft and generated a phylogenetic tree using fasttree software.
 The resulted phylogenetic tree is depicted in Figure~\ref{fig:species-tree}.
 
-\begin{figure}
-\includegraphics[scale=0.14]{../speciestree/speciestree.pdf}
-\caption{The tree is generated from all the clusters with one and only one gene per strain.}
+\begin{figure}[H]
+\includegraphics[scale=0.13]{../speciestree/speciestree-annotated.png}
+\caption{The tree is generated from all genes that are copied only once per genome using an approximately-maximum-likelihood method called fasttree.}
 \label{fig:species-tree}
@@
 If a gene is essential in a strain, the corresponding value in the matrix is 1 and if the gene  is not essential, the value is 0.
-Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree.
+Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree using ``phylip neighbor'' software which uses a neighbour joining method.
 Figure~\ref{fig:essentiality-tree} indicates that the resulting tree does not maintain the phylogenetic information of the species under study.
 
-\begin{figure}
-\includegraphics[scale=0.14]{make-essentiality-tree/essentiality-tree.pdf}
-\caption{The phylogenetic tree is generated using ``phylip neighbor'' software.}
+\begin{figure}[H]
+\includegraphics[scale=0.13]{make-essentiality-tree/essentiality-tree.pdf}
+\caption{The tree is generated from all the genes that are copied only once per genome.
+We have made a binary matrix from the essentiality of these genes.
+Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree using ``phylip neighbor'' software which uses a neighbour joining method.}
 \label{fig:essentiality-tree}
@@
 The resulted heatmaps can be seen in Figure~\ref{fig:pairwise-venn}.
-The dendograms obtained from heatmaps in this figure are consistent with the species tree to some extent.
 
@@
 \begin{tabular}{c c}
-\includegraphics[page=1, scale=0.35]{essentiality-heatmap.pdf} & 
-\includegraphics[page=2, scale=0.35]{essentiality-heatmap.pdf} \\
-\includegraphics[page=3, scale=0.35]{essentiality-heatmap.pdf} & 
-\includegraphics[page=4, scale=0.35]{essentiality-heatmap.pdf} \\
-\includegraphics[page=5, scale=0.35]{essentiality-heatmap.pdf} & 
-\\
+\includegraphics[scale=0.35]{venn/venn.pdf}&
+\includegraphics[page=5, scale=0.34]{essentiality-heatmap.pdf} \\ 
+\includegraphics[page=2, scale=0.34]{essentiality-heatmap.pdf} &
+\includegraphics[page=3, scale=0.34]{essentiality-heatmap.pdf} \\ 
+\includegraphics[page=1, scale=0.34]{essentiality-heatmap.pdf} &
+\includegraphics[page=4, scale=0.34]{essentiality-heatmap.pdf}
 \end{tabular}
-\caption{The number of genes that are essential in one strain and absent in the other (upper left), the number of genes that are essential in both strains (upper right), the number of genes that are essential in one strain and present but not essential in the other strain (middle left), the number of genes that are present in one strain and absent in the other (middle right), and the number of genes that are shared between the two strains (lower left).}
+\caption{The heat maps shows the normalised number of genes in each group.
+The genes can be present in both strains (Pres-Pres in the Venn diagram and the upper right heat map), essential in both strains (Ess-Ess in the Venn diagram and the middle left heat map), essential in one species and not essential in the other (Ess-NoEs in the Venn diagram and the middle right heat map), essential in one species and absent in the other species (Ess-Abs in the Venn diagram and the bottom left heat map), and present in one species and absent in the other (pres-Abs in the Venn diagram and the bottom right heat map).}
 \label{fig:pairwise-venn}
@@
 
+\section{Are conserved genes more likely to be essential?}
+To find out whether conserved genes are more likely to be essential or not we have counted the number of essential, sometimes essential and never essential genes in both core and accessory genomes.
+The results are summarised in Table~\ref{table:core-accessory}.
+Core genes are the ones that have at least one copy per genome, otherwise, the genes are called accessory.
+The p-values show that the essentiality of genes is not independent of their conservation.
+
+\begin{table*}
+\centering
+\begin{tabular}{| l || c | c | c || c || c | c | c || c || c |}
+\hline
+Species&\multicolumn{4}{c ||}{Core} & \multicolumn{4}{c |}{Accessory} & Fisher's test\\
+\cline{2-9}
+Name&Ess&SEss&NEss&Sum&Ess&SEss&NEss&Sum&p-value\\
+\hline
+Salmonella typhimurium A130&266&665&1621&2552&53&327&1528&1908&$< 2.2e-16$\\
+Salmonella typhimurium D23580&267&670&1624&2561&54&339&1550&1943&$< 2.2e-16$\\
+Salmonella typhimurium SL1344&266&674&1630&2570&63&350&1546&1959&$< 2.2e-16$\\
+Salmonella enteritidis P125109&266&679&1615&2560&61&311&1015&1387&$= 2.669e-14$\\
+Salmonella typhi Ty2&265&628&1564&2457&56&294&1505&1855&$< 2.2e-16$\\
+Escherichia coli UPEC ST131&299&685&1635&2619&62&323&1767&2152&$< 2.2e-16$\\
+Escherichia coli ETEC CS17&280&656&1603&2539&107&356&1586&2049&$< 2.2e-16$\\
+Escherichia coli ETEC H10407&284&656&1623&2563&89&329&1743&2161&$< 2.2e-16$\\
+%Escherichia coli K-12&279&657&1612&2548&174&278&1311&1763&$= 6.664e-16$\\
+Citrobacter rodentium ICC168&304&685&1630&2619&96&356&1868&2320&$< 2.2e-16$\\
+Enterobacter cloacae NCTC 9394&280&623&1526&2429&34&129&1128&1291&$< 2.2e-16$\\
+Klebsiella penumoniae RH201207&357&864&1780&3001&69&279&2314&2662&$< 2.2e-16$\\
+Klebsiella pneumoniae Ecl8&357&840&1770&2967&61&225&1751&2037&$< 2.2e-16$\\
+\hline
+\end{tabular}
+\caption{The number of essential (Ess), sometimes essential (SEss) and never essential (NEss) genes in both core and accessory genomes.
+Core genes are the ones that have at least one copy per genome, otherwise, the genes are called accessory.}
+\label{table:core-accessory}
+\end{table*}
+
 \end{document}

### 2016-02-29 documentation/EnTrI/EnTrI.tex
Added code to study the insertion position bias and compare clusters with EC numbers

@@
 \section{Are EnTrI results biased?}
-Since transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for two types of biases: the distance from the origin and GC content.
+Since transposon insertion biases can affect the essentiallity level inferred from transposon mutagenesis experiments, the dataset has been tested for three types of biases: the distance from the origin, GC content, and the position of insertions in genes.
 The distance bias in every individual strain is depicted in Figure~\ref{fig:distance_bias_individual}.
@@
 \end{figure}
+
+The bias towards the position of insertions within genes has been studied since the genes can tolerate lower number of insertions at the start of the gene.
+We have divided our genes into 3 parts: first 60 nucleotides, last 60 nucleotides, and the rest and plotted the normalised number of insertions in each of these groups.
+Figure~\ref{fig:insertion-position-bias} summarises the results.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[scale=.45, page = 1]{insertion-position-bias.pdf} &
+\includegraphics[scale=.45, page = 2]{insertion-position-bias.pdf} \\
+\includegraphics[scale=.45, page = 3]{insertion-position-bias.pdf} &
+\includegraphics[scale=.45, page = 4]{insertion-position-bias.pdf}
+\end{tabular}
+\caption{We have divided our genes into 3 parts: first 60 nucleotides (position 1-60), last 60 nucleotides (position 121-180), and the rest (position 61-120) and plotted the normalised number of insertion sites in each of these groups.}
+\label{fig:insertion-position-bias}
+\end{figure*}
 

### 2016-02-29 codes/cluster-check.py

@@ -1,0 +1,79 @@
+from Bio import SeqIO
+from itertools import combinations
+from re import search, findall
+from os import listdir
+from collections import defaultdict
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+def read_clusters(input_dir):
+    sequence_occurrence = defaultdict(list)
+    list_of_files = listdir(input_dir)
+    for filename in list_of_files:
+        with open('{0}/{1}'.format(input_dir, filename)) as cluster_file:
+            for line in cluster_file:
+                cells = line.split()
+                sequence_occurrence[cells[1]].append(filename)
+    return sequence_occurrence
+
+seqdb = '../sequences/fasta-dna/chromosome/seqdb.fasta'
+clustdb = '../results/homclust/EFam-clusters'
+ec_clusters = read_fasta_sequences(seqdb)
+efam_clusters = read_clusters(clustdb)
+tp, fn, fp, tn = 0, 0, 0, 0
+for key1, key2 in combinations(ec_clusters.keys(), r=2):
+    search_results1 = search('\[EC_number=([^\]]*)\]', ec_clusters[key1].description)
+    find_results1 = findall('(\d+\.\d+\.\d+)\.\d+', search_results1.group(1))
+    search_results2 = search('\[EC_number=([^\]]*)\]', ec_clusters[key2].description)
+    find_results2 = findall('(\d+\.\d+\.\d+)\.\d+', search_results2.group(1))
+    if find_results1 and find_results2:
+        ec_match = 0
+        for item1 in find_results1:
+            for item2 in find_results2:
+                if item1 == item2:
+                    ec_match = 1
+                    break
+            if ec_match:
+                break
+        efam_match = 0
+        for item1 in efam_clusters[key1]:
+            for item2 in efam_clusters[key2]:
+                if item1 == item2:
+                    efam_match = 1
+                    break
+            if efam_match:
+                break
+        if not ec_match and not efam_match:
+            tn += 1
+        elif not ec_match and efam_match:
+            fp += 1
+            print '{}\t{}\t{}\t{}\t{}\t{}'.format(key1, find_results1, efam_clusters[key1], key2, find_results2, efam_clusters[key2])
[…]

### 2016-02-29 codes/cluster_size_dist_plot.R

@@ -1,33 +1,0 @@
-list_of_files <- list.files(path="../results/homclust/EFam-clusters", pattern="*.txt", full.names=T, recursive=FALSE)
-count = c()
-clusters = matrix(nrow=length(list_of_files), ncol=2)
-i = 1
-for (filename in list_of_files)
-{
-  count = c(count, length(readLines(filename)))
-  clusters[i,1]= basename(filename)
-  clusters[i,2]=count[i]
-  i = i+1
-}
-count_table = table(count)
-
-pdf("../results/cluster-size-dist.pdf")
-
-mar.default <- c(5,4,4,2) + 0.1
-par(mar = mar.default + c(0, 1, 0, 0)) 
-n= 30
-cols = c(rep("green1", 5), rep("orange", 13), rep("gray35", 96))
-midpoints <- barplot(count_table,xlab="Cluster size",ylab="Frequency", main ="Cluster size distribution", cex.lab = 2, cex.axis = 1.5, cex.main = 2,
-        #col = rainbow(n, s = 1, v = 1, start = 0, end = max(1, n - 1)/n, alpha = 1),
-        col=cols,
-        xlim=c(0,30), xaxt="n"
-        #, yaxt="n"
-        )
-#axis(2, at=c(0,10,20,30,40), labels=c("0","100","400","900","1600"))
-axis(1, at=midpoints[seq(2,n,2)], labels=seq(2,n,2))
-#plot(ecdf(count), xlim=c(0,20))
-lines(c(6.1, 6.1), c(-100,3000), col = "red", lwd=3, lty = 2)
-lines(c(21.7, 21.7), c(-100,3000), col = "red", lwd=3, lty = 2)
-legend(7,2300, c("ORFan","Single copy", "Multiple copy"), lty=c(1,1,1),cex=1.5, bty="n", lwd=c(4,4, 4),col=c("green1","orange", "gray35"))
-
-dev.off()

### 2016-02-29 codes/insertion-position-bias.R

@@ -1,0 +1,33 @@
+insertion_positions <- read.table("../results/insertion-position-bias.out", row.names=1)
+pdf("../results/insertion-position-bias.pdf")
+toplot <- colMeans(insertion_positions[,1:ncol(insertion_positions)-1])
+midpoints <- barplot(toplot, main="All", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2)
+axis(1, at=midpoints[seq(18,180,18)], labels=seq(18,180,18), cex.axis=1.5)
+essential = c()
+nonessential = c()
+beneficialloss = c()
+for (i in seq(1,nrow(insertion_positions)))
+{
+  if (insertion_positions[i,ncol(insertion_positions)] < 0.2)
+  {
+    essential = rbind(essential, insertion_positions[i,1:ncol(insertion_positions)-1]) 
+  }
+  else if (insertion_positions[i,ncol(insertion_positions)] < 2)
+  {
+    nonessential = rbind(nonessential, insertion_positions[i,1:ncol(insertion_positions)-1])
+  }
+  else
+  {
+    beneficialloss = rbind(beneficialloss, insertion_positions[i,1:ncol(insertion_positions)-1]) 
+  }
+}
+toplot <- colMeans(essential)
+midpoints <- barplot(toplot, main="Essential", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2)
+axis(1, at=midpoints[seq(18,180,18)], labels=seq(18,180,18), cex.axis=1.5)
+toplot <- colMeans(nonessential)
+midpoints <- barplot(toplot, main="Non-essential", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2)
+axis(1, at=midpoints[seq(18,180,18)], labels=seq(18,180,18), cex.axis=1.5)
+toplot <- colMeans(beneficialloss)
+midpoints <- barplot(toplot, main="Beneficial loss", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2)
+axis(1, at=midpoints[seq(18,180,18)], labels=seq(18,180,18), cex.axis=1.5)
+dev.off()

### 2016-02-29 codes/insertion-position-bias.py

@@ -1,0 +1,67 @@
+from collections import defaultdict
+from re import match
+from os import listdir, path
+from math import floor
+
+seqdb = '../sequences/fasta-dna/chromosome/seqdb.fasta'
+plots = '../sequences/plot-files/chromosome'
+result = '../results/insertion-position-bias.out'
+genome_length = {"SL1344":4878012, "STMMW":4879400, "SEN":4685848, "t":4791961, "STM":4895639, "ETEC":5153435,
+                 "b":4641652, "CS17":4994793, "NCTC13441":5174631, "ROD":5346659, "BN373":5324709, "ERS227112":5869288,
+                 "ENC":4908759}
+list_of_files = listdir(plots)
+plots_dict = defaultdict(list)
+for filename in list_of_files:
+    name, extension = path.splitext(filename)
+    with open('{0}/{1}'.format(plots, filename), 'r') as plotfile:
+        for line in plotfile:
+            cells = line.split()
+            plots_dict[name].append(int(cells[0]) + int(cells[1]))
+
+genome_insertions = dict()
+for item in plots_dict.keys():
+    genome_insertions[item] = sum(plots_dict[item])
+
+gene_name = ''
+position_insertion = [0]*180
+with open(result, 'w') as tofile:
+    with open(seqdb, 'r') as sequencefile:
+        for line in sequencefile:
+            if line.startswith('>'):
+                if gene_name != '':
+                    tofile.write(gene_name)
+                    for item in position_insertion:
+                        tofile.write('\t' + str(item))
+                    tofile.write('\t' + str(ii) + '\n')
+                match_result = match('>\s*((\S+?)_+\S+)\s+\[\S+/(\d+)\-(\d+)\s\((\w+)\)', line)
+                if match_result is None:
+                    match_result = match('>\s*(([a-zA-Z]+)\d+)\s+\[\S+/(\d+)\-(\d+)\s\((\w+)\)', line)
+                if match_result is not None:
+                    gene_name = match_result.group(1)
+                    strain_name = match_result.group(2)
+                    start = int(match_result.group(3)) - 1
+                    end = int(match_result.group(4)) - 1
+                    strand = match_result.group(5)
+                else:
+                    strain_name = ''
+                if strain_name not in plots_dict.keys() or (end - start + 1) < 180:
+                    gene_name = ''
+                else:
+                    for i in range(0,60):
+                        if plots_dict[strain_name][start+i] > 0:
[…]

### 2016-03-04 documentation/EnTrI/EnTrI.tex
Fixed a bug in insertion-position-bias.py

@@
 
-The bias towards the position of insertions within genes has been studied since the genes can tolerate lower number of insertions at the start of the gene.
+The bias towards the position of insertions within genes has been studied since the genes can tolerate more insertions at the 3' end of the gene.
 We have divided our genes into 3 parts: first 60 nucleotides, last 60 nucleotides, and the rest and plotted the normalised number of insertions in each of these groups.

### 2016-03-04 codes/insertion-position-bias.R

@@ -1,8 +1,10 @@
 insertion_positions <- read.table("../results/insertion-position-bias.out", row.names=1)
 pdf("../results/insertion-position-bias.pdf")
 toplot <- colMeans(insertion_positions[,1:ncol(insertion_positions)-1])
-midpoints <- barplot(toplot, main="All", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2)
-axis(1, at=midpoints[seq(18,180,18)], labels=seq(18,180,18), cex.axis=1.5)
+cols = c(rep("khaki4",60), rep("slategray",60), rep("palevioletred4",60))
+midpoints <- barplot(toplot, main="All", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2, col=cols, border=NA)
+axis(1, at=midpoints[c(1,180)], labels=c('5\'','3\''), cex.axis=1.5)
+legend(61,0.7, c(expression(1 %->% 60),"internal", expression((L-60+1) %->% L)), lty=c(1,1,1),cex=1.5, lwd=c(4,4, 4),col=c(cols[1],cols[61],cols[121]), bg="white")
 essential = c()
 nonessential = c()
 beneficialloss = c()
@@ -22,12 +24,15 @@ for (i in seq(1,nrow(insertion_positions)))
   }
 }
 toplot <- colMeans(essential)
-midpoints <- barplot(toplot, main="Essential", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2)
-axis(1, at=midpoints[seq(18,180,18)], labels=seq(18,180,18), cex.axis=1.5)
+midpoints <- barplot(toplot, main="Essential", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2, col=cols, border=NA)
+axis(1, at=midpoints[c(1,180)], labels=c('5\'','3\''), cex.axis=1.5)
+legend(61,0.5, c(expression(1 %->% 60),"internal", expression((L-60+1) %->% L)), lty=c(1,1,1),cex=1.5, lwd=c(4,4, 4),col=c(cols[1],cols[61],cols[121]), bg="white")
 toplot <- colMeans(nonessential)
[…]

### 2016-03-04 codes/insertion-position-bias.py

@@ -20,7 +20,7 @@ for filename in list_of_files:
 
 genome_insertions = dict()
 for item in plots_dict.keys():
-    genome_insertions[item] = sum(plots_dict[item])
+    genome_insertions[item] = sum([1 for x in plots_dict[item] if x > 0])
 
 gene_name = ''
 position_insertion = [0]*180
@@ -64,4 +64,7 @@ with open(result, 'w') as tofile:
                         gene_insertions = sum([1 for x in plots_dict[strain_name][start:end+1] if x > 0])
                         ii = (float(gene_insertions)/(end-start+1))/(float(genome_insertions[strain_name])/genome_length[strain_name])
                     if strand == 'Complement':
-                        position_insertion.reverse()
+                        temp = position_insertion[0:(len(position_insertion)-1)]
+                        temp.reverse()
+                        temp.append(position_insertion[len(position_insertion)-1])
+                        position_insertion = temp

### 2016-03-08 codes/define-core-accessory.py
Added Salmonella typhimurium to define-core-accessory.py - Changed the colors in heatmaps

@@ -33,8 +33,8 @@ sequences = read_fasta_sequences(seqdb)
 list_of_files = listdir(clusters)
 species_names = defaultdict()
 species_names = {"all":["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441", "ROD", "SEN", "SL1344", "STM",
-    "STMMW", "t"], "salmonella":["SEN", "SL1344", "STM", "STMMW", "t"], "ecoli":["CS17", "ETEC",
-    "NCTC13441"], "klebsiella":["ERS227112", "BN373"], "citrobacter":["ROD"], "enterobacter":
+    "STMMW", "t"],"typhimurium":["STM", "SL1344", "STMMW"], "salmonella":["SEN", "SL1344", "STM", "STMMW", "t"],
+    "ecoli":["CS17", "ETEC", "NCTC13441"], "klebsiella":["ERS227112", "BN373"], "citrobacter":["ROD"], "enterobacter":
     ["ENC"], "salmonellaecoli":["SEN", "SL1344", "STM", "STMMW", "t", "CS17", "ETEC", "NCTC13441"],
     "salmonellaecolicitrobacter":["SEN", "SL1344", "STM", "STMMW", "t", "CS17", "ETEC", "NCTC13441", "ROD"],
     "klebsiellaenterobacter":["ERS227112", "BN373", "ENC"], "salmonellaty2":["t"], "salmonellap125109":["SEN"],

### 2016-05-10 documentation/EnTrI/EnTrI.tex
Added Salmonella typhimurium to define-core-accessory.py - Changed the colors in heatmaps
Added upsetr, Keio results, duplications,...

@@
 \begin{tabular}{c c c}
-\includegraphics[page=2, scale=0.22]{biases.pdf} & 
-\includegraphics[page=4, scale=0.22]{biases.pdf} &
-\includegraphics[page=6, scale=0.22]{biases.pdf} \\
-\includegraphics[page=8, scale=0.22]{biases.pdf} & 
-\includegraphics[page=10, scale=0.22]{biases.pdf} &
-\includegraphics[page=12, scale=0.22]{biases.pdf} \\
-\includegraphics[page=14, scale=0.22]{biases.pdf} & 
-\includegraphics[page=16, scale=0.22]{biases.pdf} &
-\includegraphics[page=18, scale=0.22]{biases.pdf} \\
-\includegraphics[page=20, scale=0.22]{biases.pdf} & 
-\includegraphics[page=22, scale=0.22]{biases.pdf} &
-\includegraphics[page=24, scale=0.22]{biases.pdf}
+\includegraphics[page=2, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=4, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=6, scale=0.22]{biases-without-ends.pdf} \\
+\includegraphics[page=8, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=10, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=12, scale=0.22]{biases-without-ends.pdf} \\
+\includegraphics[page=14, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=16, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=18, scale=0.22]{biases-without-ends.pdf} \\
+\includegraphics[page=20, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=22, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=24, scale=0.22]{biases-without-ends.pdf}
 \end{tabular}
@@
 \begin{tabular}{c c c}
-\includegraphics[page=3, scale=0.22]{biases.pdf} & 
-\includegraphics[page=5, scale=0.22]{biases.pdf} &
-\includegraphics[page=7, scale=0.22]{biases.pdf} \\
-\includegraphics[page=9, scale=0.22]{biases.pdf} & 
-\includegraphics[page=11, scale=0.22]{biases.pdf} &
-\includegraphics[page=13, scale=0.22]{biases.pdf} \\
-\includegraphics[page=15, scale=0.22]{biases.pdf} & 
-\includegraphics[page=17, scale=0.22]{biases.pdf} &
-\includegraphics[page=19, scale=0.22]{biases.pdf} \\
-\includegraphics[page=21, scale=0.22]{biases.pdf} & 
-\includegraphics[page=23, scale=0.22]{biases.pdf} &
-\includegraphics[page=25, scale=0.22]{biases.pdf}
+\includegraphics[page=3, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=5, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=7, scale=0.22]{biases-without-ends.pdf} \\
+\includegraphics[page=9, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=11, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=13, scale=0.22]{biases-without-ends.pdf} \\
+\includegraphics[page=15, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=17, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=19, scale=0.22]{biases-without-ends.pdf} \\
+\includegraphics[page=21, scale=0.22]{biases-without-ends.pdf} & 
+\includegraphics[page=23, scale=0.22]{biases-without-ends.pdf} &
+\includegraphics[page=25, scale=0.22]{biases-without-ends.pdf}
 \end{tabular}
@@
 \begin{tabular}{c c}
-\includegraphics[scale=0.35, page=26]{biases.pdf} &
-\includegraphics[scale=0.35, page=27]{biases.pdf}
+\includegraphics[scale=0.35, page=26]{biases-without-ends.pdf} &
+\includegraphics[scale=0.35, page=27]{biases-without-ends.pdf}
 \end{tabular}
@@
 
+%fisher test: fisher.test(matrix(c(280,623,1526,34,129,1128),nrow=2,byrow=TRUE))
 \begin{table*}

### 2016-05-10 codes/upsetr.R

@@ -1,0 +1,43 @@
+library("UpSetR")
+library(stringr)
+# listinput <- list(a=c(1,1,1,1,2,1,2,3,3,3), b=c(1,4,5), d=c(1,2,5,6,7), e=c(4,8,9))
+# upset(fromList(listinput))
+
+clusters_path <- "../results/merge-clust-plot"
+list_of_files <- list.files(path=clusters_path, full.names=T, recursive=FALSE)
+presence = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
+                "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
+essentiality = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
+                    "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
+for (filename in list_of_files)
+{
+  cluster <- as.matrix(read.table(filename))
+  cluster_size = nrow(cluster)
+  for (i in (1:cluster_size))
+  {
+    if (as.numeric(cluster[i,5]) >= 0)
+    {
+      match = str_match(cluster[i,2], "([[:graph:]]+)\\_[[:alnum:]]+")[2]
+      if (is.na(match))
+      {
+        match = str_match(cluster[i,2], "([[:alpha:]]+)[[:digit:]]+")[2]
+      }
+      if (match %in% names(presence))
+      {
+        presence[[match]] = c(presence[[match]], strsplit(basename(filename),"\\.")[[1]][1])
+        if (as.numeric(cluster[i,5]) < 0.2)
+        {
+          essentiality[[match]] = c(essentiality[[match]], strsplit(basename(filename),"\\.")[[1]][1])
+        }
+      }
+    }
+  }
+}
+names(presence) <- c("Citrobacter", "EC_ETEC_CS17", "Enterobacter", "EC_ETEC_H10407", "EC_UPEC",
+         "KP_RH201207", "KP_Ecl8", "S_enteritidis", "S_typhimurium_A130",
+         "S_typhimurium_SL1344", "S_typhimurium_D23580", "S_typhi", "EC_k12")
[…]

### 2016-05-10 codes/use-fdollop.py

@@ -1,0 +1,69 @@
+#! /bin/sh
+""":"
+exec python3 $0 ${1+"$@"}
+"""
+
+from os import listdir, path, mkdir, system
+from Bio import SeqIO
+from shutil import rmtree
+from re import match
+from collections import defaultdict
+
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+
+def makedir(dirname):
+    if path.exists(dirname):
+        input_var = 'i'
+        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
+            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
+        if input_var == 'Y' or input_var == 'y':
+            rmtree(dirname)
+        else:
+            raise SystemExit
+    mkdir(dirname)
+
+seqdb = '/home/fatemeh/EnTrI/sequences/fasta-protein/chromosome/seqdb.fasta'
+clusters = '/home/fatemeh/EnTrI/results/merge-clust-plot'
+outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-dollo'
+makedir(outdir)
+sequences = read_fasta_sequences(seqdb)
+list_of_files = listdir(clusters)
+species_names = defaultdict()
+species_names = {"all":["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441", "ROD", "SEN", "SL1344", "STM",
+    "STMMW", "t"],"typhimurium":["STM", "SL1344", "STMMW"], "salmonella":["SEN", "SL1344", "STM", "STMMW", "t"],
+    "ecoli":["CS17", "ETEC", "NCTC13441"], "klebsiella":["ERS227112", "BN373"], "citrobacter":["ROD"], "enterobacter":
+    ["ENC"], "salmonellaecoli":["SEN", "SL1344", "STM", "STMMW", "t", "CS17", "ETEC", "NCTC13441"],
+    "salmonellaecolicitrobacter":["SEN", "SL1344", "STM", "STMMW", "t", "CS17", "ETEC", "NCTC13441", "ROD"],
+    "klebsiellaenterobacter":["ERS227112", "BN373", "ENC"], "salmonellaty2":["t"], "salmonellap125109":["SEN"],
[…]

### 2016-06-07 documentation/EnTrI2/manuscript.tex
Wrote a part of introduction, tested redefining insertion indices using number of reads, tested using a mixture of gamma distributions

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Proceedings of the National Academy of Sciences (PNAS)
% LaTeX Template
% Version 1.0 (19/5/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% The PNAStwo class was created and is owned by PNAS:
% http://www.pnas.org/site/authors/LaTex.xhtml
% This template has been modified from the blank PNAS template to include
% examples of how to insert content and drastically change commenting. The
% structural integrity is maintained as in the original blank template.
%
% Original header:
%% PNAStmpl.tex
%% Template file to use for PNAS articles prepared in LaTeX
%% Version: Apr 14, 2008
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

%------------------------------------------------
% BASIC CLASS FILE
%------------------------------------------------

%% PNAStwo for two column articles is called by default.
%% Uncomment PNASone for single column articles. One column class
%% and style files are available upon request from pnas@nas.edu.

%\documentclass{pnasone}
\documentclass{pnastwo}

%------------------------------------------------
% POSITION OF TEXT
%------------------------------------------------

%% Changing position of text on physical page:
%% Since not all printers position
%% the printed page in the same place on the physical page,
%% you can change the position yourself here, if you need to:

% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
                         %  physical page; positive dimension will lower it.

%% You may set the dimension to the size that you need.

%------------------------------------------------
% GRAPHICS STYLE FILE
%------------------------------------------------

%% Requires graphics style file (graphicx.sty), used for inserting
%% .eps/image files into LaTeX articles.
%% Note that inclusion of .eps files is for your reference only;
%% when submitting to PNAS please submit figures separately.

%% Type into the square brackets the name of the driver program 
%% that you are using. If you don't know, try dvips, which is the
%% most common PC driver, or textures for the Mac. These are the options:

% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]

\usepackage{graphicx}
\graphicspath{{figs/}}
%------------------------------------------------
% OPTIONAL POSTSCRIPT FONT FILES
%------------------------------------------------

%% PostScript font files: You may need to edit the PNASoneF.sty
%% or PNAStwoF.sty file to make the font names match those on your system. 
%% Alternatively, you can leave the font style file commands commented out
%% and typeset your article using the default Computer Modern 
%% fonts (recommended). If accepted, your article will be typeset
%% at PNAS using PostScript fonts.

% Choose PNASoneF for one column; PNAStwoF for two column:
%\usepackage{PNASoneF}
%\usepackage{PNAStwoF}

%------------------------------------------------
% ADDITIONAL OPTIONAL STYLE FILES
%------------------------------------------------

%% The AMS math files are commonly used to gain access to useful features
%% like extended math fonts and math commands.

\usepackage{amssymb,amsfonts,amsmath}

%------------------------------------------------
% OPTIONAL MACRO FILES
%------------------------------------------------

%% Insert self-defined macros here.
%% \newcommand definitions are recommended; \def definitions are supported

%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
%\def\s{\sigma}

%------------------------------------------------
% DO NOT EDIT THIS SECTION
%------------------------------------------------

%% For PNAS Only:
\contributor{Submitted to Proceedings of the National Academy of Sciences of the United States of America}
\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
\copyrightyear{2008}
\issuedate{Issue Date}
\volume{Volume}
\issuenumber{Issue Number}

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TITLE AND AUTHORS
%----------------------------------------------------------------------------------------

\title{Title of the publication} % For titles, only capitalize the first letter

%------------------------------------------------

%% Enter authors via the \author command.  
%% Use \affil to define affiliations.
%% (Leave no spaces between author name and \affil command)

%% Note that the \thanks{} command has been disabled in favor of
%% a generic, reserved space for PNAS publication footnotes.

%% \author{<author name>
%% \affil{<number>}{<Institution>}} One number for each institution.
%% The same number should be used for authors that
%% are affiliated with the same institution, after the first time
%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
%% Then, before last author ...
%% \and
%% \author{<author name>
%% \affil{<number>}{}}

%% For example, assuming Garcia and Sonnery are both affiliated with
%% Universidad de Murcia:
%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
%% United Kingdom},
%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}

\author{John Smith\affil{1}{University of California},
James Smith\affil{2}{University of Oregon}
\and
Jane Smith\affil{1}{}}

\contributor{Submitted to Proceedings of the National Academy of Sciences
of the United States of America}

%----------------------------------------------------------------------------------------

\maketitle % The \maketitle command is necessary to build the title page

\begin{article}

%----------------------------------------------------------------------------------------
%	ABSTRACT, KEYWORDS AND ABBREVIATIONS
%----------------------------------------------------------------------------------------

\begin{abstract}

\end{abstract}

%------------------------------------------------

\keywords{Keyword1 | Keyword2 | Keyword3} % When adding keywords, separate each term with a straight line: |

%------------------------------------------------

%% Optional for entering abbreviations, separate the abbreviation from
%% its definition with a comma, separate each pair with a semicolon:
%% for example:
%% \abbreviations{SAM, self-assembled monolayer; OTS,
%% octadecyltrichlorosilane}

% \abbreviations{}
\abbreviations{SAM, self-assembled monolayer; OTS, octadecyltrichlorosilane}

%----------------------------------------------------------------------------------------
%	PUBLICATION CONTENT
%----------------------------------------------------------------------------------------

%% The first letter of the article should be drop cap: \dropcap{} e.g.,
%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts

\section{Introduction}

\dropcap{S}tudying the essentiality of genes helps with identifying the fundamental processes necessary for cell viability \cite{juhas_essence_2011}. So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}. The results have led to new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising new genomes \cite{hutchison_global_1999, hutchison_design_2016}. Researchers have used different methods for studying the essentility of genes in prokaryotes. Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}. Another method that is widely used due to its simplicity and accuracy is transposon mutagenesis along with high-throughput sequencing \cite{gawronski_tracking_2009, van_opijnen_tn-seq:_2009, langridge_simultaneous_2009, christen_essential_2011, goodman_identifying_2011, wetmore_rapid_2015, rubin_essential_2015}. In this method, pools of single insertion mutants are constructed using transposon mutagenesis and the effect of each mutation on the survival of mutants is evaluated by sequencing the survivors \cite{barquist_approaches_2013}. This can lead to the identification of essential genes.

Although the essentiality of genes has been studied in a variety of organisms, there is still room to study the evolutionary conservation of essentiality. Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges. We extend this research by studying 12 bacterial strains. These include \textit{Salmonella} enterica subsp.\@ enterica serovar Typhi str.\@ Ty2, \textit{Salmonella} enterica subsp.\@ enterica serovar Enteritidis str.\@ P125109, \textit{Salmonella} enterica subsp.\@ enterica serovar Typhimurium str.\@ SL1344, \textit{Salmonella} enterica subsp.\@ enterica serovar Typhimurium str.\@ A130, \textit{Salmonella} enterica subsp.\@ enterica serovar Typhimurium str.\@ D23580, \textit{Escherichia} coli UPEC ST131, \textit{Escherichia} coli ETEC CS17, \textit{Escherichia} coli ETEC H10407, \textit{Citrobacter} rodentium ICC168, \textit{Klebsiella} pneumoniae RH201207, \textit{Klebsiella} pneumoniae subsp.\@ pneumoniae Ecl8, and \textit{Enterobacter} cloacae subsp.\@ cloacae NCTC 9394. All these strains are selected from Enterobacteriaceae family.

Enterobacteriaceae is a family that includes bacteria with different host ranges and pathogenecity found in soil, water, plants, animals and humans \cite{brenner_bergeys_2006}. In humans, various strains from this family can cause diarrhoea, septicaemia, urinary tract infection, meningitis, respiratory disease, and wound and burn infection \cite{brenner_bergeys_2006}. Besides, they can infect poultry and livestocks and cause financial losses for farmers \cite{brenner_bergeys_2006}. Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality in strains from 5 different species in this family.

%------------------------------------------------

\section{Results}

We have studied the essentiality of genes in 12 strains from Enterobacteriaceae family. The species are depicted in Fig~\ref{fig:tree}.

%------------------------------------------------

\section{Discussion}

%----------------------------------------------------------------------------------------
%	MATERIALS AND METHODS
%----------------------------------------------------------------------------------------

%% Optional Materials and Methods Section
%% The Materials and Methods section header will be added automatically.

\begin{materials}

\end{materials}

%----------------------------------------------------------------------------------------
%	APPENDICES (OPTIONAL)
%----------------------------------------------------------------------------------------

%\appendix
%An appendix without a title.
%
%\appendix[Appendix title]
%An appendix with a title.

%----------------------------------------------------------------------------------------
%	ACKNOWLEDGEMENTS
%----------------------------------------------------------------------------------------

\begin{acknowledgments}

\end{acknowledgments}

%----------------------------------------------------------------------------------------
%	BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

%% PNAS does not support submission of supporting .tex files such as BibTeX.
%% Instead all references must be included in the article .tex document. 
%% If you currently use BibTeX, your bibliography is formed because the 
%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
%% .tex document. To conform to PNAS requirements, copy the reference listings
%% from your .bbl file and add them to the article .tex file, using the
%% bibliography environment described above.

%%  Contact pnas@nas.edu if you need assistance with your
%%  bibliography.

% Sample bibliography item in PNAS format:
%% \bibitem{in-text reference} comma-separated author names up to 5,
%% for more than 5 authors use first author last name et al. (year published)
%% article title  {\it Journal Name} volume #: start page-end page.
%% ie,
% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991) 
% A short C-terminal sequence is necessary and sufficient for the
% targeting of chitinases to the plant vacuole. 
% {\it Proc Natl Acad Sci USA} 88:10362-10366.

%% Enter the largest bibliography number in the facing curly brackets
%% following \begin{thebibliography}

%\begin{thebibliography}{100}
%
%\end{thebibliography}
\bibliography{references}{}
\bibliographystyle{unsrt}

%----------------------------------------------------------------------------------------

\end{article}

%----------------------------------------------------------------------------------------
%	FIGURES AND TABLES
%----------------------------------------------------------------------------------------
\begin{figure}
\includegraphics[scale=0.25]{consensus-tree.pdf}
\label{fig:tree}
\caption{Species tree. How is it made?}
\end{figure}

%----------------------------------------------------------------------------------------

\end{document}

### 2016-06-07 codes/calculate-insertion-index.R

@@ -1,5 +1,6 @@
 library(stringr)
 library("MASS")
+library(mixtools)
 fastas_dir <- "~/EnTrI/sequences/fasta-protein/chromosome"
 plots_dir <- "~/EnTrI/sequences/plot-files/chromosome"
 output_dir1 <- "~/EnTrI/results/insertion-indices/insfree/"
@@ -43,27 +44,31 @@ for (filename in list_of_files)
         if (!(is.null(plots[[locusid]])))
         {
           len = end - start + 1
-          starttrim = floor(len*5/100)
-          endtrim = floor(len*20/100)
-          ii = (sum(plots[[locusid]][start:end]>0) / len) / (sumlength[[locusid]][1] / sumlength[[locusid]][2])
-          # ii = sum(plots[[locusid]][start:end]>0) / (end-start+1)
+          starttrim = floor(len*0/100)
+          endtrim = floor(len*0/100)
           essentiality = "not"
           if (direction == "Forward")
           {
-            if (100 < len & sum(plots[[locusid]][(start+starttrim):(end-endtrim)]) == 0)
+            newstart = start + starttrim
+            newend = end - endtrim
+            if (70 < len & sum(plots[[locusid]][newstart:newend]) == 0)
             {
               essentiality = "essential"
             }
-            iitable <- rbind(iitable, c(locustag, ii, essentiality))
           }
           else
           {
-            if (100 < len & sum(plots[[locusid]][(start+endtrim):(end-starttrim)]) == 0)
+            newstart = start + endtrim
+            newend = end - starttrim
+            if (70 < len & sum(plots[[locusid]][newstart:newend]) == 0)
             {
               essentiality = "essential"
             }
-            iitable <- rbind(iitable, c(locustag, ii, essentiality))
           }
+          newlen = newend - newstart + 1
+          ii = (sum(plots[[locusid]][newstart:newend]>0) / newlen) / (sumlength[[locusid]][1] / sumlength[[locusid]][2])
[…]

### 2016-06-14 documentation/EnTrI2/manuscript.tex
Wrote a part of introduction, tested redefining insertion indices using number of reads, tested using a mixture of gamma distributions
Tested mixtools with Lars' initial values, Regenerated the species tree with raxml and bootstrap values, Ran Hieranoid, Defined core and accessory on hieranoid results

@@
 \usepackage{amssymb,amsfonts,amsmath}
+\usepackage{color}
 
@@
 Besides, they can infect poultry and livestocks and cause financial losses for farmers \cite{brenner_bergeys_2006}.
-Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality in strains from 5 different species in this family.
+Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.
+
+\textcolor{red}{\{A summary of what we have done\}}
 
@@
 \section{Results}
+\textcolor{red}{\{An explanation of the experiment\}}
 
-We have studied the essentiality of genes in 12 strains from Enterobacteriaceae family. The species are depicted in Fig~\ref{fig:tree}.
+\textbf{Clustering orthologous and paralogous genes.} To study the essentiality of genes in 12 strains depicted in Fig~\ref{fig:tree}, We needed to cluster sets of orthologous genes in these strain.
+Plenty of methods are proposed for this purpose.
+Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
+We have used Hieranoid to cluster the sets of orthologous genes.
+In addition, we intended to study the essentiality of genes in paralogous genes.
+For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
 
@@
 \begin{figure}
-\includegraphics[scale=0.25]{consensus-tree.pdf}
+\includegraphics[scale=0.25]{consensustree.pdf}
 \label{fig:tree}

### 2016-06-14 codes/calculate-insertion-index.R

@@ -91,12 +91,6 @@ for (filename in list_of_files)
   iifile = as.matrix(read.table(filename, as.is = TRUE))
   
   ii= as.numeric(iifile[,2])
-  # h <- hist(ii,breaks=(0:(max(ii)*50+1)/50), xlim=c(0,4), freq=FALSE,xlab="Insertion index", main=dict[locusid])
-  # upperbound = h$mids[min(which(h$counts<10 & h$mids>1))]
-  # iinonzero = ii+1e-3
-  # mixmdl = gammamixEM(iinonzero, k=2)
-  # essen = max(min(iinonzero[apply(mixmdl$posterior, 1, which.max)==2]), min(iinonzero[apply(mixmdl$posterior, 1, which.max)==1]))-1e-3
-  
   
   nG = length(ii)
 
@@ -116,29 +110,43 @@ for (filename in list_of_files)
   m2 = h$mids[max(which(h$counts>5))]
   I1 = ((ii < m1)&(ii > 0))
   I2 = ((ii >= m1)&(ii < m2))
-  # I3 = (ii >= m2)
 
   f1 = (sum(I1) + sum(ii == 0))/nG
   f2 = (sum(I2))/nG
-  # f3 = (sum(I3))/nG
 
   d1 = fitdistr(ii[I1], "gamma", lower=min(ii[I1]))
   d2 = fitdistr(ii[I2], "gamma", lower=min(ii[I2])) #fit curves
-  # d3 = fitdistr(ii[I3], "gamma")
+  
+  # use mixtools
+  # eps = 1e-3
+  # iinonzero = ii+eps
+  # mixmdl = gammamixEM(iinonzero, alpha = c(1,d2$estimate[1]), beta = c(1/d1$estimate[2],1/d2$estimate[2]), k=2)
 
   #plots
   hist(ii,breaks=0:(max(ii)*50+1)/50, xlim=c(0,4), freq=FALSE,xlab="Insertion index", main=dict[locusid])
   lines(0:2000/500, f1*dgamma(0:2000/500, 1, d1$estimate[2])) # was [2]
   lines(0:2000/500, f2*dgamma(0:2000/500, d2$estimate[1], d2$estimate[2]))
+  # h <- hist(ii,breaks=(0:(max(ii)*50+1)/50), xlim=c(0,4), freq=FALSE,xlab="Insertion index", main=dict[locusid])
[…]

### 2016-06-21 codes/calculate-insertion-index.R
Corrected the distribution for Salmonella Enteritidis, corrected biases, used dollo to define core essential genes

@@ -1,6 +1,6 @@
 library(stringr)
 library("MASS")
-library(mixtools)
+# library(mixtools)
 fastas_dir <- "~/EnTrI/sequences/fasta-protein/chromosome"
 plots_dir <- "~/EnTrI/sequences/plot-files/chromosome"
 output_dir1 <- "~/EnTrI/results/insertion-indices/insfree/"

### 2016-06-21 codes/check-biases-new.py

@@ -1,0 +1,53 @@
+from collections import defaultdict
+from re import match, findall
+from os import listdir, path
+
+seqdb = '../sequences/fasta-dna/chromosome/seqdb.fasta'
+iis = '../results/insertion-indices/gamma'
+results = '../results/insertion-indices/check-biases'
+genome_length = {"SL1344":4878012, "STMMW":4879400, "SEN":4685848, "t":4791961, "STM":4895639, "ETEC":5153435,
+                 "b":4641652, "CS17":4994793, "NCTC13441":5174631, "ROD":5346659, "BN373":5324709, "ERS227112":5869288,
+                 "ENC":4908759}
+
+list_of_files = listdir(iis)
+for filename in list_of_files:
+    iis_dict = defaultdict(tuple)
+    with open('{0}/{1}'.format(iis, filename), 'r') as iifile:
+        for line in iifile:
+            cells = line.split()
+            iis_dict[cells[0]] = (float(cells[1]), cells[2])
+
+    gene_name = ''
+    with open(results + '/' + filename, 'w') as tofile:
+        with open(seqdb, 'r') as sequencefile:
+            for line in sequencefile:
+                if line.startswith('>'):
+                    if gene_name != '':
+                        tofile.write('{0}\t{1}\t{2}\t{3}\t{4}\n'.format(gene_name, iis_dict[gene_name][0], iis_dict[gene_name][1], float(start[0])/genome_length[strain_name], float(gc)/gene_length))
+                    gc = 0
+                    match_result = match('>\s*((\S+?)_+\S+)\s+\[\S+/((\d+\-\d+\s)+)\(', line)
+                    if match_result is None:
+                        match_result = match('>\s*(([a-zA-Z]+)\d+)\s+\[\S+/((\d+\-\d+\s)+)\(', line)
+                    if match_result is not None:
+                        gene_name = match_result.group(1)
+                        strain_name = match_result.group(2)
+                        starts_ends = findall('\d+\-\d+\s', match_result.group(3))
+                        start = []
+                        end =  []
+                        for item in starts_ends:
+                            match_result = match('(\d+)-(\d+)', item)
+                            start.append(int(match_result.group(1)))
+                            end.append(int(match_result.group(2)))
+                    else:
+                        gene_name = ''
+                    if gene_name not in iis_dict.keys():
+                        gene_name = ''
+                    else:
+                        # gene_insertions = 0
+                        gene_length = 0
+                        for i in range(0,len(start)):
[…]

### 2016-06-21 codes/check-biases.R

@@ -1,4 +1,5 @@
 library(stringr)
+library("MASS")
 library(gdata)
 library(mgcv)
 names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t")
@@ -7,87 +8,152 @@ dict = c("Citrobacter", "Escherichia coli ETEC CS17", "Enterobacter", "Escherich
          "Salmonella typhimurium SL1344", "Salmonella typhimurium D23580", "Salmonella typhi")
 names(dict) <- names
 
-biasespath <- c("../results/check-biases/with-ends.txt", "../results/check-biases/without-ends.txt")
-for (item in biasespath)
+biasespath <- "../results/insertion-indices/check-biases/"
+outdir <- "../results/insertion-indices/normalised-insertion-indices"
+list_of_files <- list.files(path=biasespath, full.names=T, recursive=FALSE)
+iitotal = c()
+gctotal = c()
+dtotal = c()
+ii_dnormalisedtotal = c()
+pdf("../results/insertion-indices/biases.pdf")
+for (filename in list_of_files)
 {
-  biasestable = read.table(item, header = FALSE)
-  colnames(biasestable) <- c("name", "ii", "dist", "gc")
-  
-  # pdf(paste("../results/biases-", strsplit(basename(item), "\\.")[[1]][1], ".pdf",sep=''))
-  
+  biasestable = read.table(filename, header = FALSE)
+  colnames(biasestable) <- c("name", "ii", "essentiality", "dist", "gc")
   mar.default <- c(5,4,4,2) + 0.1
   par(mar = mar.default + c(0, 1, 0, 0))
   ii <- c(biasestable$ii)
-  d <- c(biasestable$dist)
-  plot(d, ii, pch = '.', ylim=c(0,3), xlab = "Gene position", ylab = "insertion index",
-       main = "Distance bias", cex.lab = 2, cex.axis = 2, cex.main =2)
-  fit <- gam(ii~s(d))
-  gamprediction <- predict.gam(fit, data.frame(d))
-  lines(loess.smooth(d,ii, span=0.2), col=2, lwd=5)
-  
-  j = 1
[…]

### 2016-06-21 codes/define-core-accessory-hieranoid-ancestralii.py

@@ -1,0 +1,179 @@
+#! /bin/sh
+""":"
+exec python3 $0 ${1+"$@"}
+"""
+
+from os import listdir, path, mkdir
+from Bio import SeqIO
+from shutil import rmtree
+from re import match, search
+from collections import defaultdict
+
+
+class Stack:
+    def __init__(self):
+        self.items = []
+
+    def isEmpty(self):
+        return self.items == []
+
+    def push(self, item):
+        self.items.append(item)
+
+    def pop(self):
+        return self.items.pop()
+
+    def size(self):
+        return len(self.items)
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+def makedir(dirname):
+    if path.exists(dirname):
+        input_var = 'i'
+        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
+            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
+        if input_var == 'Y' or input_var == 'y':
+            rmtree(dirname)
+        else:
+            raise SystemExit
+    mkdir(dirname)
+
+def read_gene_essentiality(indir):
+    list_of_files = listdir(indir)
+    iidict = {}
+    for filename in list_of_files:
+        with open(indir+'/'+filename) as from_file:
+            for line in from_file:
+                cells = line.split()
+                iidict[cells[0]] = cells[1]
+    return iidict
+
+def read_k12(inpath, iidict):
+    with open(inpath) as from_file:
+        for line in from_file:
+            cells = line.split()
+            iidict[cells[0]] = 'essential'
+    return iidict
+
+seqdb = '/home/fatemeh/EnTrI/sequences/fasta-protein/chromosome/seqdb.fasta'
+clusters = '/home/fatemeh/EnTrI/results/hieranoid-result.txt'
+insertion_indices = '/home/fatemeh/EnTrI/results/insertion-indices/gamma'
+# k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
+outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid'
+makedir(outdir)
+speciestreedir = '/home/fatemeh/EnTrI/codes/speciestrees'
+sequences = read_fasta_sequences(seqdb)
[…]

### 2016-07-12 documentation/EnTrI2/manuscript.tex
Computed ancestral insertion indices for nodes on species trees, Ran upsetr on hieranoid outputs, Added SL3261 to the sequences and started rerunning codes, Made a poster for SMBE

@@
 
-\keywords{Keyword1 | Keyword2 | Keyword3} % When adding keywords, separate each term with a straight line: |
+\keywords{essentiality | conservation | transposon insertion} % When adding keywords, separate each term with a straight line: |
 
@@
 % \abbreviations{}
-\abbreviations{SAM, self-assembled monolayer; OTS, octadecyltrichlorosilane}
+%\abbreviations{SAM, self-assembled monolayer; OTS, octadecyltrichlorosilane}
 
@@
 
-\section{Results}
-\textcolor{red}{\{An explanation of the experiment\}}
+\section{Bias}
 
-\textbf{Clustering orthologous and paralogous genes.} To study the essentiality of genes in 12 strains depicted in Fig~\ref{fig:tree}, We needed to cluster sets of orthologous genes in these strain.
-Plenty of methods are proposed for this purpose.
-Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
-We have used Hieranoid to cluster the sets of orthologous genes.
-In addition, we intended to study the essentiality of genes in paralogous genes.
-For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
+\section{Gene classes}
+\subsection{Genus specific}
+\subsection{Single copy}
+\subsection{Multi-copy}
+\section{Evolution of essentiality}
+
+\section{Case study of genes}
+\subsection{Core genes}
+\subsection{Accessory genes}
 
@@
 \begin{materials}
-
+\textbf{Clustering orthologous and paralogous genes.} To study the essentiality of genes in 12 our strains, We needed to cluster sets of orthologous genes in these strain.
+Plenty of methods are proposed for this purpose.
+Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
+We have used Hieranoid to cluster the sets of orthologous genes.
+In addition, we intended to study the essentiality of genes in paralogous genes.
+For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
 \end{materials}
@@
 %----------------------------------------------------------------------------------------
-\begin{figure}
-\includegraphics[scale=0.25]{consensustree.pdf}
-\label{fig:tree}
-\caption{Species tree. How is it made?}
-\end{figure}
 

### 2016-07-12 codes/check-biases-new.py

@@ -23,7 +23,7 @@ for filename in list_of_files:
             for line in sequencefile:
                 if line.startswith('>'):
                     if gene_name != '':
-                        tofile.write('{0}\t{1}\t{2}\t{3}\t{4}\n'.format(gene_name, iis_dict[gene_name][0], iis_dict[gene_name][1], float(start[0])/genome_length[strain_name], float(gc)/gene_length))
+                        tofile.write('{0}\t{1}\t{2}\t{3}\t{4}\t{5}\n'.format(gene_name, iis_dict[gene_name][0], iis_dict[gene_name][1], float(start[0])/genome_length[strain_name], float(gc)/gene_length, float(gene_length)))
                     gc = 0
                     match_result = match('>\s*((\S+?)_+\S+)\s+\[\S+/((\d+\-\d+\s)+)\(', line)
                     if match_result is None:
@@ -33,7 +33,7 @@ for filename in list_of_files:
                         strain_name = match_result.group(2)
                         starts_ends = findall('\d+\-\d+\s', match_result.group(3))
                         start = []
-                        end =  []
+                        end = []
                         for item in starts_ends:
                             match_result = match('(\d+)-(\d+)', item)
                             start.append(int(match_result.group(1)))

### 2016-07-12 codes/define-core-accessory-fdollop.py

@@ -1,218 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path, mkdir
-from Bio import SeqIO
-from shutil import rmtree
-from re import match, findall
-from collections import defaultdict
-
-def read_fasta_sequences(filepath):
-    with open(filepath, 'rU') as fasta_file:
-        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
-    return fasta_dict
-
-def makedir(dirname):
-    if path.exists(dirname):
-        input_var = 'i'
-        while not (input_var == 'y' or input_vawr == 'Y' or input_var == 'n' or input_var == 'N'):
-            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
-        if input_var == 'Y' or input_var == 'y':
-            rmtree(dirname)
-        else:
-            raise SystemExit
-    mkdir(dirname)
-
-def read_gene_essentiality(indir):
-    list_of_files = listdir(indir)
-    iidict = {}
-    for filename in list_of_files:
-        with open(indir+'/'+filename) as from_file:
-            for line in from_file:
-                cells = line.split()
-                iidict[cells[0]] = cells[2]
-    return iidict
-
-def read_k12(inpath, iidict):
-    with open(inpath) as from_file:
-        for line in from_file:
-            cells = line.split()
-            iidict[cells[0]] = 'essential'
-    return iidict
-
-seqdb = '/home/fatemeh/EnTrI/sequences/fasta-protein/chromosome/seqdb.fasta'
-clusters = '/home/fatemeh/EnTrI/results/hieranoid-result.txt'
-insertion_indices = '/home/fatemeh/EnTrI/results/insertion-indices/gamma'
-k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
-outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid'
-makedir(outdir)
-sequences = read_fasta_sequences(seqdb)
-gene_essentiality = read_gene_essentiality(insertion_indices)
-gene_essentiality = read_k12(k12path, gene_essentiality)
-species_names = defaultdict()
-species_names = {"all":["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441", "ROD", "SEN", "SL1344", "STM",
[…]

### 2016-07-12 codes/define-core-accessory-hieranoid-fdollop.py

@@ -84,6 +84,7 @@ for item in species_names.keys():
     gene_dict = {species_names[item][i]: 0 for i in range(num_species)}
     essentiality_dict = {species_names[item][i]: 0 for i in range(num_species)}
     nodeinfo = speciestreedir + '/nodeessentiality.txt'
+    speciestree = speciestreedir + '/' + item + '.tre'
 
     with open (clusters) as from_file:
         for line in from_file:
@@ -91,10 +92,9 @@ for item in species_names.keys():
                 gene_dict[key] = 0
                 essentiality_dict[key] = 0
             list_of_genes = []
-            speciestree = speciestreedir + '/' + item + '.tre'
 
             # cells = line.split()
-            findall_result = findall('(([a-zA-Z0-9]+?)_[a-zA-z0-9]+):', line)
+            findall_result = findall('(([a-zA-Z0-9]+?)_[a-zA-Z0-9]+):', line)
             temp_result = findall('(([a-zA-Z]+?)\d+):', line)
             for matches in temp_result:
                 if matches[1] != 'n':
@@ -108,6 +108,28 @@ for item in species_names.keys():
                         essentiality_dict[name] = 1
 
             if len(species_names[item]) > 1:
+                with open(nodeinfo, 'w') as nodefile:
+                    nodefile.write('     ' + str(len(species_names[item])) + '    1\n')
+                    for key in gene_dict:
+                        if gene_dict[key] == 1:
+                            nodefile.write(key + '          1\n')
+                        else:
+                            nodefile.write(key + '          0\n')
+
+                try:
+                    system(
+                        'fdollop -infile {0} -intreefile {1} -outfile {2} -method d -progress N -treeprint N -ancseq Y'.format(
+                            nodeinfo, speciestree, dollopout))
+                except:
+                    raise SystemExit
+                presence = 0
+                with open(dollopout, 'r') as dollopfile:
+                    for line in dollopfile:
+                        if line.startswith('root'):
+                            cells = line.split()
+                            if cells[2] == 'yes':
+                                presence = 1
+
                 with open(nodeinfo, 'w') as nodefile:
                     nodefile.write('     ' + str(len(species_names[item])) + '    1\n')
                     for key in essentiality_dict:
@@ -128,7 +150,7 @@ for item in species_names.keys():
                             if cells[2] == 'yes':
                                 essentiality = 1
 
-                if gene_dict[min(gene_dict, key=gene_dict.get)] == 1:
+                if presence:
[…]

### 2016-07-12 codes/define-core-accessory-hieranoid.py

@@ -97,7 +97,7 @@ for item in species_names.keys():
             list_of_essential_genes = []
 
             # cells = line.split()
-            findall_result = findall('(([a-zA-Z0-9]+?)_[a-zA-z0-9]+):', line)
+            findall_result = findall('(([a-zA-Z0-9]+?)_[a-zA-Z0-9]+):', line)
             temp_result = findall('(([a-zA-Z]+?)\d+):', line)
             for matches in temp_result:
                 if matches[1] != 'n':

### 2016-07-12 codes/upsetr-hieranoid.R

@@ -1,0 +1,55 @@
+library("UpSetR")
+library(stringr)
+# listinput <- list(a=c(1,1,1,1,2,1,2,3,3,3), b=c(1,4,5), d=c(1,2,5,6,7), e=c(4,8,9))
+# upset(fromList(listinput))
+
+clusters_path <- "../results/hieranoid-result.txt"
+essentiality_path <- "../results/insertion-indices/normalised-insertion-indices/"
+k12_path <- "../results/ecogene-k12.txt"
+list_of_files <- list.files(path=essentiality_path, full.names=T, recursive=FALSE)
+list_of_essential_genes = list()
+for (filename in list_of_files)
+{
+  iis <- as.matrix(read.table(filename))
+  list_of_essential_genes[strsplit(basename(filename),"\\.")[[1]][1]] <- list(iis[,1][iis[,3]=='essential'])
+}
+iis = as.matrix(read.table(k12_path))
+list_of_essential_genes["b"] <- list(iis[,1])
+presence = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
+                "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
+essentiality = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
+                    "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
+cluster <- readLines(clusters_path)
+numclusters = length(cluster)
+for (clusterindex in (1:numclusters))
+{
+  line = cluster[clusterindex]
+  match_result = str_match_all(line, '[[,(]](([[:alnum:]]+)_[[:alnum:]]+):')
+  temp_result = str_match_all(line, '[[,(]](([[:alpha:]]+)[[:digit:]]+):')
+  match_result[[1]] <- rbind(match_result[[1]], temp_result[[1]])
+  match_result <- match_result[[1]]
+  for (i in (1:nrow(match_result)))
+  {
+    name = match_result[i,3]
+    if (name %in% names(presence)) 
+    {
+      presence[[name]] <- rbind(presence[[name]], clusterindex)
+      gene = match_result[i,2]
+      if (gene %in% list_of_essential_genes[[name]])
+      {
[…]

### 2016-07-15 documents/manuscript.tex
Added SL3261, reorganised everything

--- 2016-07-12 documentation/EnTrI2/manuscript.tex
@@
 
+\usepackage{float}
 \usepackage{graphicx}
-\graphicspath{{figs/}}
+\usepackage{caption}
+\graphicspath{{../figures/}}
 %------------------------------------------------
@@
 
-\title{Title of the publication} % For titles, only capitalize the first letter
+\title{Is essentiality of genes conserved in Enterobacteriaceae?} % For titles, only capitalize the first letter
 
@@
 
-\author{John Smith\affil{1}{University of California},
-James Smith\affil{2}{University of Oregon}
+\author{Fatemeh Ashari-Ghomi\affil{1}{University of Canterbury},
+Paul P. Gardner\affil{1}{}
 \and
-Jane Smith\affil{1}{}}
+Lars Barquist\affil{2}{Wurzburg University}}
 
@@
 Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges.
-We extend this research by studying 12 bacterial strains.
-These include \textit{Salmonella} enterica subsp.\@ enterica serovar Typhi str.\@ Ty2, \textit{Salmonella} enterica subsp.\@ enterica serovar Enteritidis str.\@ P125109, \textit{Salmonella} enterica subsp.\@ enterica serovar Typhimurium str.\@ SL1344, \textit{Salmonella} enterica subsp.\@ enterica serovar Typhimurium str.\@ A130, \textit{Salmonella} enterica subsp.\@ enterica serovar Typhimurium str.\@ D23580, \textit{Escherichia} coli UPEC ST131, \textit{Escherichia} coli ETEC CS17, \textit{Escherichia} coli ETEC H10407, \textit{Citrobacter} rodentium ICC168, \textit{Klebsiella} pneumoniae RH201207, \textit{Klebsiella} pneumoniae subsp.\@ pneumoniae Ecl8, and \textit{Enterobacter} cloacae subsp.\@ cloacae NCTC 9394.
-All these strains are selected from Enterobacteriaceae family.
+We extend this research by studying 13 bacterial strains from Enterobacteriaceae.
+These strains are depicted in Fig.\@~\ref{fig:species-tree}.
 
@@
 
-\section{Bias}
-
+\section{Are transposons biased towards certain positions?}
+\begin{itemize}
+\item There is a bias towards the position of the gene (Fig.~\ref{fig:distance-bias}).
+\item There is a negligible bias towards certain motifs (Fig.~\ref{fig:logos})
+\item The G-C bias needs to be studied further (Fig.~\ref{fig:GC-bias})
+\item The number of insertions on the 3’ and 5’ ends is more than the internal region in essential genes and less than the internal region in beneficial losses.
+(Fig.~\ref{fig:insertion-position-bias})
+\end{itemize}
 \section{Gene classes}
@@
 \section{Evolution of essentiality}
-
+\begin{itemize}
+\item UpSetR results (Fig.~\ref{fig:upsetr})
+\item Stringent
+\item Dollo law
+\item Ancestral insertion index
+\end{itemize}
 \section{Case study of genes}
@@
 %----------------------------------------------------------------------------------------
+\begin{figure}
+\includegraphics[scale=0.3]{phylosift-aa-raxmlbootstrap.pdf}
+\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in Keio collection \cite{baba_construction_2006}.
+We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
+\label{fig:species-tree}
+\end{figure}
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=2, scale=0.25]{biases.pdf}&
+\includegraphics[page=5, scale=0.25]{biases.pdf}&
+\includegraphics[page=8, scale=0.25]{biases.pdf}\\
+\includegraphics[page=11, scale=0.25]{biases.pdf}&
+\includegraphics[page=14, scale=0.25]{biases.pdf}&
+\includegraphics[page=17, scale=0.25]{biases.pdf}\\
+\includegraphics[page=20, scale=0.25]{biases.pdf}&
+\includegraphics[page=23, scale=0.25]{biases.pdf}&
+\includegraphics[page=26, scale=0.25]{biases.pdf}\\
+\includegraphics[page=29, scale=0.25]{biases.pdf}&
+\includegraphics[page=32, scale=0.25]{biases.pdf}&
+\includegraphics[page=35, scale=0.25]{biases.pdf}\\
+&\includegraphics[page=38, scale=0.25]{biases.pdf}&\\
+\end{tabular}
+\caption{The plots show the position of the genes within the genome (normalised by the lengths of the genomes) versus the insertion indices of the genes}
+\label{fig:distance-bias}
+\end{figure*}
+
+\begin{figure*}
+\begin{tabular}{c c}
+\includegraphics[scale=0.6]{100logo-bits.pdf}&
+\includegraphics[scale=0.6]{100logo-prob.pdf}
+\end{tabular}
+\caption{We have generated the logos from 10 nucleotides flanking the 100 top most frequent insertion sites.}
+\label{fig:logos}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=1, scale=0.25]{biases.pdf}&
+\includegraphics[page=4, scale=0.25]{biases.pdf}&
+\includegraphics[page=7, scale=0.25]{biases.pdf}\\
+\includegraphics[page=10, scale=0.25]{biases.pdf}&
+\includegraphics[page=13, scale=0.25]{biases.pdf}&
+\includegraphics[page=16, scale=0.25]{biases.pdf}\\
+\includegraphics[page=19, scale=0.25]{biases.pdf}&
+\includegraphics[page=22, scale=0.25]{biases.pdf}&
+\includegraphics[page=25, scale=0.25]{biases.pdf}\\
+\includegraphics[page=28, scale=0.25]{biases.pdf}&
+\includegraphics[page=31, scale=0.25]{biases.pdf}&
+\includegraphics[page=34, scale=0.25]{biases.pdf}\\
+&\includegraphics[page=37, scale=0.25]{biases.pdf}&\\
+\end{tabular}
+\caption{The plots show the G-C contents of the genes (normalised by the lengths of the genes) against their insertion indices}
+\label{fig:GC-bias}
+\end{figure*}
+
+\begin{figure*}
+\begin{tabular}{c c}
+\includegraphics[scale=0.5, page=1]{insertion-position-bias.pdf}&
+\includegraphics[scale=0.5, page=2]{insertion-position-bias.pdf}\\
+\includegraphics[scale=0.5, page=3]{insertion-position-bias.pdf}&
+\includegraphics[scale=0.5, page=4]{insertion-position-bias.pdf}
+\end{tabular}
+\caption{We have divided our genes into 3 segments: 5\% of
+the genes on the 5’ end, 20\% of the genes on the 3’ end, and the rest in the middle.}
+\label{fig:insertion-position-bias}
+\end{figure*}
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.65, page=1]{upsetr.pdf}\\
+\includegraphics[scale=0.65, page=2]{upsetr.pdf}
+\caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.}
+\label{fig:upsetr}
+\end{figure}
 

### 2016-07-15 bin/calculate-insertion-index.R

@@ -1,0 +1,180 @@
+library(stringr)
+library("MASS")
+# library(mixtools)
+fastas_dir <- "~/EnTrI/data/fasta-protein/chromosome"
+plots_dir <- "~/EnTrI/data/plot-files/chromosome"
+output_dir1 <- "~/EnTrI/results/insertion-indices/insfree/"
+output_dir2 <- "~/EnTrI/results/insertion-indices/gamma/"
+list_of_files <- list.files(path=plots_dir, full.names=T, recursive=FALSE)
+plots = list()
+sumlength = list()
+for (filename in list_of_files)
+{
+  plotfile = as.matrix(read.table(filename, as.is=TRUE))
+  locusid = strsplit(basename(filename),"\\.")[[1]][1]
+  plots[[locusid]] = plotfile[,1] + plotfile[,2]
+  sumlength[[locusid]] = c(sum(plots[[locusid]]>0), length(plots[[locusid]]))
+}
+
+list_of_files <- list.files(path=fastas_dir, full.names=T, recursive=FALSE)
+list_of_files <- list_of_files[ !grepl("/home/fatemeh/EnTrI/data/fasta-protein/chromosome/U00096.fasta",list_of_files) ]
+list_of_files <- list_of_files[ !grepl("/home/fatemeh/EnTrI/data/fasta-protein/chromosome/seqdb.fasta",list_of_files) ]
+dir.create(output_dir1)
+dir.create(output_dir2)
+for (filename in list_of_files)
+{
+  iitable = c()
+  fastafile = readLines(filename)
+  for (line in fastafile)
+  {
+    if (startsWith(line, ">"))
+    {
+      matchresult = str_match(line, ">([[:graph:]]+)[[:blank:]]\\[[[:graph:]]+\\/([[:digit:]]+)\\-([[:digit:]]+)[[:blank:]]\\(([[:alpha:]]+)\\)")
+      if (!(is.na(matchresult[5])))
+      {
+        locustag = matchresult[2]
+        start = as.numeric(matchresult[3])
+        end = as.numeric(matchresult[4])
+        direction = matchresult[5]
+        locusid = str_match(locustag, "([[:graph:]]+)\\_+[[:alnum:]]+")[2]
+        if (is.na(locusid))
+        {
+          locusid = str_match(locustag, "([[:alpha:]]+)[[:digit:]]+")[2]
+        }
[…]

### 2016-07-15 bin/define-core-accessory-hieranoid-ancestralii.py

@@ -1,0 +1,297 @@
+#! /bin/sh
+""":"
+exec python3 $0 ${1+"$@"}
+"""
+
+from os import listdir, path, mkdir
+from Bio import SeqIO
+from shutil import rmtree
+from re import match, findall
+from collections import defaultdict
+
+
+class Stack:
+    def __init__(self):
+        self.items = []
+
+    def isEmpty(self):
+        return self.items == []
+
+    def push(self, item):
+        self.items.append(item)
+
+    def pop(self):
+        return self.items.pop()
+
+    def size(self):
+        return len(self.items)
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+def makedir(dirname):
+    if path.exists(dirname):
+        input_var = 'i'
+        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
+            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
+        if input_var == 'Y' or input_var == 'y':
+            rmtree(dirname)
+        else:
+            raise SystemExit
+    mkdir(dirname)
+
+def read_gene_essentiality(indir):
+    list_of_files = listdir(indir)
+    iidict = {}
+    for filename in list_of_files:
+        with open(indir+'/'+filename) as from_file:
+            for line in from_file:
+                cells = line.split()
+                iidict[cells[0]] = float(cells[1])
+    return iidict
+
+# def read_k12(inpath, iidict):
+#     with open(inpath) as from_file:
+#         for line in from_file:
+#             cells = line.split()
+#             iidict[cells[0]] = 'essential'
+#     return iidict
+
+seqdb = '/home/fatemeh/EnTrI/data/fasta-protein/chromosome/seqdb.fasta'
+clusters = '/home/fatemeh/EnTrI/results/hieranoid/hieranoid-result.txt'
+insertion_indices = '/home/fatemeh/EnTrI/results/insertion-indices/gamma'
+# k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
+outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid-ancestralii'
+makedir(outdir)
[…]

### 2016-07-15 bin/define-core-accessory-hieranoid-fdollop.py

@@ -1,0 +1,235 @@
+#! /bin/sh
+""":"
+exec python3 $0 ${1+"$@"}
+"""
+
+from os import listdir, path, mkdir, system
+from Bio import SeqIO
+from shutil import rmtree
+from re import match, findall
+from collections import defaultdict
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+def makedir(dirname):
+    if path.exists(dirname):
+        input_var = 'i'
+        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
+            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
+        if input_var == 'Y' or input_var == 'y':
+            rmtree(dirname)
+        else:
+            raise SystemExit
+    mkdir(dirname)
+
+def read_gene_essentiality(indir):
+    list_of_files = listdir(indir)
+    iidict = {}
+    for filename in list_of_files:
+        with open(indir+'/'+filename) as from_file:
+            for line in from_file:
+                cells = line.split()
+                iidict[cells[0]] = cells[2]
+    return iidict
+
+def read_k12(inpath, iidict):
+    with open(inpath) as from_file:
+        for line in from_file:
+            cells = line.split()
+            iidict[cells[0]] = 'essential'
+    return iidict
+
+seqdb = '/home/fatemeh/EnTrI/data/fasta-protein/chromosome/seqdb.fasta'
+clusters = '/home/fatemeh/EnTrI/results/hieranoid/hieranoid-result.txt'
+insertion_indices = '/home/fatemeh/EnTrI/results/insertion-indices/normalised-insertion-indices'
+k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
+outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid-fdollop'
+speciestreedir = '/home/fatemeh/EnTrI/bin/speciestrees'
+dollopout = '/home/fatemeh/EnTrI/bin/speciestrees/dollop.out'
+makedir(outdir)
+sequences = read_fasta_sequences(seqdb)
+gene_essentiality = read_gene_essentiality(insertion_indices)
+gene_essentiality = read_k12(k12path, gene_essentiality)
[…]

### 2016-07-15 bin/insertion-position-bias.R

@@ -1,0 +1,38 @@
+insertion_positions <- read.table("../results/insertion-indices/insertion-position-bias.out", row.names=1)
+pdf("../figures/insertion-position-bias.pdf")
+toplot <- colMeans(insertion_positions[,1:ncol(insertion_positions)-1])
+cols = c(rep("khaki4",5), rep("slategray",75), rep("palevioletred4",20))
+midpoints <- barplot(toplot, main="All", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2, col=cols, border=NA)
+axis(1, at=midpoints[c(1,100)], labels=c('5\'','3\''), cex.axis=1.5)
+legend(41,0.7, c("First 5%","internal", "Last 20%"), lty=c(1,1,1),cex=1.5, lwd=c(4,4, 4),col=c(cols[1],cols[6],cols[81]), bg="white")
+essential = c()
+nonessential = c()
+beneficialloss = c()
+for (i in seq(1,nrow(insertion_positions)))
+{
+  if (insertion_positions[i,ncol(insertion_positions)] < 0.2)
+  {
+    essential = rbind(essential, insertion_positions[i,1:ncol(insertion_positions)-1]) 
+  }
+  else if (insertion_positions[i,ncol(insertion_positions)] < 2)
+  {
+    nonessential = rbind(nonessential, insertion_positions[i,1:ncol(insertion_positions)-1])
+  }
+  else
+  {
+    beneficialloss = rbind(beneficialloss, insertion_positions[i,1:ncol(insertion_positions)-1]) 
+  }
+}
+toplot <- colMeans(essential)
+midpoints <- barplot(toplot, main="Essential", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2, col=cols, border=NA)
+axis(1, at=midpoints[c(1,100)], labels=c('5\'','3\''), cex.axis=1.5)
+legend(41,0.3, c("First 5%","internal", "Last 20%"), lty=c(1,1,1),cex=1.5, lwd=c(4,4, 4),col=c(cols[1],cols[6],cols[81]), bg="white")
+toplot <- colMeans(nonessential)
[…]

### 2016-07-15 bin/make-logos-top-100.py

@@ -1,0 +1,79 @@
+#! /bin/sh
+""":"
+exec python3 $0 ${1+"$@"}
+"""
+
+from os import listdir, path, remove
+import numpy as np
+
+genomespath = '../data/fasta-genome/chromosome'
+plotspath = '../data/plot-files/chromosome'
+genomextension = 'fa'
+logospath = '../results/logos/100logos.txt'
+complement = {'a': 't', 'c': 'g', 'g': 'c', 't': 'a', 'u': 'a', 'm': 'k', 'k': 'm', 'r': 'y',
+              'y': 'r', 'w': 'w', 's': 's', 'v': 'b', 'b': 'v', 'h': 'd', 'd': 'h', 'n': 'n'}
+
+if path.isfile(logospath):
+    input_var = 'i'
+    while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
+        input_var = input('File {0} already exists. Replace? [Y/n] '.format(logospath))
+        if input_var == 'Y' or input_var == 'y':
+            remove(logospath)
+        else:
+            raise SystemExit
+
+plots_f = []
+plots_b = []
+sequences = ''
+list_of_files = listdir(plotspath)
+for filename in list_of_files:
+    name, extension = path.splitext(filename)
+    with open('{0}/{1}'.format(plotspath, filename)) as plotfile:
+        for line in plotfile:
+            cells = line.split()
+            plots_f.append(int(cells[0]))
+            plots_b.append(int(cells[1]))
+    with open('{0}/{1}.{2}'.format(genomespath, name, genomextension)) as genomefile:
+        for line in genomefile:
+            if not line.startswith('>'):
+                if line.endswith('\n'):
+                    line = line[:-1]
+                sequences += line
+plots = np.array(plots_f + plots_b)
+top_indices = np.argpartition(plots, -100)[-100:] # returns the indices of 100 maximums in plots
+with open (logospath, 'a') as logosfile:
+        # for i in range(0, len(plots_f)):
+        #     if plots_f[i] > 0 or plots_b[i] > 0:
+        #         if 10 <= i < len(sequences)-10:
+        #             seq_to_write = sequences[i-10:i+11]
+        #         elif i < 10:
+        #             seq_to_write = sequences[(i-10)%len(sequences):len(sequences)]+sequences[0:i+11]
+        #         else:
+        #             seq_to_write = sequences[i-10:len(sequences)]+sequences[0:(i+11)%len(sequences)]
+        #         if plots_f[i] > 0:
[…]

### 2016-07-15 bin/upsetr-hieranoid.R

@@ -1,0 +1,61 @@
+library("UpSetR")
+library(stringr)
+# listinput <- list(a=c(1,1,1,1,2,1,2,3,3,3), b=c(1,4,5), d=c(1,2,5,6,7), e=c(4,8,9))
+# upset(fromList(listinput))
+
+clusters_path <- "../results/hieranoid/hieranoid-result.txt"
+essentiality_path <- "../results/insertion-indices/normalised-insertion-indices/"
+k12_path <- "../results/ecogene-k12.txt"
+list_of_files <- list.files(path=essentiality_path, full.names=T, recursive=FALSE)
+list_of_essential_genes = list()
+for (filename in list_of_files)
+{
+  iis <- as.matrix(read.table(filename))
+  list_of_essential_genes[strsplit(basename(filename),"\\.")[[1]][1]] <- list(iis[,1][iis[,3]=='essential'])
+}
+iis = as.matrix(read.table(k12_path))
+list_of_essential_genes["b"] <- list(iis[,1])
+presence = rev(list("BN373"=c(), "ERS227112"=c(), "ENC"=c(), "ROD"=c(), "SL1344"=c(), "SL3261"=c(), "STM"=c(), "STMMW"=c(), "SEN"=c(),
+                "t"=c(), "NCTC13441"=c(), "CS17"=c(), "ETEC"=c(), "b"=c()))
+essentiality = rev(list("BN373"=c(), "ERS227112"=c(), "ENC"=c(), "ROD"=c(), "SL1344"=c(), "SL3261"=c(), "STM"=c(), "STMMW"=c(), "SEN"=c(),
+                    "t"=c(), "NCTC13441"=c(), "CS17"=c(), "ETEC"=c(), "b"=c()))
+cluster <- readLines(clusters_path)
+numclusters = length(cluster)
+for (clusterindex in (1:numclusters))
+{
+  line = cluster[clusterindex]
+  match_result = str_match_all(line, '[[,(]](([[:alnum:]]+)_[[:alnum:]]+):')
+  temp_result = str_match_all(line, '[[,(]](([[:alpha:]]+)[[:digit:]]+):')
+  match_result[[1]] <- rbind(match_result[[1]], temp_result[[1]])
+  match_result <- match_result[[1]]
+  for (i in (1:nrow(match_result)))
+  {
+    name = match_result[i,3]
+    if (name %in% names(presence)) 
+    {
+      presence[[name]] <- rbind(presence[[name]], clusterindex)
[…]

### 2016-07-15 codes/ancestral-insertion-index.py

@@ -1,173 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path, mkdir
-from Bio import SeqIO
-from shutil import rmtree
-from re import match
-from collections import defaultdict
-
-def read_fasta_sequences(filepath):
-    with open(filepath, 'rU') as fasta_file:
-        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
-    return fasta_dict
-
-def makedir(dirname):
-    if path.exists(dirname):
-        input_var = 'i'
-        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
-            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
-        if input_var == 'Y' or input_var == 'y':
-            rmtree(dirname)
-        else:
-            raise SystemExit
-    mkdir(dirname)
-
-seqdb = '/home/fatemeh/EnTrI/sequences/fasta-protein/chromosome/seqdb.fasta'
-clusters = '/home/fatemeh/EnTrI/results/single-copy-genes'
-outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-singcopy-ancestralii'
-makedir(outdir)
-sequences = read_fasta_sequences(seqdb)
-list_of_files = listdir(clusters)
-species_names = defaultdict()
-species_names = {"all":["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441", "ROD", "SEN", "SL1344", "STM",
-    "STMMW", "t", "b"],"typhimurium":["STM", "SL1344", "STMMW"], "salmonella":["SEN", "SL1344", "STM", "STMMW", "t"],
-    "ecoli":["CS17", "ETEC", "NCTC13441", "b"], "klebsiella":["ERS227112", "BN373"], "citrobacter":["ROD"], "enterobacter":
-    ["ENC"], "salmonellaecoli":["SEN", "SL1344", "STM", "STMMW", "t", "CS17", "ETEC", "NCTC13441", "b"],
-    "salmonellaecolicitrobacter":["SEN", "SL1344", "STM", "STMMW", "t", "CS17", "ETEC", "NCTC13441", "b", "ROD"],
[…]

### 2016-07-15 codes/check-biases.R

@@ -1,173 +1,0 @@
-library(stringr)
-library("MASS")
-library(gdata)
-library(mgcv)
-names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t")
-dict = c("Citrobacter", "Escherichia coli ETEC CS17", "Enterobacter", "Escherichia coli ETEC H10407", "Escherichia coli UPEC",
-         "Klebsiella pneumoniae RH201207", "Klebsiella pneumoniae Ecl8", "Salmonella enteritidis", "Salmonella typhimurium A130",
-         "Salmonella typhimurium SL1344", "Salmonella typhimurium D23580", "Salmonella typhi")
-names(dict) <- names
-
-biasespath <- "../results/insertion-indices/check-biases/"
-outdir <- "../results/insertion-indices/normalised-insertion-indices"
-list_of_files <- list.files(path=biasespath, full.names=T, recursive=FALSE)
-iitotal = c()
-gctotal = c()
-dtotal = c()
-ii_dnormalisedtotal = c()
-# lengths=c()
-pdf("../results/insertion-indices/biases.pdf")
-for (filename in list_of_files)
-{
-  biasestable = read.table(filename, header = FALSE)
-  colnames(biasestable) <- c("name", "ii", "essentiality", "dist", "gc", "length")
-  # for (i in seq(1, nrow(biasestable), 1))
-  # {
-  #   if (biasestable$"gc"[i] < 0.45)
-  #   {
-  #     lengths[as.character(biasestable$"name"[i])] = biasestable$"length"[i]
-  #   }
-  # }
-  mar.default <- c(5,4,4,2) + 0.1
-  par(mar = mar.default + c(0, 1, 0, 0))
-  ii <- c(biasestable$ii)
-  iitotal <- c(iitotal, ii)
-  gc <- c(biasestable$gc)
-  gctotal <- c(gctotal , gc)
-  d <- c(biasestable$dist)
-  dtotal <- c(dtotal , d)
-  
-  plot(gc, ii, pch = '.', ylim=c(0,5), xlab = "GC content", ylab = "insertion index",
[…]

### 2016-07-15 codes/check-biases.py

@@ -1,106 +1,0 @@
-from collections import defaultdict
-from re import match, findall
-from os import listdir, path
-
-seqdb = '../sequences/fasta-dna/chromosome/seqdb.fasta'
-plots = '../sequences/plot-files/chromosome'
-resultwithends = '../results/check-biases/with-ends.txt'
-resultwithoutends = '../results/check-biases/without-ends.txt'
-genome_length = {"SL1344":4878012, "STMMW":4879400, "SEN":4685848, "t":4791961, "STM":4895639, "ETEC":5153435,
-                 "b":4641652, "CS17":4994793, "NCTC13441":5174631, "ROD":5346659, "BN373":5324709, "ERS227112":5869288,
-                 "ENC":4908759}
-
-list_of_files = listdir(plots)
-plots_dict = defaultdict(list)
-for filename in list_of_files:
-    name, extension = path.splitext(filename)
-    with open('{0}/{1}'.format(plots, filename), 'r') as plotfile:
-        for line in plotfile:
-            cells = line.split()
-            plots_dict[name].append(int(cells[0]) + int(cells[1]))
-
-genome_insertions = dict()
-for item in plots_dict.keys():
-    genome_insertions[item] = sum([1 for x in plots_dict[item] if x > 0])
-
-gene_name = ''
-with open(resultwithends, 'w') as tofile:
-    with open(seqdb, 'r') as sequencefile:
-        for line in sequencefile:
-            if line.startswith('>'):
-                if gene_name != '':
-                    tofile.write('{0}\t{1}\t{2}\t{3}\n'.format(gene_name, ii, float(start[0])/genome_length[strain_name], float(gc)/gene_length))
-                gc = 0
-                match_result = match('>\s*((\S+?)_+\S+)\s+\[\S+/((\d+\-\d+\s)+)\(', line)
-                if match_result is None:
-                    match_result = match('>\s*(([a-zA-Z]+)\d+)\s+\[\S+/((\d+\-\d+\s)+)\(', line)
-                if match_result is not None:
-                    gene_name = match_result.group(1)
-                    strain_name = match_result.group(2)
-                    starts_ends = findall('\d+\-\d+\s', match_result.group(3))
-                    start = []
-                    end =  []
-                    for item in starts_ends:
-                        match_result = match('(\d+)-(\d+)', item)
-                        start.append(int(match_result.group(1)))
-                        end.append(int(match_result.group(2)))
-                else:
-                    strain_name = ''
[…]

### 2016-07-15 codes/clust2plot.R

@@ -1,184 +1,0 @@
-library(stringr)
-# args <- commandArgs(trailingOnly = TRUE)
-# clusters <- args[1]
-clusters_path <- "../results/merge-clust-plot"
-#clusters_path <- c("../results/merge-clust-plot", "../results/merge-clust-plot-without-ends/")
-for (cpitem in clusters_path)
-{
-  list_of_files <- list.files(path=cpitem, full.names=T, recursive=FALSE)
-  names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t", "b")
-  genuses = c(1, 2, 3, 2, 2, 3, 3, 4, 4, 4, 4, 4, 2)
-  names(genuses) <- names
-  numspecies = length(names)
-  file_II = list()
-  file_size = list()
-  file_group = list()
-  for (filename in list_of_files)
-  {
-    clustspecies = c()
-    cluster <- as.matrix(read.table(filename))
-    i_sum = 0
-    l_sum = 0
-    cluster_size = nrow(cluster)
-    clust_with_ii_size = 0
-    for (i in (1:cluster_size))
-    {
-      if (as.numeric(cluster[i,5]) >= 0)
-      {
-        match = str_match(cluster[i,2], "([[:graph:]]+)\\_[[:alnum:]]+")[2]
-        if (is.na(match))
-        {
-          match = str_match(cluster[i,2], "([[:alpha:]]+)[[:digit:]]+")[2]
-        }
-        if (match %in% names)
-        {
-          clustspecies = c(clustspecies, match) 
-        }
-        i_sum = i_sum + as.numeric(cluster[i, 5])
-        l_sum = l_sum + (as.numeric(cluster[i, 4]) - as.numeric(cluster[i, 3]) + 1) * 3
-        clust_with_ii_size = clust_with_ii_size  + 1
-      }
-    }
-    if (l_sum > clust_with_ii_size * 60)
-    {
-      file_II[basename(filename)] = i_sum / clust_with_ii_size
-      file_size[basename(filename)] = cluster_size
-      unique_clustspecies = unique(clustspecies)
-      one_or_more = length(unique_clustspecies)
[…]

### 2016-07-15 codes/cluster-check.py

@@ -1,60 +1,0 @@
-from Bio import SeqIO
-from itertools import combinations
-from re import search, findall
-from os import listdir
-from collections import defaultdict
-
-def read_fasta_sequences(filepath):
-    with open(filepath, 'rU') as fasta_file:
-        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
-    return fasta_dict
-
-def read_clusters(input_dir):
-    sequence_occurrence = defaultdict(list)
-    list_of_files = listdir(input_dir)
-    for filename in list_of_files:
-        with open('{0}/{1}'.format(input_dir, filename)) as cluster_file:
-            for line in cluster_file:
-                cells = line.split()
-                sequence_occurrence[cells[1]].append(filename)
-    return sequence_occurrence
-
-seqdb = '../sequences/fasta-dna/chromosome/seqdb.fasta'
-clustdb = '../results/homclust/EFam-clusters'
-ec_clusters = read_fasta_sequences(seqdb)
-efam_clusters = read_clusters(clustdb)
-tp, fn, fp, tn = 0, 0, 0, 0
-for key1, key2 in combinations(ec_clusters.keys(), r=2):
-    search_results1 = search('\[EC_number=([^\]]*)\]', ec_clusters[key1].description)
-    find_results1 = findall('(\d+\.\d+\.\d+)\.\d+', search_results1.group(1))
-    search_results2 = search('\[EC_number=([^\]]*)\]', ec_clusters[key2].description)
-    find_results2 = findall('(\d+\.\d+\.\d+)\.\d+', search_results2.group(1))
-    if find_results1 and find_results2:
-        ec_match = 0
-        for item1 in find_results1:
-            for item2 in find_results2:
-                if item1 == item2:
-                    ec_match = 1
-                    break
-            if ec_match:
-                break
-        efam_match = 0
-        for item1 in efam_clusters[key1]:
-            for item2 in efam_clusters[key2]:
-                if item1 == item2:
-                    efam_match = 1
-                    break
-            if efam_match:
-                break
-        if not ec_match and not efam_match:
-            tn += 1
-        elif not ec_match and efam_match:
-            fp += 1
-            print '{}\t{}\t{}\t{}\t{}\t{}'.format(key1, find_results1, efam_clusters[key1], key2, find_results2, efam_clusters[key2])
-        elif ec_match and not efam_match:
-            fn += 1
[…]

### 2016-07-15 codes/cluster-check2.py

@@ -1,87 +1,0 @@
-from Bio import SeqIO
-from itertools import combinations
-from re import search, findall, match
-from os import listdir
-from collections import defaultdict
-
-def read_EC_files(filepath):
-    ec_dict = defaultdict(list)
-    locus_tags = {'t', 'ROD', 'ETEC', 'b', 'SL1344'}
-    list_of_files = listdir(filepath)
-    for filename in list_of_files:
-        with open(filepath+filename) as ecfile:
-            for line in ecfile:
-                if line.startswith('ID'):
-                    ec = []
-                elif line.startswith('DE'):
-                    ec += findall('\d+\.\d+\.\d+\.\d+', line)
-                elif line.startswith('GN'):
-                    search_results = search('OrderedLocusNames=([^;\{]*)', line)
-                    if search_results:
-                        find_results = findall('\w+', search_results.group(1))
-                        for item1 in find_results:
-                            for item2 in locus_tags:
-                                if item1.startswith(item2):
-                                    break
-                                else:
-                                    continue
-                            break
-                        if item1.startswith(item2):
-                            ec_dict[item1] = ec
-                            ec = []
-    return ec_dict
-
-
-def read_clusters(input_dir):
-    sequence_occurrence = defaultdict(list)
-    list_of_files = listdir(input_dir)
-    for filename in list_of_files:
-        with open('{0}/{1}'.format(input_dir, filename)) as cluster_file:
-            for line in cluster_file:
-                cells = line.split()
-                sequence_occurrence[cells[1]].append(filename)
-    return sequence_occurrence
-
-ECpath = '../sequences/EC/'
-clustdb = '../results/homclust/EFam-clusters'
-ec_clusters = read_EC_files(ECpath)
-efam_clusters = read_clusters(clustdb)
-tp, fn, fp, tn = 0, 0, 0, 0
-for key1, key2 in combinations(ec_clusters.keys(), r=2):
-    if ec_clusters[key1] and ec_clusters[key2]:
-        ecs1= []
-        ecs2 = []
-        for item in ec_clusters[key1]:
-            match_results = match('\d+\.\d+\.\d+', item)
-            ecs1.append(match_results.group(0))
-        for item in ec_clusters[key2]:
-            match_results = match('\d+\.\d+\.\d+', item)
-            ecs2.append(match_results.group(0))
-        ec_match = 0
-        for item1 in ecs1:
-            for item2 in ecs2:
[…]

### 2016-07-15 codes/cluster2dnafasta.py

@@ -1,96 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from sys import argv
-from os import path, mkdir, system, listdir, remove, getcwd
-from shutil import rmtree, copyfile, move
-from Bio import SeqIO
-from collections import defaultdict
-from re import search
-from argparse import ArgumentParser
-
-def check_directory_name(dirname):
-    if dirname.endswith('/'):
-        dirname = dirname[:-1]
-    return dirname
-
-def read_fasta_sequences(filepath):
-    with open(filepath, 'rU') as fasta_file:
-        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
-    return fasta_dict
-
-def makedir(dirname):
-    # print ('Making directory \'{0}\'..'.format(dirname))
-    if path.exists(dirname):
-        input_var = 'i'
-        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
-            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
-        if input_var == 'Y' or input_var == 'y':
-            rmtree(dirname)
-        else:
-            raise SystemExit
-    mkdir(dirname)
-
-parser = ArgumentParser(description='Get DNA fasta files of the clusters made by homclust')
-#parser.add_argument('dupthr', help='Only clusters with more number of members than this threshold are considered')
-parser.add_argument('seqdb', help='Fasta file for DNA sequences in all species')
-parser.add_argument('cldir', help='Directory of EFam clusters resulted from homclust')
-#parser.add_argument('msadir', help='Directory of EFam MSAs resulted from homclust')
-parser.add_argument('outdir', help='Directory for outputs')
-args = parser.parse_args()
-#dup_thresh = int(args.dupthr)
-seqdb = args.seqdb
-cldir = args.cldir
-outdir = args.outdir
-#msadir = args.msadir
-cldir = check_directory_name(cldir)
-#msa_indir = check_directory_name(msadir)
-outdir = check_directory_name(outdir)
-makedir(outdir)
-fasta_outdir = outdir
-#fasta_outdir = '{}/fasta'.format(outdir)
-#makedir(fasta_outdir)
-#msa_outdir = '{}/msa'.format(outdir)
-#makedir('{}/msa'.format(outdir))
-
-fastas = read_fasta_sequences(seqdb)
[…]

### 2016-07-15 codes/define-core-accessory-hieranoid-ancestralii.py

@@ -1,297 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path, mkdir
-from Bio import SeqIO
-from shutil import rmtree
-from re import match, findall
-from collections import defaultdict
-
-
-class Stack:
-    def __init__(self):
-        self.items = []
-
-    def isEmpty(self):
-        return self.items == []
-
-    def push(self, item):
-        self.items.append(item)
-
-    def pop(self):
-        return self.items.pop()
-
-    def size(self):
-        return len(self.items)
-
-def read_fasta_sequences(filepath):
-    with open(filepath, 'rU') as fasta_file:
-        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
-    return fasta_dict
-
-def makedir(dirname):
-    if path.exists(dirname):
-        input_var = 'i'
-        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
-            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
-        if input_var == 'Y' or input_var == 'y':
-            rmtree(dirname)
-        else:
-            raise SystemExit
-    mkdir(dirname)
-
-def read_gene_essentiality(indir):
-    list_of_files = listdir(indir)
-    iidict = {}
-    for filename in list_of_files:
-        with open(indir+'/'+filename) as from_file:
-            for line in from_file:
-                cells = line.split()
-                iidict[cells[0]] = float(cells[1])
-    return iidict
-
-# def read_k12(inpath, iidict):
-#     with open(inpath) as from_file:
-#         for line in from_file:
-#             cells = line.split()
-#             iidict[cells[0]] = 'essential'
-#     return iidict
-
-seqdb = '/home/fatemeh/EnTrI/sequences/fasta-protein/chromosome/seqdb.fasta'
-clusters = '/home/fatemeh/EnTrI/results/hieranoid-result.txt'
-insertion_indices = '/home/fatemeh/EnTrI/results/insertion-indices/gamma'
-# k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
-outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid-ancestralii'
-makedir(outdir)
[…]

### 2016-07-15 codes/define-core-accessory-hieranoid.py

@@ -1,218 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path, mkdir
-from Bio import SeqIO
-from shutil import rmtree
-from re import match, findall
-from collections import defaultdict
-
-def read_fasta_sequences(filepath):
-    with open(filepath, 'rU') as fasta_file:
-        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
-    return fasta_dict
-
-def makedir(dirname):
-    if path.exists(dirname):
-        input_var = 'i'
-        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
-            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
-        if input_var == 'Y' or input_var == 'y':
-            rmtree(dirname)
-        else:
-            raise SystemExit
-    mkdir(dirname)
-
-def read_gene_essentiality(indir):
-    list_of_files = listdir(indir)
-    iidict = {}
-    for filename in list_of_files:
-        with open(indir+'/'+filename) as from_file:
-            for line in from_file:
-                cells = line.split()
-                iidict[cells[0]] = cells[2]
-    return iidict
-
-def read_k12(inpath, iidict):
-    with open(inpath) as from_file:
-        for line in from_file:
-            cells = line.split()
-            iidict[cells[0]] = 'essential'
-    return iidict
-
-seqdb = '/home/fatemeh/EnTrI/sequences/fasta-protein/chromosome/seqdb.fasta'
-clusters = '/home/fatemeh/EnTrI/results/hieranoid-result.txt'
-insertion_indices = '/home/fatemeh/EnTrI/results/insertion-indices/normalised-insertion-indices'
-k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
-outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid'
-makedir(outdir)
-sequences = read_fasta_sequences(seqdb)
-gene_essentiality = read_gene_essentiality(insertion_indices)
-gene_essentiality = read_k12(k12path, gene_essentiality)
-species_names = defaultdict()
-species_names = {"all":["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441", "ROD", "SEN", "SL1344", "STM",
[…]

### 2016-07-15 codes/insertion-position-bias.R

@@ -1,38 +1,0 @@
-insertion_positions <- read.table("../results/insertion-position-bias.out", row.names=1)
-pdf("../results/insertion-position-bias.pdf")
-toplot <- colMeans(insertion_positions[,1:ncol(insertion_positions)-1])
-cols = c(rep("khaki4",5), rep("slategray",75), rep("palevioletred4",20))
-midpoints <- barplot(toplot, main="All", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2, col=cols, border=NA)
-axis(1, at=midpoints[c(1,100)], labels=c('5\'','3\''), cex.axis=1.5)
-legend(41,0.7, c("First 5%","internal", "Last 20%"), lty=c(1,1,1),cex=1.5, lwd=c(4,4, 4),col=c(cols[1],cols[6],cols[81]), bg="white")
-essential = c()
-nonessential = c()
-beneficialloss = c()
-for (i in seq(1,nrow(insertion_positions)))
-{
-  if (insertion_positions[i,ncol(insertion_positions)] < 0.2)
-  {
-    essential = rbind(essential, insertion_positions[i,1:ncol(insertion_positions)-1]) 
-  }
-  else if (insertion_positions[i,ncol(insertion_positions)] < 2)
-  {
-    nonessential = rbind(nonessential, insertion_positions[i,1:ncol(insertion_positions)-1])
-  }
-  else
-  {
-    beneficialloss = rbind(beneficialloss, insertion_positions[i,1:ncol(insertion_positions)-1]) 
-  }
-}
-toplot <- colMeans(essential)
-midpoints <- barplot(toplot, main="Essential", xaxt="n", xlab="position", ylab="mean ii", cex.axis=1.5, cex.lab=1.5, cex.main=2, col=cols, border=NA)
-axis(1, at=midpoints[c(1,100)], labels=c('5\'','3\''), cex.axis=1.5)
-legend(41,0.3, c("First 5%","internal", "Last 20%"), lty=c(1,1,1),cex=1.5, lwd=c(4,4, 4),col=c(cols[1],cols[6],cols[81]), bg="white")
-toplot <- colMeans(nonessential)
[…]

### 2016-07-15 codes/insertion-resolution-calculator.py

@@ -1,23 +1,0 @@
-from collections import defaultdict
-from re import match
-from os import listdir, path
-
-plots = '../sequences/plot-files/chromosome'
-genome_length = {"SL1344":4878012, "STMMW":4879400, "SEN":4685848, "t":4791961, "STM":4895639, "ETEC":5153435,
-                 "b":4641652, "CS17":4994793, "NCTC13441":5174631, "ROD":5346659, "BN373":5324709, "ERS227112":5869288,
-                 "ENC":4908759}
-list_of_files = listdir(plots)
-plots_dict = defaultdict(list)
-for filename in list_of_files:
-    name, extension = path.splitext(filename)
-    with open('{0}/{1}'.format(plots, filename), 'r') as plotfile:
-        for line in plotfile:
-            cells = line.split()
-            plots_dict[name].append(int(cells[0]) + int(cells[1]))
-
-genome_insertions = dict()
-for item in plots_dict.keys():
-    genome_insertions[item] = sum([1 for x in plots_dict[item] if x > 0])
-
-for item in plots_dict.keys():
-    print ('Genome name: ' + item + '\t' + 'Resolution: ' + str(float(genome_length[item])/float(genome_insertions[item])) + '\n')

### 2016-07-15 codes/list-beneficial-losses.py

@@ -1,59 +1,0 @@
-from Bio import SeqIO
-from os import listdir
-
-def read_fasta_sequences(filepath):
-    with open(filepath, 'rU') as fasta_file:
-        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
-    return fasta_dict
-
-seqdb = '../sequences/fasta-protein/chromosome/seqdb.fasta'
-iis = '../results/check-biases/without-ends.txt'
-coreesspath = '../results/define-core-accessory/all/core-essential-genomes'
-corenespath = '../results/define-core-accessory/all/core-never-essential-genomes'
-accessoryesspath = '../results/define-core-accessory/all/accessory-essential-genomes'
-accessorynespath = '../results/define-core-accessory/all/accessory-never-essential-genomes'
-benloss = '../results/beneficial-losses.fasta'
-coreess = '../results/core-essential.fasta'
-corenes = '../results/core-never-essential.fasta'
-accessoryess = '../results/accessory-essential.fasta'
-accessorynes = '../results/accessory-never-essential.fasta'
-sequence_dict = read_fasta_sequences(seqdb)
-with open(iis, 'r') as iisfile:
-    with open(benloss, 'w') as benlossfile:
-        for line in iisfile:
-            cells = line.split()
-            if float(cells[1]) > 2:
-                benlossfile.write('>')
-                benlossfile.write(sequence_dict[cells[0]].description)
-                benlossfile.write('\n')
-                benlossfile.write(str(sequence_dict[cells[0]].seq))
-                benlossfile.write('\n')
-
-
-list_of_files = listdir(coreesspath)
-with open(coreess, 'w') as outfile:
-    for filename in list_of_files:
-        with open(coreesspath+'/'+filename) as infile:
-            for line in infile:
-                outfile.write(line)
-
-list_of_files = listdir(corenespath)
-with open(corenes, 'w') as outfile:
-    for filename in list_of_files:
-        with open(corenespath+'/'+filename) as infile:
-            for line in infile:
-                outfile.write(line)
-
-list_of_files = listdir(accessoryesspath)
-with open(accessoryess, 'w') as outfile:
-    for filename in list_of_files:
-        with open(accessoryesspath+'/'+filename) as infile:
-            for line in infile:
-                outfile.write(line)
-
-list_of_files = listdir(accessorynespath)
-with open(accessorynes, 'w') as outfile:
-    for filename in list_of_files:
[…]

### 2016-07-15 codes/loess-fit.R

@@ -1,16 +1,0 @@
-indir <- "../results/per-species-ii/"
-names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t")
-dict = c("Citrobacter", "Escherichia coli ETEC CS17", "Enterobacter", "Escherichia coli ETEC H10407", "Escherichia coli UPEC",
-         "Klebsiella pneumoniae RH201207", "Klebsiella pneumoniae Ecl8", "Salmonella enteritidis", "Salmonella typhimurium A130",
-         "Salmonella typhimurium SL1344", "Salmonella typhimurium D23580", "Salmonella typhi")
-names(dict) <- names
-list_of_files <- list.files(path=indir, full.names=T, recursive=FALSE)
-for (filename in list_of_files)
-{
-  iitable = read.table(filename, header = FALSE)
-  ii = iitable[,2][iitable[,2]>=0]
-  h <- hist(ii, breaks =seq(min(ii),max(ii)+1,0.02), plot=FALSE)
-  plot(h)
-  l = loess(h$counts ~ h$mids)
-  lines(l, col="red")
-}

### 2016-07-15 codes/make-essentiality-tree.py

@@ -1,69 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path
-from argparse import ArgumentParser
-from re import match, compile
-
-parser = ArgumentParser(description='Finds core genome and makes an essentiality tree from them')
-parser.add_argument('indir', help='Directory of fasta clusters merged with essentiality information (merge-clust-plot)')
-parser.add_argument('out', help='Output file name')
-args = parser.parse_args()
-input_dir = args.indir
-out_path = args.out
-
-# input_dir = '../results/merge-clust-plot/final_clusters'
-# out_path = 'test.txt'
-species_names = ["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441", "ROD", "SEN", "SL1344", "STM", "STMMW", "t"]
-num_species = len(species_names)
-gene_dict = {species_names[i]: 0 for i in range(num_species)}
-ii_dict = {species_names[i]: i for i in range(num_species)}
-name_dict = {"BN373": "KlePn.2", "CS17": "EscCo.1", "ENC": "EntCo.1",
-             "ERS227112": "KlePn.1", "ETEC": "EscCo.2", "NCTC13441": "EscCo.3", "ROD": "CitRo.1",
-             "SEN": "SalEn.1", "SL1344": "SalEn.3", "STM": "SalEn.2", "STMMW": "SalEn.4",
-             "t": "SalEn.5"}
-iil = [None] * num_species
-
-list_of_gene_files = []
-list_of_insertion_indices = []
-
-list_of_files = listdir(input_dir)
-for filename in list_of_files:
-    for key in gene_dict.keys():
-        gene_dict[key] = 0
-        iil[ii_dict[key]] = 0
-    with open('{0}/{1}'.format(input_dir, filename)) as from_file:
-        for line in from_file:
-            cells = line.split()
-            match_result = match('([a-zA-Z0-9]+?)_\S+', cells[1])
-            if not match_result:
-                match_result = match('([a-zA-Z]+?)\d+\S*', cells[1])
-            if match_result:
-                name = match_result.group(1)
-                if name in gene_dict.keys():
-                    gene_dict[name] += 1
-                    if 0 <= float(cells[4]) < 0.2:
-                        iil[ii_dict[name]] = 1
[…]

### 2016-07-15 codes/make_marker.py

@@ -1,138 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path, mkdir
-from argparse import ArgumentParser
-from re import match, compile
-from shutil import rmtree
-
-
-def makedir(dirname):
-    # print ('Making directory \'{0}\'..'.format(dirname))
-    if path.exists(dirname):
-        input_var = 'i'
-        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
-            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
-        if input_var == 'Y' or input_var == 'y':
-            rmtree(dirname)
-        else:
-            raise SystemExit
-    mkdir(dirname)
-
-parser = ArgumentParser(description='Finds marker genes to make phylogenetic trees')
-parser.add_argument('indir', help='Directory of fasta clusters (EFam-fastas)')
-parser.add_argument('indnadir', help='Directory of DNA fasta clusters (dna-clusters)')
-parser.add_argument('outdir', help='Directory for outputs (protein fastas)')
-parser.add_argument('outdnadir', help='Directory for outputs(DNA fastas)')
-args = parser.parse_args()
-input_dir = args.indir
-input_dnadir = args.indnadir
-output_dir = args.outdir
-output_dnadir = args.outdnadir
-
-makedir(output_dir)
-makedir(output_dnadir)
-
-gene_dict = {"BN373": 0, "CS17": 0, "ENC": 0, "ERS227112": 0, "ETEC": 0, "NCTC13441": 0,
-             "ROD": 0, "SEN": 0, "SL1344": 0, "STM": 0, "STMMW": 0, "t": 0, "b":0}
-
-name_dict = {"BN373": "KlePn.2", "CS17": "EscCo.1", "ENC": "EntCo.1",
-             "ERS227112": "KlePn.1", "ETEC": "EscCo.2", "NCTC13441": "EscCo.3", "ROD": "CitRo.1",
-             "SEN": "SalEn.1", "SL1344": "SalEn.3", "STM": "SalEn.2", "STMMW": "SalEn.4",
-             "t": "SalEn.5", "b": "EscCo.4"}
-
-list_of_gene_files = []
-
-list_of_files = listdir(input_dir)
-for filename in list_of_files:
-    for key in gene_dict.keys():
-        gene_dict[key] = 0
[…]

### 2016-07-15 codes/not-run-yet/clust2plot-conserved-unconserved.R

@@ -1,140 +1,0 @@
-library(stringr)
-# args <- commandArgs(trailingOnly = TRUE)
-# clusters <- args[1]
-clusters_path <- c("../results/merge-clust-plot-without-ends/")
-for (cpitem in clusters_path)
-{
-  list_of_files <- list.files(path=cpitem, full.names=T, recursive=FALSE)
-  names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t", "b")
-  numspecies = length(names)
-  file_II = list()
-  file_size = list()
-  file_group = list()
-  for (filename in list_of_files)
-  {
-    clustspecies = c()
-    cluster <- as.matrix(read.table(filename))
-    i_sum = 0
-    l_sum = 0
-    cluster_size = nrow(cluster)
-    clust_with_ii_size = 0
-    for (i in (1:cluster_size))
-    {
-      if (as.numeric(cluster[i,5]) >= 0)
-      {
-        match = str_match(cluster[i,2], "([[:graph:]]+)\\_[[:alnum:]]+")[2]
-        if (is.na(match))
-        {
-          match = str_match(cluster[i,2], "([[:alpha:]]+)[[:digit:]]+")[2]
-        }
-        if (match %in% names)
-        {
-          clustspecies = c(clustspecies, match) 
-        }
-        i_sum = i_sum + as.numeric(cluster[i, 5])
-        l_sum = l_sum + (as.numeric(cluster[i, 4]) - as.numeric(cluster[i, 3]) + 1) * 3
-        clust_with_ii_size = clust_with_ii_size  + 1
-      }
-    }
-    if (l_sum > clust_with_ii_size * 60)
-    {
-      file_II[basename(filename)] = i_sum / clust_with_ii_size
-      file_size[basename(filename)] = cluster_size
-      one_or_more = length(unique(clustspecies))
-      greater_than_one = length(table(clustspecies)[table(clustspecies)>1])
-      if (as.numeric(one_or_more) <= 0.3 * numspecies)
-      {
-        file_group[basename(filename)] = 'ORFan'
-      }
-      else
-      {
[…]

### 2016-07-15 codes/not-run-yet/p-values.R

@@ -1,26 +1,0 @@
-ratios_conserved <- c(8*100/(320+8), 31*100/(135+31), 36*100/(159+36), 11*100/(240+11), 14*100/(186+14), 24*100/(582+21), 100*100/(488+100), 48*100/(600+48), 40*100/(791+40), 40*100/(1228+40), 30*100/(711+30), 41*100/(604+41))
-ratios_unconserved <- c(385*100/(3583+385), 363*100/(3772+363), 376*100/(3968+376), 427*100/(3767+427), 361*100/(3934+361), 354*100/(3800+354), 395*100/(3677+395), 374*100/(3728+374), 371*100/(3783+371), 376*100/(4018+376), 305*100/(3949+305), 267*100/(2811+267))
-ratios <- rbind(ratios_conserved,ratios_unconserved)
-p_values <- c(7.323e-07, 8.904e-05, 3.020307e-05, 0.001477236, 0.6000692, 3.70441e-05, 4.325686e-07, 0.1807112, 2.494271e-05, 2.776819e-12, 0.001383878, 0.05882195)
-names <- c("Salmonella typhi", "Salmonella enteritidis", "Salmonella typhimurium SL1344", "Salmonella typhimurium A130",
-           "Salmonella typhimurium D23580", "Escherichia coli UPEC", "Escherichia coli ETEC CS17", "Escherichia coli ETEC H10407",
-           "Citrobacter", "Klebsiella pneumoniae RH201207", "Klebsiella pneumoniae Ecl8", "Enterobacter")
-names(p_values) <- names
-colnames(ratios) <- names
-mar.default <- c(5,4,2,2) + 0.1
-par(mar = mar.default + c(13, 1, 0, 0))
-barplot(ratios, beside=TRUE,ylim=c(0,25),col=c("limegreen","gray17"), las=2, ylab="% Essential", cex.lab = 1.5, axes=FALSE)
-axis(2, at=c(0,25), cex.axis=1.5)
-text(1.8, max(ratios[,1])+1, "***", cex=2)
-text(4.7, max(ratios[,2])+1, "***", cex=2)
-text(7.75, max(ratios[,3])+1, "***", cex=2)
-text(10.7, max(ratios[,4])+1, "*", cex=2)
-text(13.75, max(ratios[,5])+1.5, "NS", cex=1)
-text(16.7, max(ratios[,6])+1, "***", cex=2)
[…]

### 2016-07-15 codes/pairwise-venn-diagrams.py

@@ -1,205 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path, mkdir
-from re import match
-from shutil import rmtree
-
-def makedir(dirname):
-    # print ('Making directory \'{0}\'..'.format(dirname))
-    if path.exists(dirname):
-        input_var = 'i'
-        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
-            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
-        if input_var == 'Y' or input_var == 'y':
-            rmtree(dirname)
-        else:
-            raise SystemExit
-    mkdir(dirname)
-
-class Venn(object):
-    def __init__(self, species1, species2, clusters):
-        self.sp1 = species1
-        self.sp2 = species2
-        self.clusters = clusters
-        self.sp1presp2abs = 0
-        self.sp1presp2pre = 0
-        self.sp1abssp2pre = 0
-        self.sp1esssp2abs = 0
-        self.sp1esssp2nes = 0
-        self.sp1esssp2ess = 0
-        self.sp1nessp2ess = 0
-        self.sp1abssp2ess = 0
-        self.sp1pre = 0
-        self.sp1abs = 0
-        self.sp1ess = 0
-        self.sp1nes = 0
-        self.sp2pre = 0
-        self.sp2abs = 0
-        self.sp2ess = 0
-        self.sp2nes = 0
-
-    def find_values(self):
-        list_of_files = listdir(self.clusters)
-        for filename in list_of_files:
-            sp1ispresent = 0
-            sp2ispresent = 0
-            sp1isessential = 0
-            sp2isessential = 0
-            with open(self.clusters + '/' + filename, 'r') as clusterfile:
-                for line in clusterfile:
-                    cells = line.split()
-                    match_result = match('([a-zA-Z0-9]+?)_\S+', cells[1])
-                    if not match_result:
-                        match_result = match('([a-zA-Z]+?)\d+\S*', cells[1])
-                    if match_result:
-                        name = match_result.group(1)
-                    if name == self.sp1:
-                        sp1ispresent += 1
-                        if 0 <= float(cells[4]) < 0.2:
-                            sp1isessential += 1
-                    if name == self.sp2:
-                        sp2ispresent += 1
-                        if 0 <= float(cells[4]) < 0.2:
[…]

### 2016-07-15 codes/polar-mutation-bias.py

@@ -1,27 +1,0 @@
-iipath = '../results/check-biases/without-ends.txt'
-operonpath = '../operons/SL1344-cropped.opr'
-locustag = 'SL1344'
-iidict = {}
-with open(iipath, 'r') as iifile:
-    for iiline in iifile:
-        if iiline.startswith(locustag):
-            iicells = iiline.split()
-            iidict[iicells[0]] = float(iicells[1])
-print len(iidict)
-print sum(0 <= iidict[x] < 0.2 for x in iidict.keys())
-
-
-# operonid = ''
-# operongenes = []
-# operonstrand = ''
-operongenesnum = 0
-operonessentialsnum = 0
-with open(operonpath, 'r') as operonfile:
-    for operonline in operonfile:
-        operoncells = operonline.split()
-        if operoncells[2].startswith(locustag):
-            operongenesnum += 1
-            if operoncells[2] in iidict.keys() and 0 <= iidict[operoncells[2]] < 0.2:
-                operonessentialsnum += 1
-print operongenesnum
-print operonessentialsnum

### 2016-07-15 codes/remove-biases.R

@@ -1,41 +1,0 @@
-library(stringr)
-library(gdata)
-names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t")
-dict = c("Citrobacter", "Escherichia coli ETEC CS17", "Enterobacter", "Escherichia coli ETEC H10407", "Escherichia coli UPEC",
-         "Klebsiella pneumoniae RH201207", "Klebsiella pneumoniae Ecl8", "Salmonella enteritidis", "Salmonella typhimurium A130",
-         "Salmonella typhimurium SL1344", "Salmonella typhimurium D23580", "Salmonella typhi")
-names(dict) <- names
-
-biasespath <- c("../results/check-biases/with-ends.txt", "../results/check-biases/without-ends.txt")
-for (item in biasespath)
-{
-  biasestable = read.table(item, header = FALSE)
-  colnames(biasestable) <- c("name", "ii", "dist", "gc")
-  j = 1
-  prevj = 1
-  for (i in seq(1, length(names)-1))
-  {
-    ii <- c()
-    d <- c()
-    while((i == length(names) & j <= length(biasestable$name)) | !startsWith(toString(biasestable$name[j]), names[i+1]))
-    {
-      ii <- c(ii, biasestable$ii[j])
-      d <- c(d, biasestable$dist[j])
-      j = j + 1
-    }
-    fit <- gam(ii~s(d))
-    gamprediction <- predict.gam(fit, data.frame(d))
-    ii = (ii * mean(ii)) / gamprediction
-    biasestable$ii[prevj : (j-1)] = ii
-    prevj = j
-  }
-  ii <- c(biasestable$ii)
-  gc = c(biasestable$gc)
-  fit <- gam(ii~s(gc))
-  gamprediction <- predict.gam(fit, data.frame(gc))
-  biasestable$ii = (biasestable$ii * mean(biasestable$ii))/gamprediction
-  write.table(biasestable, file = paste("../results/remove-biases/normalised-iis-", strsplit(basename(item), "\\.")[[1]][1], ".txt",sep=''), quote = FALSE, row.names = FALSE, col.names = FALSE) 
-}
-
-# hist(biasestable$ii, xlim=c(0,3),breaks=1e5)
-# http://stats.stackexchange.com/questions/132652/how-to-determine-which-distribution-fits-my-data-best

### 2016-07-15 codes/test100logos.py

@@ -1,79 +1,0 @@
-#! /bin/sh
-""":"
-exec python3 $0 ${1+"$@"}
-"""
-
-from os import listdir, path, remove
-import numpy as np
-
-genomespath = '../sequences/fasta-genome/chromosome'
-plotspath = '../sequences/plot-files/chromosome'
-genomextension = 'fa'
-logospath = '../results/logos/100logos.txt'
-complement = {'a': 't', 'c': 'g', 'g': 'c', 't': 'a', 'u': 'a', 'm': 'k', 'k': 'm', 'r': 'y',
-              'y': 'r', 'w': 'w', 's': 's', 'v': 'b', 'b': 'v', 'h': 'd', 'd': 'h', 'n': 'n'}
-
-if path.isfile(logospath):
-    input_var = 'i'
-    while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
-        input_var = input('File {0} already exists. Replace? [Y/n] '.format(logospath))
-        if input_var == 'Y' or input_var == 'y':
-            remove(logospath)
-        else:
-            raise SystemExit
-
-plots_f = []
-plots_b = []
-sequences = ''
-list_of_files = listdir(plotspath)
-for filename in list_of_files:
-    name, extension = path.splitext(filename)
-    with open('{0}/{1}'.format(plotspath, filename)) as plotfile:
-        for line in plotfile:
-            cells = line.split()
-            plots_f.append(int(cells[0]))
-            plots_b.append(int(cells[1]))
-    with open('{0}/{1}.{2}'.format(genomespath, name, genomextension)) as genomefile:
-        for line in genomefile:
-            if not line.startswith('>'):
-                if line.endswith('\n'):
-                    line = line[:-1]
-                sequences += line
-plots = np.array(plots_f + plots_b)
-top_indices = np.argpartition(plots, -100)[-100:] # returns the indices of 100 maximums in plots
-with open (logospath, 'a') as logosfile:
-        # for i in range(0, len(plots_f)):
-        #     if plots_f[i] > 0 or plots_b[i] > 0:
-        #         if 10 <= i < len(sequences)-10:
-        #             seq_to_write = sequences[i-10:i+11]
-        #         elif i < 10:
-        #             seq_to_write = sequences[(i-10)%len(sequences):len(sequences)]+sequences[0:i+11]
-        #         else:
[…]

### 2016-07-15 codes/upsetr-hieranoid.R

@@ -1,55 +1,0 @@
-library("UpSetR")
-library(stringr)
-# listinput <- list(a=c(1,1,1,1,2,1,2,3,3,3), b=c(1,4,5), d=c(1,2,5,6,7), e=c(4,8,9))
-# upset(fromList(listinput))
-
-clusters_path <- "../results/hieranoid-result.txt"
-essentiality_path <- "../results/insertion-indices/normalised-insertion-indices/"
-k12_path <- "../results/ecogene-k12.txt"
-list_of_files <- list.files(path=essentiality_path, full.names=T, recursive=FALSE)
-list_of_essential_genes = list()
-for (filename in list_of_files)
-{
-  iis <- as.matrix(read.table(filename))
-  list_of_essential_genes[strsplit(basename(filename),"\\.")[[1]][1]] <- list(iis[,1][iis[,3]=='essential'])
-}
-iis = as.matrix(read.table(k12_path))
-list_of_essential_genes["b"] <- list(iis[,1])
-presence = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
-                "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
-essentiality = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
-                    "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
-cluster <- readLines(clusters_path)
-numclusters = length(cluster)
-for (clusterindex in (1:numclusters))
-{
-  line = cluster[clusterindex]
-  match_result = str_match_all(line, '[[,(]](([[:alnum:]]+)_[[:alnum:]]+):')
-  temp_result = str_match_all(line, '[[,(]](([[:alpha:]]+)[[:digit:]]+):')
-  match_result[[1]] <- rbind(match_result[[1]], temp_result[[1]])
-  match_result <- match_result[[1]]
-  for (i in (1:nrow(match_result)))
-  {
-    name = match_result[i,3]
-    if (name %in% names(presence)) 
-    {
-      presence[[name]] <- rbind(presence[[name]], clusterindex)
-      gene = match_result[i,2]
-      if (gene %in% list_of_essential_genes[[name]])
-      {
[…]

### 2016-07-15 codes/upsetr.R

@@ -1,48 +1,0 @@
-library("UpSetR")
-library(stringr)
-# listinput <- list(a=c(1,1,1,1,2,1,2,3,3,3), b=c(1,4,5), d=c(1,2,5,6,7), e=c(4,8,9))
-# upset(fromList(listinput))
-
-clusters_path <- "../results/single-copy-genes/"
-list_of_files <- list.files(path=clusters_path, full.names=T, recursive=FALSE)
-presence = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
-                "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
-essentiality = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
-                    "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
-list_of_essential_genes = list("ROD"=c(), "CS17"=c(), "ENC"=c(), "ETEC"=c(), "NCTC13441"=c(), "ERS227112"=c(), "BN373"=c(), "SEN"=c(), "STM"=c(),
-                    "SL1344"=c(), "STMMW"=c(), "t"=c(), "b"=c())
-for (filename in list_of_files)
-{
-  cluster <- as.matrix(read.table(filename))
-  cluster_size = nrow(cluster)
-  for (i in (1:cluster_size))
-  {
-    if (as.numeric(cluster[i,5]) >= 0)
-    {
-      match = str_match(cluster[i,2], "([[:graph:]]+)\\_[[:alnum:]]+")[2]
-      if (is.na(match))
-      {
-        match = str_match(cluster[i,2], "([[:alpha:]]+)[[:digit:]]+")[2]
-      }
-      if (match %in% names(presence))
-      {
-        presence[[match]] = c(presence[[match]], strsplit(basename(filename),"\\.")[[1]][1])
-        if (as.numeric(cluster[i,5]) < 0.2)
-        {
-          essentiality[[match]] = c(essentiality[[match]], strsplit(basename(filename),"\\.")[[1]][1])
-          list_of_essential_genes[[match]] = c(list_of_essential_genes[[match]], cluster[i,2])
-        }
-      }
-    }
-  }
-}
[…]

### 2016-07-15 codes/venn-to-heatmap.R

@@ -1,26 +1,0 @@
-library("gplots")
-venspath <- "../results/pairwise-venn-diagrams/"
-list_of_files <- list.files(path=venspath, full.names=T, recursive=FALSE)
-plot_mains = c("Ess(rows)-Abs(columns)", "Ess-Ess", "Ess(rows)-NoEs(columns)", "Pres(rows)-Abs(columns)", "Pres-Pres")
-i = 1
-
-my_palette <- colorRampPalette(c("steelblue4", "slategray2"))(n = 100)
-
-pdf("../results/essentiality-heatmap.pdf")
-
-for (filename in list_of_files)
-{
-  heattable <- as.matrix(read.table(filename))
-  names <- heattable[1,2:nrow(heattable)]
-  heattable <- heattable[-1,]
-  heattable <- heattable[,-1]
-  rownames(heattable) <- names
-  colnames(heattable) <- names
-  heattable <- heattable[c(5,6,7,8,9,2,3,4,1,10,11,12),c(5,6,7,8,9,2,3,4,1,10,11,12)]
-  class(heattable) <- "numeric" 
-  heatmap.2(heattable, margins = c(15, 15), main = plot_mains[i], density.info="none", trace="none", keysize=1,
-            lmat=rbind( c(0, 3), c(2,1), c(0,4) ), lhei=c(1.5, 6, 2 ), Rowv=FALSE, Colv=FALSE, dendrogram='none', col=my_palette)
-  i = i+1
-}
-
-dev.off()

### 2016-07-15 codes/visualise-kobas.R

@@ -1,25 +1,0 @@
-library(stringr)
-library(ggplot2)
-lineset <- readLines("../KOBAS/beneficial-losses/t.kobas")
-kegg_table <- c()
-for (item in lineset)
-  {
-    if (!(startsWith(item, '#') | startsWith(item, '-') | item ==''))
-    {
-      kegg_table <- rbind(kegg_table, strsplit(item, '\t'))
-    }
-}
-names <- c()
-pval <- c()
-for (item in kegg_table)
-  {
-    names <- c(names, item[1])
-    pval <- c(pval,as.numeric(item[7]))
-  }
-df <- data.frame(pathway=names,pvalue=-log10(pval))
-df = df[-seq(21,nrow(df),1),]
-pdf("../results/pathway-enrichment-ty2-beneficiallosses.pdf")
-ggplot(data=df, aes(x=reorder(pathway,pvalue),y=pvalue))+geom_bar(stat="identity")+ coord_flip()+labs(x='Pathway',y='-log10(P-value)') +
-  geom_abline(slope = 0, intercept = -log10(0.05), color="red")+ylim(c(0,max(df[1,2], -log10(0.05))))+ggtitle("Beneficial losses") +
-  theme(text = element_text(size=15))
-dev.off()

### 2016-09-06 documents/manuscript.tex
Added a study of low GC genes, 5' end of essential genes, and sampling for essentiality definition

@@
 We extend this research by studying 13 bacterial strains from Enterobacteriaceae.
-These strains are depicted in Fig.\@~\ref{fig:species-tree}.
+These strains are depicted in Fig.\@ \ref{fig:species-tree}.
 
@@
 %------------------------------------------------
+\section{Transposon mutagenesis}
+\subsection{Strains}
+\subsection{Transposon insertion experiment}
+\subsection{Essentiality}
+\section{Are transposons biased towards certain positions?}
+Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015}.
+We have performed a thorough study of these biases.
+To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome.
+Fig.\@ \ref{fig:distance-bias} shows the results.
+The red line is a loess curve that has been fitted to the data when the smoothness parameter equals $0.2$.
+The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication.
+A possible explanation for this phenomenon is that the bacteria were under replication while being infected with transposons.
+Therefore, the number of gene copies close to the origin of replication was greater and more insertions have occurred in these genes.
 
-\section{Are transposons biased towards certain positions?}
-\begin{itemize}
-\item There is a bias towards the position of the gene (Fig.~\ref{fig:distance-bias}).
-\item There is a negligible bias towards certain motifs (Fig.~\ref{fig:logos})
-\item The G-C bias needs to be studied further (Fig.~\ref{fig:GC-bias})
-\item The number of insertions on the 3’ and 5’ ends is more than the internal region in essential genes and less than the internal region in beneficial losses.
+We have tested whether our transposons are biased towards certain motifs. For this, we have generated a logo from 10 nucleotides flanking the 100 top most
+frequent insertion sites in each genome.
+The results are depicted in Fig.\@  \ref{fig:logos}.
+The results show a slight bias towards certain combinations of bases.
+In addition, we have investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index.
+The red lines show the loess curve when the smoothness parameter is $0.2$.
+\textcolor{red}{\{To be completed\}}
+
+The other bias that we have considered is the bias towards certain locations in a gene.
+We have divided every gene into 100 bins and calculated the mean insertion index for each bin.
+Fig.\@ 5a shows almost no bias towards any location.
+We have also divided our genes into three groups - essential genes, non-essential genes, and beneficial losses - and studied the bias in each group.
+The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses.
+The case for the non-essential genes is similar to the average (Fig.\@ 5a).
+High number of insertions at the 3' end of essential genes implies that the functional part of the genes are located before the insertions.
+On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors.
+\textcolor{red}{\{To be tested\}}
+%\begin{itemize}
+%\item There is a bias towards the position of the gene (Fig.~\ref{fig:distance-bias}).
+%\item There is a negligible bias towards certain motifs (Fig.~\ref{fig:logos})
+%\item The G-C bias needs to be studied further (Fig.~\ref{fig:GC-bias})
+%\item The number of insertions on the 3' and 5' ends is more than the internal region in essential genes and less than the internal region in beneficial losses.
 (Fig.~\ref{fig:insertion-position-bias})
-\end{itemize}
+%\end{itemize}
 \section{Gene classes}
@@
 
-%------------------------------------------------
-
-\section{Discussion}
-
-%----------------------------------------------------------------------------------------
-%	MATERIALS AND METHODS
-%----------------------------------------------------------------------------------------
-
-%% Optional Materials and Methods Section
-%% The Materials and Methods section header will be added automatically.
-
-\begin{materials}
-\textbf{Clustering orthologous and paralogous genes.} To study the essentiality of genes in 12 our strains, We needed to cluster sets of orthologous genes in these strain.
+%%------------------------------------------------
+%
+%\section{Discussion}
+%
+%
+%
+%%----------------------------------------------------------------------------------------
+%%	MATERIALS AND METHODS
+%%----------------------------------------------------------------------------------------
+%
+%%% Optional Materials and Methods Section
+%%% The Materials and Methods section header will be added automatically.
+%
+%\begin{materials}
+%\textbf{Clustering orthologous and paralogous genes.} To study the essentiality of genes in our 12 strains, We needed to cluster sets of orthologous genes in these strain.
 Plenty of methods are proposed for this purpose.
@@
 For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
-\end{materials}
+%\end{materials}
 
@@
 \caption{We have divided our genes into 3 segments: 5\% of
-the genes on the 5’ end, 20\% of the genes on the 3’ end, and the rest in the middle.}
+the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle.}
 \label{fig:insertion-position-bias}

### 2016-09-06 bin/insertion-position-bias.R

@@ -36,3 +36,11 @@ midpoints <- barplot(toplot, main="Beneficial loss", xaxt="n", xlab="position", 
 axis(1, at=midpoints[c(1,100)], labels=c('5\'','3\''), cex.axis=1.5)
 legend(41,1.5, c("First 5%","internal", "Last 20%"), lty=c(1,1,1),cex=1.5, lwd=c(4,4, 4),col=c(cols[1],cols[6],cols[81]), bg="white")
 dev.off()
+
+# for (item in essential)
+# {
+#   if (mean(as.numeric(item[2:6])) >= (mean(as.numeric(item[7:81]))+0.02))
+#   {
+#     print(item)
+#   }
+# }

### 2016-09-06 bin/semisupervised.R

@@ -1,0 +1,115 @@
+library(stringr)
+library("upclass")
+datapath <- "../results/insertion-indices/check-biases/"
+essentialitypath <- "../results/insertion-indices/normalised-insertion-indices/"
+k12_path <- "../results/ecogene-k12.txt"
+clusters_path <- "../results/hieranoid/hieranoid-result.txt"
+list_of_files <- list.files(path=datapath, full.names=T, recursive=FALSE)
+datatable = data.frame()
+counter = 0
+for (filename in list_of_files)
+{
+  thistable <- read.table(filename, header = FALSE)
+  thistable[,7] = counter
+  counter = counter + 1
+  datatable = rbind(datatable, thistable)
+}
+datatable <- datatable[-3]
+row.names(datatable) <- datatable[,1]
+datatable <- datatable[,-1]
+names(datatable) <- c("ii", "distance", "gc", "length", "strain#")
+for (coli in (1:ncol(datatable)))
+{
+  
+  if (max(datatable[,coli]) > 1)
+  {
+    datatable[,coli]=datatable[,coli]/max(datatable[,coli])
+  }
+}
+classes <- sapply(row.names(datatable),function(x) NULL)
+iis=data.frame()
+list_of_essential_genes = c()
+species = c('b')
+for (filename in list_of_files)
+{
+  iis <- as.matrix(read.table(filename))
+  list_of_essential_genes <- c(list_of_essential_genes, as.vector(iis[,1][iis[,3]=='essential']))
+  species = c(species, strsplit(basename(filename),"\\.")[[1]][1])
+}
+iis = as.matrix(read.table(k12_path))[,1]
+list_of_essential_genes <- c(list_of_essential_genes,iis)
+
+cluster <- readLines(clusters_path)
+numclusters = length(cluster)
+for (clusterindex in (1:numclusters))
+{
+  line = cluster[clusterindex]
+  match_result = str_match_all(line, '[[,(]](([[:alnum:]]+)_[[:alnum:]]+):')
+  temp_result = str_match_all(line, '[[,(]](([[:alpha:]]+)[[:digit:]]+):')
+  match_result[[1]] <- rbind(match_result[[1]], temp_result[[1]])
+  match_result <- match_result[[1]]
+  if (setequal(match_result[,3], species) & length(match_result[,3]) == length(species))
+  {
+    sum = 0
[…]

### 2016-09-06 bin/study-low-gc.py

@@ -1,4 +1,6 @@
 from Bio import SeqIO
+from re import match, findall
+from os import listdir, remove
 
 
 def read_fasta_sequences(filepath):
@@ -6,11 +8,51 @@ def read_fasta_sequences(filepath):
         fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
     return fasta_dict
 
-seqdb = '~/EnTrI/data/fasta-protein/chromosome/seqdb.fasta'
+seqdb = '../data/fasta-protein/chromosome/seqdb.fasta'
 seqdict = read_fasta_sequences(seqdb)
-lowgc = '~/EnTrI/results/low-gc.txt'
-worddict = {}
+biases = '../results/insertion-indices/check-biases'
+lowgc = '../results/low-gc.txt'
+lowgcdesc = '../results/low-gc-with-description.txt'
+info = '../results/low-gc-info.txt'
+list_of_files = listdir(biases)
+with open(lowgc, 'w') as tofile:
+    for filename in list_of_files:
+        with open(biases+'/'+filename, 'r') as fromfile:
+            for line in fromfile:
+                cells = line.split()
+                if float(cells[4]) <= 0.45:
+                    tofile.write(cells[0]+'\t'+cells[5]+'\n')
 with open(lowgc, 'r') as fromfile:
-    for line in fromfile:
-        cells = line.split()
-        
+    with open(lowgcdesc, 'w') as tofile:
+        for line in fromfile:
+            cells = line.split()
+            match_result = match("\S+\s+\[\S+\s+\([a-zA-Z]+\)\]\s+\[([^\]]+)\]", seqdict[cells[0]].description)
+            tofile.write(cells[0]+'\t'+cells[1]+'\t'+match_result.group(1)+'\n')
+remove(lowgc)
+
+worddict = {}
+with open(info, 'w') as tofile:
+    with open(lowgcdesc, 'r') as fromfile:
+        for line in fromfile:
+            cells = line.split('\t')
+            description = cells[2].replace(',', ' ')
+            description = description.replace(':', ' ')
+            description = description.replace('_', ' ')
+            description = description.replace(';', ' ')
+            description = description.replace('-', ' ')
+            description = description.replace('(', ' ')
+            description = description.replace(')', ' ')
+            description = description.replace('.', ' ')
+            description = description.replace('/', ' ')
+            description = description.replace('%', ' ')
+            description = description.replace('*', ' ')
+            words = findall("\S+", description)
[…]

### 2016-09-19 documents/manuscript.tex
Added a word enrichment study on beneficial losses, adjusted p-values, changed p-values to one-tailed, changed the control for high G-C word enrichment to interquartile, changed insertion index distribution plot

@@
 We extend this research by studying 13 bacterial strains from Enterobacteriaceae.
-These strains are depicted in Fig.\@ \ref{fig:species-tree}.
+These strains and \textit{Escherichia coli} K-12 MG1655 studied by Baba et al.\@ \cite{baba_construction_2006} are depicted in Fig.\@ \ref{fig:species-tree}.
 
-Enterobacteriaceae is a family that includes bacteria with different host ranges and pathogenecity found in soil, water, plants, animals and humans \cite{brenner_bergeys_2006}.
+Enterobacteriaceae is a family that includes Gram-negative bacteria with different host ranges and pathogenecity found in soil, water, plants, animals and humans \cite{brenner_bergeys_2006}.
 In humans, various strains from this family can cause diarrhoea, septicaemia, urinary tract infection, meningitis, respiratory disease, and wound and burn infection \cite{brenner_bergeys_2006}.
@@
 \subsection{Strains}
-\subsection{Transposon insertion experiment}
+We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
+These strains are all selected from Enterobacteriaceae family.
+\subsection{Transposon insertion workflow}
+We have used Tn5 transposon to generate single inserted mutants and placed our mutants in a selective media for Tn5.
+We have picked the mutants and pooled them and used the splinkerette adapter and primers designed in \cite{barquist_tradis_2016} for PCR enrichment.
+Then we have sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
 \subsection{Essentiality}
+We have used a value called insertion index to evaluate the essentiality of a gene.
+This value is calculated by summing up the number of transposon insertion sites observed in a gene.
+Since the lengths of the genes are different, we have normalised the insertion index by dividing it by gene length.
+Our experiment has been performed on different strains and the library density is different in each experiment.
+Therefore, in order to make the insertion indices in all the strains comparable, we have normalised our insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
+We have not studied genes shorter than 100 base-pairs as they might not be targeted by any transposon due to their shortness.
+
+We have divided our genes into three groups: essential genes, non-essential genes, and beneficial losses.
+We have adapted the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
+The insertion index distribution plot has two peaks and a heavy tail Fig.\@ \ref{fig:iidist}.
+The first peak shows the genes with no or just a few insertions which are considered as essential genes.
+We have fitted an exponential distribution to the first peak and a gamma distribution to the second one.
+Then, we have calculated the log odds ratio for belonging to each of these distributions for each gene.
+The region that has log odds value between -2 and 2 is called the ambiguous region, the genes belonging to the first peak are essential and the rest of the genes are not essential.
+Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.
 \section{Are transposons biased towards certain positions?}
-Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015}.
+Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}.
 We have performed a thorough study of these biases.
@@
 Therefore, the number of gene copies close to the origin of replication was greater and more insertions have occurred in these genes.
+To overcome this bias, we have normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplying this value by the average insertion index.
 
@@
 The red lines show the loess curve when the smoothness parameter is $0.2$.
-\textcolor{red}{\{To be completed\}}
+As the figure shows, when G-C content is less than $40\%$, the insertion index is low, however when it is higher than $50\%$, the insertion index is almost constant.
+A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which causes a reduction in the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
+The other reason is that the genes with low G-C content are enriched in mobile genetic elements (corrected p-value for the number of repetition of these words in the quartile with lowest G-C and the interquartile genes: pathogenicity island: 4.55189365569093e-54, prophage: 3.20756138380406e-26, phage: 1.81663788061583e-08, bacteriophage: 0.00285196530662877) compared to the genes with high G-C content and that has caused seeing a different pattern of essentiality in that region.
 
@@
 We have divided every gene into 100 bins and calculated the mean insertion index for each bin.
-Fig.\@ 5a shows almost no bias towards any location.
+Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location.
 We have also divided our genes into three groups - essential genes, non-essential genes, and beneficial losses - and studied the bias in each group.
 The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses.
-The case for the non-essential genes is similar to the average (Fig.\@ 5a).
+The case for the non-essential genes is similar to the average (Fig.\@ \ref{fig:insertion-position-bias}).
 High number of insertions at the 3' end of essential genes implies that the functional part of the genes are located before the insertions.
 On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors.
-\textcolor{red}{\{To be tested\}}
+\textcolor{red}{\{To be tested\}} We have calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes to overcome these biases.
 %\begin{itemize}
@@
 %\end{itemize}
-\section{Gene classes}
-\subsection{Genus specific}
-\subsection{Single copy}
-\subsection{Multi-copy}
-\section{Evolution of essentiality}
+\section{Essentiality and conservation}
+\subsection{Homolog clustering algorithm (homclust)}
+To study whether each gene in our 13 organisms is conserved we have proposed a program that clusters homologous proteins.
+This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
+It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
+Then it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
+In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by relaxing Jackhmmer parameters.
+Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
+The program is summarised in Fig.\@ \ref{fig:homclust} and its results are illustrated in Fig.\@ \ref{fig:homclust-results}.
+\subsection{Gene classes}
+%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
+Plenty of methods are proposed for this purpose.
+Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
+We have used Hieranoid to cluster the sets of orthologous genes.
+In addition, we intended to study the essentiality of genes in paralogous genes.
+For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
+To study the conservation of genes, we have used homclust and divided the clusters of homologous genes into three groups.
+Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
+These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
+
+We have divided the clusters of orthologous genes into 4 groups based on their essentiality (essential, ambiguous, non-essential, and beneficial loss) and 3 groups based on their conservation (genus specific, single copy, and multi-copy).
+The results are depicted in Fig.\@ \ref{fig:iidist}.
+The figure shows that most of the essential clusters are single copy and most of the beneficial losses are genus specific.
+\textcolor{red}{\{KEGG results\}}
+\subsection{Evolution of essentiality}
+We have compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
+The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show a tree-like signal.
 \begin{itemize}
@@
 %\begin{materials}
-%\textbf{Clustering orthologous and paralogous genes.} To study the essentiality of genes in our 12 strains, We needed to cluster sets of orthologous genes in these strain.
-Plenty of methods are proposed for this purpose.
-Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
-We have used Hieranoid to cluster the sets of orthologous genes.
-In addition, we intended to study the essentiality of genes in paralogous genes.
-For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
+
 %\end{materials}
@@
 
+\begin{figure}
+\includegraphics[scale=0.7]{homclust.pdf}
+\caption{The steps of our proposed algorithm for clustering homologous genes.}
+\label{fig:homclust}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=0.7]{cluster-size-dist.pdf}
+\caption{Size distribution for all clusters of homologous genes.
+Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
+\label{fig:homclust-results}
+\end{figure}
+
 \begin{figure*}
@@
 \includegraphics[page=34, scale=0.25]{biases.pdf}\\
-&\includegraphics[page=37, scale=0.25]{biases.pdf}&\\
+\includegraphics[page=37, scale=0.25]{biases.pdf}&&
+\includegraphics[page=40, scale=0.25]{biases.pdf}\\
 \end{tabular}
@@
 \centering
+\includegraphics[scale=1, page=1]{cluster-essentiality.pdf}
+\caption{The genes have been clustered into orthologous groups using Hieranoid and paralogous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
+Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
+The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
+\label{fig:iidist}
+\end{figure}
+
+\begin{figure}
+\centering
 \includegraphics[scale=0.65, page=1]{upsetr.pdf}\\

### 2016-09-19 bin/clust2plot.R

@@ -1,0 +1,247 @@
+library(stringr)
+library("MASS")
+require(reshape2)  # this is the library that lets you flatten out data
+require(ggplot2)
+# args <- commandArgs(trailingOnly = TRUE)
+# clusters <- args[1]
+clusters_path <- "../results/merge-clust-plot"
+
+list_of_files <- list.files(path=clusters_path, full.names=T, recursive=FALSE)
+names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t", "SL3261")
+genuses = c(1, 2, 3, 2, 2, 4, 4, 5, 5, 5, 5, 5, 5)
+names(genuses) <- names
+numspecies = length(names)
+file_II = list()
+file_size = list()
+file_group = list()
+for (filename in list_of_files)
+{
+  clustspecies = c()
+  cluster <- as.matrix(read.table(filename))
+  i_sum = 0
+  l_sum = 0
+  cluster_size = nrow(cluster)
+  clust_with_ii_size = 0
+  for (i in (1:cluster_size))
+  {
+    if (as.numeric(cluster[i,5]) >= 0)
+    {
+      match = str_match(cluster[i,2], "([[:graph:]]+)\\_[[:alnum:]]+")[2]
+      if (is.na(match))
+      {
+        match = str_match(cluster[i,2], "([[:alpha:]]+)[[:digit:]]+")[2]
+      }
+      if (match %in% names)
+      {
+        clustspecies = c(clustspecies, match) 
+      }
+      i_sum = i_sum + as.numeric(cluster[i, 5])
+      l_sum = l_sum + (as.numeric(cluster[i, 4]) - as.numeric(cluster[i, 3]) + 1) * 3
+      clust_with_ii_size = clust_with_ii_size  + 1
+    }
+  }
+  if (l_sum > clust_with_ii_size * 60)
+  {
+    file_II[basename(filename)] = i_sum / clust_with_ii_size
+    file_size[basename(filename)] = cluster_size
+    unique_clustspecies = unique(clustspecies)
+    one_or_more = length(unique_clustspecies)
+    greater_than_one = length(table(clustspecies)[table(clustspecies)>1])
+    cluster_genuses = c()
+    for (item in unique_clustspecies)
+    {
[…]

### 2016-09-19 bin/merge-clust-plot.py

@@ -1,0 +1,61 @@
+#! /bin/sh
+""":"
+exec python3 $0 ${1+"$@"}
+"""
+
+from sys import argv
+from os import listdir, mkdir, path
+from shutil import rmtree
+from re import match
+
+
+def makedir(dirname):
+    # print ('Making directory \'{0}\'..'.format(dirname))
+    if path.exists(dirname):
+        input_var = 'i'
+        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
+            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
+        if input_var == 'Y' or input_var == 'y':
+            rmtree(dirname)
+        else:
+            raise SystemExit
+    mkdir(dirname)
+
+
+def add_insertion_index(in_clusters, in_iis, k12essentials, output):
+    iis_dict = {}
+    list_of_files = listdir(in_iis)
+    for filename in list_of_files:
+        with open(in_iis+'/'+filename) as iis_file:
+            for line in iis_file:
+                cells = line.split()
+                iis_dict[cells[0]] = cells[1]
+    list_of_files = listdir(in_clusters)
+    for filename in list_of_files:
+        with open('{0}/{1}'.format(in_clusters, filename)) as from_file:
+            with open('{0}/{1}'.format(output, filename), 'w') as to_file:
+                for line in from_file:
+                    cells = line.split()
+                    match_result = match('b\d\d\d\d', cells[1])
+                    for item in cells:
+                        to_file.write(item+'\t')
+                    if cells[1] in iis_dict.keys():
+                        to_file.write(iis_dict[cells[1]]+'\n')
+                    # elif match_result is not None:
+                    #     if cells[1] in k12essentials:
+                    #         to_file.write('0\n')
+                    #     else:
+                    #         to_file.write('7\n')
+                    else:
+                        to_file.write('-1\n')
+
+indir = str(argv[1])
+iifile = str(argv[2])
+k12file = str(argv[3])
+outdir = str(argv[4])
+makedir(outdir)
+k12list = list()
+with open (k12file, 'r') as infile:
+    for line in infile:
+        k12list.append(line.rstrip())
+add_insertion_index(indir, iifile, k12list, outdir)

### 2016-09-19 bin/orfan-single-multiple.R

@@ -1,0 +1,77 @@
+library(stringr)
+list_of_files <- list.files(path="../results/homclust/EFam-clusters", pattern="*.txt", full.names=T, recursive=FALSE)
+names = c("ROD", "CS17", "ENC", "ETEC", "NCTC13441", "ERS227112", "BN373", "SEN", "STM", "SL1344", "STMMW", "t", "SL3261")
+genuses = c(1, 2, 3, 2, 2, 4, 4, 5, 5, 5, 5, 5, 5)
+names(genuses) <- names
+numspecies = length(names)
+clusters = matrix(nrow=length(list_of_files), ncol=5)
+i = 1
+for (filename in list_of_files)
+{
+  clust_tbl =  read.table(filename, header = FALSE)
+  clustspecies = c()
+  for (item in clust_tbl[,2])
+  {
+    match = str_match(item, "([[:graph:]]+)\\_[[:alnum:]]+")[2]
+    if (is.na(match))
+    {
+      match = str_match(item, "([[:alpha:]]+)[[:digit:]]+")[2]
+    }
+    if (match %in% names)
+    {
+      clustspecies = c(clustspecies, match) 
+    }
+  }
+  unique_clustspecies = unique(clustspecies)
+  clusters[i,1]= basename(filename)
+  clusters[i,2]= nrow(clust_tbl)
+  clusters[i,3]= length(unique_clustspecies)
+  clusters[i,4]= length(table(clustspecies)[table(clustspecies)>1])
+  cluster_genuses = c()
+  for (item in unique_clustspecies)
+  {
+    cluster_genuses <- c(cluster_genuses, genuses[unique_clustspecies])
+  }
+  clusters[i,5]= length(unique(cluster_genuses))
+  i = i+1
+}
+
+orfancount = 0
+orfan = c()
+singcopycount = 0
+singcopy = c()
+multicopycount = 0
+multicopy = c()
+for (i in seq(1,nrow(clusters)))
+{
+  if (as.numeric(clusters[i,5]) <= 1)
+  {
+    orfancount = orfancount + 1
+    orfan = c(orfan, as.numeric(clusters[i,2]))
+  }
+  else if (as.numeric(clusters[i,3]) - as.numeric(clusters[i,4]) >= 0.7 * as.numeric(clusters[i,3]))
+  {
+    singcopycount = singcopycount + 1
+    singcopy = c(singcopy, as.numeric(clusters[i,2]))
+  }
+  else
+  {
[…]

### 2016-09-19 bin/study-beneficial-losses.py

@@ -1,0 +1,89 @@
+from Bio import SeqIO
+from os import listdir
+from re import match, findall
+from collections import defaultdict
+
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+seqdb = '../data/fasta-protein/chromosome/seqdb.fasta'
+seqdict = read_fasta_sequences(seqdb)
+essentiality = '../results/insertion-indices/normalised-insertion-indices'
+hloss = '../results/not-beneficial-loss-description.txt'
+bloss = '../results/beneficial-loss-description.txt'
+lossinfo = '../results/essentiality-info.txt'
+
+list_of_files = listdir(essentiality)
+with open(hloss, 'w') as hfile:
+    with open(bloss, 'w') as bfile:
+        for filename in list_of_files:
+            with open(essentiality + '/' + filename, 'r') as fromfile:
+                for line in fromfile:
+                    cells = line.split()
+                    match_result = match("\S+\s+\[\S+\s+\([a-zA-Z]+\)\]\s+\[([^\]]+)\]", seqdict[cells[0]].description)
+                    if cells[2] == 'beneficial-loss':
+                        bfile.write(cells[0]+'\t'+match_result.group(1)+'\n')
+                    else:
+                        hfile.write(cells[0] + '\t' + match_result.group(1) + '\n')
+
+worddict = defaultdict(list)
+
+with open(hloss, 'r') as fromfile:
+    for line in fromfile:
+        cells = line.split('\t')
+        description = cells[1].replace(',', ' ')
+        description = description.replace(':', ' ')
+        description = description.replace('_', ' ')
+        description = description.replace(';', ' ')
+        description = description.replace('-', ' ')
+        description = description.replace('(', ' ')
+        description = description.replace(')', ' ')
+        description = description.replace('.', ' ')
+        description = description.replace('/', ' ')
+        description = description.replace('%', ' ')
+        description = description.replace('*', ' ')
+        description = description.replace('\\', ' ')
+        description = description.replace('\'', ' ')
+        words = findall("\S+", description)
+        for item in words:
+            if len(item) > 3:
+                word = item.lower()
+                if word in worddict.keys():
+                    worddict[word][0] += 1
+                else:
+                    worddict[word] = [1, 0]
+
+with open(bloss, 'r') as fromfile:
+    for line in fromfile:
[…]

### 2016-09-19 bin/study-low-gc.R

@@ -1,0 +1,12 @@
+codons <- "../results/low-gc-info.txt"
+codon_table <- read.table(codons, header = TRUE)
+fishertable =data.frame()
+for (i in seq(1,(nrow(codon_table)-1)))
+{
+  fisherresult = fisher.test(matrix(c(codon_table[i,2],codon_table[i,3],codon_table[nrow(codon_table),2],codon_table[nrow(codon_table),3]),nrow=2,byrow=TRUE), alternative = "greater")
+  fishertable[i,1] = codon_table[i,1]
+  fishertable[i,2] = fisherresult$p.value
+}
+fishertable[,3] = p.adjust(fishertable[,2], method = "BH")
+fishertable = fishertable[order(fishertable[,3], fishertable[,2]),]
+write.table(fishertable, file = "../results/low-gc-pval.txt", quote = FALSE, col.names = c("word", "p-value", "corrected-p-value"), row.names = FALSE, sep="\t")

### 2016-09-19 documents/manuscript.tex
Added word enrichment analysis figure for beneficial losses

@@
 The figure shows that most of the essential clusters are single copy and most of the beneficial losses are genus specific.
-\textcolor{red}{\{KEGG results\}}
+
+\textcolor{red}{\{KEGG results\}} We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}.
+It shows that most of the beneficial losses are mobile elements.
+Most of these mobile elements are not included in KEGG database.
 \subsection{Evolution of essentiality}
@@
 
+\begin{figure}
+\includegraphics[scale=0.7]{essentiality-pval.pdf}
+\caption{The word enrichment analysis for beneficial losses compared to other genes. The red line shows p-value = 0.05.}
+\label{fig:essentiality-pval}
+\end{figure}
+
 \begin{figure*}

### 2016-10-01 documents/manuscript.tex
Updated document, adopted codes from previous versions, added word enrichment analysis, changed ancestralii method

--- 2016-09-19 documents/manuscript.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Proceedings of the National Academy of Sciences (PNAS)
-% LaTeX Template
-% Version 1.0 (19/5/13)
-%
-% This template has been downloaded from:
-% http://www.LaTeXTemplates.com
-%
-% Original author:
-% The PNAStwo class was created and is owned by PNAS:
-% http://www.pnas.org/site/authors/LaTex.xhtml
-% This template has been modified from the blank PNAS template to include
-% examples of how to insert content and drastically change commenting. The
-% structural integrity is maintained as in the original blank template.
-%
-% Original header:
-%% PNAStmpl.tex
-%% Template file to use for PNAS articles prepared in LaTeX
-%% Version: Apr 14, 2008
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\documentclass[12pt,letterpaper]{article}
+\usepackage[top=0.85in,left=1in,footskip=0.75in,marginparwidth=2in]{geometry}
 
-%----------------------------------------------------------------------------------------
-%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
-%----------------------------------------------------------------------------------------
+% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
+\usepackage[utf8]{inputenc}
 
-%------------------------------------------------
-% BASIC CLASS FILE
-%------------------------------------------------
+% clean citations
+\usepackage{cite}
 
-%% PNAStwo for two column articles is called by default.
-%% Uncomment PNASone for single column articles. One column class
-%% and style files are available upon request from pnas@nas.edu.
+% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
+\usepackage{nameref,hyperref}
 
-%\documentclass{pnasone}
-\documentclass{pnastwo}
+% line numbers
+\usepackage[right]{lineno}
 
-%------------------------------------------------
-% POSITION OF TEXT
-%------------------------------------------------
+% improves typesetting in LaTeX
+\usepackage{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
 
-%% Changing position of text on physical page:
-%% Since not all printers position
-%% the printed page in the same place on the physical page,
-%% you can change the position yourself here, if you need to:
+% text layout - change as needed
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 6in 
+\textheight 8.75in
+\renewcommand{\baselinestretch}{1.5}
 
-% \advance\voffset -.5in % Minus dimension will raise the printed page on the 
-                         %  physical page; positive dimension will lower it.
+% use adjustwidth environment to exceed text width (see examples in text)
+\usepackage{changepage}
 
-%% You may set the dimension to the size that you need.
+% adjust caption style
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
 
-%------------------------------------------------
-% GRAPHICS STYLE FILE
-%------------------------------------------------
+% remove brackets from references
+\makeatletter
+\renewcommand{\@biblabel}[1]{\quad#1.}
+\makeatother
 
-%% Requires graphics style file (graphicx.sty), used for inserting
-%% .eps/image files into LaTeX articles.
-%% Note that inclusion of .eps files is for your reference only;
-%% when submitting to PNAS please submit figures separately.
+% headrule, footrule and page numbers
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
 
-%% Type into the square brackets the name of the driver program 
-%% that you are using. If you don't know, try dvips, which is the
-%% most common PC driver, or textures for the Mac. These are the options:
-
-% [dvips], [xdvi], [dvipdf], [dvipdfm], [dvipdfmx], [pdftex], [dvipsone],
-% [dviwindo], [emtex], [dviwin], [pctexps], [pctexwin], [pctexhp], [pctex32],
-% [truetex], [tcidvi], [vtex], [oztex], [textures], [xetex]
-
-\usepackage{float}
-\usepackage{graphicx}
-\usepackage{caption}
-\graphicspath{{../figures/}}
-%------------------------------------------------
-% OPTIONAL POSTSCRIPT FONT FILES
-%------------------------------------------------
-
-%% PostScript font files: You may need to edit the PNASoneF.sty
-%% or PNAStwoF.sty file to make the font names match those on your system. 
-%% Alternatively, you can leave the font style file commands commented out
-%% and typeset your article using the default Computer Modern 
-%% fonts (recommended). If accepted, your article will be typeset
-%% at PNAS using PostScript fonts.
-
-% Choose PNASoneF for one column; PNAStwoF for two column:
-%\usepackage{PNASoneF}
-%\usepackage{PNAStwoF}
-
-%------------------------------------------------
-% ADDITIONAL OPTIONAL STYLE FILES
-%------------------------------------------------
-
-%% The AMS math files are commonly used to gain access to useful features
-%% like extended math fonts and math commands.
-
-\usepackage{amssymb,amsfonts,amsmath}
+% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
 \usepackage{color}
 
-%------------------------------------------------
-% OPTIONAL MACRO FILES
-%------------------------------------------------
+% define custom colors (this one is for figure captions)
+\definecolor{Gray}{gray}{.25}
 
-%% Insert self-defined macros here.
-%% \newcommand definitions are recommended; \def definitions are supported
+% this is required to include graphics
+\usepackage{graphicx}
+\graphicspath{{../figures/}}
 
-%\newcommand{\mfrac}[2]{\frac{\displaystyle #1}{\displaystyle #2}}
-%\def\s{\sigma}
+% use if you want to put caption to the side of the figure - see example in text
+\usepackage{sidecap}
 
-%------------------------------------------------
-% DO NOT EDIT THIS SECTION
-%------------------------------------------------
+% use for have text wrap around figures
+\usepackage{wrapfig}
+\usepackage[pscoord]{eso-pic}
+\usepackage[fulladjust]{marginnote}
+\reversemarginpar
 
-%% For PNAS Only:
-\contributor{Submitted to Proceedings of the National Academy of Sciences of the United States of America}
-\url{www.pnas.org/cgi/doi/10.1073/pnas.0709640104}
-\copyrightyear{2008}
-\issuedate{Issue Date}
-\volume{Volume}
-\issuenumber{Issue Number}
+% document begins here
+\begin{document}
+\vspace*{0.35in}
 
-%----------------------------------------------------------------------------------------
+% title goes here:
+\begin{flushleft}
+{\Large
+\textbf\newline{Is essentiality of genes conserved in Enterobacteriaceae?}
+}
+\newline
+% authors go here:
+\\
+%Fatemeh Ashari Ghomi\textsuperscript{1},
+%Paul P. Gardner\textsuperscript{1,*},
+%Lars Barquist\textsuperscript{2}
+%\\
+%\bigskip
+%\bf{1} University of Canterbury
+%\\
+%\bf{2} Wurzburg University
+%\\
+%\bigskip
+%* paul.gardner@canterbury.ac.nz
 
-\begin{document}
+\end{flushleft}
 
-%----------------------------------------------------------------------------------------
-%	TITLE AND AUTHORS
-%----------------------------------------------------------------------------------------
+\section*{Abstract}
+%Fitting multiple figures into very tight manuscripts while keeping it pleasant to read is challenging.
+Therefore figures are often simply attached to the very end of a manuscript file.
+While easier for the authors, this practice is inconvenient for readers.
+This \LaTeX template shows how to generate a compiled PDF with figures embedded into the text.
+It provides several examples of how to embed figures or tables directly into the text thus giving you a range of options from which you should choose the one best suited for your manuscript.
+Check out Schlegel et al., (2016) as example of use \cite{Schlegel2016}.
+%
+% now start line numbers
+%\linenumbers
 
-\title{Is essentiality of genes conserved in Enterobacteriaceae?} % For titles, only capitalize the first letter
-
-%------------------------------------------------
-
-%% Enter authors via the \author command.  
-%% Use \affil to define affiliations.
-%% (Leave no spaces between author name and \affil command)
-
-%% Note that the \thanks{} command has been disabled in favor of
-%% a generic, reserved space for PNAS publication footnotes.
-
-%% \author{<author name>
-%% \affil{<number>}{<Institution>}} One number for each institution.
-%% The same number should be used for authors that
-%% are affiliated with the same institution, after the first time
-%% only the number is needed, ie, \affil{number}{text}, \affil{number}{}
-%% Then, before last author ...
-%% \and
-%% \author{<author name>
-%% \affil{<number>}{}}
-
-%% For example, assuming Garcia and Sonnery are both affiliated with
-%% Universidad de Murcia:
-%% \author{Roberta Graff\affil{1}{University of Cambridge, Cambridge,
-%% United Kingdom},
-%% Javier de Ruiz Garcia\affil{2}{Universidad de Murcia, Bioquimica y Biologia
-%% Molecular, Murcia, Spain}, \and Franklin Sonnery\affil{2}{}}
-
-\author{Fatemeh Ashari-Ghomi\affil{1}{University of Canterbury},
-Paul P. Gardner\affil{1}{}
-\and
-Lars Barquist\affil{2}{Wurzburg University}}
-
-\contributor{Submitted to Proceedings of the National Academy of Sciences
-of the United States of America}
-
-%----------------------------------------------------------------------------------------
-
-\maketitle % The \maketitle command is necessary to build the title page
-
-\begin{article}
-
-%----------------------------------------------------------------------------------------
-%	ABSTRACT, KEYWORDS AND ABBREVIATIONS
-%----------------------------------------------------------------------------------------
-
-\begin{abstract}
-
-\end{abstract}
-
-%------------------------------------------------
-
-\keywords{essentiality | conservation | transposon insertion} % When adding keywords, separate each term with a straight line: |
-
-%------------------------------------------------
-
-%% Optional for entering abbreviations, separate the abbreviation from
-%% its definition with a comma, separate each pair with a semicolon:
-%% for example:
-%% \abbreviations{SAM, self-assembled monolayer; OTS,
-%% octadecyltrichlorosilane}
-
-% \abbreviations{}
-%\abbreviations{SAM, self-assembled monolayer; OTS, octadecyltrichlorosilane}
-
-%----------------------------------------------------------------------------------------
-%	PUBLICATION CONTENT
-%----------------------------------------------------------------------------------------
-
-%% The first letter of the article should be drop cap: \dropcap{} e.g.,
-%\dropcap{I}n this article we study the evolution of ''almost-sharp'' fronts
-
-\section{Introduction}
-
-\dropcap{S}tudying the essentiality of genes helps with identifying the fundamental processes necessary for cell viability \cite{juhas_essence_2011}.
+% the * after section prevents numbering
+\section*{Introduction}
+Studying the essentiality of genes helps with identifying the fundamental processes necessary for cell viability \cite{juhas_essence_2011}.
 So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}.
@@
 Although the essentiality of genes has been studied in a variety of organisms, there is still room to study the evolutionary conservation of essentiality.
-Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges.
+Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens}.
+Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium and \textit{Salmonella} Typhi.
+In a similar study, Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges.
 We extend this research by studying 13 bacterial strains from Enterobacteriaceae.
@@
 
+\begin{figure}
+\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
+\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in Keio collection \cite{baba_construction_2006}.
+We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
+\label{fig:species-tree}
+\end{figure}
+
+\textcolor{red}{\{A summary of what we have done\}}
+
+\section{Transposon mutagenesis}
+We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
+These strains are all selected from Enterobacteriaceae family.
 Enterobacteriaceae is a family that includes Gram-negative bacteria with different host ranges and pathogenecity found in soil, water, plants, animals and humans \cite{brenner_bergeys_2006}.
@@
 
-\textcolor{red}{\{A summary of what we have done\}}
-
-%------------------------------------------------
-\section{Transposon mutagenesis}
-\subsection{Strains}
-We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
-These strains are all selected from Enterobacteriaceae family.
-\subsection{Transposon insertion workflow}
 We have used Tn5 transposon to generate single inserted mutants and placed our mutants in a selective media for Tn5.
@@
 Then we have sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
-\subsection{Essentiality}
+The number of insertions in a gene can imply the degree of essentiality for that gene.
+
+%\subsection{Essentiality}
 We have used a value called insertion index to evaluate the essentiality of a gene.
@@
 Our experiment has been performed on different strains and the library density is different in each experiment.
-Therefore, in order to make the insertion indices in all the strains comparable, we have normalised our insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
+Therefore, in order to make the insertion indices comparable in all the strains, we have normalised our insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
 We have not studied genes shorter than 100 base-pairs as they might not be targeted by any transposon due to their shortness.
 
-We have divided our genes into three groups: essential genes, non-essential genes, and beneficial losses.
+We have divided the genes into three groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
 We have adapted the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
-The insertion index distribution plot has two peaks and a heavy tail Fig.\@ \ref{fig:iidist}.
+The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
 The first peak shows the genes with no or just a few insertions which are considered as essential genes.
@@
 Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=1, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=2, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=3, scale=0.25]{essentiality.pdf}\\
+\includegraphics[page=4, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=5, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=6, scale=0.25]{essentiality.pdf}\\
+\includegraphics[page=7, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=8, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=9, scale=0.25]{essentiality.pdf}\\
+\includegraphics[page=10, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=11, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=12, scale=0.25]{essentiality.pdf}\\
+&\includegraphics[page=13, scale=0.25]{essentiality.pdf}&\\
+\end{tabular}
+\caption{Plots show the insertion index distribution for each genome.
+The plots are divided into 4 regions using red lines.
+These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
+\label{fig:iidist-species}
+\end{figure*}
+
 \section{Are transposons biased towards certain positions?}
@@
 To overcome this bias, we have normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplying this value by the average insertion index.
-
-We have tested whether our transposons are biased towards certain motifs. For this, we have generated a logo from 10 nucleotides flanking the 100 top most
-frequent insertion sites in each genome.
-The results are depicted in Fig.\@  \ref{fig:logos}.
-The results show a slight bias towards certain combinations of bases.
-In addition, we have investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index.
-The red lines show the loess curve when the smoothness parameter is $0.2$.
-As the figure shows, when G-C content is less than $40\%$, the insertion index is low, however when it is higher than $50\%$, the insertion index is almost constant.
-A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which causes a reduction in the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
-The other reason is that the genes with low G-C content are enriched in mobile genetic elements (corrected p-value for the number of repetition of these words in the quartile with lowest G-C and the interquartile genes: pathogenicity island: 4.55189365569093e-54, prophage: 3.20756138380406e-26, phage: 1.81663788061583e-08, bacteriophage: 0.00285196530662877) compared to the genes with high G-C content and that has caused seeing a different pattern of essentiality in that region.
-
-The other bias that we have considered is the bias towards certain locations in a gene.
-We have divided every gene into 100 bins and calculated the mean insertion index for each bin.
-Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location.
-We have also divided our genes into three groups - essential genes, non-essential genes, and beneficial losses - and studied the bias in each group.
-The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses.
-The case for the non-essential genes is similar to the average (Fig.\@ \ref{fig:insertion-position-bias}).
-High number of insertions at the 3' end of essential genes implies that the functional part of the genes are located before the insertions.
-On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors.
-\textcolor{red}{\{To be tested\}} We have calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes to overcome these biases.
-%\begin{itemize}
-%\item There is a bias towards the position of the gene (Fig.~\ref{fig:distance-bias}).
-%\item There is a negligible bias towards certain motifs (Fig.~\ref{fig:logos})
-%\item The G-C bias needs to be studied further (Fig.~\ref{fig:GC-bias})
-%\item The number of insertions on the 3' and 5' ends is more than the internal region in essential genes and less than the internal region in beneficial losses.
-(Fig.~\ref{fig:insertion-position-bias})
-%\end{itemize}
-\section{Essentiality and conservation}
-\subsection{Homolog clustering algorithm (homclust)}
-To study whether each gene in our 13 organisms is conserved we have proposed a program that clusters homologous proteins.
-This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
-It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
-Then it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
-In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by relaxing Jackhmmer parameters.
-Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
-The program is summarised in Fig.\@ \ref{fig:homclust} and its results are illustrated in Fig.\@ \ref{fig:homclust-results}.
-\subsection{Gene classes}
-%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
-Plenty of methods are proposed for this purpose.
-Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
-We have used Hieranoid to cluster the sets of orthologous genes.
-In addition, we intended to study the essentiality of genes in paralogous genes.
-For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
-To study the conservation of genes, we have used homclust and divided the clusters of homologous genes into three groups.
-Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
-These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
-
-We have divided the clusters of orthologous genes into 4 groups based on their essentiality (essential, ambiguous, non-essential, and beneficial loss) and 3 groups based on their conservation (genus specific, single copy, and multi-copy).
-The results are depicted in Fig.\@ \ref{fig:iidist}.
-The figure shows that most of the essential clusters are single copy and most of the beneficial losses are genus specific.
-\textcolor{red}{\{KEGG results\}}
-\subsection{Evolution of essentiality}
-We have compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
-The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show a tree-like signal.
-\begin{itemize}
-\item UpSetR results (Fig.~\ref{fig:upsetr})
-\item Stringent
-\item Dollo law
-\item Ancestral insertion index
-\end{itemize}
-\section{Case study of genes}
-\subsection{Core genes}
-\subsection{Accessory genes}
-
-%%------------------------------------------------
-%
-%\section{Discussion}
-%
-%
-%
-%%----------------------------------------------------------------------------------------
-%%	MATERIALS AND METHODS
-%%----------------------------------------------------------------------------------------
-%
-%%% Optional Materials and Methods Section
-%%% The Materials and Methods section header will be added automatically.
-%
-%\begin{materials}
-
-%\end{materials}
-
-%----------------------------------------------------------------------------------------
-%	APPENDICES (OPTIONAL)
-%----------------------------------------------------------------------------------------
-
-%\appendix
-%An appendix without a title.
-%
-%\appendix[Appendix title]
-%An appendix with a title.
-
-%----------------------------------------------------------------------------------------
-%	ACKNOWLEDGEMENTS
-%----------------------------------------------------------------------------------------
-
-\begin{acknowledgments}
-
-\end{acknowledgments}
-
-%----------------------------------------------------------------------------------------
-%	BIBLIOGRAPHY
-%----------------------------------------------------------------------------------------
-
-%% PNAS does not support submission of supporting .tex files such as BibTeX.
-%% Instead all references must be included in the article .tex document. 
-%% If you currently use BibTeX, your bibliography is formed because the 
-%% command \verb+\bibliography{}+ brings the <filename>.bbl file into your
-%% .tex document. To conform to PNAS requirements, copy the reference listings
-%% from your .bbl file and add them to the article .tex file, using the
-%% bibliography environment described above.
-
-%%  Contact pnas@nas.edu if you need assistance with your
-%%  bibliography.
-
-% Sample bibliography item in PNAS format:
-%% \bibitem{in-text reference} comma-separated author names up to 5,
-%% for more than 5 authors use first author last name et al. (year published)
-%% article title  {\it Journal Name} volume #: start page-end page.
-%% ie,
-% \bibitem{Neuhaus} Neuhaus J-M, Sitcher L, Meins F, Jr, Boller T (1991) 
-% A short C-terminal sequence is necessary and sufficient for the
-% targeting of chitinases to the plant vacuole. 
-% {\it Proc Natl Acad Sci USA} 88:10362-10366.
-
-%% Enter the largest bibliography number in the facing curly brackets
-%% following \begin{thebibliography}
-
-%\begin{thebibliography}{100}
-%
-%\end{thebibliography}
-\bibliography{references}{}
-\bibliographystyle{unsrt}
-
-%----------------------------------------------------------------------------------------
-
-\end{article}
-
-%----------------------------------------------------------------------------------------
-%	FIGURES AND TABLES
-%----------------------------------------------------------------------------------------
-\begin{figure}
-\includegraphics[scale=0.3]{phylosift-aa-raxmlbootstrap.pdf}
-\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in Keio collection \cite{baba_construction_2006}.
-We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
-\label{fig:species-tree}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.7]{homclust.pdf}
-\caption{The steps of our proposed algorithm for clustering homologous genes.}
-\label{fig:homclust}
-\end{figure}
-
-\begin{figure}
-\includegraphics[scale=0.7]{cluster-size-dist.pdf}
-\caption{Size distribution for all clusters of homologous genes.
-Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
-\label{fig:homclust-results}
-\end{figure}
 
@@
 
+We have tested whether our transposons are biased towards certain motifs.
+For this, we have generated a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
+The results are depicted in Fig.\@  \ref{fig:logos}.
+The results show a slight bias towards certain combinations of bases.
+In addition, we have investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index Fig.\@  \ref{fig:GC-bias}.
+The red lines show the loess curve when the smoothness parameter is $0.2$.
+As the figure shows, when G-C content is less than $40\%$, the insertion index is low, however when it is higher than $50\%$, the insertion index is almost constant.
+A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which causes a reduction in the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
+The other reason is that the genes with low G-C content are enriched in mobile genetic elements (corrected p-value for the number of repetition of these words in the quartile with lowest G-C and the interquartile genes: pathogenicity island: 4.55189365569093e-54, prophage: 3.20756138380406e-26, phage: 1.81663788061583e-08, bacteriophage: 0.00285196530662877) compared to the genes with high G-C content and that has caused seeing a different pattern of essentiality in that region.
+
 \begin{figure*}
 \begin{tabular}{c c}
-\includegraphics[scale=0.6]{100logo-bits.pdf}&
-\includegraphics[scale=0.6]{100logo-prob.pdf}
+\includegraphics[scale=0.55]{100logo-bits.pdf}&
+\includegraphics[scale=0.55]{100logo-prob.pdf}
 \end{tabular}
@@
 
+The other bias that we have considered is the bias towards certain locations in a gene.
+We have divided every gene into 100 bins and calculated the mean insertion index for each bin.
+Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location.
+We have also studied the bias in each of the groups: essential genes, non-essential genes, and beneficial losses.
+The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses.
+The case for the non-essential genes is similar to the average (Fig.\@ \ref{fig:insertion-position-bias}).
+High number of insertions at the 3' end of essential genes implies that the functional part of the genes are located before the insertions.
+On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors.
+\textcolor{red}{\{To be tested\}} We have calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes to overcome these biases.
+The insertion index distribution for each genome after correcting for all distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
+
 \begin{figure*}
 \begin{tabular}{c c}
-\includegraphics[scale=0.5, page=1]{insertion-position-bias.pdf}&
-\includegraphics[scale=0.5, page=2]{insertion-position-bias.pdf}\\
-\includegraphics[scale=0.5, page=3]{insertion-position-bias.pdf}&
-\includegraphics[scale=0.5, page=4]{insertion-position-bias.pdf}
+\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
+\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
+\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
+\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
 \end{tabular}
-\caption{We have divided our genes into 3 segments: 5\% of
+\caption{We have divided the genes into 3 segments: 5\% of
 the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle.}
@@
 
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=1, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=2, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=3, scale=0.25]{essentiality-normalised.pdf}\\
+\includegraphics[page=4, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=5, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=6, scale=0.25]{essentiality-normalised.pdf}\\
+\includegraphics[page=7, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=8, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=9, scale=0.25]{essentiality-normalised.pdf}\\
+\includegraphics[page=10, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=11, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=12, scale=0.25]{essentiality-normalised.pdf}\\
+&\includegraphics[page=13, scale=0.25]{essentiality-normalised.pdf}&\\
+\end{tabular}
+\caption{Plots show the insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes.
+The plots are divided into 4 regions using red lines.
+These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
+\label{fig:iidist-species-normalised}
+\end{figure*}
+%\begin{itemize}
+%\item There is a bias towards the position of the gene (Fig.~\ref{fig:distance-bias}).
+%\item There is a negligible bias towards certain motifs (Fig.~\ref{fig:logos})
+%\item The G-C bias needs to be studied further (Fig.~\ref{fig:GC-bias})
+%\item The number of insertions on the 3' and 5' ends is more than the internal region in essential genes and less than the internal region in beneficial losses.
+(Fig.~\ref{fig:insertion-position-bias})
+%\end{itemize}
+\section{Essentiality and conservation}
+To study whether each gene in our 13 organisms is conserved we have proposed a program that clusters homologous proteins.
+This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
+It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
+Then it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
+In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by relaxing Jackhmmer parameters.
+Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
+The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 13 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
+
+\begin{figure}
+\includegraphics[scale=0.2]{homclust.pdf}
+\caption{The steps of our proposed algorithm for clustering homologous genes.}
+\label{fig:homclust}
+\end{figure}
+
+\subsection{Gene classes}
+%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
+Plenty of methods are proposed for this purpose.
+Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
+We have used Hieranoid to cluster the sets of orthologous genes.
+In addition, we intended to study the essentiality of genes in paralogous genes.
+For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
+To study the conservation of genes, we have used homclust and divided the clusters of homologous genes into three groups.
+Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
+These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
+
+\begin{figure}
+\includegraphics[scale=0.3]{cluster-size-dist.pdf}
+\caption{Size distribution for all clusters of homologous genes.
+Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
+\label{fig:homclust-results}
+\end{figure}
+
+We have divided the clusters of orthologous genes into 4 groups based on their essentiality (essential, ambiguous, non-essential, and beneficial loss) and 3 groups based on their conservation (genus specific, single copy, and multi-copy).
+The results are depicted in Fig.\@ \ref{fig:iidist}.
+The figure shows that most of the essential clusters are single copy and most of the beneficial losses are genus specific.
+
 \begin{figure}
 \centering
-\includegraphics[scale=1, page=1]{cluster-essentiality.pdf}
+\includegraphics[scale=0.5]{cluster-essentiality.pdf}
 \caption{The genes have been clustered into orthologous groups using Hieranoid and paralogous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
@@
 
+\textcolor{red}{\{KEGG results\}} To study which functions are enriched in each class of essentiality, we have gathered the``note" section for each gene from their embl files.
+Then, we have counted the repeat number of each word for the genes in each class and the genes that do not belong to that class and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
+The P-values are then corrected using Benjamini-Hochberg-Yekutieli procedure.
+Fig.\@ \ref{fig:essentiality-pval} shows the top 20 enriched words for each essentiality class.
+The results show an enrichment of the genes related to replication, transcription, translation, division, and rod shape determining proteins in essential class.
+The non-essential genes are mostly transport proteins, flagellar proteins, ATPase, and DNA repair proteins.
+Beneficial losses are enriched in transposase enzymes, putative and hypothetical proteins, and mobile elements.
+Beneficial losses also contain many fimbrial proteins which probably has occurred because these proteins are not needed in a rich lab medium \textcolor{red}{\{TRUE?\}}.
+%We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}.
+It shows that most of the beneficial losses are mobile elements.
+Most of these mobile elements are not included in KEGG database.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[scale=0.4]{essential-pval.pdf}&
+\includegraphics[scale=0.4]{non-essential-pval.pdf}\\
+\includegraphics[scale=0.4]{beneficialloss-pval.pdf}&
+\end{tabular}
+\caption{The word enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
+The red line shows P-value = 0.05.
+The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
+\label{fig:essentiality-pval}
+\end{figure*}
+
+\subsection{The evolution of essentiality}
+%We have compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
+The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show a tree-like signal.
+%\begin{itemize}
+%\item UpSetR results (Fig.~\ref{fig:upsetr})
+%\item Stringent
+%\item Dollo law
+%\item Ancestral insertion index
+%\end{itemize}
+In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
+If a gene is essential in a genome, its corresponding cell in the matrix has value 1 and 0 otherwise.
+Then we have used the Bray-Curtis distance to generate a distance matrix for the species under study.
+Finally, a tree has been generated using PHYLIP Neighbor.
+The resulting tree is not similar to the species tree.
+
 \begin{figure}
 \centering
-\includegraphics[scale=0.65, page=1]{upsetr.pdf}\\
-\includegraphics[scale=0.65, page=2]{upsetr.pdf}
+\includegraphics[scale=0.2]{neighbor-joining-essentiality-tree.pdf}
+\caption{The tree is generated from all the genes that are copied only once per genome.
+We have made a binary matrix from the essentiality of these genes.
+Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree using PHYLIP Neighbor which uses a neighbour joining method.}
+\label{fig:neighborjoining}
+\end{figure}
+
+We have also compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
+For this, we have counted the number of genes that are core between every combination of our strains, and the number of genes that are core essential between those combinations.
+We have used UpSetR package \cite{conway_upsetr:_2016} in R to visualise the results in Fig.\@ \ref{fig:upsetr}.
+As shown in the figures, among 1908 genes that are core between all the strains under study, only 184 are core essential.
+The results propose that although conservation of genes follows a tree-like trend (800 genes are core in Klebsiellas, 273 genes are core in Salmonellas, 208 genes are core in Klebsiellas and Enterobacter, ...), the essentiality does not show a tree-like signal and most of the large sets of core essential genes belong to only one strain.
+We believe this happens due to the small number of essential genes.
+
+\begin{figure*}
+\centering
+\includegraphics[scale=0.6, page=1]{upsetr.pdf}\\
+\includegraphics[scale=0.6, page=2]{upsetr.pdf}
 \caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.}
 \label{fig:upsetr}
+\end{figure*}
+
+To test if essential genes are more likely to be conserved, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
+For the essential genes to be more likely to be conserved, this ratio should increase as we go up in the tree.
+We have used three different methods for this.
+The first method is intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are present in all the strains in its child nodes and are core essential if and only if they are essential in all the strains in its child nodes.
+
+The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
+In this method, we have averaged over the insertion indices of the pair of closest children of the node that we are calculating the core genes for.
+We have repeated this and averaged the averages until we reached the node under study.
+Then, we have plotted the insertion indices and fitted an exponential and a gamma distribution to the plot and found the essential genes at that level.
+
+The third method is using Dollo law to define core genes and core essential genes.
+This method, assumes that the gain of genes (essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (essentiality) and minimise the number of times that the genes (the essentiality) has been lost.
+Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
+The numbers predicted using these three methods are shown in Fig.\@ \ref{fig:annotated-speciestree}.
+As this figure shows, the ratio between core essential and core genes is almost constant using the intersection and dollo method; however, this ratio increases as we go higher in the tree using the ancestral insertion index method.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap-annotated.pdf}
+\caption{The tree shows the species tree in Fig.\@\ref{fig:species-tree} annotated by the number of core essential genes and core genes at each node.
+We have used three methods for defining core essential genes and core genes.
+The numbers at the leaves are the same using all these three methods.
+At the internal nodes, red shows the numbers using the intersection method, green shows the ancestral insertion index method, and turquoise shows fdollop method.}
+\label{fig:annotated-speciestree}
 \end{figure}
 
-%----------------------------------------------------------------------------------------
+%\section*{Acknowledgments}
+%
+%
+%\nolinenumbers
+
+%This is where your bibliography is generated. Make sure that your .bib file is actually called library.bib
+\bibliography{references}
+
+%This defines the bibliographies style. Search online for a list of available styles.
+\bibliographystyle{abbrv}
 

### 2016-10-01 bin/adopt/duplication_distance_calculator.py

@@ -1,0 +1,126 @@
+#! /bin/sh
+""":"
+exec python3 $0 ${1+"$@"}
+"""
+
+from Bio import Phylo
+from os import listdir, system, path, mkdir, remove, stat
+from shutil import rmtree
+from argparse import ArgumentParser
+from re import match
+from csv import reader
+
+def makedir(dirname):
+    # print ('Making directory \'{0}\'..'.format(dirname))
+    if path.exists(dirname):
+        input_var = 'i'
+        while not (input_var == 'y' or input_var == 'Y' or input_var == 'n' or input_var == 'N'):
+            input_var = input('Directory {0} already exists. Replace? [Y/n] '.format(dirname))
+        if input_var == 'Y' or input_var == 'y':
+            rmtree(dirname)
+        else:
+            raise SystemExit
+    mkdir(dirname)
+
+parser = ArgumentParser(description='Plots distance vs. delta insertion index')
+parser.add_argument('msadir', help='Directory of nucleotide multiple sequence alignments (pal2nal)')
+parser.add_argument('iidir', help='Directory of clusters with insertion indices (merge-clust-plot output)')
+parser.add_argument('distdir', help='Directory for saving the distances between sequences')
+parser.add_argument('deltaiidir', help='Directory for saving the distance between insertion indices')
+args = parser.parse_args()
+msadir = args.msadir
+iidir = args.iidir
+distdir = args.distdir
+deltaiidir = args.deltaiidir
+
+makedir(distdir)
+makedir(deltaiidir)
+
+list_of_files = listdir(msadir)
+for filename in list_of_files:
+    clust_size = 0
+    with open(iidir + '/' + filename[:-8] + '.txt') as iifile:
+        for line in iifile:
+            clust_size += 1
+    if clust_size >= 23:
+        if stat('{0}/{1}'.format(msadir, filename)).st_size:
+            try:
+                system('fdnadist -sequence {0}/{1} -method F84 -outfile {2}/{3}.temp'.format(msadir, filename, distdir, path.splitext(filename)[0]))
+            except:
+                raise SystemExit
+            with open('{0}/{1}'.format(msadir, filename)) as msafile:
+                with open('{0}/{1}.temp'.format(distdir, path.splitext(filename)[0])) as distfile_from:
+                    with open('{0}/{1}.txt'.format(distdir, path.splitext(filename)[0]), 'w') as distfile_to:
+                        flag = 1
+                        distfile_from.readline()
+                        while True:
+                            msaline = msafile.readline()
[…]

### 2016-10-01 bin/adopt/plot-distance-vs-deltaii.R

@@ -1,0 +1,77 @@
+require(ggplot2)
+require(stringr)
+# require(dplyr)
+diipath <- "../results/duplication_distance_calculator/deltaiis/"
+dspath <- "../results/duplication_distance_calculator/distances/"
+list_of_files <- list.files(path=dspath, pattern="*.txt", full.names=FALSE, recursive=FALSE)
+dii <- c()
+ds <- c()
+genes <- c()
+for (filename in list_of_files)
+{
+  dii_tbl <- read.table(paste(diipath,'/',filename,sep = ''), sep = '\t', header = FALSE)
+  ds_tbl <- read.table(paste(dspath,'/',filename,sep = ''), sep = '\t', header = FALSE)
+  if (nrow(dii_tbl) > 1)
+  {
+    for (i in seq(1,(nrow(dii_tbl)-1)))
+    {
+      for (j in seq((i+2), ncol(dii_tbl)))
+      {
+        genei = str_match(dii_tbl[i,1], "([[:graph:]]*)\\_[[:digit:]]+\\-[[:digit:]]+")[2]
+        genomei = str_match(genei, "([[:alnum:]]*)\\_[[:graph:]]+")[2]
+        if (is.na(genomei))
+        {
+          genomei = str_match(genei, "([[:alpha:]]*)[[:digit:]]+")[2]
+        }
+        genej = str_match(dii_tbl[j-1,1], "([[:graph:]]*)\\_[[:digit:]]+\\-[[:digit:]]+")[2]
+        genomej = str_match(genej, "([[:alnum:]]*)\\_[[:graph:]]+")[2]
+        if (is.na(genomej))
+        {
+          genomej = str_match(genej, "([[:alpha:]]*)[[:digit:]]+")[2]
+        }
+        if (dii_tbl[i,j] != -1 & genomei  == genomej)
+        {
+          dii <- c(dii, dii_tbl[i,j])
+          ds <- c(ds, ds_tbl[i,j])
+          genes <- rbind(genes, c(genei, genej))
+        }
+      }
+    }
+  }
+}
+
+pdf("../results/distance-vs-deltaii.pdf")
+
+# plot(ds, dii, ylim=c(0,2))
+# lines(lowess(ds, dii, f = .2), col = 2, lwd=5)
+# regln <- lm(dii~ds)
+# abline(regln, col="red")
+
+ds_quartiles <- c()
+thresholds <- quantile(ds)
+for (value in ds)
+{
+  if (value < thresholds[2])
+  {
[…]

### 2016-10-01 bin/define-core-accessory-hieranoid.py

@@ -9,10 +9,12 @@ from shutil import rmtree
 from re import match, findall
 from collections import defaultdict
 
+
 def read_fasta_sequences(filepath):
     with open(filepath, 'rU') as fasta_file:
         fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
     return fasta_dict
+
 
 def makedir(dirname):
     if path.exists(dirname):
@@ -25,6 +27,7 @@ def makedir(dirname):
             raise SystemExit
     mkdir(dirname)
 
+
 def read_gene_essentiality(indir):
     list_of_files = listdir(indir)
     iidict = {}
@@ -35,6 +38,7 @@ def read_gene_essentiality(indir):
                 iidict[cells[0]] = cells[2]
     return iidict
 
+
 def read_k12(inpath, iidict):
     with open(inpath) as from_file:
         for line in from_file:
@@ -43,7 +47,7 @@ def read_k12(inpath, iidict):
     return iidict
 
 seqdb = '/home/fatemeh/EnTrI/data/fasta-protein/chromosome/seqdb.fasta'
-clusters = '/home/fatemeh/EnTrI/results/hieranoid/hieranoid-result.txt'
+clusters = '/home/fatemeh/EnTrI/results/hieranoid/clusters.txt'
 insertion_indices = '/home/fatemeh/EnTrI/results/insertion-indices/normalised-insertion-indices'
 k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
 outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid'
@@ -60,6 +64,9 @@ species_names = {"all":["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441"
     "klebsiellaenterobacter":["ERS227112", "BN373", "ENC"], "salmonellaty2":["t"], "salmonellap125109":["SEN"],
     "salmonellasl1344":["SL1344"], "salmonellasl3261":["SL3261"], "salmonellaa130":["STM"], "salmonellad23580":["STMMW"], "ecolist131":["NCTC13441"],
     "ecolics17":["CS17"], "ecolih10407":["ETEC"], "ecolik12":["b"], "klebsiellarh201207":["ERS227112"], "klebsiellaecl8":["BN373"]}
+
+with open('/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid/info.txt', 'w') as infofile:
[…]

### 2016-10-01 bin/upsetr-hieranoid.R

@@ -3,7 +3,7 @@ library(stringr)
 # listinput <- list(a=c(1,1,1,1,2,1,2,3,3,3), b=c(1,4,5), d=c(1,2,5,6,7), e=c(4,8,9))
 # upset(fromList(listinput))
 
-clusters_path <- "../results/hieranoid/hieranoid-result.txt"
+clusters_path <- "../results/hieranoid/clusters.txt"
 essentiality_path <- "../results/insertion-indices/normalised-insertion-indices/"
 k12_path <- "../results/ecogene-k12.txt"
 list_of_files <- list.files(path=essentiality_path, full.names=T, recursive=FALSE)
@@ -24,17 +24,20 @@ numclusters = length(cluster)
 for (clusterindex in (1:numclusters))
 {
   line = cluster[clusterindex]
-  match_result = str_match_all(line, '[[,(]](([[:alnum:]]+)_[[:alnum:]]+):')
-  temp_result = str_match_all(line, '[[,(]](([[:alpha:]]+)[[:digit:]]+):')
-  match_result[[1]] <- rbind(match_result[[1]], temp_result[[1]])
-  match_result <- match_result[[1]]
-  for (i in (1:nrow(match_result)))
+  genes = unlist(strsplit(line , "\t"))
+  match_result = str_match_all(line, '([[:alnum:]]+_|[[:alpha:]]+)[[:alnum:]]+')
+  species = gsub('_','',match_result[[1]][,2])
+  # match_result = str_match_all(line, '[[,(]](([[:alnum:]]+)_[[:alnum:]]+):')
+  # temp_result = str_match_all(line, '[[,(]](([[:alpha:]]+)[[:digit:]]+):')
+  # match_result[[1]] <- rbind(match_result[[1]], temp_result[[1]])
+  # match_result <- match_result[[1]]
+  for (i in (1:length(species)))
   {
-    name = match_result[i,3]
+    name = species[i]
     if (name %in% names(presence)) 
     {
       presence[[name]] <- rbind(presence[[name]], clusterindex)
-      gene = match_result[i,2]
+      gene = genes[i]
       if (gene %in% list_of_essential_genes[[name]])
       {
         essentiality[[name]] <- rbind(essentiality[[name]], clusterindex)
@@ -54,8 +57,8 @@ names(essentiality) <- rev(c("Klebsiella pneumoniae Ecl8", "Klebsiella pneumonia
[…]

### 2016-10-03 documents/manuscript.tex
Added KEGG pathway enrichment analysis

@@
 \setlength{\parindent}{0.5cm}
-\textwidth 6in 
-\textheight 8.75in
+\textwidth 6.5in 
+\textheight 9in
 \renewcommand{\baselinestretch}{1.5}
@@
 A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which causes a reduction in the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
-The other reason is that the genes with low G-C content are enriched in mobile genetic elements (corrected p-value for the number of repetition of these words in the quartile with lowest G-C and the interquartile genes: pathogenicity island: 4.55189365569093e-54, prophage: 3.20756138380406e-26, phage: 1.81663788061583e-08, bacteriophage: 0.00285196530662877) compared to the genes with high G-C content and that has caused seeing a different pattern of essentiality in that region.
+The other reason is that the genes with low G-C content are enriched in mobile genetic elements compared to the genes with average G-C content (\ref{fig:gc-pval}) and this has caused seeing a different pattern of essentiality in that region.
 
@@
 \label{fig:GC-bias}
+\end{figure*}
+
+\begin{figure*}
+\includegraphics[scale=0.55]{lowgc-pval.pdf}
+\caption{Word enrichment analysis for low G-C genes compared to genes with interquartile G-C level.
+The red line shows P-value = 0.05.
+The P-values have been calculated using Fisher's exact test and corrected using Benjamini-Hochberg-Yekutieli.}
+\label{fig:gc-pval}
 \end{figure*}
@@
 
-\textcolor{red}{\{KEGG results\}} To study which functions are enriched in each class of essentiality, we have gathered the``note" section for each gene from their embl files.
+To study which functions are enriched in each class of essentiality, we have gathered the``note" section for each gene from their embl files.
 Then, we have counted the repeat number of each word for the genes in each class and the genes that do not belong to that class and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
@@
 
+We have also conducted a pathway enrichment analysis for these three groups.
+For this, we have downloaded pathway datasets for strains that were available in KEGG database.
+This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.
+Then we have merged these databases and used a hypergeometric test to find which pathways are enriched in each essentiality class.
+Finally, we have corrected the P-values using Benjamini-Hochberg-Yekutieli.
+The results are depicted in ~\ref{fig:essentiality-pathway}.
+
 \begin{figure*}
@@
 \end{tabular}
-\caption{The word enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
+\caption{Word enrichment analysis for essential genes, non-essential genes, and beneficial losses compared to other genes.
 The red line shows P-value = 0.05.
@@
 \label{fig:essentiality-pval}
+\end{figure*}
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[scale=0.4]{essential-pathways.pdf}&
+\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
+\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
+\end{tabular}
+\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
+The red line shows P-value = 0.05.
+The P-values have been calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
+\label{fig:essentiality-pathway}
 \end{figure*}

### 2016-10-03 bin/pathanalysis.R

@@ -1,7 +1,39 @@
 library(ggplot2)
-db <- as.matrix(read.table("~/EnTrI/results/KEGG/cirobacter_rodentium_ICC168.dat", header=TRUE))
-genes <- as.matrix(read.table("~/EnTrI/results/KEGG/ROD-essential.txt", header=TRUE))
-enrichment = data.frame(matrix(nrow=length(unique(db[,2])),ncol=4),row.names =unique(db[,2]) )
+library(stringr)
+dbs <- "~/EnTrI/results/KEGG/"
+essentiality <- "~/EnTrI/results/insertion-indices/normalised-insertion-indices/"
+list_of_files <- list.files(path=dbs, full.names=T, recursive=FALSE)
+db=matrix(,nrow=0,ncol=3)
+for (filename in list_of_files)
+{
+  dbfile = as.matrix(read.table(filename, as.is=TRUE, header=TRUE, sep="\t"))
+  for (i in seq(1,nrow(dbfile)))
+  {
+    dbfile[i,3] = str_match(dbfile[i,3], '([[:print:]]+)-[[ Citrobacter rodentium]]|[[ Enterobacter cloacae subsp. cloacae NCTC 9394]]|
+                            [[ Escherichia coli O78:H11:K80 H10407 (ETEC)]]|[[ Escherichia coli K\\-12 MG1655]]|
+                            [[ Salmonella enterica subsp. enterica serovar Enteritidis P125109]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhimurium D23580]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhimurium SL1344]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhi Ty2]]')[2]
+    
+  }
+  db = rbind(db,dbfile)
+}
+list_of_files <- list.files(path=essentiality, full.names=T, recursive=FALSE)
+genes = c()
+for (filename in list_of_files)
+{
+  essentialityfile = as.matrix(read.table(filename, as.is=TRUE, header=TRUE))
+  for (i in seq(1,nrow(essentialityfile)))
+  {
+    if (essentialityfile[i,3] == "non-essential")
+    {
+      genes=c(genes,essentialityfile[i,1])
+    }
+  }
+}
+
[…]

### 2016-10-03 bin/study-low-gc.R

@@ -3,10 +3,22 @@ codon_table <- read.table(codons, header = TRUE)
 fishertable =data.frame()
 for (i in seq(1,(nrow(codon_table)-1)))
 {
-  fisherresult = fisher.test(matrix(c(codon_table[i,2],codon_table[i,3],codon_table[nrow(codon_table),2],codon_table[nrow(codon_table),3]),nrow=2,byrow=TRUE), alternative = "greater")
+  fisherresult = fisher.test(matrix(c(codon_table[i,2],codon_table[i,3],(codon_table[nrow(codon_table),2]-codon_table[i,2]),
+                                      (codon_table[nrow(codon_table),3]-codon_table[i,3])),nrow=2,byrow=TRUE), alternative = "greater")
   fishertable[i,1] = codon_table[i,1]
   fishertable[i,2] = fisherresult$p.value
 }
 fishertable[,3] = p.adjust(fishertable[,2], method = "BH")
 fishertable = fishertable[order(fishertable[,3], fishertable[,2]),]
 write.table(fishertable, file = "../results/low-gc-pval.txt", quote = FALSE, col.names = c("word", "p-value", "corrected-p-value"), row.names = FALSE, sep="\t")
+
+df = fishertable[-seq(25,nrow(fishertable),1),]
+df[,3] = -log10(df[,3])
+df = df[,-2]
+df = df[c(-2,-3,-8,-9),]
+names(df) = c("word", "pvalue")
+pdf("../figures/lowgc-pval.pdf")
+ggplot(data=df, aes(x=reorder(word,pvalue),y=pvalue))+geom_bar(stat="identity")+ coord_flip()+labs(x='Word',y='-log10(P-value)') +
+  geom_abline(slope = 0, intercept = -log10(0.05), color="red")+ylim(c(0,max(df[1,2], -log10(0.05))))+ggtitle("Low G-C genes") +
+  theme(text = element_text(size=15))
+dev.off()

### 2016-10-04 documents/manuscript.tex
Added some comments to manuscript. (II) - stupid git.

@@
 {\Large
-\textbf\newline{Is essentiality of genes conserved in Enterobacteriaceae?}
+\textbf\newline{Is the essentiality of genes conserved in Enterobacteriaceae?}
+\newline{Is gene essentiality conserved?}
+\newline{Are essential genes conserved?}
 }
@@
 \section{Transposon mutagenesis}
+
+{\bf overview paragraph.}
+
 We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
@@
 
-\section{Are transposons biased towards certain positions?}
+%\section{Are transposons biased towards certain positions?}
+\section{Are there biases in transposon mutagenesis data?}
+
+{\bf OVERVIEW PARAGRAPH:}
+
 Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}.
-We have performed a thorough study of these biases.
-To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome.
+We have performed a detailed study of these biases.
+{\bf LIST TYPES OF BIAS HERE.}
+
+{\bf ORIGIN OF REPLICATION BIAS:} To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome.
 Fig.\@ \ref{fig:distance-bias} shows the results.
@@
 To overcome this bias, we have normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplying this value by the average insertion index.
+
+{\bf mention possible confounding factor for this study.
+Essential genes may (are???) non-uniformly distributed in the genome (I presume they are clustered near the origin).
+This would be expected to lower the insertion index near the origin, however, the number of non-essential genes is $\approx 10\times$ the number of essential genes, therefore the signature of biased numbers of insertions near the origin remains.
+}
 
@@
 
-We have tested whether our transposons are biased towards certain motifs.
+{\bf PREFERRED INSERTION MOTIF BIAS:} We have tested whether our transposons are biased towards certain motifs.
 For this, we have generated a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
@@
 
+\begin{itemize}
+\item model H-NS binding sites? CGWTWHWww Lang et al (2007)
+\item seems unlikely -- show bulk of genes are around 50\% G+C (add box-whisker plots to scatter diagrams?)
+\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? 
+\end{itemize}
+
 \begin{figure*}
@@
 \end{tabular}
-\caption{We have generated the logos from 10 nucleotides flanking the 100 top most frequent insertion sites.}
+\caption{We have generated sequence logos plots using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
+On the left the height of each character corresponds to a bit score for that character (i.e.
+$\log(\frac{f^{obs}_{i}}{f^{exp}_{i}}$, where espected frequencies ($f^{exp}_{i}$) are derived from ?????).
+On the right ...
+}
 \label{fig:logos}

### 2016-10-05 documents/manuscript.tex
annotated upsetr results with cross and tick marks, made small changes to manuscript

@@
 \end{tabular}
-\caption{The plots show the position of the genes within the genome (normalised by the lengths of the genomes) versus the insertion indices of the genes}
+\caption{The plots show the distance of the genes from DnaA gene normalised by the lengths of the genomes versus the insertion indices of the genes.
+The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
+The red curves show the loess curve when the smoothness parameter is $0.2$.}
 \label{fig:distance-bias}
@@
 \end{tabular}
-\caption{We have generated sequence logos plots using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
+\caption{We have generated sequence logo plots using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
 On the left the height of each character corresponds to a bit score for that character (i.e.
-$\log(\frac{f^{obs}_{i}}{f^{exp}_{i}}$, where espected frequencies ($f^{exp}_{i}$) are derived from ?????).
-On the right ...
-}
+$2-\sum f_a\times\log_2f_a-\frac{1}{\ln2}\times\frac{3}{2\times n}$, where $f_a$ is the relative frequency of base $a$ and $n$ is the number of sequences).
+To put it in simple words, the height of the set of characters shows how biased that position is and the height of each character shows the amount of bias towards that character.
+On the right the height of each character shows the relative frequency of that character.}
 \label{fig:logos}
@@
 \end{tabular}
-\caption{The plots show the G-C contents of the genes (normalised by the lengths of the genes) against their insertion indices}
+\caption{The plots show the ratio of G-C bases in the genes normalised by the lengths of the genes against their insertion indices.
+The red curves show the loess curve when the smoothness parameter is $0.2$.}
 \label{fig:GC-bias}
@@
 \end{tabular}
-\caption{We have divided the genes into 3 segments: 5\% of
-the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle.}
+\caption{The plots show the average insertion index in over the genes in all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
+Each bin shows the average insertion index for $1\%$ of the genes.
+The genes have been divided into 3 segments: 5\% of
+the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
 \label{fig:insertion-position-bias}
@@
 %\end{itemize}
-In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
+%In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
 If a gene is essential in a genome, its corresponding cell in the matrix has value 1 and 0 otherwise.
@@
 The resulting tree is not similar to the species tree.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.2]{neighbor-joining-essentiality-tree.pdf}
-\caption{The tree is generated from all the genes that are copied only once per genome.
+%
+%\begin{figure}
+%\centering
+%\includegraphics[scale=0.2]{neighbor-joining-essentiality-tree.pdf}
+%\caption{The tree is generated from all the genes that are copied only once per genome.
 We have made a binary matrix from the essentiality of these genes.
 Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree using PHYLIP Neighbor which uses a neighbour joining method.}
-\label{fig:neighborjoining}
-\end{figure}
+%\label{fig:neighborjoining}
+%\end{figure}
 
-We have also compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
+In order to test if the essentiality of genes follows a tree-like trend, we have compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
 For this, we have counted the number of genes that are core between every combination of our strains, and the number of genes that are core essential between those combinations.
@@
 \centering
-\includegraphics[scale=0.6, page=1]{upsetr.pdf}\\
-\includegraphics[scale=0.6, page=2]{upsetr.pdf}
-\caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.}
+\includegraphics[scale=0.6]{upsetr1-edited.pdf}\\
+\includegraphics[scale=0.6]{upsetr2-edited.pdf}
+\caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.
+The bars show the number of genes that are core between the strains marked with black circles.
+The tick marks show phylogenetically informative columns and the cross marks show non-informative columns.}
 \label{fig:upsetr}

### 2016-10-11 documents/manuscript.tex
Added the difference between different core-essential definition methods, added the difference core-essential genes in root and branches

@@
 {\Large
-\textbf\newline{Is the essentiality of genes conserved in Enterobacteriaceae?}
-\newline{Is gene essentiality conserved?}
-\newline{Are essential genes conserved?}
+\textbf\newline{Are essential genes conserved?}
 }
@@
 \section{Transposon mutagenesis}
-
-{\bf overview paragraph.}
-
+During evolution species can gain or lose genes.
+We have investigated if these gene gain and losses are related to the essentiality of the genes themselves.
+For this, we have evaluated the essentiality of genes and their conservation and the relationship between these two.
 We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
 These strains are all selected from Enterobacteriaceae family.
-Enterobacteriaceae is a family that includes Gram-negative bacteria with different host ranges and pathogenecity found in soil, water, plants, animals and humans \cite{brenner_bergeys_2006}.
-In humans, various strains from this family can cause diarrhoea, septicaemia, urinary tract infection, meningitis, respiratory disease, and wound and burn infection \cite{brenner_bergeys_2006}.
-Besides, they can infect poultry and livestocks and cause financial losses for farmers \cite{brenner_bergeys_2006}.
+Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
 Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.
 
+Transposon mutagenesis is frequently used for the study of essentiality.
 We have used Tn5 transposon to generate single inserted mutants and placed our mutants in a selective media for Tn5.
-We have picked the mutants and pooled them and used the splinkerette adapter and primers designed in \cite{barquist_tradis_2016} for PCR enrichment.
-Then we have sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
-The number of insertions in a gene can imply the degree of essentiality for that gene.
+We have picked the mutants and pooled them and performed PCR enrichment using the method described in \cite{barquist_tradis_2016}.
+Then, we have sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
+The number of insertions in a gene implies the degree of essentiality for that gene.
 
@@
 We have divided the genes into three groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
-We have adapted the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
+We have utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
 The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
@@
 \section{Are there biases in transposon mutagenesis data?}
-
-{\bf OVERVIEW PARAGRAPH:}
-
+To evaluate the essentiality of a gene, we use the number of insertions within that gene.
+However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis.
 Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}.
 We have performed a detailed study of these biases.
-{\bf LIST TYPES OF BIAS HERE.}
+The biases that we have studied include: gene length bias, origin of replication bias, preferred insertion motif bias, and positional bias within genes.
 
-{\bf ORIGIN OF REPLICATION BIAS:} To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome.
+\subsection{Length bias}
+Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions.
+As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes.
+In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index.
+It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
+\begin{figure*}
+\centering
+\includegraphics[page=42, scale=0.3]{biases.pdf}
+\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
+\label{fig:length-bias}
+\end{figure*}
+
+\subsection{Origin of replication bias}
+To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome.
 Fig.\@ \ref{fig:distance-bias} shows the results.
@@
 
+As these three methods lead to different results, we have compared the differences between the genes found in these three methods.
+For this, we have compared the set of core essential genes resulted from intersection and ancestral insertion index methods and performed word enrichment analysis that was explained before on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method.
+Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure.
+The results are depicted in Fig.\@\ref{fig:method-difference}.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[scale=0.4]{ancestral-intersection-pval.pdf}&
+\includegraphics[scale=0.4]{dollo-ancestral-pval.pdf}\\
+\includegraphics[scale=0.4]{dollo-intersection-pval.pdf}&
+\end{tabular}
+\caption{The figure shows the difference between core essential genes in intersection and ancestral insertion index methods, ancestral insertion index and fdollop methods, and intersection and fdollop methods.
+The red line shows P-value = 0.05.
+The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure.
+The top left figure shows words that are enriched in core essential genes found using ancestral insertion index method but not enriched in core essential genes found using intersection method.
+The top right figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using ancestral insertion index method.
+The bottom left figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using intersection method.}
+\label{fig:method-difference}
+\end{figure*}
+
 %\section*{Acknowledgments}

### 2016-10-11 bin/branch-root-difference.py

@@ -1,0 +1,565 @@
+from Bio import SeqIO
+from re import match, findall
+from collections import defaultdict
+from os import listdir
+
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+rootpath = '../results/define-core-accessory-hieranoid/all/core-essential-genomes'
+ecolipath = '../results/define-core-accessory-hieranoid/ecoli/core-essential-genomes'
+salmonellapath = '../results/define-core-accessory-hieranoid/salmonella/core-essential-genomes'
+klebsiellapath = '../results/define-core-accessory-hieranoid/klebsiella/core-essential-genomes'
+klebsiellaenterobacterpath = '../results/define-core-accessory-hieranoid/klebsiellaenterobacter/core-essential-genomes'
+salmonellacitrobacterpath = '../results/define-core-accessory-hieranoid/salmonellacitrobacter/core-essential-genomes'
+salmonellaecolicitrobacterpath = '../results/define-core-accessory-hieranoid/salmonellaecolicitrobacter/core-essential-genomes'
+genomepath = '../data/fasta-protein/chomosome/seqdb.fasta'
+
+ecodes = '../results/branch-root/root-ecoli-description.txt'
+necodes = '../results/branch-root/not-root-ecoli-description.txt'
+ecoinfo = '../results/branch-root/ecoli-info.txt'
+
+saldes = '../results/branch-root/root-salmonella-description.txt'
+nsaldes = '../results/branch-root/not-root-salmonella-description.txt'
+salinfo = '../results/branch-root/salmonella-info.txt'
+
+kledes = '../results/branch-root/root-klebsiella-description.txt'
+nkledes = '../results/branch-root/not-root-klebsiella-description.txt'
+kleinfo = '../results/branch-root/klebsiella-info.txt'
+
+kleentdes = '../results/branch-root/root-klebsiellaenterobacter-description.txt'
+nkleentdes = '../results/branch-root/not-root-klebsiellaenterobacter-description.txt'
+kleentinfo = '../results/branch-root/klebsiellaenterobacter-info.txt'
+
+salcitdes = '../results/branch-root/root-salmonellacitrobacter-description.txt'
+nsalcitdes = '../results/branch-root/not-root-salmonellacitrobacter-description.txt'
[…]

### 2016-10-11 bin/core-definition-method-difference.py

@@ -1,0 +1,266 @@
+from Bio import SeqIO
+from re import match, findall
+from collections import defaultdict
+from os import listdir
+
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+coreintersectpath = '../results/define-core-accessory-hieranoid/all/core-essential-genomes'
+coreancestraliipath = '../results/define-core-accessory-hieranoid-ancestralii/all/core-essential-genomes'
+coredollopath = '../results/define-core-accessory-hieranoid-fdollop/all/core-essential-genomes'
+genomepath = '../data/fasta-protein/chomosome/seqdb.fasta'
+
+ancintdes = '../results/word-enrichment/ancestral-intersection-description.txt'
+nancintdes = '../results/word-enrichment/not-ancestral-intersection-description.txt'
+ancintinfo = '../results/word-enrichment/ancestral-intersection-info.txt'
+
+dolancdes = '../results/word-enrichment/dollo-ancestral-description.txt'
+ndolancdes = '../results/word-enrichment/not-dollo-ancestral-description.txt'
+dolancinfo = '../results/word-enrichment/dollo-ancestral-info.txt'
+
+dolintdes = '../results/word-enrichment/dollo-intersection-description.txt'
+ndolintdes = '../results/word-enrichment/not-dollo-intersection-description.txt'
+dolintinfo = '../results/word-enrichment/dollo-intersection-info.txt'
+
+list_of_files = listdir(coreintersectpath)
+coreintersect = {}
+for filename in list_of_files:
+    if filename != "b.fasta":
+        coreintersect.update(read_fasta_sequences(coreintersectpath+'/'+filename))
+
+list_of_files = listdir(coreancestraliipath)
+coreancestralii = {}
+for filename in list_of_files:
+    if filename != "b.fasta":
+        coreancestralii.update(read_fasta_sequences(coreancestraliipath+'/'+filename))
+
+list_of_files = listdir(coredollopath)
+coredollo = {}
+for filename in list_of_files:
+    if filename != "b.fasta":
+        coredollo.update(read_fasta_sequences(coredollopath+'/'+filename))
+
+with open(nancintdes, 'w') as nfile:
+    with open(ancintdes, 'w') as yfile:
+        for item in coreancestralii.keys():
[…]

### 2016-10-11 bin/montecarlo.R

@@ -8,18 +8,27 @@ list_of_files <- list.files(path=plots_dir, full.names=T, recursive=FALSE)
 plots = list()
 sampledplots = list(list())
 sumlength = list()
-numsamples = 2
+weights = list()
+numsamples = 100
+locusid = c()
 for (filename in list_of_files)
 {
   plotfile = as.matrix(read.table(filename, as.is=TRUE))
-  locusid = strsplit(basename(filename),"\\.")[[1]][1]
-  plots[[locusid]] = plotfile[,1] + plotfile[,2]
-  for (i in 1:numsamples)
-  {
-    sampledplots[[locusid]][[i]] = sample(plots[[locusid]])
-  }
-  sumlength[[locusid]] = c(sum(plots[[locusid]]), length(plots[[locusid]]))
+  lid = strsplit(basename(filename),"\\.")[[1]][1]
+  locusid = c(locusid,lid)
+  plots[[lid]] = plotfile[,1] + plotfile[,2]
+  # plot(plots[[lid]], pch = '.', ylim=c(0,10))
+  # lines(loess.smooth(1:length(plots[[lid]]), plots[[lid]], span=0.2, family = "gaussian"), col=2, lwd=5)
+  plotsdata = as.data.frame(cbind(1:length(plots[[lid]]), as.vector(plots[[lid]])))
+  names(plotsdata) <- c("position", "num_inserts")
+  mdl <- loess(num_inserts~position, plotsdata, span=0.2, family = "gaussian")
+  # lo <- loess(num_inserts ~ position, plotsdata)
+  weights[[lid]] = predict(mdl)
+  sumlength[[lid]] = c(sum(plots[[lid]]), length(plots[[lid]]))
 }
+sampledplots = lapply(locusid, function(filename) replicate(numsamples, sample(plots[[filename]]), simplify=FALSE))
+# sampledplots = lapply(locusid, function(filename) lapply( 1:numsamples, function(i) sample(plots[[filename]], prob=weights[[lid]])))
+names(sampledplots) <- locusid
 
 list_of_files <- list.files(path=fastas_dir, full.names=T, recursive=FALSE)
 list_of_files <- list_of_files[ !grepl("/home/fatemeh/EnTrI/data/fasta-protein/chromosome/U00096.fasta",list_of_files) ]

### 2016-10-13 documents/manuscript.tex
Edited the manuscript, Started separating results and methods

@@
 
-\section{Transposon mutagenesis}
-During evolution species can gain or lose genes.
-We have investigated if these gene gain and losses are related to the essentiality of the genes themselves.
-For this, we have evaluated the essentiality of genes and their conservation and the relationship between these two.
-We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
-These strains are all selected from Enterobacteriaceae family.
-Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
-Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.
-
-Transposon mutagenesis is frequently used for the study of essentiality.
-We have used Tn5 transposon to generate single inserted mutants and placed our mutants in a selective media for Tn5.
-We have picked the mutants and pooled them and performed PCR enrichment using the method described in \cite{barquist_tradis_2016}.
-Then, we have sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
-The number of insertions in a gene implies the degree of essentiality for that gene.
-
-%\subsection{Essentiality}
-We have used a value called insertion index to evaluate the essentiality of a gene.
-This value is calculated by summing up the number of transposon insertion sites observed in a gene.
-Since the lengths of the genes are different, we have normalised the insertion index by dividing it by gene length.
-Our experiment has been performed on different strains and the library density is different in each experiment.
-Therefore, in order to make the insertion indices comparable in all the strains, we have normalised our insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
-We have not studied genes shorter than 100 base-pairs as they might not be targeted by any transposon due to their shortness.
-
-We have divided the genes into three groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
-We have utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
-The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
-The first peak shows the genes with no or just a few insertions which are considered as essential genes.
-We have fitted an exponential distribution to the first peak and a gamma distribution to the second one.
-Then, we have calculated the log odds ratio for belonging to each of these distributions for each gene.
-The region that has log odds value between -2 and 2 is called the ambiguous region, the genes belonging to the first peak are essential and the rest of the genes are not essential.
-Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c c}
-\includegraphics[page=1, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=2, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=3, scale=0.25]{essentiality.pdf}\\
-\includegraphics[page=4, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=5, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=6, scale=0.25]{essentiality.pdf}\\
-\includegraphics[page=7, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=8, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=9, scale=0.25]{essentiality.pdf}\\
-\includegraphics[page=10, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=11, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=12, scale=0.25]{essentiality.pdf}\\
-&\includegraphics[page=13, scale=0.25]{essentiality.pdf}&\\
-\end{tabular}
-\caption{Plots show the insertion index distribution for each genome.
-The plots are divided into 4 regions using red lines.
-These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-\label{fig:iidist-species}
-\end{figure*}
+\section{Results}
 
 %\section{Are transposons biased towards certain positions?}
-\section{Are there biases in transposon mutagenesis data?}
-To evaluate the essentiality of a gene, we use the number of insertions within that gene.
+\subsection{Are there biases in transposon mutagenesis data?}
+To evaluate the essentiality of a gene, the number of insertions within that gene was measured.
 However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis.
 Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}.
-We have performed a detailed study of these biases.
-The biases that we have studied include: gene length bias, origin of replication bias, preferred insertion motif bias, and positional bias within genes.
+We performed a detailed study of these biases.
+The biases that we studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
 
-\subsection{Length bias}
-Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions.
+%\subsection{Length bias}
+%Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions.
 As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes.
-In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index.
+%In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index in short genes.
 It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
-\begin{figure*}
-\centering
-\includegraphics[page=42, scale=0.3]{biases.pdf}
-\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
-\label{fig:length-bias}
-\end{figure*}
+%\begin{figure*}
+%\centering
+%\includegraphics[page=42, scale=0.3]{biases.pdf}
+%\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
+%\label{fig:length-bias}
+%\end{figure*}
 
-\subsection{Origin of replication bias}
+\subsubsection{Origin of replication bias}
+One possible source of bias is the distance from origin of replication.
+When the bacteria are under replication during the transposn insertion process, there are more copies of the genes close to the origin of replication than the genes further away.
+This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
+The other factor that we considered was if essential genes are not uniformly distributed in the genome and are clustered near the origin.
+Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication.
+However, the essential genes are more frequent in the leading strand than the lagging one.
+
 To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome.
 Fig.\@ \ref{fig:distance-bias} shows the results.
-The red line is a loess curve that has been fitted to the data when the smoothness parameter equals $0.2$.
 The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication.
-A possible explanation for this phenomenon is that the bacteria were under replication while being infected with transposons.
-Therefore, the number of gene copies close to the origin of replication was greater and more insertions have occurred in these genes.
-To overcome this bias, we have normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplying this value by the average insertion index.
-
-{\bf mention possible confounding factor for this study.
-Essential genes may (are???) non-uniformly distributed in the genome (I presume they are clustered near the origin).
-This would be expected to lower the insertion index near the origin, however, the number of non-essential genes is $\approx 10\times$ the number of essential genes, therefore the signature of biased numbers of insertions near the origin remains.
-}
+To overcome this bias, we normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplied this value by the average insertion index.
 
@@
 
-{\bf PREFERRED INSERTION MOTIF BIAS:} We have tested whether our transposons are biased towards certain motifs.
-For this, we have generated a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
-The results are depicted in Fig.\@  \ref{fig:logos}.
-The results show a slight bias towards certain combinations of bases.
-In addition, we have investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index Fig.\@  \ref{fig:GC-bias}.
-The red lines show the loess curve when the smoothness parameter is $0.2$.
+\subsubsection{Preferred insertion motif bias}
+Another concern is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, rather than their essentiality level.
+For this, we generated a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
+The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases.
+
+In addition, we investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index Fig.\@  \ref{fig:GC-bias}.
 As the figure shows, when G-C content is less than $40\%$, the insertion index is low, however when it is higher than $50\%$, the insertion index is almost constant.
-A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which causes a reduction in the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
+A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
 The other reason is that the genes with low G-C content are enriched in mobile genetic elements compared to the genes with average G-C content (\ref{fig:gc-pval}) and this has caused seeing a different pattern of essentiality in that region.
@@
 \item seems unlikely -- show bulk of genes are around 50\% G+C (add box-whisker plots to scatter diagrams?)
-\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? 
+\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? It is not mentioned in the paper.
 \end{itemize}
@@
 \end{tabular}
-\caption{We have generated sequence logo plots using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
+\caption{Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
 On the left the height of each character corresponds to a bit score for that character (i.e.
@@
 
-The other bias that we have considered is the bias towards certain locations in a gene.
-We have divided every gene into 100 bins and calculated the mean insertion index for each bin.
+The other question that we tried to answer was whether insertions are tolerated in some regions in a gene?
+For example, can essential genes tolerate insertions at their 3' end without losing their funtionality?
+To address this question, we divided every gene into 100 bins and calculated the mean insertion index for each bin.
 Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location.
-We have also studied the bias in each of the groups: essential genes, non-essential genes, and beneficial losses.
+We also studied the bias in each of the groups: essential genes, non-essential genes, and beneficial losses.
 The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses.
@@
 On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors.
-\textcolor{red}{\{To be tested\}} We have calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes to overcome these biases.
-The insertion index distribution for each genome after correcting for all distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
+\textcolor{red}{\{To be tested\}} We calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes to overcome these biases.
+The insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
 
@@
 \end{tabular}
-\caption{The plots show the average insertion index in over the genes in all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
+\caption{The plots show the average insertion index in 100 percentiles in all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
 Each bin shows the average insertion index for $1\%$ of the genes.
@@
 %\end{itemize}
-\section{Essentiality and conservation}
-To study whether each gene in our 13 organisms is conserved we have proposed a program that clusters homologous proteins.
+\subsection{Essentiality and conservation}
+Essential genes are needed for the growth of organisms.
+Because of that, one might think that essential genes should not be lost in a short period of time throughout evolution, unless they are no longer needed in new organisms or they are replaced by new pathways.
+Therefore, it is expected that most of the essential genes are conserved in different organisms from the same family.
+We have tested this idea by comparing the essentiality and conservation of genes in Enterobacteriaceae family.
+
+To study whether each gene in our 13 organisms is conserved we proposed a program that clusters homologous proteins.
 This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
 It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
-Then it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
-In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by relaxing Jackhmmer parameters.
+Then, it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
+In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by running Jackhmmer with more permissive parameters.
 Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
@@
 
-\subsection{Gene classes}
+\subsubsection{Gene classes}
 %We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
@@
 For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
-To study the conservation of genes, we have used homclust and divided the clusters of homologous genes into three groups.
+We needed to evaluate essentiality and conservation of genes to study the relationship between this two.
+So, we divided our genes into different levels of essentiality using the method explained in Section~\ref{sec:essentiality}.
+Moreover, we used our proposed algorithm to define different levels of conservation.
+
+We divided the clusters of homologous genes into three groups based on their conservation.
 Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
 These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
+The results for comparing these three groups and the four levels of essentiality (essential, ambiguous, non-essential, and beneficial losses) are depicted in Fig.\@ \ref{fig:iidist}.
+The figure shows that most of the essential clusters are single copy and most of the beneficial losses are genus specific.
+Besides, most of the multi-copy genes are non-essential.
 
@@
 \end{figure}
-
-We have divided the clusters of orthologous genes into 4 groups based on their essentiality (essential, ambiguous, non-essential, and beneficial loss) and 3 groups based on their conservation (genus specific, single copy, and multi-copy).
-The results are depicted in Fig.\@ \ref{fig:iidist}.
-The figure shows that most of the essential clusters are single copy and most of the beneficial losses are genus specific.
 
@@
 
-To study which functions are enriched in each class of essentiality, we have gathered the``note" section for each gene from their embl files.
-Then, we have counted the repeat number of each word for the genes in each class and the genes that do not belong to that class and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
+To study which functions are enriched in each class of essentiality, we gathered the``note" section for each gene from their embl files.
+Then, we counted the repeat number of each word for the genes in each class and the genes that do not belong to that class and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
 The P-values are then corrected using Benjamini-Hochberg-Yekutieli procedure.
@@
 The results show an enrichment of the genes related to replication, transcription, translation, division, and rod shape determining proteins in essential class.
-The non-essential genes are mostly transport proteins, flagellar proteins, ATPase, and DNA repair proteins.
+The non-essential genes are mostly membrane associated proteins, flagellar proteins, ATPase, and DNA repair proteins.
 Beneficial losses are enriched in transposase enzymes, putative and hypothetical proteins, and mobile elements.
@@
 
-We have also conducted a pathway enrichment analysis for these three groups.
-For this, we have downloaded pathway datasets for strains that were available in KEGG database.
+We also conducted a pathway enrichment analysis for these three groups.
+For this, we downloaded pathway datasets for strains that were available in KEGG database.
 This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.
-Then we have merged these databases and used a hypergeometric test to find which pathways are enriched in each essentiality class.
-Finally, we have corrected the P-values using Benjamini-Hochberg-Yekutieli.
-The results are depicted in ~\ref{fig:essentiality-pathway}.
+Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class.
+Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.
+The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes.
+However, as mobile genetic elements are not stored in KEGG database, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.
 
@@
 The red line shows P-value = 0.05.
-The P-values have been calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
+The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
 \label{fig:essentiality-pathway}
@@
 
-\subsection{The evolution of essentiality}
+\subsubsection{The evolution of essentiality}
 %We have compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
@@
 %\end{figure}
+If essential genes are more likely to be conserved than non-essential genes, we expect to see species that are closer together have more essential genes in common than other species.
+Moreover, we expect the ratio of core essential genes to core genes to increase as we go up in the phylogenetic tree.
+We have put these two ideas to test in this chapter.
 
-In order to test if the essentiality of genes follows a tree-like trend, we have compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
-For this, we have counted the number of genes that are core between every combination of our strains, and the number of genes that are core essential between those combinations.
-We have used UpSetR package \cite{conway_upsetr:_2016} in R to visualise the results in Fig.\@ \ref{fig:upsetr}.
+In order to test if the essentiality of genes follows a tree-like trend, we compared the number of genes that were conserved in different bacteria in our study and the number of genes that were essential in these bacteria.
+For this, we counted the number of genes that were core between every combination of our bacteria, and the number of genes that were core essential between those combinations.
+We used UpSetR package \cite{conway_upsetr:_2016} in R to visualise the results in Fig.\@ \ref{fig:upsetr}.
 As shown in the figures, among 1908 genes that are core between all the strains under study, only 184 are core essential.
-The results propose that although conservation of genes follows a tree-like trend (800 genes are core in Klebsiellas, 273 genes are core in Salmonellas, 208 genes are core in Klebsiellas and Enterobacter, ...), the essentiality does not show a tree-like signal and most of the large sets of core essential genes belong to only one strain.
+We looked at subsets of the genes that were essential (core) in every combination of our bacterial strains to see whether they were phylogenetically informative or not.
+A phylogenetically informative subset is a subset that is essential (core) in two or more bacteria but not in all bacteria.
+We have marked the phylogenetically informative sets of genes with ticks and the uninformative ones with crosses.
+The results propose that although conservation of genes follows a tree-like trend with many phylogenetically informative sets of genes with high cardinality, the essentiality does not show a tree-like signal and most of the large sets of core essential genes belong to only one bacteria.
 We believe this happens due to the small number of essential genes.
@@
 
-To test if essential genes are more likely to be conserved, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
-For the essential genes to be more likely to be conserved, this ratio should increase as we go up in the tree.
-We have used three different methods for this.
-The first method is intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are present in all the strains in its child nodes and are core essential if and only if they are essential in all the strains in its child nodes.
+Furthermore, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
+We used three different methods for this.
+The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
 
 The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
-In this method, we have averaged over the insertion indices of the pair of closest children of the node that we are calculating the core genes for.
-We have repeated this and averaged the averages until we reached the node under study.
-Then, we have plotted the insertion indices and fitted an exponential and a gamma distribution to the plot and found the essential genes at that level.
+In this method, we averaged over the insertion indices of the pair of closest children of the ancestral node.
+We repeated this and averaged the averages until we reached the ancestral node.
+Then, we plotted the insertion indices and fitted an exponential and a gamma distribution to the plot as described in Section~\ref{sec:essentiality} and found the essential genes at that level.
 
 The third method is using Dollo law to define core genes and core essential genes.
-This method, assumes that the gain of genes (essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (essentiality) and minimise the number of times that the genes (the essentiality) has been lost.
+This method, assumes that the gain of genes (gain of essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (gain of essentiality) and minimise the number of times that a gene (the essentiality of a gene) has been lost.
 Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
@@
 \caption{The tree shows the species tree in Fig.\@\ref{fig:species-tree} annotated by the number of core essential genes and core genes at each node.
-We have used three methods for defining core essential genes and core genes.
+We used three methods to define core essential genes and core genes.
 The numbers at the leaves are the same using all these three methods.
@@
 
-As these three methods lead to different results, we have compared the differences between the genes found in these three methods.
-For this, we have compared the set of core essential genes resulted from intersection and ancestral insertion index methods and performed word enrichment analysis that was explained before on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method.
+As these three methods lead to different results, we compared the differences between the genes found in these three methods.
+For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods.
+Then, we performed word enrichment analysis that was explained before on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method.
 Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure.
@@
 
+\section{Materials and Methods}
+\subsection{Transposon mutagenesis}
+Throughout time, species can gain or lose genes.
+We have investigated if these gene gain and losses are related to the essentiality of the genes themselves.
+For this, we have evaluated the essentiality of genes and their conservation and the relationship between these two.
+We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
+These strains are all selected from Enterobacteriaceae family.
+Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
+Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.
+
+We used transposon mutagenesis which is a frequently used method for the study of essentiality.
+We generated single inserted mutants using Tn5 transposon and placed our mutants in a selective media for Tn5.
+Then, we picked the mutants and pooled them and performed PCR enrichment using the method described in \cite{barquist_tradis_2016}.
+We sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
+The number of insertions in a gene implies the degree of essentiality for that gene.
+
+\subsection{Essentiality}\label{sec:essentiality}
+The more transposon insertions we observe in a gene after sequencing the genomes, the less essential the gene is.
+In order to quantify the essentiality of genes, we used a measure named insertion index which is proportional to the number of insertions in a gene.
+
+To calculate the insertion index for each gene, we summed up the number of transposon insertion sites observed in that gene.
+Since the lengths of the genes are different, we normalised the insertion index by dividing it by gene length.
+Our experiment is performed on different strains and the library density is different in each experiment.
+Therefore, in order to make the insertion indices comparable in all the strains, we normalised our insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
+We did not genes shorter than 100 base-pairs as they might not be targeted by any transposon due to their shortness.
+
+Based on insertion indices, we divided the genes into four groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
+We utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
+The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
+The first peak shows the genes with no or just a few insertions which are considered as essential genes.
+We fitted an exponential distribution to the first peak and a gamma distribution to the second one.
+Then, we calculated the log odds ratio for belonging to each of these distributions for each gene.
+The region that has log odds value between -2 and 2 is called the ambiguous region, the genes belonging to the first peak are essential and the rest of the genes are not essential.
+Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=1, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=2, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=3, scale=0.25]{essentiality.pdf}\\
+\includegraphics[page=4, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=5, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=6, scale=0.25]{essentiality.pdf}\\
+\includegraphics[page=7, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=8, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=9, scale=0.25]{essentiality.pdf}\\
+\includegraphics[page=10, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=11, scale=0.25]{essentiality.pdf}&
+\includegraphics[page=12, scale=0.25]{essentiality.pdf}\\
+&\includegraphics[page=13, scale=0.25]{essentiality.pdf}&\\
+\end{tabular}
+\caption{Plots show the insertion index distribution for each genome.
+The plots are divided into 4 regions using red lines.
+These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
+\label{fig:iidist-species}
+\end{figure*}
 %\section*{Acknowledgments}

### 2016-10-16 documents/manuscript.tex
Wrote the introduction

@@
 \section*{Introduction}
-Studying the essentiality of genes helps with identifying the fundamental processes necessary for cell viability \cite{juhas_essence_2011}.
+%Studying the essentiality of genes helps with identifying the fundamental processes necessary for cell viability \cite{juhas_essence_2011}.
 So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}.
 The results have led to new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising new genomes \cite{hutchison_global_1999, hutchison_design_2016}.
-Researchers have used different methods for studying the essentility of genes in prokaryotes.
+
+Different methods have been used so far for studying the essentility of genes in prokaryotes.
 Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}.
@@
 
-Although the essentiality of genes has been studied in a variety of organisms, there is still room to study the evolutionary conservation of essentiality.
-Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens}.
-Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium and \textit{Salmonella} Typhi.
+Now that the essentiality of genes can be evaluated using different methods, it is possible to compare the essentiality data in different organisms and investigate the differentiation of essentiality of their genes.
+Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
+Freed et al.\@ \cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that some genes have gained essentiality in \textit{Shigella flexneri} while there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}.
+Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium and \textit{Salmonella} Typhi and found that the essentiality of genes differs in different organisms.
 In a similar study, Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges.
-We extend this research by studying 13 bacterial strains from Enterobacteriaceae.
-These strains and \textit{Escherichia coli} K-12 MG1655 studied by Baba et al.\@ \cite{baba_construction_2006} are depicted in Fig.\@ \ref{fig:species-tree}.
+Although there are many studies on differentiation of essentiality in different organisms, these studies usually include two or three strains.
+
+Our aim is to study the essentiality of genes in an evolutionary framework in 13 different organisms from Enterobacteriaceae family.
+In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study.
+In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential.
+We first performed a detailed study of biases than can influence the inference of essentiality.
+Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome).
+We have also investigated how essentiality changes in the phylogenetic tree for these organisms.
 
@@
 
-\textcolor{red}{\{A summary of what we have done\}}
+\section{Results}
+Throughout time, species can gain or lose genes.
+We have investigated if these gene gain and losses are related to the essentiality of the genes themselves.
+In this section, we have first described the biases that can affect our study, and then evaluated the essentiality of genes and their conservation and the relationship between these two.
 
-\section{Results}
-
-%\section{Are transposons biased towards certain positions?}
 \subsection{Are there biases in transposon mutagenesis data?}
@@
 The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication.
-To overcome this bias, we normalised our insertion indices by dividing the value of the insertion index by the predicted value by loess for that position and then multiplied this value by the average insertion index.
 
@@
 Another concern is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, rather than their essentiality level.
-For this, we generated a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
+For this, we used Weblogo~\cite{crooks_weblogo:_2004} to generated a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
 The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases.
@@
 For example, can essential genes tolerate insertions at their 3' end without losing their funtionality?
-To address this question, we divided every gene into 100 bins and calculated the mean insertion index for each bin.
+To address this question, we divided every gene into percentiles and calculated the mean insertion index for each percentile.
 Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location.
@@
 On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors.
-\textcolor{red}{\{To be tested\}} We calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes to overcome these biases.
-The insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
+\textcolor{red}{\{To be tested\}}
 
@@
 \end{tabular}
-\caption{The plots show the average insertion index in 100 percentiles in all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
-Each bin shows the average insertion index for $1\%$ of the genes.
-The genes have been divided into 3 segments: 5\% of
+\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
+The genes are divided into 3 segments: 5\% of
 the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
@@
 
-\begin{figure*}
-\centering
-\begin{tabular}{c c c}
-\includegraphics[page=1, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=2, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=3, scale=0.25]{essentiality-normalised.pdf}\\
-\includegraphics[page=4, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=5, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=6, scale=0.25]{essentiality-normalised.pdf}\\
-\includegraphics[page=7, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=8, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=9, scale=0.25]{essentiality-normalised.pdf}\\
-\includegraphics[page=10, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=11, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=12, scale=0.25]{essentiality-normalised.pdf}\\
-&\includegraphics[page=13, scale=0.25]{essentiality-normalised.pdf}&\\
-\end{tabular}
-\caption{Plots show the insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes.
-The plots are divided into 4 regions using red lines.
-These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-\label{fig:iidist-species-normalised}
-\end{figure*}
 %\begin{itemize}
@@
 
-To study whether each gene in our 13 organisms is conserved we proposed a program that clusters homologous proteins.
-This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
-It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
-Then, it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
-In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by running Jackhmmer with more permissive parameters.
-Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
-The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 13 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
-
-\begin{figure}
-\includegraphics[scale=0.2]{homclust.pdf}
-\caption{The steps of our proposed algorithm for clustering homologous genes.}
-\label{fig:homclust}
-\end{figure}
-
 \subsubsection{Gene classes}
@@
 
-We divided the clusters of homologous genes into three groups based on their conservation.
-Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
-These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
 The results for comparing these three groups and the four levels of essentiality (essential, ambiguous, non-essential, and beneficial losses) are depicted in Fig.\@ \ref{fig:iidist}.
@@
 Besides, most of the multi-copy genes are non-essential.
-
-\begin{figure}
-\includegraphics[scale=0.3]{cluster-size-dist.pdf}
-\caption{Size distribution for all clusters of homologous genes.
-Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
-\label{fig:homclust-results}
-\end{figure}
 
@@
 We used three different methods for this.
-The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
+These methods are explained in...
 
-The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
-In this method, we averaged over the insertion indices of the pair of closest children of the ancestral node.
-We repeated this and averaged the averages until we reached the ancestral node.
-Then, we plotted the insertion indices and fitted an exponential and a gamma distribution to the plot as described in Section~\ref{sec:essentiality} and found the essential genes at that level.
-
-The third method is using Dollo law to define core genes and core essential genes.
-This method, assumes that the gain of genes (gain of essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (gain of essentiality) and minimise the number of times that a gene (the essentiality of a gene) has been lost.
-Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
 The numbers predicted using these three methods are shown in Fig.\@ \ref{fig:annotated-speciestree}.
@@
 \subsection{Transposon mutagenesis}
-Throughout time, species can gain or lose genes.
-We have investigated if these gene gain and losses are related to the essentiality of the genes themselves.
-For this, we have evaluated the essentiality of genes and their conservation and the relationship between these two.
-We have studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
+We studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
 These strains are all selected from Enterobacteriaceae family.
 Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
-Here, we perform a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.
+Here, we performed a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.
 
@@
 
-\subsection{Essentiality}\label{sec:essentiality}
+\subsection{Essentiality levels}\label{sec:essentiality}
 The more transposon insertions we observe in a gene after sequencing the genomes, the less essential the gene is.
@@
 Therefore, in order to make the insertion indices comparable in all the strains, we normalised our insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
-We did not genes shorter than 100 base-pairs as they might not be targeted by any transposon due to their shortness.
 
@@
 The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
+We fitted a loess curve to this distribution to find where the two peaks separate from each other.
 The first peak shows the genes with no or just a few insertions which are considered as essential genes.
@@
 \end{figure*}
+
+\subsection{Bias correction}
+We observed a distance from origin of replication bias and also a positional bias within genes.
+Hence, we needed to correct for these biases before inferring the essentiality of genes from their insertion indices.
+We did not include genes shorter than 100 base-pairs in our study as they might not be targeted by any transposon due to their shortness.
+
+To overcome the distance from origin of replication bias, we divided the value of insertion index for a gene in a specific position by the predicted value by loess for that position.
+This value was then multiplied by the average insertion index.
+To overcome the positional bias within genes, we calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes.
+The insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
+
+\begin{figure*}
+\centering
+\begin{tabular}{c c c}
+\includegraphics[page=1, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=2, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=3, scale=0.25]{essentiality-normalised.pdf}\\
+\includegraphics[page=4, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=5, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=6, scale=0.25]{essentiality-normalised.pdf}\\
+\includegraphics[page=7, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=8, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=9, scale=0.25]{essentiality-normalised.pdf}\\
+\includegraphics[page=10, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=11, scale=0.25]{essentiality-normalised.pdf}&
+\includegraphics[page=12, scale=0.25]{essentiality-normalised.pdf}\\
+&\includegraphics[page=13, scale=0.25]{essentiality-normalised.pdf}&\\
+\end{tabular}
+\caption{Plots show the insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes.
+The plots are divided into 4 regions using red lines.
+These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
+\label{fig:iidist-species-normalised}
+\end{figure*}
+
+\subsection{Conservation levels}
+To study whether each gene in our 13 organisms is conserved we proposed a program that clusters homologous proteins.
+This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
+It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
+Then, it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
+In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by running Jackhmmer with more permissive parameters.
+Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
+The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 13 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
+
+\begin{figure}
+\includegraphics[scale=0.2]{homclust.pdf}
+\caption{The steps of our proposed algorithm for clustering homologous genes.}
+\label{fig:homclust}
+\end{figure}
+
+We divided the clusters of homologous genes into three groups based on their conservation.
+Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
+These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
+
+\begin{figure}
+\includegraphics[scale=0.3]{cluster-size-dist.pdf}
+\caption{Size distribution for all clusters of homologous genes.
+Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
+\label{fig:homclust-results}
+\end{figure}
+
+\subsection{Defining core and core essential genes}
+The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
+
+The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
+In this method, we averaged over the insertion indices of the pair of closest children of the ancestral node.
+We repeated this and averaged the averages until we reached the ancestral node.
+Then, we plotted the insertion indices and fitted an exponential and a gamma distribution to the plot as described in Section~\ref{sec:essentiality} and found the essential genes at that level.
+
+The third method is using Dollo law to define core genes and core essential genes.
+This method, assumes that the gain of genes (gain of essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (gain of essentiality) and minimise the number of times that a gene (the essentiality of a gene) has been lost.
+Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
 %\section*{Acknowledgments}

### 2016-10-17 documents/manuscript.tex
Revised the manuscript

@@
 % the * after section prevents numbering
-\section*{Introduction}
-%Studying the essentiality of genes helps with identifying the fundamental processes necessary for cell viability \cite{juhas_essence_2011}.
+\section{Introduction}
+With the advent of sequencing technologies and genome annotation methods many genes have been identified.
+However, not all of these genes are of the same importance for the growth of an organism.
 So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}.
-The results have led to new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising new genomes \cite{hutchison_global_1999, hutchison_design_2016}.
-
-Different methods have been used so far for studying the essentility of genes in prokaryotes.
+These studies can give us new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising minimal genomes \cite{hutchison_global_1999, hutchison_design_2016, reus_blueprint_2016}.
+Different methods have been used for studying the essentility of genes in prokaryotes.
 Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}.
@@
 
-Our aim is to study the essentiality of genes in an evolutionary framework in 13 different organisms from Enterobacteriaceae family.
+Our aim is to study the essentiality of genes in an evolutionary framework in 13 different organisms from Enterobacteriaceae family (Fig.\@ \ref{fig:species-tree}).
+Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
 In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study.
@@
 
-\section{Results}
+\section{Results and discussion}
 Throughout time, species can gain or lose genes.
-We have investigated if these gene gain and losses are related to the essentiality of the genes themselves.
+We investigated if these gene gain and losses are related to the essentiality of the genes.
 In this section, we have first described the biases that can affect our study, and then evaluated the essentiality of genes and their conservation and the relationship between these two.
@@
 \subsection{Are there biases in transposon mutagenesis data?}
-To evaluate the essentiality of a gene, the number of insertions within that gene was measured.
+To evaluate the essentiality of a gene, the number of insertions within that gene was measured as explained in~\ref{sec:essentiality}.
 However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis.
@@
 We performed a detailed study of these biases.
-The biases that we studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
+The biases that we have studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
 
@@
 This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
-The other factor that we considered was if essential genes are not uniformly distributed in the genome and are clustered near the origin.
+The other factor that can affect the results is if essential genes are clustered near the origin by nature.
 Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication.
@@
 
-To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome.
-Fig.\@ \ref{fig:distance-bias} shows the results.
+To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Fig.\@ \ref{fig:distance-bias}.
 The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication.
@@
 \subsubsection{Preferred insertion motif bias}
-Another concern is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, rather than their essentiality level.
-For this, we used Weblogo~\cite{crooks_weblogo:_2004} to generated a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
+Another concern while inferring essentiality from transposon mutagenesis data is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, instead of their essentiality level.
+For this, we used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
 The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases.
 
-In addition, we investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index Fig.\@  \ref{fig:GC-bias}.
+In addition, we investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index (Fig.\@  \ref{fig:GC-bias}).
 As the figure shows, when G-C content is less than $40\%$, the insertion index is low, however when it is higher than $50\%$, the insertion index is almost constant.
 A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
-The other reason is that the genes with low G-C content are enriched in mobile genetic elements compared to the genes with average G-C content (\ref{fig:gc-pval}) and this has caused seeing a different pattern of essentiality in that region.
+The other reason is that the genes with low G-C content are enriched in mobile genetic elements compared to the genes with average G-C content (Fig.\@ \ref{fig:gc-pval}) and this has caused seeing a different pattern of essentiality in that region.
 
@@
 
-The other question that we tried to answer was whether insertions are tolerated in some regions in a gene?
+\subsubsection{Positional bias within genes}
+The other question that we tried to answer was whether insertions are tolerated in some regions in a gene.
 For example, can essential genes tolerate insertions at their 3' end without losing their funtionality?
 To address this question, we divided every gene into percentiles and calculated the mean insertion index for each percentile.
-Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location.
-We also studied the bias in each of the groups: essential genes, non-essential genes, and beneficial losses.
+Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location when considering all genes together.
+We also studied the bias in three of the groups defined in Section~\ref{sec:essentiality}: essential genes, non-essential genes, and beneficial losses.
 The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses.
-The case for the non-essential genes is similar to the average (Fig.\@ \ref{fig:insertion-position-bias}).
+The case for the non-essential genes is similar to all genes.
 High number of insertions at the 3' end of essential genes implies that the functional part of the genes are located before the insertions.
@@
 For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
-We needed to evaluate essentiality and conservation of genes to study the relationship between this two.
-So, we divided our genes into different levels of essentiality using the method explained in Section~\ref{sec:essentiality}.
-Moreover, we used our proposed algorithm to define different levels of conservation.
+In order to study the relationship between essentiality and conservation, we needed to evaluate the essentiality and conservation of genes.
+For this, we divided the genes into different levels of essentiality (essential genes, ambiguous, non-essential genes, beneficial losses).
+We first normalised the biases that exist in the data using the method explained in Section~\ref{sec:biascorrection} and then defined the essentiality level of genes using the method explained in Section~\ref{sec:essentiality}.
+Moreover, we grouped the genes into different classes of conservation (genus specific, single copy, multi-copy) using the method explaind in Section~\ref{sec:conservationclasses}.
 
-The results for comparing these three groups and the four levels of essentiality (essential, ambiguous, non-essential, and beneficial losses) are depicted in Fig.\@ \ref{fig:iidist}.
-The figure shows that most of the essential clusters are single copy and most of the beneficial losses are genus specific.
-Besides, most of the multi-copy genes are non-essential.
+The results for comparing four levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
+The high number of single copy genes in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality.
+However, the relatively high number of essential genes in genus specific class implies that each genus has a set of essential genes that makes it distinct.
+The figure also shows that beneficial losses are over represented in genus specific class.
+Therefore, beneficial losses are mostly recent genes that the organism tends to lose in the long run.
+Besides, most of the multi-copy clusters are non-essential and there are only a few multi-copy clusters that are essential.
+This can be explained by the redundancy that genes can keep even after $\sim100$ million years \cite{dean_pervasive_2008}.
+In the presence of two redundant variations of one gene, if we knock out one copy by transposon mutagenesis, the other copy compensates and the organism can still survive.
 
@@
 
-To study which functions are enriched in each class of essentiality, we gathered the``note" section for each gene from their embl files.
+To study which functions are enriched in each class of essentiality, we gathered the descriptions of genes from their embl files.
 Then, we counted the repeat number of each word for the genes in each class and the genes that do not belong to that class and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
@@
 \subsubsection{The evolution of essentiality}
-%We have compared the number of genes that are conserved in different strains in our study and the number of genes that are essential in these strains.
-The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show a tree-like signal.
-%\begin{itemize}
-%\item UpSetR results (Fig.~\ref{fig:upsetr})
-%\item Stringent
-%\item Dollo law
-%\item Ancestral insertion index
-%\end{itemize}
 %In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
@@
 %\end{figure}
-If essential genes are more likely to be conserved than non-essential genes, we expect to see species that are closer together have more essential genes in common than other species.
-Moreover, we expect the ratio of core essential genes to core genes to increase as we go up in the phylogenetic tree.
-We have put these two ideas to test in this chapter.
+
+In this section, we have counted the genes that are essential and conserved at each level of the phylogenetic tree.
+We were interested to see if the trend of the essentiality variation between different organisms is in agreement with the phylogenetic tree.
+In other words, we have tested if organisms close together have more essential genes in common than organisms that have separated earlier in the phylogenetic tree.
+
+We needed to cluster sets of orthologous genes in the bacteria that we were studying.
+Homologous clusters introduced in ~\ref{sec:conservationclasses} were not useful for this purpose as sets of paralogous genes with different essentiality levels can make essentiality inference ambiguous.
+Plenty of methods are proposed for this purpose.
+Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
+We used Hieranoid for clustering orthologous genes.
 
 In order to test if the essentiality of genes follows a tree-like trend, we compared the number of genes that were conserved in different bacteria in our study and the number of genes that were essential in these bacteria.
-For this, we counted the number of genes that were core between every combination of our bacteria, and the number of genes that were core essential between those combinations.
+The genes that are conserved in all bacteria being studied are called core genes and the genes that are conserved and essential are called core essential genes.
+We counted the number of genes that were core between every combination of bacteria, and the number of core essential genes between those combinations.
 We used UpSetR package \cite{conway_upsetr:_2016} in R to visualise the results in Fig.\@ \ref{fig:upsetr}.
-As shown in the figures, among 1908 genes that are core between all the strains under study, only 184 are core essential.
-We looked at subsets of the genes that were essential (core) in every combination of our bacterial strains to see whether they were phylogenetically informative or not.
-A phylogenetically informative subset is a subset that is essential (core) in two or more bacteria but not in all bacteria.
+As shown in the figures, among 1908 genes that are core between all the bacteria under study, only 184 are core essential.
+We looked at subsets of the genes that were core essential (core) in every combination of our bacterial strains to see whether they were phylogenetically informative or not.
+A phylogenetically informative subset is a subset that is core essential (core) in two or more bacteria but not in all bacteria.
 We have marked the phylogenetically informative sets of genes with ticks and the uninformative ones with crosses.
 The results propose that although conservation of genes follows a tree-like trend with many phylogenetically informative sets of genes with high cardinality, the essentiality does not show a tree-like signal and most of the large sets of core essential genes belong to only one bacteria.
-We believe this happens due to the small number of essential genes.
+We believe this is due to the small number of essential genes.
+Each bacterium has about 300 to 500 essential genes among which 184 is core essential between all bacteria.
+A portion of the remaining essential genes are specific to each bacterium and many are shared between all bacteria except one.
+The number of remaining essential genes is so low that causes the essentiality trend not to be tree-like.
+Furthermore, some of the predicted essential genes might be artefacts of transposon mutagenesis method.
 
@@
 
-Furthermore, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
-We used three different methods for this.
-These methods are explained in...
-
+To further study the trend of essentiality changes, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
+We used three different methods for inferring core genes and core essential genes.
+These methods are explained in Section~\ref{sec:core}.
 The numbers predicted using these three methods are shown in Fig.\@ \ref{fig:annotated-speciestree}.
@@
 
-As these three methods lead to different results, we compared the differences between the genes found in these three methods.
+As these three methods lead to conflicting results, we compared the differences between the genes found in these three methods.
 For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods.
@@
 \section{Materials and Methods}
-\subsection{Transposon mutagenesis}
-We studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
-These strains are all selected from Enterobacteriaceae family.
-Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
-Here, we performed a transposon-directed insertion-site sequencing experiment to study the conservation of essentiality of genes in strains from 5 different species in this family.
-
-We used transposon mutagenesis which is a frequently used method for the study of essentiality.
-We generated single inserted mutants using Tn5 transposon and placed our mutants in a selective media for Tn5.
+\subsection{Transposon mutagenesis}\label{sec:transposoninsertion}
+We studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains (Fig.\@ \ref{fig:species-tree}) and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
+These strains are all selected from Enterobacteriaceae family and a transposon mutagenesis study have been performed on them.
+We generated single inserted mutants using Tn5 transposon and placed the mutants in a selective media for Tn5.
 Then, we picked the mutants and pooled them and performed PCR enrichment using the method described in \cite{barquist_tradis_2016}.
@@
 To calculate the insertion index for each gene, we summed up the number of transposon insertion sites observed in that gene.
-Since the lengths of the genes are different, we normalised the insertion index by dividing it by gene length.
+Since the lengths of the genes are different, the insertion indices were then normalised by dividing them by gene length.
 Our experiment is performed on different strains and the library density is different in each experiment.
-Therefore, in order to make the insertion indices comparable in all the strains, we normalised our insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
+Therefore, in order to make the insertion indices comparable in all the strains, we normalised the insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
 
-Based on insertion indices, we divided the genes into four groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
+Based on insertion indices, the genes were divided into four groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
 We utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
 The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
-We fitted a loess curve to this distribution to find where the two peaks separate from each other.
+A loess curve was fitted to this distribution to find where the two peaks separate from each other.
 The first peak shows the genes with no or just a few insertions which are considered as essential genes.
@@
 
-\subsection{Bias correction}
+\subsection{Bias correction}\label{sec:biascorrection}
 We observed a distance from origin of replication bias and also a positional bias within genes.
-Hence, we needed to correct for these biases before inferring the essentiality of genes from their insertion indices.
+Hence, these biases needed to be corrected before inferring the essentiality of genes from their insertion indices.
 We did not include genes shorter than 100 base-pairs in our study as they might not be targeted by any transposon due to their shortness.
@@
 
-\subsection{Conservation levels}
-To study whether each gene in our 13 organisms is conserved we proposed a program that clusters homologous proteins.
+\subsection{Conservation classes}\label{sec:conservationclasses}
+To study whether each gene in the 13 organisms is conserved we proposed a program that clusters homologous proteins.
 This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
@@
 
-\subsection{Defining core and core essential genes}
+\subsection{Defining core and core essential genes}\label{sec:core}
+We used three different methods for defining core and core essential genes: intersection, ancestral insertion index, and Dollo law. These three methods are explained in what follows.
+
 The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
@@
 Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
-%\section*{Acknowledgments}
+
+\section*{Acknowledgments}
 %

### 2016-10-17 documents/manuscript.tex
Wrote abstract and conclusion

@@
 \section*{Abstract}
-%Fitting multiple figures into very tight manuscripts while keeping it pleasant to read is challenging.
-Therefore figures are often simply attached to the very end of a manuscript file.
-While easier for the authors, this practice is inconvenient for readers.
-This \LaTeX template shows how to generate a compiled PDF with figures embedded into the text.
-It provides several examples of how to embed figures or tables directly into the text thus giving you a range of options from which you should choose the one best suited for your manuscript.
-Check out Schlegel et al., (2016) as example of use \cite{Schlegel2016}.
-%
+Many genes have been identified with advances in sequencing technology and genome annotation methods.
+However, not all of these genes are of the same importance.
+We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae.
+We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes.
+After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level.
+Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved.
+However, on average, essential genes are more likely to be conserved.
+
 % now start line numbers
@@
 Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}.
-Another method that is widely used due to its simplicity and accuracy is transposon mutagenesis along with high-throughput sequencing \cite{gawronski_tracking_2009, van_opijnen_tn-seq:_2009, langridge_simultaneous_2009, christen_essential_2011, goodman_identifying_2011, wetmore_rapid_2015, rubin_essential_2015}.
+Another method that is widely used due to its simplicity is transposon mutagenesis along with high-throughput sequencing \cite{gawronski_tracking_2009, van_opijnen_tn-seq:_2009, langridge_simultaneous_2009, christen_essential_2011, goodman_identifying_2011, wetmore_rapid_2015, rubin_essential_2015}.
 In this method, pools of single insertion mutants are constructed using transposon mutagenesis and the effect of each mutation on the survival of mutants is evaluated by sequencing the survivors \cite{barquist_approaches_2013}.
@@
 
-Our aim is to study the essentiality of genes in an evolutionary framework in 13 different organisms from Enterobacteriaceae family (Fig.\@ \ref{fig:species-tree}).
+Our aim is to study the essentiality of genes in an evolutionary framework in 14 different organisms from Enterobacteriaceae family (Fig.\@ \ref{fig:species-tree}).
 Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
@@
 \includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
-\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in Keio collection \cite{baba_construction_2006}.
+\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in EcoGene \cite{zhou_ecogene_2013}.
 We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
@@
 
-To study which functions are enriched in each class of essentiality, we gathered the descriptions of genes from their embl files.
-Then, we counted the repeat number of each word for the genes in each class and the genes that do not belong to that class and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
-The P-values are then corrected using Benjamini-Hochberg-Yekutieli procedure.
+To study which functions are enriched in each class of essentiality, we used the word enrichment method explained in Section~\ref{sec:enrichment}.
 Fig.\@ \ref{fig:essentiality-pval} shows the top 20 enriched words for each essentiality class.
@@
 
-We also conducted a pathway enrichment analysis for these three groups.
-For this, we downloaded pathway datasets for strains that were available in KEGG database.
-This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.
-Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class.
-Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.
+We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}.
 The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes.
@@
 
-In this section, we have counted the genes that are essential and conserved at each level of the phylogenetic tree.
+In this section, we compared the number of genes that were essential and conserved at each level of the phylogenetic tree.
 We were interested to see if the trend of the essentiality variation between different organisms is in agreement with the phylogenetic tree.
@@
 For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods.
-Then, we performed word enrichment analysis that was explained before on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method.
+Then, we performed word enrichment analysis explained in Section~\ref{sec:enrichment} on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method.
 Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure.
@@
 \subsection{Conservation classes}\label{sec:conservationclasses}
-To study whether each gene in the 13 organisms is conserved we proposed a program that clusters homologous proteins.
+To study whether each gene in the 14 organisms is conserved we proposed a program that clusters homologous proteins.
 This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
@@
 Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
-The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 13 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
+The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 14 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
 
@@
 
+\subsection{Word and pathway enrichment analysis}\label{sec:enrichment}
+To study the enrichment of words in a group of genes compared to a second group, We gathered the descriptions of genes from their embl files.
+Then, we counted the repeat number of each word for the genes in each group and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
+The P-values were then corrected using Benjamini-Hochberg-Yekutieli procedure.
+
+For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database.
+This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.
+Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class.
+Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.
+
 \subsection{Defining core and core essential genes}\label{sec:core}
@@
 
+\section{Conclusion}
+In this paper, we studied the relationship between the essentiality and conservation of genes in 14 bacteria from Enterobacteriaceae family.
+We first studied the biases that can affect our results and showed that transposon insertions are more abundant near the origin of replication.
+In addition, there is a slight preferred insertion motif bias and a G-C bias in A-T rich genes.
+Moreover we showed that transposon insertions are more abundant near the ends of essential genes compared to the internal region, while it is the opposite in beneficial losses.
+
+After correcting biases, we studied the essentiality versus conservation by dividing the genes into three classes of conservation: genus specific, single copy, and multi-copy, dividing them into four levels of essentiality: essential, ambiguous, non-essential, and benefical losses.
+We found that essential genes are mostly single copy, however there is a considerable number of genus specific genes in essential level which help to distinguish between genera.
+Furthermore, beneficial losses are mostly genus specific which means they are mostly new genes.
+The other finding was that multi-copy genes are mostly non-essential which can happen due to redundancy.
+
+We also found that the pattern of essentiality changes is not in agreement with the phylogenetic tree due to the small number of essential genes.
+Also the ratio between core essential genes and core genes can be either constant or increasing depending on the method that we use for defining core and core essential genes.
+\textcolor{red}{\{WHAT?\}}
+
+Overall, we have found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.
+
 \section*{Acknowledgments}

### 2016-10-23 documents/manuscript.tex
Developed a new protein clustering algorithm using MCL

@@
 However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis.
-Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}.
+Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, green_insertion_2012, rubin_essential_2015, kimura_nucleoid_2016}.
 We performed a detailed study of these biases.
@@
 The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases.
+\textcolor{red}{\{Our genomes are G-C rich, so it makes sense to see more G-Cs in this plot.
+Use BLogo\}}
 
@@
 The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes.
-However, as mobile genetic elements are not stored in KEGG database, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.
+However, as mobile genetic elements are not stored in KEGG database \cite{kanehisa_kegg:_2000}, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.
 
@@
 We used Hieranoid for clustering orthologous genes.
+Hieranoid need a species tree for clustering genes.
+We used the method explained in Section~\ref{sec:phylogenetic} to generate this tree.
 
@@
 The number of insertions in a gene implies the degree of essentiality for that gene.
+
+\subsection{Generating the species tree}\label{sec:phylogenetic}
+We first ran search and align commands from PhyloSift package \cite{darling_phylosift:_2014} to select gene markers for generating a phylogenetic tree and aligning them.
+Then, we concatenated the protein alignments for all 14 genomes and ran RaxML \cite{stamatakis_raxml_2014} with "PROTGAMMALG4M" amino acid substitution model, "a" algorithm and 100 alternative runs on distinct starting trees.
 
@@
 
-For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database.
+For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database \cite{kanehisa_kegg:_2000}.
 This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.

### 2016-10-23 bin/homclust2.py

@@ -1,0 +1,221 @@
+# USAGE: homclust.py [-options] <seq_fasta_file> <seqdb_fasta_file> <output_directory>
+
+# Needs Python, HMMER 3, and mafft installed.
+
+# Clusters homologous proteins in different species. The first input file is a protein fasta file from one
+# species and the second input file is a file resulted from merging all fasta files (including the first input file)
+# for all given species.
+
+# For merging fasta files, one can use:
+# cat <file1> <file2> ... <filen> > seqdb.fasta
+
+#! /bin/sh
+""":"
+exec python $0 ${1+"$@"}
+"""
+
+from os import path, mkdir, system, listdir
+from shutil import rmtree
+from Bio import SeqIO
+from argparse import ArgumentParser
+from math import sqrt
+from timeit import default_timer
+
+
+def mergefiles(inpath, outpath):
+    list_of_files = listdir(inpath)
+    with open(outpath, 'w') as tofile:
+        write = tofile.write
+        for filename in list_of_files:
+            with open(inpath + '/' + filename, 'r') as fromfile:
+                for line in fromfile:
+                    write(line)
+    numspecies = len(list_of_files)
+    return numspecies
+
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+
+def extractfeatures(inpath, outpath):
+    sequences = read_fasta_sequences(inpath)
+    with open(outpath, 'w') as tofile:
+        keys = sequences.keys()
+        write = tofile.write
+        for key in keys:
+            write(sequences[key].id)
+            seq = sequences[key].seq.upper()
+            aadict = {'A': 0, 'R': 0, 'N': 0, 'D': 0, 'C': 0, 'Q': 0, 'E': 0, 'G': 0, 'H': 0, 'I': 0, 'L': 0,
+                      'K': 0, 'M': 0, 'F': 0, 'P': 0, 'S': 0, 'T': 0, 'W': 0, 'Y': 0, 'V': 0}
+            length = len(seq)
+            for item in seq:
+                if item in aadict.keys():
+                    aadict[item] += 1
+                elif item == 'B':
+                    aadict['N'] += 0.5
+                    aadict['D'] += 0.5
+                elif item == 'Z':
+                    aadict['Q'] += 0.5
+                    aadict['E'] += 0.5
[…]

### 2016-10-23 bin/montecarlo.R

@@ -9,7 +9,7 @@ plots = list()
 sampledplots = list(list())
 sumlength = list()
 weights = list()
-numsamples = 100
+numsamples = 2
 locusid = c()
 for (filename in list_of_files)
 {
@@ -22,6 +22,7 @@ for (filename in list_of_files)
   plotsdata = as.data.frame(cbind(1:length(plots[[lid]]), as.vector(plots[[lid]])))
   names(plotsdata) <- c("position", "num_inserts")
   mdl <- loess(num_inserts~position, plotsdata, span=0.2, family = "gaussian")
+  # mdl <- loess.smooth(plotsdata$position, plotsdata$num_inserts, span=0.2)
   weights[[lid]] = predict(mdl)
   weights[[lid]] = weights[[lid]] / sum(weights[[lid]])
   sumlength[[lid]] = c(sum(plots[[lid]]), length(plots[[lid]]))

### 2016-11-08 bin/montecarlo.py
Ran Monte Carlo, edited protein clustering alg

@@ -1,0 +1,26 @@
+from os import listdir
+from collections import defaultdict
+import numpy as np
+
+fastas_dir = '../data/fasta-protein/chromosome'
+plots_dir = '../data/plot-files/chromosome'
+list_of_files = listdir(plots_dir)
+plots = defaultdict(list)
+sampledplots = []
+sumlength = []
+weights = []
+numsamples = 1000
+locusid = []
+list_of_files = ['t.plot']
+
+for filename in list_of_files:
+    plotfile = []
+    with open(plots_dir + '/' + filename, 'r') as fromfile:
+        for line in fromfile:
+            cells = line.split()
+            plotfile.append(cells)
+    lid = filename.split('.')[0]
+    locusid.append(lid)
+    plots[lid] = [int(plotfile[i][0])+int(plotfile[i][1]) for i in range(len(plotfile))]
+    for i in range(numsamples):
+        sampledplots.append(np.random.choice(plots[lid], len(plots[lid])))

### 2016-11-08 bin/montecarlo_window.R

@@ -1,33 +1,49 @@
 library(stringr)
 library("MASS")
 library("DESeq2")
-library("mclust")
+library(bigmemory)
+#library("mclust")
 fastas_dir <- "~/EnTrI/data/fasta-protein/chromosome"
 plots_dir <- "~/EnTrI/data/plot-files/chromosome"
-outdir <- "~/EnTrI/results/monte-carlo/"
 list_of_files <- list.files(path=plots_dir, full.names=T, recursive=FALSE)
-plots = list()
 sampledplots = list(list())
+plots = list()
 sumlength = list()
-numsamples = 10
+weights = list()
+windowsize = 50
+numsamples = 2
+locusid = c()
 for (filename in list_of_files)
 {
   plotfile = as.matrix(read.table(filename, as.is=TRUE))
-  locusid = strsplit(basename(filename),"\\.")[[1]][1]
-  plots[[locusid]] = plotfile[,1] + plotfile[,2]
-  plotsdata = as.data.frame(cbind(1:length(plots[[locusid]]), as.vector(plots[[locusid]])))
+  lastwindowsize = nrow(plotfile) %% windowsize
+  lid = strsplit(basename(filename),"\\.")[[1]][1]
+  locusid = c(locusid,lid)
+  # plots[[lid]] = plotfile[,1] + plotfile[,2]
+  for (i in seq(1, (nrow(plotfile)%/%windowsize)))
+  {
+    plots[[lid]][i] = sum(plotfile[((i-1)*windowsize+1):(i*windowsize),1] + plotfile[((i-1)*windowsize+1):(i*windowsize),2])
+  }
+  i =+ 1
+  plots[[lid]][i] = sum(plotfile[((i-1)*windowsize+1):((i-1)*windowsize+lastwindowsize),1] + plotfile[((i-1)*windowsize+1):((i-1)*windowsize+lastwindowsize),2])
+  # plot(plots[[lid]], pch = '.', ylim=c(0,10))
+  # lines(loess.smooth(1:length(plots[[lid]]), plots[[lid]], span=0.2, family = "gaussian"), col=2, lwd=5)
+  plotsdata = as.data.frame(cbind(1:length(plots[[lid]]), as.vector(plots[[lid]])))
   names(plotsdata) <- c("position", "num_inserts")
   mdl <- loess(num_inserts~position, plotsdata, span=0.2, family = "gaussian",
[…]

### 2016-11-08 bin/parsetreetest_del.py

@@ -1,0 +1,427 @@
+from os import path, mkdir, system, listdir
+from shutil import rmtree
+#from Bio import SeqIO
+from argparse import ArgumentParser
+from random import randint
+from re import findall
+from timeit import default_timer
+
+
+def addtostring(mainstring, position, addingstring):
+    newstring = mainstring[:position] + addingstring + mainstring[position:]
+    return newstring
+
+
+def randomndigits(n):
+    start = 10 ** (n - 1)
+    end = (10 ** n) - 1
+    return randint(start, end)
+
+
+def addnodenames(tree):
+    n = 10
+    i = 0
+    prevchar = ''
+    while i < len(tree):
+        char = tree[i]
+        if char == ':' and prevchar == ')':
+            nodename = 'n' + str(randomndigits(n))
+            tree = addtostring(tree, i, nodename)
+            i += len(nodename)
+        elif char == ';' and prevchar == ')':
+            nodename = 'n' + str(randomndigits(n)) + ':0'
+            tree = addtostring(tree, i, nodename)
+            i += len(nodename)
+        prevchar = tree[i]
+        i += 1
+    return tree
+
+
+def findnodes(tree):
+    # findall_results = findall('[\(\),]([^\(^\)^,^:]*):([^,^\)^;]*)', tree)
+    nodes = findall('[\(\),]([^\(^\)^,^:]*):', tree)
+    return nodes
+
+
+def calculaterootdistance(tree, nodes):
+    rootdistance = {}
+    for item in nodes:
+        i = 0
+        numpar = 0
+        nodedistance = 0
+        numparsaver = 0
+        distance = ''
+        nodename = ''
+        char = tree[i]
+        while i < len(tree):
+            if char == '(':
+                numpar += 1
+                i += 1
+                char = tree[i]
+                while char != ':' and char != '(':
+                    nodename += char
+                    i += 1
+                    char = tree[i]
+            elif char == ')':
+                if item == nodename or numpar < numparsaver:
+                    nodedistance += float(distance)
+                    numparsaver = numpar
+                numpar -= 1
+                i += 1
+                char = tree[i]
+                distance = ''
+                nodename = ''
+                while char != ':' and char != '(':
+                    nodename += char
+                    i += 1
+                    char = tree[i]
+            elif char == ':':
+                i += 1
+                char = tree[i]
+                while char != ',' and char != ')' and char != ';':
+                    distance += char
+                    i += 1
+                    char = tree[i]
+            elif char == ',':
+                if item == nodename or numpar < numparsaver:
[…]

### 2016-11-08 bin/tree-data.R

@@ -1,0 +1,102 @@
+strain =c(
+374/4699,
+438/5055,
+328/3220,
+353/4216,
+463/4518,
+468/4515,
+512/4374,
+389/4468,
+337/4199,
+380/4094,
+447/4317,
+520/4399,
+447/4494,
+298/4149
+)
+
+
+speciesInt = c( 
+325/4564,
+328/3220,
+353/4216,
+289/3588,
+253/3461
+)
+
+
+speciesAII = c( 
+328/3220,
+353/4216,
+346/4564,
+319/3588,
+337/3601
+)
+
+speciesDollo = c( 
+328/3220,
+353/4216,
+325/4564,
+354/3892,
+365/3959
+)
+
+
+subFamilyInt = c(
+280/2960,
+233/2604,
+251/2668
+)
+
+subFamilyAII = c(
+323/2960,
+327/2605,
+274/2668
+
+)
+
+subFamilyDollo = c(
+336/3408,
+409/4120,
+284/2856
+)
+
+familyInt = c(
+184/1908
+)
+
+familyAII = c(
+273/1908
+)
+
+familyDollo = c(
+410/4344
+)
+
+
+pdf(file = '~/EnTrI/figures/method-comparison.pdf')
+
+plot(4*strain/strain, strain, xlim=c(1,4), ylim=c(0.07,0.15), col="olivedrab4", ylab="proportion of ancestral genes", xaxt="n", xlab="",pch=19)
+
+points(3*speciesInt/speciesInt,speciesInt, col="red4", pch=19)
+points(3*speciesAII/speciesAII,speciesAII, col="olivedrab4", pch=19)
+points(3*speciesDollo/speciesDollo,speciesDollo, col="cyan3", pch=19)
+
+
+points(2*subFamilyInt/subFamilyInt,subFamilyInt, col="red4", pch=19)
+points(2*subFamilyAII/subFamilyAII,subFamilyAII, col="olivedrab4", pch=19)
+points(2*subFamilyDollo/subFamilyDollo,subFamilyDollo, col="cyan3", pch=19)
+
+points(1*familyInt/familyInt,familyInt, col="red4", pch=19)
+points(1*familyAII/familyAII,familyAII, col="olivedrab4", pch=19)
+points(1*familyDollo/familyDollo,familyDollo, col="cyan3", pch=19)
+
+lines(1:4, c(median(familyInt), median(subFamilyInt), median(speciesInt), median(strain) ), col="red4", lwd=3)
+
[…]

### 2016-12-06 documents/manuscript.tex
Ran Monte Carlo, edited protein clustering alg
Summed over ii, largest uninterrupted fraction, and monte carlo results

@@
 {\Large
-\textbf\newline{Are essential genes conserved?}
+\textbf\newline{Title}
 }
@@
 % now start line numbers
-%\linenumbers
+\linenumbers
 
@@
 Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium and \textit{Salmonella} Typhi and found that the essentiality of genes differs in different organisms.
-In a similar study, Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges.
+In a similar study, Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes and ncRNAs in \textit{Salmonella} serovars Typhi and Typhimurium.
 Although there are many studies on differentiation of essentiality in different organisms, these studies usually include two or three strains.
@@
 %This defines the bibliographies style. Search online for a list of available styles.
-\bibliographystyle{abbrv}
+\bibliographystyle{apalike}
 

### 2016-12-06 bin/montecarlo.R

@@ -102,6 +102,7 @@ for (filename in list_of_files)
   dds <- DESeqDataSetFromMatrix(counttable, DataFrame(condition), ~ condition)
   res <- DESeq(dds)
   result<-results(res)
+  
   # mod = Mclust(result$log2FoldChange, G=1:2)
   essentiality = c()
   for (i in 1:length(result@rownames))

### 2017-02-13 documents/manuscript.tex
Edited manuscript, found some interesting genes that are essential in some clades

--- 2016-10-17 documents/manuscript.tex
@@
 % adjust caption style
-\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
+\usepackage[font=small,aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
 
@@
 \usepackage{graphicx}
+\usepackage{subcaption}
 \graphicspath{{../figures/}}
@@
 \reversemarginpar
+\usepackage{amsmath}
 
@@
 \section*{Abstract}
-Many genes have been identified with advances in sequencing technology and genome annotation methods.
+%Many genes have been identified with advances in sequencing technology and genome annotation methods.
 However, not all of these genes are of the same importance.
@@
 \section{Introduction}
-With the advent of sequencing technologies and genome annotation methods many genes have been identified.
-However, not all of these genes are of the same importance for the growth of an organism.
-So far, scientists have studied the essential genes in organisms from different domains of life \cite{luo_deg_2014}.
-These studies can give us new insights for developing new antibiotics that target essential genes of pathogenic bacteria \cite{clatworthy_targeting_2007, peters_comprehensive_2016} and synthesising minimal genomes \cite{hutchison_global_1999, hutchison_design_2016, reus_blueprint_2016}.
-Different methods have been used for studying the essentility of genes in prokaryotes.
-Baba et al.\@ \cite{baba_construction_2006} have made a library of single gene deletions using phage lambda Red recombination system to screen essential genes while another group have used antisense RNA knockdowns for this purpose \cite{xu_staphylococcus_2010}.
-Another method that is widely used due to its simplicity is transposon mutagenesis along with high-throughput sequencing \cite{gawronski_tracking_2009, van_opijnen_tn-seq:_2009, langridge_simultaneous_2009, christen_essential_2011, goodman_identifying_2011, wetmore_rapid_2015, rubin_essential_2015}.
-In this method, pools of single insertion mutants are constructed using transposon mutagenesis and the effect of each mutation on the survival of mutants is evaluated by sequencing the survivors \cite{barquist_approaches_2013}.
-This can lead to the identification of essential genes.
+%The study of cells can give us new insights for understanding the foundation of life.
+However, the naturally existing cells are very complex which hampers the study of their constituents.
+Restricting these studies to the genes that are required for the growth of cells can reduce this complexity and enable us to study life in its simplest form \cite{juhas_essence_2011}.
 
-Now that the essentiality of genes can be evaluated using different methods, it is possible to compare the essentiality data in different organisms and investigate the differentiation of essentiality of their genes.
-Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
-Freed et al.\@ \cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that some genes have gained essentiality in \textit{Shigella flexneri} while there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}.
-Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium and \textit{Salmonella} Typhi and found that the essentiality of genes differs in different organisms.
-In a similar study, Barquist et al.\@ \cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to study the differentiation of the essentiality of genes in \textit{Salmonella} serovars Typhi and Typhimurium which has led to divergence in their pathogenecity and host ranges.
-Although there are many studies on differentiation of essentiality in different organisms, these studies usually include two or three strains.
+Generating genomic variants that carry genes responsible for essential cellular processes can open up new research directions.
+Adding genes or metabolic pathways to cells provides new variants with specialised phenotypes.
+Modified cells have different potential applications in biotechnology \cite{juhas_bacillus_2014}, fuel production \cite{seo_synthetic_2013}, healthcare, and food production \cite{juhas_meeting_2013}.
+Another important application for studying essential genes is in drug discovery \cite{juhas_essential_2012}.
+Infectious diseases are among the top major causes of mortality worldwide.
+Even though antibiotic resistance is growing among bacteria, the antibiotic discovery and development rate is diminishing \cite{fischbach_antibiotics_2009, nathan_antibiotics_2004}.
+Therefore, it is urgent to find new drugs for infectious diseases.
+New antibiotics target genes that are essential for the survival of pathogenic bacteria in order to control disease.
+Some well-known antibiotics that target essential functions are tetracyclines that bind to small ribosomal subunit and interfere with protein translation \cite{brodersen_structural_2000}, penicillins that target peptidoglycan and inhibit the cell wall synthesis \cite{chung_rapid_2009}, and quinolones that target DNA Gyrase \cite{marcusson_interplay_2009}.
+%Some candidate antibiotic targets found by investigating essential genes are trans-translation genes \textit{ssrA} and \textit{smpB} from \textit{helicobacter pylori} \cite{thibonnier_trans-translation_2008}, \textit{CcrM} and \textit{MenH} which are bacterial methyltransferases, and \textit{lptD} (previously known as \textit{ostA}) in \textit{Pseudomonas aeruginosa} \cite{srinivas_peptidomimetic_2010} whose essential role is yet under investigation \cite{werneburg_inhibition_2012}.[SOME ANTIBIOTICS THAT TARGET ESSENTIAL GENES ARE...]
 
-Our aim is to study the essentiality of genes in an evolutionary framework in 14 different organisms from Enterobacteriaceae family (Fig.\@ \ref{fig:species-tree}).
-Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
+In the earliest attempt for the identification of essential genes, Mushegian and Koonin compared the genomes of \textit{Haemophilus influenzae} and \textit{Mycoplasma genitalium}, assuming that the genes that are shared in these two phylogenetically distant bacteria are indispensable and reported 256 genes fulfilling this requirement \cite{mushegian_minimal_1996}.
+With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes declined to less than 50 which is not enough for performing all essential functions in a cell \cite{charlebois_computing_2004}.
+Therefore, the use of experimental methods for the identification of essential genes is vital.
+Researchers have now studied the essential genes in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods.
+Baba et al.\@ \cite{baba_construction_2006} made a library of single gene deletions for \textit{Escherichia coli} K-12.
+The 303 genes where viable \textit{Escherichia coli} colonies failed to grow are the candidate essential genes.
+Another group used an antisense RNA knockdown approach to study gene essentiality in \textit{Staphylococcus} aureus \cite{forsyth_genome-wide_2002}.
+In this method, if the expression of an antisense RNA hinders the growth of the cell, its cognate gene is known as essential.
+Both of these methods are labour intensive and are dependent on the accuracy of genome annotations.
+Another widely used procedure is transposon mutagenesis combined with high-throughput sequencing \cite{chao_design_2016, van_opijnen_transposon_2013, barquist_approaches_2013} which includes different approaches namely, Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}.
+These procedures differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}.
+Nonetheless, all share the same workflow: pools of single insertion mutants are constructed using transposon mutagenesis.
+After a growth phase, mutants that are fitter outnumber the less fit ones.
+Using high-throughput sequencing and tallying transposon junctions gives an indication of whether a genomic region is essential or not.
+A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium and conversely, a low number of transposon insertions indicates that a gene is essential in the medium.
+
+There are two groups of essential genes: core essential genes are indispensable for all cells, and accessory essential genes that are required for some organisms.
+Core essential genes can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003} and have been used for the synthesis of minimal cells \cite{hutchison_design_2016}.
+On the other hand, accessory essential genes are helpful in the study of specific lineages.
+Accessory essential genes may be useful in species-specific antibiotic discovery.
+Antibiotics that target core essential genes in pathogens may not be ideal, as they may attack homologous genes in their hosts or commensal bacteria.
+
+%Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
+Freed et al.
+\cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}, while some genes are only essential in \textit{Shigella flexneri}.
+These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and soxR-reducing complex, many of which are essential in \textit{Shigella flexneri} due to the absence of paralogs or other alternative systems that exist in \textit{Escherichia coli}.
+%The difference in essentiality stems from four reasons: the disruption of many of the genes in \textit{Shigella flexneri} 2a 2457T impede the growth, but these genes are not actually essential; the conditions were different in the two experiments done on \textit{Escherichia coli} K12 BW25113 and \textit{Shigella flexneri} 2a 2457T; many of the genes were not actually essential for growth, but essential for transposon insertion and antibiotic resistance; and
+
+%PPG: Reduce, summarise main findings of Canals & Barquist. e.g. Most essential genes are also essential in related species, yet there are some serovar-specific essential genes [GIVE SOME NUMBERS]... 
+Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium ATCC 14028, two isolates of \textit{Salmonella typhi} Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{Escherichia coli} K12 BW25113 and found that these cells share 268 essential genes.
+Nine genes are essential in \textit{Escherichia coli} and not essential in \textit{Salmonella}s, three of which can tolerate insertions only in some parts and one has distant paralogs in \textit{Salmonella}s that is missing in \textit{Escherichia coli}.
+For other genes, there are a few insertions in \textit{Salmonella}s, but the number is not small enough to call those genes essential.
+Moreover, 159 genes were almost essential in all \textit{Salmonella}s but not in \textit{Escherichia coli}.
+These include genes involved in replication and genes related to ribosome and its accessory proteins.
+The authors also found 26 genes that are under greater selection in \textit{Salmonella} Typhimurium compared to \textit{S.
+typhi} and 10 genes vice versa.
+Barquist et al.
+\cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{Salmonella} Typhi Ty2, \textit{Salmonella} Typhimurium SL1344, and \textit{Escherichia coli} K12 BW25113.
+%THIS IS IMPORTANT:
+These genomes share 228 essential genes which are mostly involved in cell division, transcription, translation, and fatty acid and peptidoglycan biosynthesis.
+Additionally, many of the serovar-specific essential genes in \textit{Salmonella}s are phage repressors.
+Another key difference between these two serovars is that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{Salmonella} Typhimurium and not essential in \textit{salmonella} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.
+
+Although there are some studies on differentiation of essentiality in different organisms, these studies usually include a few genomes.
+We are going to extend these studies to a larger scale in Enterobacteriaceae family.
+Our aim is to study the essentiality of genes in in 16 different organisms.
+These include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{Klebsiella pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{Salmonella} Typhimurium SL1344, \textit{Salmonella} Typhimurium SL3261, \textit{Salmonella} Typhimurium D23580, \textit{Salmonella} Typhimurium A130, \textit{Salmonella} Enteritidis P125109, \textit{Salmonella} Typhi Ty2, \textit{Escherichia coli} ST131 EC958, \textit{Escherichia coli} UPEC ST131, \textit{Escherichia coli} ETEC CS17, \textit{Escherichia coli} ETEC H10407, \textit{Escherichia coli} BW25113, and \textit{Escherichia coli} K-12 MG1655.
+The phylogenetic tree containing all these species is depicted in Fig.\@ \ref{fig:species-tree}.
+
+%WHAT FATEMEH DID....
+
+%Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
 In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study.
@@
 \includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
-\caption{The species tree containing the 13 strains under study and Escherichia \textit{coli} K-12 MG1655 studied in EcoGene \cite{zhou_ecogene_2013}.
+\caption{The species tree containing the 16 genomes used in this study.
 We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
@@
 
-\section{Results and discussion}
-Throughout time, species can gain or lose genes.
+\section{Results and Discussion}
+\subsection{Incorporating multiple measures of gene essentiality improves prediction}
+In this section we will compare a number of methods of quantifying the essentiality of genes based on transposon insertion data and introduce our own method.
+\subsubsection{Comparison of old methods}
+A number of methods have been used for evaluating the essentiality of genes using transposon insertion data.
+Freed et al.
+\cite{freed_combining_2016} have compared eleven features that can quantify the essentiality of genes.
+These include: the number of insertion sites within genes; the mean distance between insertion sites, the median distance between insertion sites, the number of base pairs before the first insertion in the $5^\prime$ end, the ratio between the number of base pairs before the first insertion in the $5^\prime$ end and gene length, the number of base pairs before the second insertion in the $5^\prime$ end, the ratio between the number of base pairs before the second insertion in the $5^\prime$ end and gene length, the length of the longest uninterrupted region in the gene, the ratio between the length of the longest uninterrupted region in the gene and gene length, the length of the longest region in the gene containing at most one insertion divided by gene length, and the number of insertion sites divided by gene length.
+Among these, the average distance between insertion sites and the length of the longest uninterrupted region were the best predictors.
+
+Barquist et al.
+\cite{barquist_tradis_2016} have used insertion indices which are calculated by dividing the number of insertion sites by gene length for defining the essentiality of genes.
+Plotting all insertion indices for all genes in a genome gives a bimodal plot, each mode representing a group of genes (essential or non-essential).
+They have fitted two gamma distributions to these modes and calculated the log odds value for each gene to test which mode it belongs to.
+
+Turner et al.
+\cite{turner_essential_2015} have randomly sampled insertions in a genome and using DESeq package \cite{anders_differential_2010} calculated log fold changes and P-values for the actual number of insertions compared to the expected number of insertions obtained from the sampling method.
+
+We have compared the predictive power of the average distance between insertion sites in a gene, the length of the longest uninterrupted region, insertion index, log odds value obtained from insertion index value, and P-value and log fold change calculated using DESeq package after sampling.
+To evaluate the accuracy of these methods, we have compared the predicted essential genes to the essential genes in \textit{Escherichia coli} K-12 MG1655 in EcoGene database \cite{zhou_ecogene_2013}.
+The results are depicted in Figure~\ref{fig:benchmark}.
+Among all methods, those that try to fit a distribution to data and predict essentiality based on that, namely log odds value from insertion indices and P-value from DESeq after sampling, are the least accurate methods.
+Sampling and then calculating log fold changes using DESeq package is the most accurate predictor and the other three methods are also good predictors despite their simplicity.
+\subsubsection{PCA based method}\label{sec:essentiality-call}
+We selected four of the predictors in previous section: insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region.
+These predictors were selected as they were more accurate and less dependent on each other.
+Afterwards, we ran principal component analysis on these predictors using \emph{prcomp} function in R and selected the first principal component.
+As Figure~\ref{fig:benchmark} shows, PCA results gave us the most powerful tool for quantifying the essentiality of genes compared to other methods.
+We have plotted the results for one sample genome and the results for other genomes are shown in {\color{red}SUPPLEMENTARY}.
+Figure~\ref{fig:pca} shows the distribution of PCA results after zero mean unit variance normalisation.
+We call this value NPEQ (Normalised Pca based Essentiality Quantifier) in this paper.
+The genes whose NPEQ values are less than -1.644854 after normalisation are beneficial losses and the genes with NPEQ greater than 1.644854 are essential.
+The rest are non-essential genes.
+The cutoff 1.644854 has been selected because if we assume that NPEQ distribution is normal, this cutoff shows the P-value 0.95.
+Moreover, by trying to define a cutoff that maximises Matthews correlation coefficient for each genome using NPEQ, we get 1.650449 as the average cutoff which is very close to this value.
+
+\begin{figure}
+\captionsetup[subfigure]{justification=centering}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[page=1, scale=0.4]{essential-call-comparison-ETEC.pdf}
+  \caption{}
+  \label{fig:benchmark}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[page=10, scale=0.4]{essential-call-comparison-ETEC.pdf}
+  \caption{}
+  \label{fig:pca}
+\end{subfigure}
+\caption{(a) The accuracy of 7 different prediction methods for quantifying the essentiality of genes.
+The higher the area under the curve, the more accurate the method is.
+(b) The distribution for our proposed method (NPEQ)}
+\label{fig:essentiality-call}
+\end{figure}
+
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c c}
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BN373.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BW25113.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-CS17.pdf}\\
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-EC958.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ENC.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ERS227112.pdf}\\
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ETEC.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-NCTC13441.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ROD.pdf}\\
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SEN.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL1344.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL3261.pdf}\\
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STM.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STMMW.pdf}&
+%\includegraphics[page=1, scale=0.23]{essential-call-comparison-t.pdf}\\
+%\end{tabular}
+%\caption{The plots show the accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is.}
+%\label{fig:benchmark}
+%\end{figure*}
+
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c c}
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BN373.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BW25113.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-CS17.pdf}\\
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-EC958.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ENC.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ERS227112.pdf}\\
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ETEC.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-NCTC13441.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ROD.pdf}\\
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SEN.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL1344.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL3261.pdf}\\
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STM.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STMMW.pdf}&
+%\includegraphics[page=10, scale=0.22]{essential-call-comparison-t.pdf}\\
+%\end{tabular}
+%\caption{The plots show the distribution of our new prediction method which is calculated by running PCA on insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region and selecting first principal component.
+This method is called NPEQ.}
+%\label{fig:pca}
+%\end{figure*}
+\subsection{TraDIS data is biased}
+In Section~\ref{sec:essentiality-call} we proposed NPEQ method that quantifies the level of essentiality of a gene.
+However, if the transposon insertion is biased to specific regions in the genome, it can increase/decrease the values predicted by our method for genes and put them in a different level of essentiality.
+Different articles have reported biases in transposon insertion using Tn5 \cite{barquist_comparison_2013, green_insertion_2012, rubin_essential_2015, canals_high-throughput_2012, langridge_simultaneous_2009}.
+We performed a detailed study of these biases.
+The biases that we studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
+\subsubsection{Distance from the origin of replication bias}
+While a study has reported Tn5 insertion bias near the origin of replication \cite{barquist_comparison_2013}, another study has reported no bias \cite{rubin_essential_2015}.
+To study the bias towards the position of gene within genome, we plotted NPEQ for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Fig.\@ \ref{fig:distance-bias}.
+The figure indicates that NPEQ increases when the genes are located further from the origin of replication.
+When the bacteria are under replication during the transposn insertion process, there are more copies of the genes close to the origin of replication than the genes further away due to the initiation of different replication forks.
+This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
+%One might argue that the bias that we have observed may have happened due to the abundance of essential genes near the origin of replication.
+Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication.
+
+\begin{figure}
+\captionsetup[subfigure]{justification=centering}
+\begin{subfigure}{.49\textwidth}
+  \centering
+  \includegraphics[page=26, scale=0.25]{biases.pdf}
+  \caption{}
+  \label{fig:distance-bias}
+\end{subfigure}
+\begin{subfigure}{.49\textwidth}
+  \centering
+  \includegraphics[page=25, scale=0.25]{biases.pdf}
+  \caption{}
+  \label{fig:GC-bias}
+\end{subfigure}
+\begin{subfigure}{\textwidth}
+  \centering
+  \includegraphics[scale=0.6]{100logo-prob-edited.pdf}
+  \caption{}
+  \label{fig:logos}
+\end{subfigure}
+\begin{subfigure}{.49\textwidth}
+  \centering
+  \includegraphics[scale=0.25, page=2]{insertion-position-bias.pdf}
+  \caption{}
+  \label{fig:insertion-position-bias-ess}
+\end{subfigure}
+\begin{subfigure}{.49\textwidth}
+  \centering
+  \includegraphics[scale=0.25, page=4]{insertion-position-bias.pdf}
+  \caption{}
+  \label{fig:insertion-position-bias-ben}
+\end{subfigure}
+\caption{(a)The distance of the genes from DnaA gene normalised by the lengths of the genomes versus NPEQ.
+The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
+The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias.
+(b) G-C content of genes against their NPEQ values.
+The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias.
+(c) Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
+The height of each character shows the relative frequency of that character.
+(d) and (e) The plots show the average insertion index in percentiles of all essential genes (d) and beneficial losses (e).
+The genes are divided into 3 segments: 5\% of the genes on the $5^\prime$ end, 20\% of the genes on the $3^\prime$ end, and the rest in the middle.
+These are shown by khaki, slate gray, and violet red respectively.}
+\label{fig:biases}
+\end{figure}
+
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c c}
+%\includegraphics[page=2, scale=0.22]{biases.pdf}&
+%\includegraphics[page=6, scale=0.22]{biases.pdf}&
+%\includegraphics[page=10, scale=0.22]{biases.pdf}\\
+%\includegraphics[page=14, scale=0.22]{biases.pdf}&
+%\includegraphics[page=18, scale=0.22]{biases.pdf}&
+%\includegraphics[page=22, scale=0.22]{biases.pdf}\\
+%\includegraphics[page=26, scale=0.22]{biases.pdf}&
+%\includegraphics[page=30, scale=0.22]{biases.pdf}&
+%\includegraphics[page=34, scale=0.22]{biases.pdf}\\
+%\includegraphics[page=38, scale=0.22]{biases.pdf}&
+%\includegraphics[page=42, scale=0.22]{biases.pdf}&
+%\includegraphics[page=46, scale=0.22]{biases.pdf}\\
+%\includegraphics[page=50, scale=0.22]{biases.pdf}&
+%\includegraphics[page=54, scale=0.22]{biases.pdf}&
+%\includegraphics[page=58, scale=0.22]{biases.pdf}\\
+%\end{tabular}
+%\caption{The plots show the distance of the genes from DnaA gene normalised by the lengths of the genomes versus the insertion indices of the genes.
+The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
+The red curves show the loess curve when the smoothness parameter is $0.2$.}
+%\label{fig:distance-bias}
+%\end{figure*}
+\subsubsection{Preferred nucleotide composition bias}
+Another concern while inferring essentiality from transposon insertion data is that transposon insertion is biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposon insertion is biased towards, instead of their essentiality level.
+During Tn5 transposition, a sequence of 9 nucleotides is duplicated.
+Lodge et al.
+\cite{lodge_transposon_1988} showed that these duplicated regions have G-C pairs at two ends and replacing these G-C pairs with A-T pairs reduces the number of transposon insertions by more than fivefold.
+Goryshin et al.
+\cite{goryshin_tn5/is50_1998} reported the palindromic sequence A-GNTYWRANC-T as the consensus target site for Tn5 transposition where the 9 letters in middle show the consensus sequence in the duplicated region.
+%They have also showed that insertion sites are usually grouped together and have 5 base pairs distance from each other.
+In some other research a similar consensus motif has been found for Tn5 \cite{canals_high-throughput_2012} which is CGCGCA-GTTYWRAAC-TGCGCG.
+Others \cite{green_insertion_2012,rubin_essential_2015} have not found such a sequence but shown that the duplicated regions are G-C rich.
+We used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from duplicated regions and 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
+The results in Figure \ref{fig:logos} show the consensus motif that we have found is very similar to \cite{green_insertion_2012} and the only difference is in positions 3 and 7 within the duplicated region.
+
+%\begin{figure*}
+%\includegraphics[scale=0.9]{100logo-prob-edited.pdf}
+%\caption{Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
+The height of each character shows the relative frequency of that character.}
+%\label{fig:logos}
+%\end{figure*}
+
+The other possible source of bias is if transpositions are more inclined to G-C or A-T rich regions.
+Rubin et al.
+\cite{rubin_essential_2015} have reported that the number of Tn5 insertions rises with the increase of G-C content and Green et al.
+\cite{green_insertion_2012} have shown that the highest number of insertions occur in high G-C content regions.
+On the other hand, Langridge et al.
+\cite{langridge_simultaneous_2009} have seen an increase in the number of Tn5 insertions in 40\% G-C content.
+In Figure \ref{fig:GC-bias}, we plotted the G-C content of genes versus their NPEQ values.
+The red curves are loess curves with smoothness parameter 0.2 and the green line shows no bias.
+%In most cases (BW25113, EC958, ENC, ETEC, NCTC13441, ROD, SEN, STM, and t)
+NPEQ decreases gradually (which means an increase in the number of insertions) as G-C content decreases and then somewhere between 40\% and 50\% G-C content, NPEQ starts to rise again in almost all genomes.
+%In three cases (SL1344, SL3261, and STMMW) NPEQ loess curve starts to fall off again at about 60\% G-C content which is probably due to a small number of genes with low NPEQ values in that region.
+The loess curve for BN373 is almost flat.
+In CS17, the loess curve goes down until it reaches the minimum value (at about 48\%) and then remains unchanged and in ERS227112, the NPEQ value is low at the beginning and then it increases at around 50\% G-C content.
+In the region where most of the genes are packed, the loess curve is almost flat.
+On the left side of this flat region, there are genes with different G-C content which are enriched in mobile genetic elements.
+So, we expect to see more insertions (smaller NPEQ values) in this region.
+We have low NPEQ values between 40\% and 50\% G-C content which is expected.
+However, in most cases when we have less than 40\% G-C content, NPEQ value is high.
+A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions.
+This has been shown for Tn10 transposon \cite{kimura_nucleoid_2016}, but not for Tn5 transposon, yet.
+Overall, the results are consistent with Langridge et al.
+\cite{langridge_simultaneous_2009}.
+The G-C content of the most of the genes with large NPEQ values is between 50\% and 60\% which is inconsistent with Green et al.
+\cite{green_insertion_2012} as this region contains most of the genes and is not considered as a high G-C content region but rather an average G-C content region.
+%Even though most of the regression lines are consistent with Rubin et al.
+results \cite{rubin_essential_2015}, the figures show that a linear regression cannot predict G-C bias as its slope changes dramatically in different datasets.
+
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c c}
+%\includegraphics[page=1, scale=0.23]{biases.pdf}&
+%\includegraphics[page=5, scale=0.23]{biases.pdf}&
+%\includegraphics[page=9, scale=0.23]{biases.pdf}\\
+%\includegraphics[page=13, scale=0.23]{biases.pdf}&
+%\includegraphics[page=17, scale=0.23]{biases.pdf}&
+%\includegraphics[page=21, scale=0.23]{biases.pdf}\\
+%\includegraphics[page=25, scale=0.23]{biases.pdf}&
+%\includegraphics[page=29, scale=0.23]{biases.pdf}&
+%\includegraphics[page=33, scale=0.23]{biases.pdf}\\
+%\includegraphics[page=37, scale=0.23]{biases.pdf}&
+%\includegraphics[page=41, scale=0.23]{biases.pdf}&
+%\includegraphics[page=45, scale=0.23]{biases.pdf}\\
+%\includegraphics[page=49, scale=0.23]{biases.pdf}&
+%\includegraphics[page=53, scale=0.23]{biases.pdf}&
+%\includegraphics[page=57, scale=0.23]{biases.pdf}\\
+%\end{tabular}
+%\caption{The plots show G-C content of genes against their NPEQ values.
+The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows the linear regression line.}
+%\label{fig:GC-bias}
+%\end{figure*}
+
+\subsubsection{Positional bias within genes}
+Some research has indicated that himar1 transposons are more probable to get inserted into the two ends of a gene compared to the middle \cite{griffin_high-resolution_2011}.
+We have tested this hypothesis using our TraDIS data.
+We divided every gene into 100 fragments with equal lengths (percentiles) and calculated the mean insertion index for each percentile.
+Insertion index is calculated using $\frac{\frac{n_p}{l_p}}{\frac{n_g}{l_g}}$, where $n_p$ is the number of insertion sites in a specific percentile, $l_p$ is the length of that percentile, $n_g$ is the number of insertion sites in the whole genome and $l_g$ is genome length.
+Mean insertion index for each percentile is calculated by averaging over all insertion indices for that specific percentile of genes.
+We saw almost no bias towards any location when considering all genes together (Fig.\@ \ref{fig:insertion-position-bias} {\color{red}SUPPLEMENTARY}).
+We studied the bias in three different groups of genes: essential genes which have no or just a few insertions, non-essential genes which have an intermediate number of insertions, and beneficial losses which have a high number of insertions.
+The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the $5^\prime$ and $3^\prime$ ends (Fig.\@ \ref{fig:insertion-position-bias-ess}) while it is the opposite in beneficial losses (Fig.\@ \ref{fig:insertion-position-bias-ben}).
+High number of insertions at the $3^\prime$ end of essential genes implies that the functional domains are located before the insertions and the insertions are not interfering with them.
+On the other hand, high number of insertions at the $5^\prime$ end of the essential genes indicates there might be alternative start codons in the $5^\prime$ end or it might be because of annotation errors that have predicted the start codon in an incorrect place before the actual start codon.
+
+%\begin{figure*}
+%\begin{tabular}{c c}
+%\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
+%\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
+%\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
+%\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
+%\end{tabular}
+%\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
+The genes are divided into 3 segments: 5\% of
+%the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
+%\label{fig:insertion-position-bias}
+%\end{figure*}
+\subsection{Most essential genes are ubiquitously essential in Enterobacteriaceae}
+Previous studies of gene essentiality in Enterobacteriaceae family have compared essential genes in different genomes in this family and studied the sets of core essential genes and accessory essential genes \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}.
+Core essential genes are responsible for essential processes such as cell division, DNA replication, transcription and translation and some important pathways like peptidoglycan and fatty acid biosynthesis \cite{barquist_comparison_2013}.
+Accessory essential genes differ in genomes due to different reasons such as niche adaptation, gene duplication and the existence of alternative pathways \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}.
+Another group of accessory essential genes are phage repressors \cite{barquist_comparison_2013}.
+Even though these genes are not essential for the growth of a cell, once phages are introduced to a cell, they become essential as long as the phage remains in the cell.
+
+In this study, we have compared the genes in 16 bacteria from Enterobacteriaceae family.
+For this we needed to study core and accessory sets of genes.
+Moreover, in the presence of two redundant variations of one gene, if we knock out one copy using TraDIS, the other copy compensates and the organism can still survive \cite{dean_pervasive_2008}.
+This leads to a different essentiality inference using TraDIS.
+Therefore, in addition to core and accessory genes, we have studied duplicate genes.
+To study whether each gene in the 16 organisms is core we used Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to iteratively search for homologous proteins in our dataset and cluster them.
+We divided the clusters of homologous genes into three groups based on their conservation.
+Genus specific class contain genes that are present only in one genus, the genes in single copy class are present in more than one genus and more than 70\% of them are not duplicated (core genes), and  the genes in multi-copy class are present in more than one genus and more than 30\% of them are duplicated (duplicate genes).
+
+We also evaluated the essentiality of genes and divided the genes into different levels of essentiality using NPEQ values.
+If NPEQ is less than -1.644854, it means that the gene has tolerated many insertions, so it is beneficial for the organism to lose this gene in the rich medium that we used.
+However, if NPEQ is greater that 1.644854, the gene can tolerate very few or no insertions indicating that the gene is essential for cell viability in our test medium.
+Any other NPEQ value shows an intermediate number of insertions in genes meaning that it is not beneficial for the organism to lose these genes, but they are not essential, too.
+We have called this group of genes non-essential.
+
+The results for comparing three levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
+The high number of single copy clusters in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality.
+However, there are also many essential genus specific genes.
+The figure also shows that beneficial losses are over represented in genus specific class.
+Therefore, beneficial losses are mostly recent genes that the organism tends to lose in the long run.
+Besides, most of the multi-copy clusters are non-essential and there are only a few multi-copy clusters that are essential.
+This can be explained by the redundancy that duplicate genes can keep even after $\sim100$ million years \cite{dean_pervasive_2008}.
+
+\begin{figure}
+\captionsetup[subfigure]{justification=centering}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{cluster-essentiality.pdf}
+  \caption{}
+  \label{fig:iidist}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{essential-pathways.pdf}
+  \caption{}
+  \label{fig:essentiality-pathway-ess}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{non-essential-pathways.pdf}
+  \caption{}
+  \label{fig:essentiality-pathway-nes}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{beneficialloss-pval.pdf}
+  \caption{}
+  \label{fig:essentiality-pathway-ben}
+\end{subfigure}
+\caption{(a) The genes have been clustered into homologous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
+Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
+The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.
+(b) and (c) KEGG Pathways enriched in essential genes (b) and non-essential genes (c).
+(d) The words enriched in the description of beneficial losses in their embl files compared to other genes.
+The red line shows P-value = 0.05.
+The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
+\label{fig:gene-classes}
+\end{figure}
+
+%\begin{figure}
+%\centering
+%\includegraphics[scale=0.5]{cluster-essentiality.pdf}
+%\caption{The genes have been clustered into homologous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
+Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
+The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
+%\label{fig:iidist}
+%\end{figure}
+
+To study which functions are enriched in each class of essentiality, we used KEGG pathway enrichment analysis \cite{kanehisa_kegg:_2000} and compared the genes in each class with the databases that were available for the genomes in this study.
+The results show that essential genes are enriched in pathways related to genetic information processing such as replication (DNA replication, homologous recombination and mismatch repair), transcription (RNA polymerase), translation (ribosome), and protein export.
+These are the essential functions that every cell needs for its viability.
+Other enriched pathways are mostly metabolism related.
+These include fatty acid biosynthesis that produce cell membrane, peptidoglycan and lipopolysaccharide biosynthesis that are essential components of cell wall, terpenoid backbone biosynthesis which feeds peptidoglycan biosynthesis, nucleotide and amino acid metabolism, and the metabolism of important cofactors and vitamins like riboflavin, biotin,porphyrin and chlorophyll (Figure~\ref{fig:essentiality-pathway-ess}).
+{\color{red} CAN WE HAVE SOMETHING LIKE THIS? http://www.nature.com/articles/srep00125}
+
+Non-essential genes are mostly involved in carbohydrate, energy, and lipid metabolism, membrane transport, cell motility and cellular community.
+Even though these functions are important, they are not essential in a rich medium in lab.
+
+As most of beneficial losses do not have homologs in other genomes, most of them are not studied and therefore there is no KEGG pathway for them.
+Because of this reason, we studied the description of these genes in their embl files and found the words that were enriched.
+These are shown in Figure~\ref{fig:essentiality-pathway-ben}.
+The results show that most of beneficial losses are mobile genetic elements like transposases and insertion elements which are not essential in their host genomes, genes that are not essential in a rich medium like stress protein and acid-resistance protein, and hypothetical proteins.
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c}
+%\includegraphics[scale=0.4]{essential-pathways.pdf}&
+%\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
+%\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
+%\includegraphics[scale=0.4]{beneficialloss-pval.pdf}
+%\end{tabular}
+%\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
+The red line shows P-value = 0.05.
+The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
+%\label{fig:essentiality-pathway}
+%\end{figure*}
+
+\subsection{phylogenetic analysis of gene essentiality identifies ...}
+%\section{Discussion}
+\section{Materials and Methods}
+%\subsection{Transposon insertion}
+%\section{Results and discussion}
+%Throughout time, species can gain or lose genes.
 We investigated if these gene gain and losses are related to the essentiality of the genes.
 In this section, we have first described the biases that can affect our study, and then evaluated the essentiality of genes and their conservation and the relationship between these two.
+%
+%
+%%\subsection{Length bias}
+%%Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions.
+As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes.
+%%In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index in short genes.
+It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
+%%\begin{figure*}
+%%\centering
+%%\includegraphics[page=42, scale=0.3]{biases.pdf}
+%%\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
+%%\label{fig:length-bias}
+%%\end{figure*}
+%
+%
+%\subsubsection{Preferred insertion motif bias}
 
-\subsection{Are there biases in transposon mutagenesis data?}
-To evaluate the essentiality of a gene, the number of insertions within that gene was measured as explained in~\ref{sec:essentiality}.
-However, if the transposons are biased to specific regions in the genome, it results in false predictions and influences the accuracy of our analysis.
-Different articles have reported biases in transposon mutagenesis \cite{barquist_comparison_2013, rubin_essential_2015, kimura_nucleoid_2016}.
-We performed a detailed study of these biases.
-The biases that we have studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
-
-%\subsection{Length bias}
-%Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions.
-As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes.
-%In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index in short genes.
-It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
+%
+%\begin{itemize}
+%\item model H-NS binding sites? CGWTWHWww Lang et al (2007)
+%\item seems unlikely -- show bulk of genes are around 50\% G+C (add box-whisker plots to scatter diagrams?)
+%\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? It is not mentioned in the paper.
+%\end{itemize}
+%
 %\begin{figure*}
-%\centering
-%\includegraphics[page=42, scale=0.3]{biases.pdf}
-%\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
-%\label{fig:length-bias}
-%\end{figure*}
-
-\subsubsection{Origin of replication bias}
-One possible source of bias is the distance from origin of replication.
-When the bacteria are under replication during the transposn insertion process, there are more copies of the genes close to the origin of replication than the genes further away.
-This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
-The other factor that can affect the results is if essential genes are clustered near the origin by nature.
-Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication.
-However, the essential genes are more frequent in the leading strand than the lagging one.
-
-To study the bias towards the position of the genes, we plotted the insertion index for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Fig.\@ \ref{fig:distance-bias}.
-The figure indicates that the insertion indices decrease when the genes are located further from the origin of replication.
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c c}
-\includegraphics[page=2, scale=0.25]{biases.pdf}&
-\includegraphics[page=5, scale=0.25]{biases.pdf}&
-\includegraphics[page=8, scale=0.25]{biases.pdf}\\
-\includegraphics[page=11, scale=0.25]{biases.pdf}&
-\includegraphics[page=14, scale=0.25]{biases.pdf}&
-\includegraphics[page=17, scale=0.25]{biases.pdf}\\
-\includegraphics[page=20, scale=0.25]{biases.pdf}&
-\includegraphics[page=23, scale=0.25]{biases.pdf}&
-\includegraphics[page=26, scale=0.25]{biases.pdf}\\
-\includegraphics[page=29, scale=0.25]{biases.pdf}&
-\includegraphics[page=32, scale=0.25]{biases.pdf}&
-\includegraphics[page=35, scale=0.25]{biases.pdf}\\
-&\includegraphics[page=38, scale=0.25]{biases.pdf}&\\
-\end{tabular}
-\caption{The plots show the distance of the genes from DnaA gene normalised by the lengths of the genomes versus the insertion indices of the genes.
-The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
-The red curves show the loess curve when the smoothness parameter is $0.2$.}
-\label{fig:distance-bias}
-\end{figure*}
-
-\subsubsection{Preferred insertion motif bias}
-Another concern while inferring essentiality from transposon mutagenesis data is that transposons are biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposons are inclined to, instead of their essentiality level.
-For this, we used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
-The results in Fig.\@  \ref{fig:logos} show a slight bias towards certain combinations of bases.
-
-In addition, we investigated if the G-C content of genes can change the number of insertions by plotting the number of G-C bases in a gene normalised by the length of the gene versus insertion index (Fig.\@  \ref{fig:GC-bias}).
-As the figure shows, when G-C content is less than $40\%$, the insertion index is low, however when it is higher than $50\%$, the insertion index is almost constant.
-A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions \cite{kimura_nucleoid_2016}.
-The other reason is that the genes with low G-C content are enriched in mobile genetic elements compared to the genes with average G-C content (Fig.\@ \ref{fig:gc-pval}) and this has caused seeing a different pattern of essentiality in that region.
-
-\begin{itemize}
-\item model H-NS binding sites? CGWTWHWww Lang et al (2007)
-\item seems unlikely -- show bulk of genes are around 50\% G+C (add box-whisker plots to scatter diagrams?)
-\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? It is not mentioned in the paper.
-\end{itemize}
-
-\begin{figure*}
-\begin{tabular}{c c}
-\includegraphics[scale=0.55]{100logo-bits.pdf}&
-\includegraphics[scale=0.55]{100logo-prob.pdf}
-\end{tabular}
-\caption{Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
-On the left the height of each character corresponds to a bit score for that character (i.e.
-$2-\sum f_a\times\log_2f_a-\frac{1}{\ln2}\times\frac{3}{2\times n}$, where $f_a$ is the relative frequency of base $a$ and $n$ is the number of sequences).
-To put it in simple words, the height of the set of characters shows how biased that position is and the height of each character shows the amount of bias towards that character.
-On the right the height of each character shows the relative frequency of that character.}
-\label{fig:logos}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c c}
-\includegraphics[page=1, scale=0.25]{biases.pdf}&
-\includegraphics[page=4, scale=0.25]{biases.pdf}&
-\includegraphics[page=7, scale=0.25]{biases.pdf}\\
-\includegraphics[page=10, scale=0.25]{biases.pdf}&
-\includegraphics[page=13, scale=0.25]{biases.pdf}&
-\includegraphics[page=16, scale=0.25]{biases.pdf}\\
-\includegraphics[page=19, scale=0.25]{biases.pdf}&
-\includegraphics[page=22, scale=0.25]{biases.pdf}&
-\includegraphics[page=25, scale=0.25]{biases.pdf}\\
-\includegraphics[page=28, scale=0.25]{biases.pdf}&
-\includegraphics[page=31, scale=0.25]{biases.pdf}&
-\includegraphics[page=34, scale=0.25]{biases.pdf}\\
-\includegraphics[page=37, scale=0.25]{biases.pdf}&&
-\includegraphics[page=40, scale=0.25]{biases.pdf}\\
-\end{tabular}
-\caption{The plots show the ratio of G-C bases in the genes normalised by the lengths of the genes against their insertion indices.
-The red curves show the loess curve when the smoothness parameter is $0.2$.}
-\label{fig:GC-bias}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[scale=0.55]{lowgc-pval.pdf}
-\caption{Word enrichment analysis for low G-C genes compared to genes with interquartile G-C level.
+%\includegraphics[scale=0.55]{lowgc-pval.pdf}
+%\caption{Word enrichment analysis for low G-C genes compared to genes with interquartile G-C level.
 The red line shows P-value = 0.05.
 The P-values have been calculated using Fisher's exact test and corrected using Benjamini-Hochberg-Yekutieli.}
-\label{fig:gc-pval}
-\end{figure*}
+%\label{fig:gc-pval}
+%\end{figure*}
+%
 
-\subsubsection{Positional bias within genes}
-The other question that we tried to answer was whether insertions are tolerated in some regions in a gene.
-For example, can essential genes tolerate insertions at their 3' end without losing their funtionality?
-To address this question, we divided every gene into percentiles and calculated the mean insertion index for each percentile.
-Fig.\@ \ref{fig:insertion-position-bias} shows almost no bias towards any location when considering all genes together.
-We also studied the bias in three of the groups defined in Section~\ref{sec:essentiality}: essential genes, non-essential genes, and beneficial losses.
-The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the 5' and 3' ends while it is the opposite in beneficial losses.
-The case for the non-essential genes is similar to all genes.
-High number of insertions at the 3' end of essential genes implies that the functional part of the genes are located before the insertions.
-On the other hand, high number of insertions at the 5' end of the essential genes indicates there might be alternative start codons in the 5' end or it might be because of alignment errors.
-\textcolor{red}{\{To be tested\}}
-
-\begin{figure*}
-\begin{tabular}{c c}
-\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
-\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
-\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
-\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
-\end{tabular}
-\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
-The genes are divided into 3 segments: 5\% of
-the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
-\label{fig:insertion-position-bias}
-\end{figure*}
-
-%\begin{itemize}
-%\item There is a bias towards the position of the gene (Fig.~\ref{fig:distance-bias}).
-%\item There is a negligible bias towards certain motifs (Fig.~\ref{fig:logos})
-%\item The G-C bias needs to be studied further (Fig.~\ref{fig:GC-bias})
-%\item The number of insertions on the 3' and 5' ends is more than the internal region in essential genes and less than the internal region in beneficial losses.
-(Fig.~\ref{fig:insertion-position-bias})
-%\end{itemize}
-\subsection{Essentiality and conservation}
-Essential genes are needed for the growth of organisms.
+%\subsection{Essentiality and conservation}
+%Essential genes are needed for the growth of organisms.
 Because of that, one might think that essential genes should not be lost in a short period of time throughout evolution, unless they are no longer needed in new organisms or they are replaced by new pathways.
@@
 We have tested this idea by comparing the essentiality and conservation of genes in Enterobacteriaceae family.
-
-\subsubsection{Gene classes}
-%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
+%
+%
+%\subsubsection{Gene classes}
+%%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
 Plenty of methods are proposed for this purpose.
@@
 For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
-In order to study the relationship between essentiality and conservation, we needed to evaluate the essentiality and conservation of genes.
+%In order to study the relationship between essentiality and conservation, we needed to evaluate the essentiality and conservation of genes.
 For this, we divided the genes into different levels of essentiality (essential genes, ambiguous, non-essential genes, beneficial losses).
@@
 Moreover, we grouped the genes into different classes of conservation (genus specific, single copy, multi-copy) using the method explaind in Section~\ref{sec:conservationclasses}.
-
-The results for comparing four levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
+%
+%The results for comparing four levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
 The high number of single copy genes in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality.
@@
 In the presence of two redundant variations of one gene, if we knock out one copy by transposon mutagenesis, the other copy compensates and the organism can still survive.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.5]{cluster-essentiality.pdf}
-\caption{The genes have been clustered into orthologous groups using Hieranoid and paralogous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
+%
+%\begin{figure}
+%\centering
+%\includegraphics[scale=0.5]{cluster-essentiality.pdf}
+%\caption{The genes have been clustered into orthologous groups using Hieranoid and paralogous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
 Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
 The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
-\label{fig:iidist}
-\end{figure}
-
-To study which functions are enriched in each class of essentiality, we used the word enrichment method explained in Section~\ref{sec:enrichment}.
+%\label{fig:iidist}
+%\end{figure}
+%
+%To study which functions are enriched in each class of essentiality, we used the word enrichment method explained in Section~\ref{sec:enrichment}.
 Fig.\@ \ref{fig:essentiality-pval} shows the top 20 enriched words for each essentiality class.
@@
 Beneficial losses also contain many fimbrial proteins which probably has occurred because these proteins are not needed in a rich lab medium \textcolor{red}{\{TRUE?\}}.
-%We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}.
+%%We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}.
 It shows that most of the beneficial losses are mobile elements.
 Most of these mobile elements are not included in KEGG database.
-
-We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}.
+%
+%We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}.
 The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes.
-However, as mobile genetic elements are not stored in KEGG database, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c}
-\includegraphics[scale=0.4]{essential-pval.pdf}&
-\includegraphics[scale=0.4]{non-essential-pval.pdf}\\
-\includegraphics[scale=0.4]{beneficialloss-pval.pdf}&
-\end{tabular}
-\caption{Word enrichment analysis for essential genes, non-essential genes, and beneficial losses compared to other genes.
+However, as mobile genetic elements are not stored in KEGG database \cite{kanehisa_kegg:_2000}, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.
+%
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c}
+%\includegraphics[scale=0.4]{essential-pval.pdf}&
+%\includegraphics[scale=0.4]{non-essential-pval.pdf}\\
+%\includegraphics[scale=0.4]{beneficialloss-pval.pdf}&
+%\end{tabular}
+%\caption{Word enrichment analysis for essential genes, non-essential genes, and beneficial losses compared to other genes.
 The red line shows P-value = 0.05.
 The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-\label{fig:essentiality-pval}
-\end{figure*}
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c}
-\includegraphics[scale=0.4]{essential-pathways.pdf}&
-\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
-\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
-\end{tabular}
-\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
+%\label{fig:essentiality-pval}
+%\end{figure*}
+%
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c}
+%\includegraphics[scale=0.4]{essential-pathways.pdf}&
+%\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
+%\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
+%\end{tabular}
+%\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
 The red line shows P-value = 0.05.
 The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-\label{fig:essentiality-pathway}
-\end{figure*}
-
-\subsubsection{The evolution of essentiality}
-%In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
+%\label{fig:essentiality-pathway}
+%\end{figure*}
+%
+%\subsubsection{The evolution of essentiality}
+%%In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
 If a gene is essential in a genome, its corresponding cell in the matrix has value 1 and 0 otherwise.
@@
 The resulting tree is not similar to the species tree.
-%
-%\begin{figure}
-%\centering
-%\includegraphics[scale=0.2]{neighbor-joining-essentiality-tree.pdf}
-%\caption{The tree is generated from all the genes that are copied only once per genome.
+%%
+%%\begin{figure}
+%%\centering
+%%\includegraphics[scale=0.2]{neighbor-joining-essentiality-tree.pdf}
+%%\caption{The tree is generated from all the genes that are copied only once per genome.
 We have made a binary matrix from the essentiality of these genes.
 Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree using PHYLIP Neighbor which uses a neighbour joining method.}
-%\label{fig:neighborjoining}
-%\end{figure}
-
-In this section, we compared the number of genes that were essential and conserved at each level of the phylogenetic tree.
+%%\label{fig:neighborjoining}
+%%\end{figure}
+%
+%In this section, we compared the number of genes that were essential and conserved at each level of the phylogenetic tree.
 We were interested to see if the trend of the essentiality variation between different organisms is in agreement with the phylogenetic tree.
 In other words, we have tested if organisms close together have more essential genes in common than organisms that have separated earlier in the phylogenetic tree.
-
-We needed to cluster sets of orthologous genes in the bacteria that we were studying.
+%
+%We needed to cluster sets of orthologous genes in the bacteria that we were studying.
 Homologous clusters introduced in ~\ref{sec:conservationclasses} were not useful for this purpose as sets of paralogous genes with different essentiality levels can make essentiality inference ambiguous.
@@
 We used Hieranoid for clustering orthologous genes.
-
-In order to test if the essentiality of genes follows a tree-like trend, we compared the number of genes that were conserved in different bacteria in our study and the number of genes that were essential in these bacteria.
+Hieranoid need a species tree for clustering genes.
+We used the method explained in Section~\ref{sec:phylogenetic} to generate this tree.
+%
+%In order to test if the essentiality of genes follows a tree-like trend, we compared the number of genes that were conserved in different bacteria in our study and the number of genes that were essential in these bacteria.
 The genes that are conserved in all bacteria being studied are called core genes and the genes that are conserved and essential are called core essential genes.
@@
 Furthermore, some of the predicted essential genes might be artefacts of transposon mutagenesis method.
-
-\begin{figure*}
-\centering
-\includegraphics[scale=0.6]{upsetr1-edited.pdf}\\
-\includegraphics[scale=0.6]{upsetr2-edited.pdf}
-\caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.
+%
+%\begin{figure*}
+%\centering
+%\includegraphics[scale=0.6, page=2]{upsetr-maximiseMCC.pdf}\\
+%\includegraphics[scale=0.6, page=3]{upsetr-maximiseMCC.pdf}
+%\caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.
 The bars show the number of genes that are core between the strains marked with black circles.
 The tick marks show phylogenetically informative columns and the cross marks show non-informative columns.}
-\label{fig:upsetr}
-\end{figure*}
-
-To further study the trend of essentiality changes, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
+%\label{fig:upsetr}
+%\end{figure*}
+%
+%To further study the trend of essentiality changes, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
 We used three different methods for inferring core genes and core essential genes.
@@
 As this figure shows, the ratio between core essential and core genes is almost constant using the intersection and dollo method; however, this ratio increases as we go higher in the tree using the ancestral insertion index method.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap-annotated.pdf}
-\caption{The tree shows the species tree in Fig.\@\ref{fig:species-tree} annotated by the number of core essential genes and core genes at each node.
+%
+%%\begin{figure}
+%%\centering
+%%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap-annotated.pdf}
+%%\caption{The tree shows the species tree in Fig.\@\ref{fig:species-tree} annotated by the number of core essential genes and core genes at each node.
 We used three methods to define core essential genes and core genes.
@@
 At the internal nodes, red shows the numbers using the intersection method, green shows the ancestral insertion index method, and turquoise shows fdollop method.}
-\label{fig:annotated-speciestree}
-\end{figure}
-
-As these three methods lead to conflicting results, we compared the differences between the genes found in these three methods.
+%%\label{fig:annotated-speciestree}
+%%\end{figure}
+%
+%As these three methods lead to conflicting results, we compared the differences between the genes found in these three methods.
 For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods.
@@
 The results are depicted in Fig.\@\ref{fig:method-difference}.
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c}
-\includegraphics[scale=0.4]{ancestral-intersection-pval.pdf}&
-\includegraphics[scale=0.4]{dollo-ancestral-pval.pdf}\\
-\includegraphics[scale=0.4]{dollo-intersection-pval.pdf}&
-\end{tabular}
-\caption{The figure shows the difference between core essential genes in intersection and ancestral insertion index methods, ancestral insertion index and fdollop methods, and intersection and fdollop methods.
+%
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c}
+%\includegraphics[scale=0.4]{ancestral-intersection-pval.pdf}&
+%\includegraphics[scale=0.4]{dollo-ancestral-pval.pdf}\\
+%\includegraphics[scale=0.4]{dollo-intersection-pval.pdf}&
+%\end{tabular}
+%\caption{The figure shows the difference between core essential genes in intersection and ancestral insertion index methods, ancestral insertion index and fdollop methods, and intersection and fdollop methods.
 The red line shows P-value = 0.05.
@@
 The bottom left figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using intersection method.}
-\label{fig:method-difference}
-\end{figure*}
-
-\section{Materials and Methods}
-\subsection{Transposon mutagenesis}\label{sec:transposoninsertion}
-We studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains (Fig.\@ \ref{fig:species-tree}) and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
+%\label{fig:method-difference}
+%\end{figure*}
+%
+%\section{Materials and Methods}
+%\subsection{Transposon mutagenesis}\label{sec:transposoninsertion}
+%We studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains (Fig.\@ \ref{fig:species-tree}) and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
 These strains are all selected from Enterobacteriaceae family and a transposon mutagenesis study have been performed on them.
@@
 The number of insertions in a gene implies the degree of essentiality for that gene.
-
-\subsection{Essentiality levels}\label{sec:essentiality}
-The more transposon insertions we observe in a gene after sequencing the genomes, the less essential the gene is.
+%
+%\subsection{Generating the species tree}\label{sec:phylogenetic}
+%We first ran search and align commands from PhyloSift package \cite{darling_phylosift:_2014} to select gene markers for generating a phylogenetic tree and aligning them.
+Then, we concatenated the protein alignments for all 14 genomes and ran RaxML \cite{stamatakis_raxml_2014} with "PROTGAMMALG4M" amino acid substitution model, "a" algorithm and 100 alternative runs on distinct starting trees.
+%
+%\subsection{Essentiality levels}\label{sec:essentiality}
+%The more transposon insertions we observe in a gene after sequencing the genomes, the less essential the gene is.
 In order to quantify the essentiality of genes, we used a measure named insertion index which is proportional to the number of insertions in a gene.
-
-To calculate the insertion index for each gene, we summed up the number of transposon insertion sites observed in that gene.
+%
+%To calculate the insertion index for each gene, we summed up the number of transposon insertion sites observed in that gene.
 Since the lengths of the genes are different, the insertion indices were then normalised by dividing them by gene length.
@@
 Therefore, in order to make the insertion indices comparable in all the strains, we normalised the insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
-
-Based on insertion indices, the genes were divided into four groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
+%
+%Based on insertion indices, the genes were divided into four groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
 We utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
@@
 Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c c}
-\includegraphics[page=1, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=2, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=3, scale=0.25]{essentiality.pdf}\\
-\includegraphics[page=4, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=5, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=6, scale=0.25]{essentiality.pdf}\\
-\includegraphics[page=7, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=8, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=9, scale=0.25]{essentiality.pdf}\\
-\includegraphics[page=10, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=11, scale=0.25]{essentiality.pdf}&
-\includegraphics[page=12, scale=0.25]{essentiality.pdf}\\
-&\includegraphics[page=13, scale=0.25]{essentiality.pdf}&\\
-\end{tabular}
-\caption{Plots show the insertion index distribution for each genome.
+%
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c c}
+%\includegraphics[page=1, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=2, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=3, scale=0.25]{essentiality.pdf}\\
+%\includegraphics[page=4, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=5, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=6, scale=0.25]{essentiality.pdf}\\
+%\includegraphics[page=7, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=8, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=9, scale=0.25]{essentiality.pdf}\\
+%\includegraphics[page=10, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=11, scale=0.25]{essentiality.pdf}&
+%\includegraphics[page=12, scale=0.25]{essentiality.pdf}\\
+%&\includegraphics[page=13, scale=0.25]{essentiality.pdf}&\\
+%\end{tabular}
+%\caption{Plots show the insertion index distribution for each genome.
 The plots are divided into 4 regions using red lines.
 These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-\label{fig:iidist-species}
-\end{figure*}
-
-\subsection{Bias correction}\label{sec:biascorrection}
-We observed a distance from origin of replication bias and also a positional bias within genes.
+%\label{fig:iidist-species}
+%\end{figure*}
+%
+%\subsection{Bias correction}\label{sec:biascorrection}
+%We observed a distance from origin of replication bias and also a positional bias within genes.
 Hence, these biases needed to be corrected before inferring the essentiality of genes from their insertion indices.
 We did not include genes shorter than 100 base-pairs in our study as they might not be targeted by any transposon due to their shortness.
-
-To overcome the distance from origin of replication bias, we divided the value of insertion index for a gene in a specific position by the predicted value by loess for that position.
+%
+%To overcome the distance from origin of replication bias, we divided the value of insertion index for a gene in a specific position by the predicted value by loess for that position.
 This value was then multiplied by the average insertion index.
@@
 The insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
-
-\begin{figure*}
-\centering
-\begin{tabular}{c c c}
-\includegraphics[page=1, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=2, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=3, scale=0.25]{essentiality-normalised.pdf}\\
-\includegraphics[page=4, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=5, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=6, scale=0.25]{essentiality-normalised.pdf}\\
-\includegraphics[page=7, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=8, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=9, scale=0.25]{essentiality-normalised.pdf}\\
-\includegraphics[page=10, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=11, scale=0.25]{essentiality-normalised.pdf}&
-\includegraphics[page=12, scale=0.25]{essentiality-normalised.pdf}\\
-&\includegraphics[page=13, scale=0.25]{essentiality-normalised.pdf}&\\
-\end{tabular}
-\caption{Plots show the insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes.
+%
+%\begin{figure*}
+%\centering
+%\begin{tabular}{c c c}
+%\includegraphics[page=1, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=2, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=3, scale=0.25]{essentiality-normalised.pdf}\\
+%\includegraphics[page=4, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=5, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=6, scale=0.25]{essentiality-normalised.pdf}\\
+%\includegraphics[page=7, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=8, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=9, scale=0.25]{essentiality-normalised.pdf}\\
+%\includegraphics[page=10, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=11, scale=0.25]{essentiality-normalised.pdf}&
+%\includegraphics[page=12, scale=0.25]{essentiality-normalised.pdf}\\
+%&\includegraphics[page=13, scale=0.25]{essentiality-normalised.pdf}&\\
+%\end{tabular}
+%\caption{Plots show the insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes.
 The plots are divided into 4 regions using red lines.
 These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-\label{fig:iidist-species-normalised}
-\end{figure*}
-
-\subsection{Conservation classes}\label{sec:conservationclasses}
-To study whether each gene in the 14 organisms is conserved we proposed a program that clusters homologous proteins.
+%\label{fig:iidist-species-normalised}
+%\end{figure*}
+%
+%\subsection{Conservation classes}\label{sec:conservationclasses}
+%To study whether each gene in the 14 organisms is conserved we proposed a program that clusters homologous proteins.
 This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
@@
 The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 14 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
-
-\begin{figure}
-\includegraphics[scale=0.2]{homclust.pdf}
-\caption{The steps of our proposed algorithm for clustering homologous genes.}
-\label{fig:homclust}
-\end{figure}
-
-We divided the clusters of homologous genes into three groups based on their conservation.
+%
+%\begin{figure}
+%\includegraphics[scale=0.2]{homclust.pdf}
+%\caption{The steps of our proposed algorithm for clustering homologous genes.}
+%\label{fig:homclust}
+%\end{figure}
+%
+%We divided the clusters of homologous genes into three groups based on their conservation.
 Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
 These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
-
-\begin{figure}
-\includegraphics[scale=0.3]{cluster-size-dist.pdf}
-\caption{Size distribution for all clusters of homologous genes.
+%
+%\begin{figure}
+%\includegraphics[scale=0.3]{cluster-size-dist.pdf}
+%\caption{Size distribution for all clusters of homologous genes.
 Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
-\label{fig:homclust-results}
-\end{figure}
-
-\subsection{Word and pathway enrichment analysis}\label{sec:enrichment}
-To study the enrichment of words in a group of genes compared to a second group, We gathered the descriptions of genes from their embl files.
+%\label{fig:homclust-results}
+%\end{figure}
+%
+%\subsection{Word and pathway enrichment analysis}\label{sec:enrichment}
+%To study the enrichment of words in a group of genes compared to a second group, We gathered the descriptions of genes from their embl files.
 Then, we counted the repeat number of each word for the genes in each group and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
 The P-values were then corrected using Benjamini-Hochberg-Yekutieli procedure.
-
-For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database.
+%
+%For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database \cite{kanehisa_kegg:_2000}.
 This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.
@@
 Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.
-
-\subsection{Defining core and core essential genes}\label{sec:core}
-We used three different methods for defining core and core essential genes: intersection, ancestral insertion index, and Dollo law. These three methods are explained in what follows.
-
-The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
-
-The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
+%
+%\subsection{Defining core and core essential genes}\label{sec:core}
+%We used three different methods for defining core and core essential genes: intersection, ancestral insertion index, and Dollo law. These three methods are explained in what follows.
+%
+%The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
+%
+%The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
 In this method, we averaged over the insertion indices of the pair of closest children of the ancestral node.
@@
 Then, we plotted the insertion indices and fitted an exponential and a gamma distribution to the plot as described in Section~\ref{sec:essentiality} and found the essential genes at that level.
-
-The third method is using Dollo law to define core genes and core essential genes.
+%
+%The third method is using Dollo law to define core genes and core essential genes.
 This method, assumes that the gain of genes (gain of essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (gain of essentiality) and minimise the number of times that a gene (the essentiality of a gene) has been lost.
 Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
-
-\section{Conclusion}
-In this paper, we studied the relationship between the essentiality and conservation of genes in 14 bacteria from Enterobacteriaceae family.
+%
+%\section{Conclusion}
+%In this paper, we studied the relationship between the essentiality and conservation of genes in 14 bacteria from Enterobacteriaceae family.
 We first studied the biases that can affect our results and showed that transposon insertions are more abundant near the origin of replication.
@@
 Moreover we showed that transposon insertions are more abundant near the ends of essential genes compared to the internal region, while it is the opposite in beneficial losses.
-
-After correcting biases, we studied the essentiality versus conservation by dividing the genes into three classes of conservation: genus specific, single copy, and multi-copy, dividing them into four levels of essentiality: essential, ambiguous, non-essential, and benefical losses.
+%
+%After correcting biases, we studied the essentiality versus conservation by dividing the genes into three classes of conservation: genus specific, single copy, and multi-copy, dividing them into four levels of essentiality: essential, ambiguous, non-essential, and benefical losses.
 We found that essential genes are mostly single copy, however there is a considerable number of genus specific genes in essential level which help to distinguish between genera.
@@
 The other finding was that multi-copy genes are mostly non-essential which can happen due to redundancy.
-
-We also found that the pattern of essentiality changes is not in agreement with the phylogenetic tree due to the small number of essential genes.
+%
+%We also found that the pattern of essentiality changes is not in agreement with the phylogenetic tree due to the small number of essential genes.
 Also the ratio between core essential genes and core genes can be either constant or increasing depending on the method that we use for defining core and core essential genes.
 \textcolor{red}{\{WHAT?\}}
-
-Overall, we have found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.
-
-\section*{Acknowledgments}
 %
+%Overall, we have found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.
 %
+%\section*{Acknowledgments}
+%%
+%%
 %\nolinenumbers

### 2017-02-13 bin/embl2peptides.py

@@ -105,7 +105,7 @@ with open(source, 'r') as from_file:
                     if not t:
                         t = ID[:3] + 'LocTag_' + str(locus_tag_counter)
                         locus_tag_counter += 1
-                    to_file.write('>{0} [{1}/{2} ({3})] {4} {5} {6} [EC_number={7}]'.format(t, ID, se, comment, p, nt, g, ec))
+                    to_file.write('>{0} [{1}/{2} ({3})] [{4}] [{5}] [{6}] [EC_number={7}]'.format(t, ID, se, comment, p, nt, g, ec))
                     to_file.write(tr)
                     to_file.write('\n')
                 s, e = [], []
@@ -126,7 +126,7 @@ with open(source, 'r') as from_file:
                     if not t:
                         t = ID[:3] + 'LocTag_' + str(locus_tag_counter)
                         locus_tag_counter += 1
-                    to_file.write('>{0} [{1}/{2} ({3})] {4} {5} {6} [EC_number={7}]'.format(t, ID, se, comment, p, nt, g, ec))
+                    to_file.write('>{0} [{1}/{2} ({3})] [{4}] [{5}] [{6}] [EC_number={7}]'.format(t, ID, se, comment, p, nt, g, ec))
                     to_file.write(tr)
                     to_file.write('\n')
                 s, e = [], []
@@ -147,7 +147,7 @@ with open(source, 'r') as from_file:
                     if not t:
                         t = ID[:3] + 'LocTag_' + str(locus_tag_counter)
                         locus_tag_counter += 1
-                    to_file.write('>{0} [{1}/{2} ({3})] {4} {5} {6} [EC_number={7}]'.format(t, ID, se, comment, p, nt, g, ec))
+                    to_file.write('>{0} [{1}/{2} ({3})] [{4}] [{5}] [{6}] [EC_number={7}]'.format(t, ID, se, comment, p, nt, g, ec))
                     to_file.write(tr)
                     to_file.write('\n')
                 find_result = findall('>?<?\d+\.\.>?<?\d+', line)
@@ -174,7 +174,7 @@ with open(source, 'r') as from_file:
                     if not t:
                         t = ID[:3] + 'LocTag_' + str(locus_tag_counter)
                         locus_tag_counter += 1
[…]

### 2017-02-13 bin/find-k12-counterparts.py

@@ -2,7 +2,7 @@ from os import listdir
 
 k12 = '../results/ecogene-k12.txt'
 clusters = '../results/hieranoid/clusters.txt'
-genes = '../results/insertion-indices/normalised-insertion-indices-with-logodds/'
+genes = '../results/insertion-indices/gamma/'
 output = '../results/ecogenecounterparts/'
 list_of_files = listdir(genes)
 genenames = []

### 2017-02-13 bin/interesting_genes.py

@@ -1,0 +1,81 @@
+from os import listdir
+from Bio import SeqIO
+from re import match
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+clustersfile = '../results/hieranoid/clusters.txt'
+seqdbfile = '../data/fasta-protein/chromosome/seqdb.fasta'
+essentialitydir = '../results/biases/normalised-pca'
+k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
+intereseting_genes = '../results/interesting_genes.csv'
+
+seqdb = read_fasta_sequences(seqdbfile)
+essentiality = {i: 0 for i in seqdb.keys()}
+ii = {i: 0 for i in seqdb.keys()}
+
+list_of_files = listdir(essentialitydir)
+for filename in list_of_files:
+    with open (essentialitydir + '/' + filename, 'r') as fromfile:
+        for line in fromfile:
+            cells = line.split()
+            ii[cells[0]] = cells[1]
+            if cells[2] == 'essential':
+                essentiality[cells[0]] = 1
+
+with open(k12path, 'r') as fromfile:
+    for line in fromfile:
+        cells = line.split()
+        essentiality[cells[0]] = 1
+        ii[cells[0]] = 7
+
+species_names = {"salmonella":["SEN", "SL1344", "STM", "STMMW", "t", "SL3261"],
+    "ecoli":["CS17", "ETEC", "NCTC13441", "b","BW25113", "EC958"], "klebsiella":["ERS227112", "BN373"], "citrobacter":["ROD"], "enterobacter":
+    ["ENC"]}
+
+with open(clustersfile, 'r') as fromfile:
+    with open(intereseting_genes, 'w') as tofile:
+        for line in fromfile:
+            cells = line.split()
+            # if (len(cells) >= 2 and 0 < sum([essentiality[i] for i in cells]) < len(cells)) or (len(cells) < 16 and
+            # 0 < sum([essentiality[i] for i in cells])): # all 500 genes with differentiating essentiality
+
+            # if (len(cells) == 16 and 14 < sum([essentiality[i] for i in cells]) < 16) or\
+            #         (14 < len(cells) < 16 and 14 < sum([essentiality[i] for i in cells])): # all clusters with only 1 exception
+
+            newcells = list(cells)
[…]

### 2017-02-13 bin/pathway-gc-40-50.R

@@ -1,0 +1,100 @@
+library(ggplot2)
+library(stringr)
+dbs <- "~/EnTrI/results/KEGG/"
+essentiality <- "~/EnTrI/results/biases/check-biases/"
+list_of_files <- list.files(path=dbs, full.names=T, recursive=FALSE)
+db=matrix(,nrow=0,ncol=3)
+for (filename in list_of_files)
+{
+  dbfile = as.matrix(read.table(filename, as.is=TRUE, header=TRUE, sep="\t"))
+  for (i in seq(1,nrow(dbfile)))
+  {
+    dbfile[i,3] = str_match(dbfile[i,3], '([[:print:]]+)-[[ Citrobacter rodentium]]|[[ Enterobacter cloacae subsp. cloacae NCTC 9394]]|
+                            [[ Escherichia coli O78:H11:K80 H10407 (ETEC)]]|[[ Escherichia coli K\\-12 MG1655]]|
+                            [[ Salmonella enterica subsp. enterica serovar Enteritidis P125109]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhimurium D23580]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhimurium SL1344]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhi Ty2]]|
+                            [[ Escherichia coli O25b:K100:H4-ST131 EC958 (UPEC)]]')[2]
+    
+  }
+  db = rbind(db,dbfile)
+}
+list_of_files <- list.files(path=essentiality, full.names=T, recursive=FALSE)
+genes = c()
+for (filename in list_of_files)
+{
+  essentialityfile = as.matrix(read.table(filename, as.is=TRUE, header=TRUE))
+  for (i in seq(1,nrow(essentialityfile)))
+  {
+    if (essentialityfile[i,5] >= 0.4 & essentialityfile[i,5]<= 0.5)
+    {
+      genes=c(genes,essentialityfile[i,1])
+    }
+  }
+}
+
+enrichment = data.frame(matrix(nrow=length(unique(db[,3])),ncol=4),row.names =unique(db[,3]) )
+for (i in seq(1,nrow(enrichment)))
+{
+  for (j in seq(1,ncol(enrichment)))
+  {
+    enrichment[i,j]=0
+  }
+}
+
+len_genes = 0
[…]

### 2017-02-13 bin/study-beneficial-losses.R

@@ -16,12 +16,12 @@ write.table(fishertable, file = "../results/word-enrichment/beneficialloss-pval.
 df = fishertable[-seq(30,nrow(fishertable),1),]
 df[,3] = -log10(df[,3])
 df = df[,-2]
-#df = df[c(-3,-9, -10,-11,-12,-14,-15,-18, -24,-25,-26,-29),]
+df = df[c(-3,-4,-12,-14,-25,-26,-27,-28,-29),]
 names(df) = c("word", "pvalue")
 pdf("../figures/beneficialloss-pval.pdf")
 ggplot(data=df, aes(x=reorder(word,pvalue),y=pvalue))+geom_bar(stat="identity")+ coord_flip()+labs(x='Word',y='-log10(P-value)') +
   geom_abline(slope = 0, intercept = -log10(0.05), color="red")+ylim(c(0,max(df[1,2], -log10(0.05))))+ggtitle("Beneficial losses") +
-  theme(text = element_text(size=24)) +theme(axis.text=element_text(size=20), axis.title=element_text(size=24,face="bold"))
+  theme(text = element_text(size=15))
 dev.off()
 
 

### 2017-02-13 bin/study-beneficial-losses.py

@@ -32,10 +32,11 @@ with open(nloss, 'w') as nfile:
                 for line in fromfile:
                     cells = line.split()
                     match_result = match("\S+\s+\[\S+\s+\([a-zA-Z]+\)\]\s+\[([^\]]+)\]", seqdict[cells[0]].description)
-                    if cells[2] == 'beneficial-loss':
-                        yfile.write(cells[0]+'\t'+match_result.group(1)+'\n')
-                    else:
-                        nfile.write(cells[0] + '\t' + match_result.group(1) + '\n')
+                    if match_result:
+                        if cells[2] == 'beneficial-loss':
+                            yfile.write(cells[0]+'\t'+match_result.group(1)+'\n')
+                        else:
+                            nfile.write(cells[0] + '\t' + match_result.group(1) + '\n')
 
 worddict = defaultdict(list)
 
@@ -110,10 +111,11 @@ with open(nessen, 'w') as nfile:
                 for line in fromfile:
                     cells = line.split()
                     match_result = match("\S+\s+\[\S+\s+\([a-zA-Z]+\)\]\s+\[([^\]]+)\]", seqdict[cells[0]].description)
-                    if cells[2] == 'essential':
-                        yfile.write(cells[0]+'\t'+match_result.group(1)+'\n')
-                    else:
-                        nfile.write(cells[0] + '\t' + match_result.group(1) + '\n')
+                    if match_result:
+                        if cells[2] == 'essential':
+                            yfile.write(cells[0]+'\t'+match_result.group(1)+'\n')
+                        else:
+                            nfile.write(cells[0] + '\t' + match_result.group(1) + '\n')
 
 worddict = defaultdict(list)
 
@@ -189,10 +191,11 @@ with open(nnone, 'w') as nfile:
                 for line in fromfile:
                     cells = line.split()
                     match_result = match("\S+\s+\[\S+\s+\([a-zA-Z]+\)\]\s+\[([^\]]+)\]", seqdict[cells[0]].description)
-                    if cells[2] == 'non-essential':
-                        yfile.write(cells[0]+'\t'+match_result.group(1)+'\n')
-                    else:
-                        nfile.write(cells[0] + '\t' + match_result.group(1) + '\n')
+                    if match_result:
+                        if cells[2] == 'non-essential':
+                            yfile.write(cells[0]+'\t'+match_result.group(1)+'\n')
+                        else:
+                            nfile.write(cells[0] + '\t' + match_result.group(1) + '\n')
 
 worddict = defaultdict(list)
 

### 2017-02-19 documents/manuscript.tex
Ran prokka on Enterobacter and reran everything, corrected parsimony code

@@
 These include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{Klebsiella pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{Salmonella} Typhimurium SL1344, \textit{Salmonella} Typhimurium SL3261, \textit{Salmonella} Typhimurium D23580, \textit{Salmonella} Typhimurium A130, \textit{Salmonella} Enteritidis P125109, \textit{Salmonella} Typhi Ty2, \textit{Escherichia coli} ST131 EC958, \textit{Escherichia coli} UPEC ST131, \textit{Escherichia coli} ETEC CS17, \textit{Escherichia coli} ETEC H10407, \textit{Escherichia coli} BW25113, and \textit{Escherichia coli} K-12 MG1655.
-The phylogenetic tree containing all these species is depicted in Fig.\@ \ref{fig:species-tree}.
 
@@
 
-\begin{figure}
-\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
-\caption{The species tree containing the 16 genomes used in this study.
+%\begin{figure}
+%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
+%\caption{The species tree containing the 16 genomes used in this study.
 We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
-\label{fig:species-tree}
-\end{figure}
+%\label{fig:species-tree}
+%\end{figure}
 
@@
   \centering
-  \includegraphics[page=1, scale=0.4]{essential-call-comparison-ETEC.pdf}
+  \includegraphics[page=1, scale=0.4]{essential-call-comparison-BW25113.pdf}
   \caption{}
@@
   \centering
-  \includegraphics[page=10, scale=0.4]{essential-call-comparison-ETEC.pdf}
+  \includegraphics[page=10, scale=0.4]{essential-call-comparison-BW25113.pdf}
   \caption{}
@@
   \centering
-  \includegraphics[page=26, scale=0.25]{biases.pdf}
+  \includegraphics[page=6, scale=0.25]{biases.pdf}
   \caption{}
@@
   \centering
-  \includegraphics[page=25, scale=0.25]{biases.pdf}
+  \includegraphics[page=5, scale=0.25]{biases.pdf}
   \caption{}
@@
 
-\subsection{phylogenetic analysis of gene essentiality identifies ...}
+\subsection{phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
+
+\begin{figure}
+\captionsetup[subfigure]{justification=centering}
+\begin{subfigure}{.6\textwidth}
+  \centering
+  \includegraphics[scale=0.12]{phylosift-aa-raxmlbootstrap-annotated.pdf}
+  \caption{}
+  \label{fig:tree}
+\end{subfigure}
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{fitch.pdf}
+  \caption{}
+  \label{fig:fitch}
+\end{subfigure}
+
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%  \includegraphics[page=2, scale=0.45]{upsetr.pdf}
+%  \caption{}
+%  \label{fig:upsetr1}
+%\end{subfigure}
+%\begin{subfigure}{1\textwidth}
+%  \centering
+%  \includegraphics[page=3, scale=0.65]{upsetr.pdf}
+%  \caption{}
+%  \label{fig:upsetr2}
+%\end{subfigure}
+\caption{(a) The species tree for all genomes in this study.
+Numbers in red show the number of core essential genes at each level and numbers in blue show the number of core genes at each level.
+(b) The ratio between core essential genes and core genes at each level in the species tree.
+The line connects the medians in each level.}
+\label{fig:essentiality-phylogeny}
+\end{figure}
 %\section{Discussion}

### 2017-02-19 bin/define-core-accessory-hieranoid-fitch.py

@@ -58,29 +58,32 @@ def read_gene_essentiality(indir):
                     iidict[cells[0]] = {0}
     return iidict
 
-# def read_k12(inpath, iidict):
-#     with open(inpath) as from_file:
-#         for line in from_file:
-#             cells = line.split()
-#             iidict[cells[0]] = 'essential'
-#     return iidict
+def read_k12(inpath, iidict):
+    with open(inpath) as from_file:
+        for line in from_file:
+            cells = line.split()
+            iidict[cells[0]] = {1}
+    return iidict
 
 seqdb = '/home/fatemeh/EnTrI/data/fasta-protein/chromosome/seqdb.fasta'
 clusters = '/home/fatemeh/EnTrI/results/hieranoid/clusters.txt'
 insertion_indices = '/home/fatemeh/EnTrI/results/biases/normalised-pca'
+k12path = '/home/fatemeh/EnTrI/results/ecogene-k12.txt'
 outdir = '/home/fatemeh/EnTrI/results/define-core-accessory-hieranoid-fitch'
 makedir(outdir)
 speciestreedir = '/home/fatemeh/EnTrI/bin/speciestrees-no-k12'
 sequences = read_fasta_sequences(seqdb)
 gene_essentiality = read_gene_essentiality(insertion_indices)
+gene_essentiality = read_k12(k12path, gene_essentiality)
 species_names = {"all":["BN373", "CS17", "ENC", "ERS227112", "ETEC", "NCTC13441", "ROD", "SEN", "SL1344", "STM",
-    "STMMW", "t", "SL3261","BW25113", "EC958"],"typhimurium":["STM", "SL1344", "STMMW", "SL3261"], "salmonella":["SEN", "SL1344", "STM", "STMMW", "t", "SL3261"],
-    "ecoli":["CS17", "ETEC", "NCTC13441","BW25113", "EC958"], "klebsiella":["ERS227112", "BN373"], "citrobacter":["ROD"], "enterobacter":
+    "STMMW", "t", "SL3261","BW25113", "EC958", "b"],"typhimurium":["STM", "SL1344", "STMMW", "SL3261"], "salmonella":["SEN", "SL1344", "STM", "STMMW", "t", "SL3261"],
[…]

### 2017-02-19 bin/pathanalysis.R

@@ -27,7 +27,7 @@ for (filename in list_of_files)
   essentialityfile = as.matrix(read.table(filename, as.is=TRUE, header=TRUE))
   for (i in seq(1,nrow(essentialityfile)))
   {
-    if (essentialityfile[i,3] == "non-essential")
+    if (essentialityfile[i,3] == "essential")
     {
       genes=c(genes,essentialityfile[i,1])
     }
@@ -94,8 +94,8 @@ while(enrichment[i,4] < 0.05)
 colnames(pathways) = c('word', 'pvalue')
 pathways = pathways[1:20,]
 
-pdf("../figures/non-essential-pathways.pdf")
+pdf("../figures/essential-pathways.pdf")
 ggplot(data=pathways, aes(x=reorder(word,pvalue),y=pvalue))+geom_bar(stat="identity")+ coord_flip()+labs(x='Pathway',y='-log10(P-value)') +
-  geom_abline(slope = 0, intercept = -log10(0.05), color="red")+ggtitle("Non-essential genes") +
+  geom_abline(slope = 0, intercept = -log10(0.05), color="red")+ggtitle("Essential pathways") +
   theme(text = element_text(size=15))
 dev.off()

### 2017-02-19 bin/tree-data.R

@@ -1,0 +1,77 @@
+strain =c(
+  357/4571,
+  315/4988,
+  373/5672,
+  325/4957,
+  314/4520,
+  303/4517,
+  316/4492,
+  307/4453,
+  302/4312,
+  308/4322,
+  308/4980,
+  321/4805,
+  327/4621,
+  345/4735,
+  296/4306,
+  299/4317
+)
+
+# strain =c(
+# 318/4988,
+# 290/4306,
+# 327/4621,
+# 311/4980,
+# 282/3725,
+# 370/5672,
+# 344/4735,
+# 325/4805,
+# 329/4957,
+# 303/4312,
+# 314/4520,
+# 304/4517,
+# 317/4492,
+# 301/4453,
+# 306/4322
+# )
+
+
+speciesFitch = c( 
+357/4571,
+287/4563,
+325/4957,
+269/3843,
+273/3829
+)
+
+
+subFamilyFitch = c(
+268/3251,
+280/3339,
+279/3411
+)
+
+
+familyFitch = c(
+288/3169
+)
+
+
+
+pdf(file = '~/EnTrI/figures/fitch.pdf')
+
+plot(4*strain/strain, strain, xlim=c(1,4), ylim=c(0.06,0.092), col="red4", ylab="proportion of ancestral genes", xaxt="n", xlab="",pch=19,
+     cex.lab=1.5, cex.axis=1.5)
+
+points(3*speciesFitch/speciesFitch,speciesFitch, col="red4", pch=19)
+
+points(2*subFamilyFitch/subFamilyFitch,subFamilyFitch, col="red4", pch=19)
+
+points(1*familyFitch/familyFitch,familyFitch, col="red4", pch=19)
+
+lines(1:4, c(median(familyFitch), median(subFamilyFitch), median(speciesFitch), median(strain) ), col="red4", lwd=3)
+
+#legend("topright", c("intersection", "ancestral II", "Dollo"), fil=c("red4", "olivedrab4", "cyan3"))
+axis(1, at = 1:4, labels=c("family", "subfam.", "species", "strain"), cex.axis=1.5)
+
+dev.off()

### 2017-02-19 bin/upsetr-hieranoid.R

@@ -47,23 +47,18 @@ for (clusterindex in (1:numclusters))
   }
 }
 
-names(presence) <- c("Enterobacter cloacae NCTC 9394", "Klebsiella pneumoniae Ecl8", "Klebsiella pneumoniae RH201207",
+names(presence) <- rev(c("Enterobacter cloacae NCTC 9394", "Klebsiella pneumoniae Ecl8", "Klebsiella pneumoniae RH201207",
                      "Citrobacter rodentium ICC168", "Salmonella typhimurium SL1344", "Salmonella typhimurium SL3261",
                      "Salmonella typhimurium D23580", "Salmonella typhimurium A130", "Salmonella enteritidis P125109",
                      "Salmonella typhi Ty2", "Escherichia coli ST131 EC958", "Escherichia coli UPEC ST131", "Escherichia coli ETEC CS17",
-                     "Escherichia coli ETEC H10407", "Escherichia coli BW25113", "Escherichia coli K-12 MG1655")
-names(essentiality) <- c("Enterobacter cloacae NCTC 9394", "Klebsiella pneumoniae Ecl8", "Klebsiella pneumoniae RH201207",
+                     "Escherichia coli ETEC H10407", "Escherichia coli BW25113", "Escherichia coli K-12 MG1655"))
+names(essentiality) <- rev(c("Enterobacter cloacae NCTC 9394", "Klebsiella pneumoniae Ecl8", "Klebsiella pneumoniae RH201207",
                      "Citrobacter rodentium ICC168", "Salmonella typhimurium SL1344", "Salmonella typhimurium SL3261",
                      "Salmonella typhimurium D23580", "Salmonella typhimurium A130", "Salmonella enteritidis P125109",
                      "Salmonella typhi Ty2", "Escherichia coli ST131 EC958", "Escherichia coli UPEC ST131", "Escherichia coli ETEC CS17",
-                     "Escherichia coli ETEC H10407", "Escherichia coli BW25113", "Escherichia coli K-12 MG1655")
+                     "Escherichia coli ETEC H10407", "Escherichia coli BW25113", "Escherichia coli K-12 MG1655"))
 
-names(presence) <- c("Klebsiella pneumoniae Ecl8", "Klebsiella pneumoniae RH201207", "Enterobacter cloacae NCTC 9394",
-                     "Citrobacter rodentium ICC168", "Salmonella Typhimurium SL1344", "Salmonella Typhimurium SL3261",
[…]

### 2017-02-27 documents/manuscript.tex
Edited manuscript

@@
 \usepackage{subcaption}
+\usepackage{pdflscape}
 \graphicspath{{../figures/}}
@@
 Therefore, in addition to core and accessory genes, we have studied duplicate genes.
-To study whether each gene in the 16 organisms is core we used Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to iteratively search for homologous proteins in our dataset and cluster them.
+To study whether each gene in the 16 organisms is core, we used Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to iteratively search for homologous proteins in our dataset and cluster them.
 We divided the clusters of homologous genes into three groups based on their conservation.
-Genus specific class contain genes that are present only in one genus, the genes in single copy class are present in more than one genus and more than 70\% of them are not duplicated (core genes), and  the genes in multi-copy class are present in more than one genus and more than 30\% of them are duplicated (duplicate genes).
+Genus specific class contains genes that are present only in one genus, the genes in single copy class are present in more than one genus and more than 70\% of them are not duplicated (core genes), and  the genes in multi-copy class are present in more than one genus and more than 30\% of them are duplicated (duplicate genes).
 
@@
 If NPEQ is less than -1.644854, it means that the gene has tolerated many insertions, so it is beneficial for the organism to lose this gene in the rich medium that we used.
-However, if NPEQ is greater that 1.644854, the gene can tolerate very few or no insertions indicating that the gene is essential for cell viability in our test medium.
+However, if NPEQ is greater than 1.644854, the gene can tolerate very few or no insertions indicating that the gene is essential for cell viability in our test medium.
 Any other NPEQ value shows an intermediate number of insertions in genes meaning that it is not beneficial for the organism to lose these genes, but they are not essential, too.
@@
 Other enriched pathways are mostly metabolism related.
-These include fatty acid biosynthesis that produce cell membrane, peptidoglycan and lipopolysaccharide biosynthesis that are essential components of cell wall, terpenoid backbone biosynthesis which feeds peptidoglycan biosynthesis, nucleotide and amino acid metabolism, and the metabolism of important cofactors and vitamins like riboflavin, biotin,porphyrin and chlorophyll (Figure~\ref{fig:essentiality-pathway-ess}).
+These include fatty acid biosynthesis that produces cell membrane, peptidoglycan and lipopolysaccharide biosynthesis that are essential components of cell wall, terpenoid backbone biosynthesis which feeds peptidoglycan biosynthesis, nucleotide and amino acid metabolism, and the metabolism of important cofactors and vitamins like riboflavin, biotin,porphyrin and chlorophyll (Figure~\ref{fig:essentiality-pathway-ess}).
 {\color{red} CAN WE HAVE SOMETHING LIKE THIS? http://www.nature.com/articles/srep00125}
@@
 
-\subsection{phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
+\subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
+When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}.
+As mutations in essential genes are less likely to be tolerable, less mutations should be fixed in these genes, and therefore essential genes should evolve more slowly than non-essential genes.
+This is known as knockout rate hypothesis \cite{krylov_gene_2003}.
+Different studies have verified this hypothesis in prokaryotes \cite{jordan_essential_2002, gerdes_experimental_2003, rocha_analysis_2004, silander_constancy_2009}.
+
+Here, we have tested this hypothesis by comparing the number of core essential genes and core genes in the phylogenetic tree as we go up to the root.
+We have used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes) to define if a gene is core essential or core at each level in the phylogenetic tree.
+If we get 1 at the top of the tree while studying essentiality of a gene using Fitch algorithm, it means that the gene is core essential and otherwise, the gene is non-essential.
+The same applies to the study of core genes.
+The phylogenetic tree has been annotated with the number of core essential genes (red) and the number of core genes (blue) at each level in Figure~\ref{fig:tree}.
+We then plotted the ratios at each level in the phylogenetic tree in Figure~\ref{fig:fitch} and connected the medians in each level.
+The connecting line shows that the ratio between core essential genes and core genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.
 
@@
 \end{subfigure}
-
-%\begin{subfigure}{.5\textwidth}
-%  \centering
-%  \includegraphics[page=2, scale=0.45]{upsetr.pdf}
-%  \caption{}
-%  \label{fig:upsetr1}
-%\end{subfigure}
-%\begin{subfigure}{1\textwidth}
-%  \centering
-%  \includegraphics[page=3, scale=0.65]{upsetr.pdf}
-%  \caption{}
-%  \label{fig:upsetr2}
-%\end{subfigure}
 \caption{(a) The species tree for all genomes in this study.
@@
 (b) The ratio between core essential genes and core genes at each level in the species tree.
+The dots in the strain level show the ratios for all 16 bacteria; the dots in the species level show the ratios for Enterobacter, Klebsiella, Citrobacter, Salmonella, and Escherichia; the subfam.
+level shows the ratios for the common ancestor of Enterobacter and klebsiella, the common ancestor of Citrobacter and Salmonella, and the common ancestor of Citrobacter, Salmonella, and Escherichia; and finally the dot in the family level shows the ratio for the root.
 The line connects the medians in each level.}
@@
 \end{figure}
+
+We have studied the essentiality status of genes involved in important biological processes in Table~\ref{tab:biological-processes}.
+These processes include cell division, DNA replication, transcription, translation, and important metabolic pathways such as peptidoglycan and fatty acid biosynthesis.
+If a gene was not core essential using Fitch algorithm, we looked at its essentiality in every genome: if it was essential in some of the genomes we classified it as accessory essential and if it was not essential in any genome we classified it as non-essential.
+FtsA and zipA genes are essential for cell division.
+FtsA is essential in all the genomes in our study, however zipA is essential everywhere except for Salmonella typhimurium A130 in which this gene is near essential.
+We investigated if this gene has a homolog in Salmonella typhimurium A130 and found no homologs.
+It has been shown that 21 mutations in ftsA gene can make it independent of zipA gene so that the zipA gene is no longer essential \cite{geissler_gain--function_2003, pichoff_ftsa_2012}.
+However, ftsA gene does not have any of these mutations in Salmonella typhimurium A130, so there might be another factor that is causing zipA not to be essential.
+\textcolor{red}{ADD MORE EXAMPLES HERE}
+
+So far we have shown that essential genes are less likely to be lost during evolution.
+The other question is if essential genes tend to keep their essentiality in different genomes.
+Bergmiller et al.
+\cite{bergmiller_patterns_2012} have shown that only essential genes whose functions cannot be taken over by other genes keep their essentiality and others are possible to be lost or non-essential in other genomes.
+The genes that can replace other genes are not necessarily homologous to them, but they have similar functions.
+
+We have compared the number of genes that are shared between our bacterial genomes (\ref{fig:upsetr-conservation}) and the number of genes essential in them (\ref{fig:upsetr-essentiality}) and used UpSetR package \cite{conway_upsetr:_2016} to visualise the results in Fig.\@ \ref{fig:upsetr}.
+As shown in the figures, among 2162 genes that are shared between all the bacteria under study, only 135 are essential everywhere.
+We looked at subsets of the genes that were shared in every combination of our bacterial strains to see whether they show a trend similar to the phylogenetic tree.
+The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show such a trend.
+The reason is that many of the essential genes that are not shared in all genomes are probably compensable by other genes and so the essentiality does not look tree-like.
+
+\begin{landscape}
+\begin{table}
+\small
+\caption{The essentiality status of genes involved in important biological processes}
+\label{tab:biological-processes}
+\begin{tabular}[t]{l l l l l}
+\hline
+\textbf{Biological process} & \textbf{Subprocess} & \textbf{Core essential} & \textbf{Accessory essential} & \textbf{Non-essential}\\
+\hline
+Cell Division & & \begin{tabular}[t]{@{}l@{}}ftsAHLQWYZ, minE,\\ mukB, zipA \end{tabular}& ftsKNX, minD & \begin{tabular}[t]{@{}l@{}}CedA, ftsJ, minC,\\ sdiA, sulA\end{tabular}\\
+DNA replication & Polymerases I, II, III & dnaENQX, holABD, polA & holC & holE, polB\\	
+& Supercoiling & gyrAB, parCE & & \\
+& Primosome-associated & dnaBCGT, priA, ssb & priB, rep & priC\\
+Transcription & RNA polymerase & rpoABC &  & \\
+& \begin{tabular}[t]{@{}l@{}}Sigma, elongation, anti- \\ and termination factors\end{tabular} & nusABG, rho, rpoDH & rpoEN & rpoS\\	
+Translation & tRNA-synthetases & \begin{tabular}[t]{@{}l@{}}alaS, argS, asnS, aspS, cysS, \\ glnS, gltX, glyQS, hisS, ileS, \\ leuS, lysS, metG, pheST, \\ proS, serS, thrS, tyrS, valS \end{tabular} & trpS	 & \\
+& Ribosome components & \begin{tabular}[t]{@{}l@{}}rplBCDEFJKLMNOPQRSTUV,\\ rplWXY, rpmABCD,\\ rpsABCDEFGHJKLMNPQRSTU\end{tabular} & rplA, rpmEGHI, rpsIO & rplI, rpmF\\
+& \begin{tabular}[t]{@{}l@{}}Initiation, elongation \\ and peptide chain \\ release factors\end{tabular} & fusA, infABC, prfAB, tsf & efp & prfC, selB, tufAB \\
+Biosynthetic pathways & & & & \\
+Peptidoglycan & & MraY, murABCDEFGI & & ddlAB\\	
+Fatty acids & & accABCD, fabABDGHIZ & & \\
+\hline
+\end{tabular}
+\end{table}
+\end{landscape}
+
+\begin{figure}
+\captionsetup[subfigure]{justification=centering}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.45,page=2]{upsetr.pdf}
+  \caption{}
+  \label{fig:upsetr-conservation}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+  \includegraphics[scale=0.45,page=3]{upsetr.pdf}
+  \caption{}
+  \label{fig:upsetr-essentiality}
+\end{subfigure}
+\caption{(a) The number of genes and (b) essential genes shared between different groups of bacteria.
+The bars show the number of genes and the filled circles show which bacteria are sharing those genes.}
+\label{fig:upsetr}
+\end{figure}
+
 %\section{Discussion}

### 2017-02-27 bin/define-core-accessory-hieranoid-fitch.py

@@ -3,12 +3,10 @@
 exec python3 $0 ${1+"$@"}
 """
 
-from os import listdir, path, mkdir, system
+from os import listdir, path, mkdir
 from Bio import SeqIO
 from shutil import rmtree
-from re import match, findall
-from collections import defaultdict
-from numpy import mean, std
+from re import match
 
 essen = 1.644854
 
@@ -95,7 +93,7 @@ for item in species_names.keys():
     makedir(esscoredir)
     nesscoredir = speciesdir + '/core-never-essential-genomes'
     makedir(nesscoredir)
-    aesscoredir = speciesdir + '/core-ambig-essential-genomes'
+    aesscoredir = speciesdir + '/core-sometimes-essential-genomes'
     makedir(aesscoredir)
     accessorydir = speciesdir + '/accessory-genomes'
     makedir(accessorydir)
@@ -122,14 +120,8 @@ for item in species_names.keys():
             list_of_genes = []
             stack_of_species = Stack()
             stack_of_genes = Stack()
-            # list_of_essential_genes = []
 
             genes = line.split()
-            # findall_result = findall('(([a-zA-Z0-9]+?)_[a-zA-Z0-9]+):', line)
-            # temp_result = findall('(([a-zA-Z]+?)\d+):', line)
-            # for matches in temp_result:
-            #     if matches[1] != 'n':
-            #         findall_result.append(matches)
             for g in genes:
                 s = match('([a-zA-Z0-9]+_|[a-zA-Z]+)[a-zA-Z0-9]+', g).group(1).strip('_')
                 if s in gene_dict.keys():
@@ -187,29 +179,16 @@ for item in species_names.keys():
                 ii.append(list(essentiality_dict.values())[0])
                 presence.append(list(gene_dict.values())[0])
 
-            # if sum(gene_dict.values()) == num_species and num_species > 1:
-            #
-            # elif sum(gene_dict.values()) == num_species and num_species == 1:
-            #     ii.append(list(essentiality_dict.values())[0])
-
-    # with open(Rin, 'w') as tofile:
-    #     for inin in ii:
-    #         tofile.write(str(inin)+'\n')
-    # system('Rscript '+Rcode)
-    # with open(Rout, 'r') as fromfile:
-    #     essen = float(fromfile.readline())
-
     gene_dict = {species_names[item][i]: {0} for i in range(num_species)}
     essentiality_dict = {species_names[item][i]: {0} for i in range(num_species)}
     with open(clusters) as from_file:
         index = 0
[…]

### 2017-03-15 documents/manuscript.tex
Corrected the annotations and reran the codes, listed some candidate genes for further study

@@
 \subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
-When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}.
+%When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}.
 As mutations in essential genes are less likely to be tolerable, less mutations should be fixed in these genes, and therefore essential genes should evolve more slowly than non-essential genes.

### 2017-03-15 bin/correct-annotations.py

@@ -1,0 +1,75 @@
+from re import match
+
+old_annot = '../data/embl/chromosome_old/Enterobacter_cloacae_subsp_cloacae_NCTC_9394_v1.embl'
+prokka_annot = '../results/prokka/ENC/ENC.embl'
+new_annot = '../data/embl/chromosome/Enterobacter_cloacae_subsp_cloacae_NCTC_9394_v1.embl'
+
+flag = 0
+with open(old_annot, 'r') as fromfile:
+    with open(new_annot, 'w') as tofile:
+        for line in fromfile:
+            if not line.startswith('SQ'):
+                tofile.write(line)
+                if not flag and match('FT\s+CDS\s+', line):
+                    flag = 1
+                if flag == 1 and match('FT\s+/locus_tag=\"(\S+)\"', line):
+                    gene = match('FT\s+/locus_tag=\"(\S+)\"', line).group(1)
+                    locus_tag = match('([a-zA-Z0-9]+_|[a-zA-Z]+)[a-zA-Z0-9]+', gene).group(1).strip('_')
+                    flag = 2
+            else:
+                break
+
+oldline = ''
+addline = ''
+counter = 1
+with open(old_annot, 'r') as oldfile:
+    with open(prokka_annot, 'r') as addfile:
+        with open(new_annot, 'a') as newfile:
+            while not oldline.startswith('SQ') and not addline.startswith('SQ'):
+                while not match('FT\s+CDS\s+', oldline) and not oldline.startswith('SQ'):
+                    oldline = oldfile.readline()
+                while not match('FT\s+CDS\s+', addline) and not addline.startswith('SQ'):
+                    addline = addfile.readline()
+                if oldline.startswith('SQ') or addline.startswith('SQ'):
+                    break
+                start_end = match('FT\s+CDS\s+[^\d]*(\d+)\.\.>?(\d+)', oldline)
+                startold = int(start_end.group(1))
+                endold = int(start_end.group(2))
+                start_end = match('FT\s+CDS\s+[^\d]*(\d+)\.\.(\d+)', addline)
+                startadd = int(start_end.group(1))
+                endadd = int(start_end.group(2))
+                # overlap = (min(endadd, endold) - max(endadd, endold)) /
+                if startold == startadd or endold == endadd:
+                    oldline = oldfile.readline()
+                    addline = addfile.readline()
+                elif startold < startadd:
+                    oldline = oldfile.readline()
+                else:
+                    newfile.write(addline)
+                    addline = addfile.readline()
[…]

### 2017-03-15 bin/study-beneficial-losses.R

@@ -16,7 +16,7 @@ write.table(fishertable, file = "../results/word-enrichment/beneficialloss-pval.
 df = fishertable[-seq(25,nrow(fishertable),1),]
 df[,3] = -log10(df[,3])
 df = df[,-2]
-df = df[c(-2,-5,-9,-13),]
+# df = df[c(-2,-5,-9,-13),]
 names(df) = c("word", "pvalue")
 pdf("../figures/beneficialloss-pval.pdf")
 ggplot(data=df, aes(x=reorder(word,pvalue),y=pvalue))+geom_bar(stat="identity")+ coord_flip()+labs(x='Word',y='-log10(P-value)') +

### 2017-03-15 bin/universally-conserved-sometimes-essential.R

@@ -1,0 +1,11 @@
+library(gplots)
+ucse <- read.csv("~/EnTrI/results/interesting_genes/universally-conserved_sometimes-essential.tsv", header=TRUE, sep='\t')
+to.plot <- as.matrix(ucse[,3:18])
+rownames(to.plot) = ucse[,1]
+breaks <- seq(min(to.plot), max(to.plot), length.out=51)
+my_palette <- colorRampPalette(c("cyan", "palevioletred4"))(n = 50)
+pdf("~/EnTrI/figures/heatmap-ucse.pdf", width = 15, height = 30)
+heatmap.2(to.plot, margins = c(10, 10), main = "Universally conserved sometimes essential", density.info="none", trace="none", keysize=1,
+          lmat=rbind( c(0, 3), c(2,1), c(0,4)), lhei=c(0.03, 1, 0.05 ), lwid = c(0.05,1), Rowv=FALSE, Colv=FALSE, dendrogram='none',
+          col=my_palette, breaks = breaks)
+dev.off()

### 2017-03-29 documents/manuscript.tex
Made a table of unannotated genes that we had annotated using prokka, added duplicated genes to heatmap

@@
 Core essential genes are responsible for essential processes such as cell division, DNA replication, transcription and translation and some important pathways like peptidoglycan and fatty acid biosynthesis \cite{barquist_comparison_2013}.
-Accessory essential genes differ in genomes due to different reasons such as niche adaptation, gene duplication and the existence of alternative pathways \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}.
+Accessory essential genes differ in genomes due to different reasons such as niche adaptation, functional redundancy and the existence of alternative pathways \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013, bergmiller_patterns_2012}.
 Another group of accessory essential genes are phage repressors \cite{barquist_comparison_2013}.
@@
 For this we needed to study core and accessory sets of genes.
-Moreover, in the presence of two redundant variations of one gene, if we knock out one copy using TraDIS, the other copy compensates and the organism can still survive \cite{dean_pervasive_2008}.
+Moreover, in the presence of two redundant variations of one gene, if we knock out one copy using TraDIS, the other copy compensates and the organism can still survive \cite{bergmiller_patterns_2012,dean_pervasive_2008}.
 This leads to a different essentiality inference using TraDIS.
@@
 
-\subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
+%\subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
+
 %When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}.
@@
 
-Here, we have tested this hypothesis by comparing the number of core essential genes and core genes in the phylogenetic tree as we go up to the root.
-We have used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes) to define if a gene is core essential or core at each level in the phylogenetic tree.
-If we get 1 at the top of the tree while studying essentiality of a gene using Fitch algorithm, it means that the gene is core essential and otherwise, the gene is non-essential.
-The same applies to the study of core genes.
-The phylogenetic tree has been annotated with the number of core essential genes (red) and the number of core genes (blue) at each level in Figure~\ref{fig:tree}.
+We have shown that many of the essential genes are conserved.
+But are essential genes more likely to be conserved?
+To answer this question, we have compared the number of ancestrally essential genes and ancestrally present genes in the phylogenetic tree as we go up to the root.
+We have used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes) to define if a gene is ancestral essentially or present at each level in the phylogenetic tree.
+If we get 1 at the top of the tree while studying essentiality of a gene using Fitch's algorithm, it means that the gene is ancestrally essential and otherwise, the gene is ancestrally non-essential.
+The same applies to the study of ancestrally present genes.
+The phylogenetic tree has been annotated with the number of ancestrally essential genes (red) and the number of ancestral genes (blue) at each level in Figure~\ref{fig:tree}.
 We then plotted the ratios at each level in the phylogenetic tree in Figure~\ref{fig:fitch} and connected the medians in each level.
-The connecting line shows that the ratio between core essential genes and core genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.
+The connecting line shows that the ratio between ancestrally essential genes and ancestral genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.
 
@@
 \caption{(a) The species tree for all genomes in this study.
-Numbers in red show the number of core essential genes at each level and numbers in blue show the number of core genes at each level.
-(b) The ratio between core essential genes and core genes at each level in the species tree.
+Numbers in red show the number of ancestrally essential genes at each level and numbers in blue show the number of ancestral genes at each level.
+(b) The ratio between ancestrally essential genes and ancestral genes at each level in the species tree.
 The dots in the strain level show the ratios for all 16 bacteria; the dots in the species level show the ratios for Enterobacter, Klebsiella, Citrobacter, Salmonella, and Escherichia; the subfam.
@@
 These processes include cell division, DNA replication, transcription, translation, and important metabolic pathways such as peptidoglycan and fatty acid biosynthesis.
-If a gene was not core essential using Fitch algorithm, we looked at its essentiality in every genome: if it was essential in some of the genomes we classified it as accessory essential and if it was not essential in any genome we classified it as non-essential.
-FtsA and zipA genes are essential for cell division.
+If a gene was not ancestrally essential using Fitch's algorithm, we looked at its essentiality in every genome: if it was essential in some of the genomes we classified it as ambiguous and if it was not essential in any genome we classified it as ancestrally non-essential.
+%FtsA and zipA genes are essential for cell division.
 FtsA is essential in all the genomes in our study, however zipA is essential everywhere except for Salmonella typhimurium A130 in which this gene is near essential.
@@
 \textcolor{red}{ADD MORE EXAMPLES HERE}
-
-So far we have shown that essential genes are less likely to be lost during evolution.
-The other question is if essential genes tend to keep their essentiality in different genomes.
-Bergmiller et al.
-\cite{bergmiller_patterns_2012} have shown that only essential genes whose functions cannot be taken over by other genes keep their essentiality and others are possible to be lost or non-essential in other genomes.
-The genes that can replace other genes are not necessarily homologous to them, but they have similar functions.
-
-We have compared the number of genes that are shared between our bacterial genomes (\ref{fig:upsetr-conservation}) and the number of genes essential in them (\ref{fig:upsetr-essentiality}) and used UpSetR package \cite{conway_upsetr:_2016} to visualise the results in Fig.\@ \ref{fig:upsetr}.
-As shown in the figures, among 2162 genes that are shared between all the bacteria under study, only 135 are essential everywhere.
-We looked at subsets of the genes that were shared in every combination of our bacterial strains to see whether they show a trend similar to the phylogenetic tree.
-The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show such a trend.
-The reason is that many of the essential genes that are not shared in all genomes are probably compensable by other genes and so the essentiality does not look tree-like.
 
@@
 \hline
-\textbf{Biological process} & \textbf{Subprocess} & \textbf{Core essential} & \textbf{Accessory essential} & \textbf{Non-essential}\\
+\textbf{Biological process} & \textbf{Subprocess} & \textbf{Ancestrally essential} & \textbf{Ambiguous} & \textbf{Ancestrally non-essential}\\
 \hline
@@
 Translation & tRNA-synthetases & \begin{tabular}[t]{@{}l@{}}alaS, argS, asnS, aspS, cysS, \\ glnS, gltX, glyQS, hisS, ileS, \\ leuS, lysS, metG, pheST, \\ proS, serS, thrS, tyrS, valS \end{tabular} & trpS	 & \\
-& Ribosome components & \begin{tabular}[t]{@{}l@{}}rplBCDEFJKLMNOPQRSTUV,\\ rplWXY, rpmABCD,\\ rpsABCDEFGHJKLMNPQRSTU\end{tabular} & rplA, rpmEGHI, rpsIO & rplI, rpmF\\
+& Ribosome components & \begin{tabular}[t]{@{}l@{}}rplBCDEFJKLMNOPQRSTUV,\\ rplWXY, rpmABCD,\\ rpsABCDEFGHJKLMNPQRSTU\end{tabular} & \begin{tabular}[t]{@{}l@{}}rplA, rpmEGHI,\\ rpsIO\end{tabular} & rplI, rpmF\\
 & \begin{tabular}[t]{@{}l@{}}Initiation, elongation \\ and peptide chain \\ release factors\end{tabular} & fusA, infABC, prfAB, tsf & efp & prfC, selB, tufAB \\
@@
 \end{landscape}
+
+%So far we have shown that essential genes are less likely to be lost during evolution.
+The other question is if essential genes tend to keep their essentiality in different genomes.
+Bergmiller et al.
+\cite{bergmiller_patterns_2012} have shown that only essential genes whose functions cannot be taken over by other genes keep their essentiality and others are possible to be lost or non-essential in other genomes.
+The genes that can replace other genes are not necessarily homologous to them, but they have similar functions.
+Genes that are essential in one organism can be non-essential or absent in other organisms.
+Bergmiller et al.
+\cite{bergmiller_patterns_2012} have studied 26 genes that are essential in E.
+\textit{coli}.
+Some of these genes are essential in other bacteria and some are non-essential or not present in other bacteria.
+In 10 cases, they were able to find genes that compensate for these essential genes if they are overexpressed and called these genes high copy suppressors.
+High copy suppressors are not necessarily similar to their counterpart essential genes in structure or sequence, but they have similar functions.
+They have shown that genes that are not present or essential everywhere are more likely to be compensated for by overexpression of another gene.
+The other reason for genes to lose their essentiality in some strains is changes in their physiology or environment \cite{barquist_comparison_2013}.
+
+We have compared the number of genes that are shared between our bacterial genomes (\ref{fig:upsetr-conservation}) and the number of genes essential in them (\ref{fig:upsetr-essentiality}) and used UpSetR package \cite{conway_upsetr:_2016} to visualise the results in Fig.\@ \ref{fig:upsetr}.
+As shown in the figures, among 2162 genes that are shared between all the bacteria under study, only 135 are essential everywhere and many of the essential genes are not conserved or essential everywhere probably because of their compensability, physiological, or environmental changes.
+%We looked at subsets of the genes that were shared in every combination of our bacterial strains to see whether they show a trend similar to the phylogenetic tree.
+The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show such a trend.
+The reason is that many of the essential genes that are not shared in all genomes are probably compensable by other genes and so the essentiality does not look tree-like.
 

### 2017-03-29 bin/unannotated-genes.py

@@ -1,0 +1,34 @@
+from re import match
+from os import listdir, mkdir
+from Bio import SeqIO
+
+
+def read_fasta_sequences(seq):
+    sequences = {}
+    with open(seq, 'rU') as fasta_file:
+            sequences.update(SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta')))
+    return sequences
+
+fastas = '../data/fasta-protein/chromosome/seqdb.fasta'
+essentialities = '../results/biases/normalised-pca/'
+newannot = '../results/unannotated/'
+mkdir(newannot)
+sequences = read_fasta_sequences(fastas)
+
+list_of_files = listdir(essentialities)
+for filename in list_of_files:
+    with open(essentialities + filename, 'r') as fromfile:
+        with open(newannot + filename, 'w') as tofile:
+            tofile.write('ID\tname\tfunction\tcoord\tNPEQ\tessentiality\n')
+            for line in fromfile:
+                cells = line.split()
+                if cells[0].endswith('added'):
+                    desc = sequences[cells[0]].description
+                    print(desc)
+                    match_result = match(
+                        '[^_]+_\d+added\s+\[[^/]+/(\S+)[^\]]+\]+\s+\[(.*)\]\s+\[.*\]\s+\[(.*)\]\s+\[.*\]', desc)
+                    coord = match_result.group(1)
+                    func = match_result.group(2)
+                    name = match_result.group(3)
+                    tofile.write(cells[0] + '\t' + name + '\t' + func + '\t' + coord + '\t' + cells[1] + '\t' +
+                                 cells[2] + '\n')

### 2017-06-20 documents/method-comparison.tex
Added an ambiguous region between non-essential and beneficial loss, used eggNOG to study beneficial losses, generated a table containing all KEGG pathways with core essential and ancestrally essential genes

--- 2017-02-27 documents/manuscript.tex
@@
 % adjust caption style
-\usepackage[font=small,aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,singlelinecheck=off]{caption}
 
@@
 \usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{pdflscape}
 \graphicspath{{../figures/}}
@@
 \reversemarginpar
-\usepackage{amsmath}
 
@@
 \begin{document}
-\vspace*{0.35in}
+In this document, I have compared different essential gene predictors.
+These include BioTraDIS \cite{barquist_tradis_2016} (blue), log fold changes from Monte Carlo method \cite{turner_essential_2015} (green), P-values from running DESeq and Monte Carlo sampling (red), the ratio between the length of the longest uninterruped segment of the gene and the length of the gene (orange), and the average distance between insertion sites.
+The last two methods are selected from \cite{freed_combining_2016} Figure S3.
+To calculate one tailed P-values from two tailed P-values obtained from DESeq, I have divided the P-values by two, and if their corresponding logFC value was less than zero, pval = 1 - pval.
+The set of essential genes are obtained by comparing the genes in each strain to the essential genes in E.coli K-12 from ecogene study \cite{zhou_ecogene_2013}.
+The results are shown in Figure~\ref{fig:roc}
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[page=1, scale=0.5]{essential-call-comparison-NCTC13441.pdf} &
+\includegraphics[page=1, scale=0.5]{essential-call-comparison-ETEC.pdf} \\
+\includegraphics[page=1, scale=0.5]{essential-call-comparison-CS17.pdf} &\\
+\end{tabular}
+\caption{The comparison of different methods for calling essential genes in three E.coli strains.}
+\label{fig:roc}
+\end{figure*}
 
-% title goes here:
-\begin{flushleft}
-{\Large
-\textbf\newline{Title}
-}
-\newline
-% authors go here:
-\\
-%Fatemeh Ashari Ghomi\textsuperscript{1},
-%Paul P. Gardner\textsuperscript{1,*},
-%Lars Barquist\textsuperscript{2}
-%\\
-%\bigskip
-%\bf{1} University of Canterbury
-%\\
-%\bf{2} Wurzburg University
-%\\
-%\bigskip
-%* paul.gardner@canterbury.ac.nz
+In Figure~\ref{fig:dotplot}, I have compared the logodds value obtained from BioTraDIS to both logFC and P-values from Monte Carlo method for CS17 strain.
 
-\end{flushleft}
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[page=2, scale=0.5]{essential-call-comparison-CS17.pdf} &
+\includegraphics[page=3, scale=0.5]{essential-call-comparison-CS17.pdf} \\
+\end{tabular}
+\caption{The figure shows the results of BioTraDIS vs. logFC from Monte Carlo method at left and BioTraDIS vs. DESeq P-values at right.}
+\label{fig:dotplot}
+\end{figure*}
 
-\section*{Abstract}
-%Many genes have been identified with advances in sequencing technology and genome annotation methods.
-However, not all of these genes are of the same importance.
-We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae.
-We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes.
-After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level.
-Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved.
-However, on average, essential genes are more likely to be conserved.
+In Figure~\ref{fig:hist}, I have plotted the distribution of each method.
+The essential and non-essential genes are well separated in Monte Carlo logFC, largest uninterrupted fraction, and mean distance between inserts.
+However, it is hard to distinguish between non-essential genes and beneficial losses in all these methods.
 
-% now start line numbers
-\linenumbers
+\begin{figure*}
+\centering
+\begin{tabular}{c c}
+\includegraphics[page=4, scale=0.5]{essential-call-comparison-CS17.pdf} &
+\includegraphics[page=5, scale=0.5]{essential-call-comparison-CS17.pdf} \\
+\includegraphics[page=6, scale=0.5]{essential-call-comparison-CS17.pdf} &
+\includegraphics[page=7, scale=0.5]{essential-call-comparison-CS17.pdf} \\
+\end{tabular}
+\caption{The figure shows the distribution of Monte Carlo logFC, Monte Carlo P-value, log2(largest uninterrupted fraction), and log2(mean distance between inserts)}
+\label{fig:hist}
+\end{figure*}
 
-% the * after section prevents numbering
-\section{Introduction}
-%The study of cells can give us new insights for understanding the foundation of life.
-However, the naturally existing cells are very complex which hampers the study of their constituents.
-Restricting these studies to the genes that are required for the growth of cells can reduce this complexity and enable us to study life in its simplest form \cite{juhas_essence_2011}.
+So far, Monte Carlo with logFC values is the best method.
+Interestingly, even though the mean distance between inserts and largest uninterrupted fraction methods are very simple, they work pretty well.
 
-Generating genomic variants that carry genes responsible for essential cellular processes can open up new research directions.
-Adding genes or metabolic pathways to cells provides new variants with specialised phenotypes.
-Modified cells have different potential applications in biotechnology \cite{juhas_bacillus_2014}, fuel production \cite{seo_synthetic_2013}, healthcare, and food production \cite{juhas_meeting_2013}.
-Another important application for studying essential genes is in drug discovery \cite{juhas_essential_2012}.
-Infectious diseases are among the top major causes of mortality worldwide.
-Even though antibiotic resistance is growing among bacteria, the antibiotic discovery and development rate is diminishing \cite{fischbach_antibiotics_2009, nathan_antibiotics_2004}.
-Therefore, it is urgent to find new drugs for infectious diseases.
-New antibiotics target genes that are essential for the survival of pathogenic bacteria in order to control disease.
-Some well-known antibiotics that target essential functions are tetracyclines that bind to small ribosomal subunit and interfere with protein translation \cite{brodersen_structural_2000}, penicillins that target peptidoglycan and inhibit the cell wall synthesis \cite{chung_rapid_2009}, and quinolones that target DNA Gyrase \cite{marcusson_interplay_2009}.
-%Some candidate antibiotic targets found by investigating essential genes are trans-translation genes \textit{ssrA} and \textit{smpB} from \textit{helicobacter pylori} \cite{thibonnier_trans-translation_2008}, \textit{CcrM} and \textit{MenH} which are bacterial methyltransferases, and \textit{lptD} (previously known as \textit{ostA}) in \textit{Pseudomonas aeruginosa} \cite{srinivas_peptidomimetic_2010} whose essential role is yet under investigation \cite{werneburg_inhibition_2012}.[SOME ANTIBIOTICS THAT TARGET ESSENTIAL GENES ARE...]
-
-In the earliest attempt for the identification of essential genes, Mushegian and Koonin compared the genomes of \textit{Haemophilus influenzae} and \textit{Mycoplasma genitalium}, assuming that the genes that are shared in these two phylogenetically distant bacteria are indispensable and reported 256 genes fulfilling this requirement \cite{mushegian_minimal_1996}.
-With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes declined to less than 50 which is not enough for performing all essential functions in a cell \cite{charlebois_computing_2004}.
-Therefore, the use of experimental methods for the identification of essential genes is vital.
-Researchers have now studied the essential genes in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods.
-Baba et al.\@ \cite{baba_construction_2006} made a library of single gene deletions for \textit{Escherichia coli} K-12.
-The 303 genes where viable \textit{Escherichia coli} colonies failed to grow are the candidate essential genes.
-Another group used an antisense RNA knockdown approach to study gene essentiality in \textit{Staphylococcus} aureus \cite{forsyth_genome-wide_2002}.
-In this method, if the expression of an antisense RNA hinders the growth of the cell, its cognate gene is known as essential.
-Both of these methods are labour intensive and are dependent on the accuracy of genome annotations.
-Another widely used procedure is transposon mutagenesis combined with high-throughput sequencing \cite{chao_design_2016, van_opijnen_transposon_2013, barquist_approaches_2013} which includes different approaches namely, Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}.
-These procedures differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}.
-Nonetheless, all share the same workflow: pools of single insertion mutants are constructed using transposon mutagenesis.
-After a growth phase, mutants that are fitter outnumber the less fit ones.
-Using high-throughput sequencing and tallying transposon junctions gives an indication of whether a genomic region is essential or not.
-A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium and conversely, a low number of transposon insertions indicates that a gene is essential in the medium.
-
-There are two groups of essential genes: core essential genes are indispensable for all cells, and accessory essential genes that are required for some organisms.
-Core essential genes can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003} and have been used for the synthesis of minimal cells \cite{hutchison_design_2016}.
-On the other hand, accessory essential genes are helpful in the study of specific lineages.
-Accessory essential genes may be useful in species-specific antibiotic discovery.
-Antibiotics that target core essential genes in pathogens may not be ideal, as they may attack homologous genes in their hosts or commensal bacteria.
-
-%Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
-Freed et al.
-\cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}, while some genes are only essential in \textit{Shigella flexneri}.
-These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and soxR-reducing complex, many of which are essential in \textit{Shigella flexneri} due to the absence of paralogs or other alternative systems that exist in \textit{Escherichia coli}.
-%The difference in essentiality stems from four reasons: the disruption of many of the genes in \textit{Shigella flexneri} 2a 2457T impede the growth, but these genes are not actually essential; the conditions were different in the two experiments done on \textit{Escherichia coli} K12 BW25113 and \textit{Shigella flexneri} 2a 2457T; many of the genes were not actually essential for growth, but essential for transposon insertion and antibiotic resistance; and
-
-%PPG: Reduce, summarise main findings of Canals & Barquist. e.g. Most essential genes are also essential in related species, yet there are some serovar-specific essential genes [GIVE SOME NUMBERS]... 
-Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium ATCC 14028, two isolates of \textit{Salmonella typhi} Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{Escherichia coli} K12 BW25113 and found that these cells share 268 essential genes.
-Nine genes are essential in \textit{Escherichia coli} and not essential in \textit{Salmonella}s, three of which can tolerate insertions only in some parts and one has distant paralogs in \textit{Salmonella}s that is missing in \textit{Escherichia coli}.
-For other genes, there are a few insertions in \textit{Salmonella}s, but the number is not small enough to call those genes essential.
-Moreover, 159 genes were almost essential in all \textit{Salmonella}s but not in \textit{Escherichia coli}.
-These include genes involved in replication and genes related to ribosome and its accessory proteins.
-The authors also found 26 genes that are under greater selection in \textit{Salmonella} Typhimurium compared to \textit{S.
-typhi} and 10 genes vice versa.
-Barquist et al.
-\cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{Salmonella} Typhi Ty2, \textit{Salmonella} Typhimurium SL1344, and \textit{Escherichia coli} K12 BW25113.
-%THIS IS IMPORTANT:
-These genomes share 228 essential genes which are mostly involved in cell division, transcription, translation, and fatty acid and peptidoglycan biosynthesis.
-Additionally, many of the serovar-specific essential genes in \textit{Salmonella}s are phage repressors.
-Another key difference between these two serovars is that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{Salmonella} Typhimurium and not essential in \textit{salmonella} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.
-
-Although there are some studies on differentiation of essentiality in different organisms, these studies usually include a few genomes.
-We are going to extend these studies to a larger scale in Enterobacteriaceae family.
-Our aim is to study the essentiality of genes in in 16 different organisms.
-These include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{Klebsiella pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{Salmonella} Typhimurium SL1344, \textit{Salmonella} Typhimurium SL3261, \textit{Salmonella} Typhimurium D23580, \textit{Salmonella} Typhimurium A130, \textit{Salmonella} Enteritidis P125109, \textit{Salmonella} Typhi Ty2, \textit{Escherichia coli} ST131 EC958, \textit{Escherichia coli} UPEC ST131, \textit{Escherichia coli} ETEC CS17, \textit{Escherichia coli} ETEC H10407, \textit{Escherichia coli} BW25113, and \textit{Escherichia coli} K-12 MG1655.
-
-%WHAT FATEMEH DID....
-
-%Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
-In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study.
-In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential.
-We first performed a detailed study of biases than can influence the inference of essentiality.
-Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome).
-We have also investigated how essentiality changes in the phylogenetic tree for these organisms.
-
-%\begin{figure}
-%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
-%\caption{The species tree containing the 16 genomes used in this study.
-We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
-%\label{fig:species-tree}
-%\end{figure}
-
-\section{Results and Discussion}
-\subsection{Incorporating multiple measures of gene essentiality improves prediction}
-In this section we will compare a number of methods of quantifying the essentiality of genes based on transposon insertion data and introduce our own method.
-\subsubsection{Comparison of old methods}
-A number of methods have been used for evaluating the essentiality of genes using transposon insertion data.
-Freed et al.
-\cite{freed_combining_2016} have compared eleven features that can quantify the essentiality of genes.
-These include: the number of insertion sites within genes; the mean distance between insertion sites, the median distance between insertion sites, the number of base pairs before the first insertion in the $5^\prime$ end, the ratio between the number of base pairs before the first insertion in the $5^\prime$ end and gene length, the number of base pairs before the second insertion in the $5^\prime$ end, the ratio between the number of base pairs before the second insertion in the $5^\prime$ end and gene length, the length of the longest uninterrupted region in the gene, the ratio between the length of the longest uninterrupted region in the gene and gene length, the length of the longest region in the gene containing at most one insertion divided by gene length, and the number of insertion sites divided by gene length.
-Among these, the average distance between insertion sites and the length of the longest uninterrupted region were the best predictors.
-
-Barquist et al.
-\cite{barquist_tradis_2016} have used insertion indices which are calculated by dividing the number of insertion sites by gene length for defining the essentiality of genes.
-Plotting all insertion indices for all genes in a genome gives a bimodal plot, each mode representing a group of genes (essential or non-essential).
-They have fitted two gamma distributions to these modes and calculated the log odds value for each gene to test which mode it belongs to.
-
-Turner et al.
-\cite{turner_essential_2015} have randomly sampled insertions in a genome and using DESeq package \cite{anders_differential_2010} calculated log fold changes and P-values for the actual number of insertions compared to the expected number of insertions obtained from the sampling method.
-
-We have compared the predictive power of the average distance between insertion sites in a gene, the length of the longest uninterrupted region, insertion index, log odds value obtained from insertion index value, and P-value and log fold change calculated using DESeq package after sampling.
-To evaluate the accuracy of these methods, we have compared the predicted essential genes to the essential genes in \textit{Escherichia coli} K-12 MG1655 in EcoGene database \cite{zhou_ecogene_2013}.
-The results are depicted in Figure~\ref{fig:benchmark}.
-Among all methods, those that try to fit a distribution to data and predict essentiality based on that, namely log odds value from insertion indices and P-value from DESeq after sampling, are the least accurate methods.
-Sampling and then calculating log fold changes using DESeq package is the most accurate predictor and the other three methods are also good predictors despite their simplicity.
-\subsubsection{PCA based method}\label{sec:essentiality-call}
-We selected four of the predictors in previous section: insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region.
-These predictors were selected as they were more accurate and less dependent on each other.
-Afterwards, we ran principal component analysis on these predictors using \emph{prcomp} function in R and selected the first principal component.
-As Figure~\ref{fig:benchmark} shows, PCA results gave us the most powerful tool for quantifying the essentiality of genes compared to other methods.
-We have plotted the results for one sample genome and the results for other genomes are shown in {\color{red}SUPPLEMENTARY}.
-Figure~\ref{fig:pca} shows the distribution of PCA results after zero mean unit variance normalisation.
-We call this value NPEQ (Normalised Pca based Essentiality Quantifier) in this paper.
-The genes whose NPEQ values are less than -1.644854 after normalisation are beneficial losses and the genes with NPEQ greater than 1.644854 are essential.
-The rest are non-essential genes.
-The cutoff 1.644854 has been selected because if we assume that NPEQ distribution is normal, this cutoff shows the P-value 0.95.
-Moreover, by trying to define a cutoff that maximises Matthews correlation coefficient for each genome using NPEQ, we get 1.650449 as the average cutoff which is very close to this value.
-
-\begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[page=1, scale=0.4]{essential-call-comparison-BW25113.pdf}
-  \caption{}
-  \label{fig:benchmark}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[page=10, scale=0.4]{essential-call-comparison-BW25113.pdf}
-  \caption{}
-  \label{fig:pca}
-\end{subfigure}
-\caption{(a) The accuracy of 7 different prediction methods for quantifying the essentiality of genes.
-The higher the area under the curve, the more accurate the method is.
-(b) The distribution for our proposed method (NPEQ)}
-\label{fig:essentiality-call}
-\end{figure}
-
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BN373.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BW25113.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-CS17.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-EC958.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ENC.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ERS227112.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ETEC.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-NCTC13441.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ROD.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SEN.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL1344.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL3261.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STM.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STMMW.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-t.pdf}\\
-%\end{tabular}
-%\caption{The plots show the accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is.}
-%\label{fig:benchmark}
-%\end{figure*}
-
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BN373.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BW25113.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-CS17.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-EC958.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ENC.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ERS227112.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ETEC.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-NCTC13441.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ROD.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SEN.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL1344.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL3261.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STM.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STMMW.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-t.pdf}\\
-%\end{tabular}
-%\caption{The plots show the distribution of our new prediction method which is calculated by running PCA on insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region and selecting first principal component.
-This method is called NPEQ.}
-%\label{fig:pca}
-%\end{figure*}
-\subsection{TraDIS data is biased}
-In Section~\ref{sec:essentiality-call} we proposed NPEQ method that quantifies the level of essentiality of a gene.
-However, if the transposon insertion is biased to specific regions in the genome, it can increase/decrease the values predicted by our method for genes and put them in a different level of essentiality.
-Different articles have reported biases in transposon insertion using Tn5 \cite{barquist_comparison_2013, green_insertion_2012, rubin_essential_2015, canals_high-throughput_2012, langridge_simultaneous_2009}.
-We performed a detailed study of these biases.
-The biases that we studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
-\subsubsection{Distance from the origin of replication bias}
-While a study has reported Tn5 insertion bias near the origin of replication \cite{barquist_comparison_2013}, another study has reported no bias \cite{rubin_essential_2015}.
-To study the bias towards the position of gene within genome, we plotted NPEQ for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Fig.\@ \ref{fig:distance-bias}.
-The figure indicates that NPEQ increases when the genes are located further from the origin of replication.
-When the bacteria are under replication during the transposn insertion process, there are more copies of the genes close to the origin of replication than the genes further away due to the initiation of different replication forks.
-This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
-%One might argue that the bias that we have observed may have happened due to the abundance of essential genes near the origin of replication.
-Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication.
-
-\begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[page=6, scale=0.25]{biases.pdf}
-  \caption{}
-  \label{fig:distance-bias}
-\end{subfigure}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[page=5, scale=0.25]{biases.pdf}
-  \caption{}
-  \label{fig:GC-bias}
-\end{subfigure}
-\begin{subfigure}{\textwidth}
-  \centering
-  \includegraphics[scale=0.6]{100logo-prob-edited.pdf}
-  \caption{}
-  \label{fig:logos}
-\end{subfigure}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[scale=0.25, page=2]{insertion-position-bias.pdf}
-  \caption{}
-  \label{fig:insertion-position-bias-ess}
-\end{subfigure}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[scale=0.25, page=4]{insertion-position-bias.pdf}
-  \caption{}
-  \label{fig:insertion-position-bias-ben}
-\end{subfigure}
-\caption{(a)The distance of the genes from DnaA gene normalised by the lengths of the genomes versus NPEQ.
-The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
-The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias.
-(b) G-C content of genes against their NPEQ values.
-The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias.
-(c) Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
-The height of each character shows the relative frequency of that character.
-(d) and (e) The plots show the average insertion index in percentiles of all essential genes (d) and beneficial losses (e).
-The genes are divided into 3 segments: 5\% of the genes on the $5^\prime$ end, 20\% of the genes on the $3^\prime$ end, and the rest in the middle.
-These are shown by khaki, slate gray, and violet red respectively.}
-\label{fig:biases}
-\end{figure}
-
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=2, scale=0.22]{biases.pdf}&
-%\includegraphics[page=6, scale=0.22]{biases.pdf}&
-%\includegraphics[page=10, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=14, scale=0.22]{biases.pdf}&
-%\includegraphics[page=18, scale=0.22]{biases.pdf}&
-%\includegraphics[page=22, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=26, scale=0.22]{biases.pdf}&
-%\includegraphics[page=30, scale=0.22]{biases.pdf}&
-%\includegraphics[page=34, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=38, scale=0.22]{biases.pdf}&
-%\includegraphics[page=42, scale=0.22]{biases.pdf}&
-%\includegraphics[page=46, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=50, scale=0.22]{biases.pdf}&
-%\includegraphics[page=54, scale=0.22]{biases.pdf}&
-%\includegraphics[page=58, scale=0.22]{biases.pdf}\\
-%\end{tabular}
-%\caption{The plots show the distance of the genes from DnaA gene normalised by the lengths of the genomes versus the insertion indices of the genes.
-The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
-The red curves show the loess curve when the smoothness parameter is $0.2$.}
-%\label{fig:distance-bias}
-%\end{figure*}
-\subsubsection{Preferred nucleotide composition bias}
-Another concern while inferring essentiality from transposon insertion data is that transposon insertion is biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposon insertion is biased towards, instead of their essentiality level.
-During Tn5 transposition, a sequence of 9 nucleotides is duplicated.
-Lodge et al.
-\cite{lodge_transposon_1988} showed that these duplicated regions have G-C pairs at two ends and replacing these G-C pairs with A-T pairs reduces the number of transposon insertions by more than fivefold.
-Goryshin et al.
-\cite{goryshin_tn5/is50_1998} reported the palindromic sequence A-GNTYWRANC-T as the consensus target site for Tn5 transposition where the 9 letters in middle show the consensus sequence in the duplicated region.
-%They have also showed that insertion sites are usually grouped together and have 5 base pairs distance from each other.
-In some other research a similar consensus motif has been found for Tn5 \cite{canals_high-throughput_2012} which is CGCGCA-GTTYWRAAC-TGCGCG.
-Others \cite{green_insertion_2012,rubin_essential_2015} have not found such a sequence but shown that the duplicated regions are G-C rich.
-We used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from duplicated regions and 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
-The results in Figure \ref{fig:logos} show the consensus motif that we have found is very similar to \cite{green_insertion_2012} and the only difference is in positions 3 and 7 within the duplicated region.
-
-%\begin{figure*}
-%\includegraphics[scale=0.9]{100logo-prob-edited.pdf}
-%\caption{Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
-The height of each character shows the relative frequency of that character.}
-%\label{fig:logos}
-%\end{figure*}
-
-The other possible source of bias is if transpositions are more inclined to G-C or A-T rich regions.
-Rubin et al.
-\cite{rubin_essential_2015} have reported that the number of Tn5 insertions rises with the increase of G-C content and Green et al.
-\cite{green_insertion_2012} have shown that the highest number of insertions occur in high G-C content regions.
-On the other hand, Langridge et al.
-\cite{langridge_simultaneous_2009} have seen an increase in the number of Tn5 insertions in 40\% G-C content.
-In Figure \ref{fig:GC-bias}, we plotted the G-C content of genes versus their NPEQ values.
-The red curves are loess curves with smoothness parameter 0.2 and the green line shows no bias.
-%In most cases (BW25113, EC958, ENC, ETEC, NCTC13441, ROD, SEN, STM, and t)
-NPEQ decreases gradually (which means an increase in the number of insertions) as G-C content decreases and then somewhere between 40\% and 50\% G-C content, NPEQ starts to rise again in almost all genomes.
-%In three cases (SL1344, SL3261, and STMMW) NPEQ loess curve starts to fall off again at about 60\% G-C content which is probably due to a small number of genes with low NPEQ values in that region.
-The loess curve for BN373 is almost flat.
-In CS17, the loess curve goes down until it reaches the minimum value (at about 48\%) and then remains unchanged and in ERS227112, the NPEQ value is low at the beginning and then it increases at around 50\% G-C content.
-In the region where most of the genes are packed, the loess curve is almost flat.
-On the left side of this flat region, there are genes with different G-C content which are enriched in mobile genetic elements.
-So, we expect to see more insertions (smaller NPEQ values) in this region.
-We have low NPEQ values between 40\% and 50\% G-C content which is expected.
-However, in most cases when we have less than 40\% G-C content, NPEQ value is high.
-A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions.
-This has been shown for Tn10 transposon \cite{kimura_nucleoid_2016}, but not for Tn5 transposon, yet.
-Overall, the results are consistent with Langridge et al.
-\cite{langridge_simultaneous_2009}.
-The G-C content of the most of the genes with large NPEQ values is between 50\% and 60\% which is inconsistent with Green et al.
-\cite{green_insertion_2012} as this region contains most of the genes and is not considered as a high G-C content region but rather an average G-C content region.
-%Even though most of the regression lines are consistent with Rubin et al.
-results \cite{rubin_essential_2015}, the figures show that a linear regression cannot predict G-C bias as its slope changes dramatically in different datasets.
-
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.23]{biases.pdf}&
-%\includegraphics[page=5, scale=0.23]{biases.pdf}&
-%\includegraphics[page=9, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=13, scale=0.23]{biases.pdf}&
-%\includegraphics[page=17, scale=0.23]{biases.pdf}&
-%\includegraphics[page=21, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=25, scale=0.23]{biases.pdf}&
-%\includegraphics[page=29, scale=0.23]{biases.pdf}&
-%\includegraphics[page=33, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=37, scale=0.23]{biases.pdf}&
-%\includegraphics[page=41, scale=0.23]{biases.pdf}&
-%\includegraphics[page=45, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=49, scale=0.23]{biases.pdf}&
-%\includegraphics[page=53, scale=0.23]{biases.pdf}&
-%\includegraphics[page=57, scale=0.23]{biases.pdf}\\
-%\end{tabular}
-%\caption{The plots show G-C content of genes against their NPEQ values.
-The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows the linear regression line.}
-%\label{fig:GC-bias}
-%\end{figure*}
-
-\subsubsection{Positional bias within genes}
-Some research has indicated that himar1 transposons are more probable to get inserted into the two ends of a gene compared to the middle \cite{griffin_high-resolution_2011}.
-We have tested this hypothesis using our TraDIS data.
-We divided every gene into 100 fragments with equal lengths (percentiles) and calculated the mean insertion index for each percentile.
-Insertion index is calculated using $\frac{\frac{n_p}{l_p}}{\frac{n_g}{l_g}}$, where $n_p$ is the number of insertion sites in a specific percentile, $l_p$ is the length of that percentile, $n_g$ is the number of insertion sites in the whole genome and $l_g$ is genome length.
-Mean insertion index for each percentile is calculated by averaging over all insertion indices for that specific percentile of genes.
-We saw almost no bias towards any location when considering all genes together (Fig.\@ \ref{fig:insertion-position-bias} {\color{red}SUPPLEMENTARY}).
-We studied the bias in three different groups of genes: essential genes which have no or just a few insertions, non-essential genes which have an intermediate number of insertions, and beneficial losses which have a high number of insertions.
-The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the $5^\prime$ and $3^\prime$ ends (Fig.\@ \ref{fig:insertion-position-bias-ess}) while it is the opposite in beneficial losses (Fig.\@ \ref{fig:insertion-position-bias-ben}).
-High number of insertions at the $3^\prime$ end of essential genes implies that the functional domains are located before the insertions and the insertions are not interfering with them.
-On the other hand, high number of insertions at the $5^\prime$ end of the essential genes indicates there might be alternative start codons in the $5^\prime$ end or it might be because of annotation errors that have predicted the start codon in an incorrect place before the actual start codon.
-
-%\begin{figure*}
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
-%\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
-%\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
-%\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
-%\end{tabular}
-%\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
-The genes are divided into 3 segments: 5\% of
-%the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
-%\label{fig:insertion-position-bias}
-%\end{figure*}
-\subsection{Most essential genes are ubiquitously essential in Enterobacteriaceae}
-Previous studies of gene essentiality in Enterobacteriaceae family have compared essential genes in different genomes in this family and studied the sets of core essential genes and accessory essential genes \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}.
-Core essential genes are responsible for essential processes such as cell division, DNA replication, transcription and translation and some important pathways like peptidoglycan and fatty acid biosynthesis \cite{barquist_comparison_2013}.
-Accessory essential genes differ in genomes due to different reasons such as niche adaptation, gene duplication and the existence of alternative pathways \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}.
-Another group of accessory essential genes are phage repressors \cite{barquist_comparison_2013}.
-Even though these genes are not essential for the growth of a cell, once phages are introduced to a cell, they become essential as long as the phage remains in the cell.
-
-In this study, we have compared the genes in 16 bacteria from Enterobacteriaceae family.
-For this we needed to study core and accessory sets of genes.
-Moreover, in the presence of two redundant variations of one gene, if we knock out one copy using TraDIS, the other copy compensates and the organism can still survive \cite{dean_pervasive_2008}.
-This leads to a different essentiality inference using TraDIS.
-Therefore, in addition to core and accessory genes, we have studied duplicate genes.
-To study whether each gene in the 16 organisms is core, we used Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to iteratively search for homologous proteins in our dataset and cluster them.
-We divided the clusters of homologous genes into three groups based on their conservation.
-Genus specific class contains genes that are present only in one genus, the genes in single copy class are present in more than one genus and more than 70\% of them are not duplicated (core genes), and  the genes in multi-copy class are present in more than one genus and more than 30\% of them are duplicated (duplicate genes).
-
-We also evaluated the essentiality of genes and divided the genes into different levels of essentiality using NPEQ values.
-If NPEQ is less than -1.644854, it means that the gene has tolerated many insertions, so it is beneficial for the organism to lose this gene in the rich medium that we used.
-However, if NPEQ is greater than 1.644854, the gene can tolerate very few or no insertions indicating that the gene is essential for cell viability in our test medium.
-Any other NPEQ value shows an intermediate number of insertions in genes meaning that it is not beneficial for the organism to lose these genes, but they are not essential, too.
-We have called this group of genes non-essential.
-
-The results for comparing three levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
-The high number of single copy clusters in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality.
-However, there are also many essential genus specific genes.
-The figure also shows that beneficial losses are over represented in genus specific class.
-Therefore, beneficial losses are mostly recent genes that the organism tends to lose in the long run.
-Besides, most of the multi-copy clusters are non-essential and there are only a few multi-copy clusters that are essential.
-This can be explained by the redundancy that duplicate genes can keep even after $\sim100$ million years \cite{dean_pervasive_2008}.
-
-\begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{cluster-essentiality.pdf}
-  \caption{}
-  \label{fig:iidist}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{essential-pathways.pdf}
-  \caption{}
-  \label{fig:essentiality-pathway-ess}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{non-essential-pathways.pdf}
-  \caption{}
-  \label{fig:essentiality-pathway-nes}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{beneficialloss-pval.pdf}
-  \caption{}
-  \label{fig:essentiality-pathway-ben}
-\end{subfigure}
-\caption{(a) The genes have been clustered into homologous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
-Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
-The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.
-(b) and (c) KEGG Pathways enriched in essential genes (b) and non-essential genes (c).
-(d) The words enriched in the description of beneficial losses in their embl files compared to other genes.
-The red line shows P-value = 0.05.
-The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-\label{fig:gene-classes}
-\end{figure}
-
-%\begin{figure}
-%\centering
-%\includegraphics[scale=0.5]{cluster-essentiality.pdf}
-%\caption{The genes have been clustered into homologous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
-Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
-The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
-%\label{fig:iidist}
-%\end{figure}
-
-To study which functions are enriched in each class of essentiality, we used KEGG pathway enrichment analysis \cite{kanehisa_kegg:_2000} and compared the genes in each class with the databases that were available for the genomes in this study.
-The results show that essential genes are enriched in pathways related to genetic information processing such as replication (DNA replication, homologous recombination and mismatch repair), transcription (RNA polymerase), translation (ribosome), and protein export.
-These are the essential functions that every cell needs for its viability.
-Other enriched pathways are mostly metabolism related.
-These include fatty acid biosynthesis that produces cell membrane, peptidoglycan and lipopolysaccharide biosynthesis that are essential components of cell wall, terpenoid backbone biosynthesis which feeds peptidoglycan biosynthesis, nucleotide and amino acid metabolism, and the metabolism of important cofactors and vitamins like riboflavin, biotin,porphyrin and chlorophyll (Figure~\ref{fig:essentiality-pathway-ess}).
-{\color{red} CAN WE HAVE SOMETHING LIKE THIS? http://www.nature.com/articles/srep00125}
-
-Non-essential genes are mostly involved in carbohydrate, energy, and lipid metabolism, membrane transport, cell motility and cellular community.
-Even though these functions are important, they are not essential in a rich medium in lab.
-
-As most of beneficial losses do not have homologs in other genomes, most of them are not studied and therefore there is no KEGG pathway for them.
-Because of this reason, we studied the description of these genes in their embl files and found the words that were enriched.
-These are shown in Figure~\ref{fig:essentiality-pathway-ben}.
-The results show that most of beneficial losses are mobile genetic elements like transposases and insertion elements which are not essential in their host genomes, genes that are not essential in a rich medium like stress protein and acid-resistance protein, and hypothetical proteins.
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{essential-pathways.pdf}&
-%\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
-%\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
-%\includegraphics[scale=0.4]{beneficialloss-pval.pdf}
-%\end{tabular}
-%\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
-The red line shows P-value = 0.05.
-The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-%\label{fig:essentiality-pathway}
-%\end{figure*}
-
-\subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
-When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}.
-As mutations in essential genes are less likely to be tolerable, less mutations should be fixed in these genes, and therefore essential genes should evolve more slowly than non-essential genes.
-This is known as knockout rate hypothesis \cite{krylov_gene_2003}.
-Different studies have verified this hypothesis in prokaryotes \cite{jordan_essential_2002, gerdes_experimental_2003, rocha_analysis_2004, silander_constancy_2009}.
-
-Here, we have tested this hypothesis by comparing the number of core essential genes and core genes in the phylogenetic tree as we go up to the root.
-We have used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes) to define if a gene is core essential or core at each level in the phylogenetic tree.
-If we get 1 at the top of the tree while studying essentiality of a gene using Fitch algorithm, it means that the gene is core essential and otherwise, the gene is non-essential.
-The same applies to the study of core genes.
-The phylogenetic tree has been annotated with the number of core essential genes (red) and the number of core genes (blue) at each level in Figure~\ref{fig:tree}.
-We then plotted the ratios at each level in the phylogenetic tree in Figure~\ref{fig:fitch} and connected the medians in each level.
-The connecting line shows that the ratio between core essential genes and core genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.
-
-\begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.6\textwidth}
-  \centering
-  \includegraphics[scale=0.12]{phylosift-aa-raxmlbootstrap-annotated.pdf}
-  \caption{}
-  \label{fig:tree}
-\end{subfigure}
-\begin{subfigure}{.4\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{fitch.pdf}
-  \caption{}
-  \label{fig:fitch}
-\end{subfigure}
-\caption{(a) The species tree for all genomes in this study.
-Numbers in red show the number of core essential genes at each level and numbers in blue show the number of core genes at each level.
-(b) The ratio between core essential genes and core genes at each level in the species tree.
-The dots in the strain level show the ratios for all 16 bacteria; the dots in the species level show the ratios for Enterobacter, Klebsiella, Citrobacter, Salmonella, and Escherichia; the subfam.
-level shows the ratios for the common ancestor of Enterobacter and klebsiella, the common ancestor of Citrobacter and Salmonella, and the common ancestor of Citrobacter, Salmonella, and Escherichia; and finally the dot in the family level shows the ratio for the root.
-The line connects the medians in each level.}
-\label{fig:essentiality-phylogeny}
-\end{figure}
-
-We have studied the essentiality status of genes involved in important biological processes in Table~\ref{tab:biological-processes}.
-These processes include cell division, DNA replication, transcription, translation, and important metabolic pathways such as peptidoglycan and fatty acid biosynthesis.
-If a gene was not core essential using Fitch algorithm, we looked at its essentiality in every genome: if it was essential in some of the genomes we classified it as accessory essential and if it was not essential in any genome we classified it as non-essential.
-FtsA and zipA genes are essential for cell division.
-FtsA is essential in all the genomes in our study, however zipA is essential everywhere except for Salmonella typhimurium A130 in which this gene is near essential.
-We investigated if this gene has a homolog in Salmonella typhimurium A130 and found no homologs.
-It has been shown that 21 mutations in ftsA gene can make it independent of zipA gene so that the zipA gene is no longer essential \cite{geissler_gain--function_2003, pichoff_ftsa_2012}.
-However, ftsA gene does not have any of these mutations in Salmonella typhimurium A130, so there might be another factor that is causing zipA not to be essential.
-\textcolor{red}{ADD MORE EXAMPLES HERE}
-
-So far we have shown that essential genes are less likely to be lost during evolution.
-The other question is if essential genes tend to keep their essentiality in different genomes.
-Bergmiller et al.
-\cite{bergmiller_patterns_2012} have shown that only essential genes whose functions cannot be taken over by other genes keep their essentiality and others are possible to be lost or non-essential in other genomes.
-The genes that can replace other genes are not necessarily homologous to them, but they have similar functions.
-
-We have compared the number of genes that are shared between our bacterial genomes (\ref{fig:upsetr-conservation}) and the number of genes essential in them (\ref{fig:upsetr-essentiality}) and used UpSetR package \cite{conway_upsetr:_2016} to visualise the results in Fig.\@ \ref{fig:upsetr}.
-As shown in the figures, among 2162 genes that are shared between all the bacteria under study, only 135 are essential everywhere.
-We looked at subsets of the genes that were shared in every combination of our bacterial strains to see whether they show a trend similar to the phylogenetic tree.
-The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show such a trend.
-The reason is that many of the essential genes that are not shared in all genomes are probably compensable by other genes and so the essentiality does not look tree-like.
-
-\begin{landscape}
-\begin{table}
-\small
-\caption{The essentiality status of genes involved in important biological processes}
-\label{tab:biological-processes}
-\begin{tabular}[t]{l l l l l}
-\hline
-\textbf{Biological process} & \textbf{Subprocess} & \textbf{Core essential} & \textbf{Accessory essential} & \textbf{Non-essential}\\
-\hline
-Cell Division & & \begin{tabular}[t]{@{}l@{}}ftsAHLQWYZ, minE,\\ mukB, zipA \end{tabular}& ftsKNX, minD & \begin{tabular}[t]{@{}l@{}}CedA, ftsJ, minC,\\ sdiA, sulA\end{tabular}\\
-DNA replication & Polymerases I, II, III & dnaENQX, holABD, polA & holC & holE, polB\\	
-& Supercoiling & gyrAB, parCE & & \\
-& Primosome-associated & dnaBCGT, priA, ssb & priB, rep & priC\\
-Transcription & RNA polymerase & rpoABC &  & \\
-& \begin{tabular}[t]{@{}l@{}}Sigma, elongation, anti- \\ and termination factors\end{tabular} & nusABG, rho, rpoDH & rpoEN & rpoS\\	
-Translation & tRNA-synthetases & \begin{tabular}[t]{@{}l@{}}alaS, argS, asnS, aspS, cysS, \\ glnS, gltX, glyQS, hisS, ileS, \\ leuS, lysS, metG, pheST, \\ proS, serS, thrS, tyrS, valS \end{tabular} & trpS	 & \\
-& Ribosome components & \begin{tabular}[t]{@{}l@{}}rplBCDEFJKLMNOPQRSTUV,\\ rplWXY, rpmABCD,\\ rpsABCDEFGHJKLMNPQRSTU\end{tabular} & rplA, rpmEGHI, rpsIO & rplI, rpmF\\
-& \begin{tabular}[t]{@{}l@{}}Initiation, elongation \\ and peptide chain \\ release factors\end{tabular} & fusA, infABC, prfAB, tsf & efp & prfC, selB, tufAB \\
-Biosynthetic pathways & & & & \\
-Peptidoglycan & & MraY, murABCDEFGI & & ddlAB\\	
-Fatty acids & & accABCD, fabABDGHIZ & & \\
-\hline
-\end{tabular}
-\end{table}
-\end{landscape}
-
-\begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.45,page=2]{upsetr.pdf}
-  \caption{}
-  \label{fig:upsetr-conservation}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.45,page=3]{upsetr.pdf}
-  \caption{}
-  \label{fig:upsetr-essentiality}
-\end{subfigure}
-\caption{(a) The number of genes and (b) essential genes shared between different groups of bacteria.
-The bars show the number of genes and the filled circles show which bacteria are sharing those genes.}
-\label{fig:upsetr}
-\end{figure}
-
-%\section{Discussion}
-\section{Materials and Methods}
-%\subsection{Transposon insertion}
-%\section{Results and discussion}
-%Throughout time, species can gain or lose genes.
-We investigated if these gene gain and losses are related to the essentiality of the genes.
-In this section, we have first described the biases that can affect our study, and then evaluated the essentiality of genes and their conservation and the relationship between these two.
-%
-%
-%%\subsection{Length bias}
-%%Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions.
-As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes.
-%%In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index in short genes.
-It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
-%%\begin{figure*}
-%%\centering
-%%\includegraphics[page=42, scale=0.3]{biases.pdf}
-%%\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
-%%\label{fig:length-bias}
-%%\end{figure*}
-%
-%
-%\subsubsection{Preferred insertion motif bias}
-
-%
-%\begin{itemize}
-%\item model H-NS binding sites? CGWTWHWww Lang et al (2007)
-%\item seems unlikely -- show bulk of genes are around 50\% G+C (add box-whisker plots to scatter diagrams?)
-%\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? It is not mentioned in the paper.
-%\end{itemize}
-%
-%\begin{figure*}
-%\includegraphics[scale=0.55]{lowgc-pval.pdf}
-%\caption{Word enrichment analysis for low G-C genes compared to genes with interquartile G-C level.
-The red line shows P-value = 0.05.
-The P-values have been calculated using Fisher's exact test and corrected using Benjamini-Hochberg-Yekutieli.}
-%\label{fig:gc-pval}
-%\end{figure*}
-%
-
-%\subsection{Essentiality and conservation}
-%Essential genes are needed for the growth of organisms.
-Because of that, one might think that essential genes should not be lost in a short period of time throughout evolution, unless they are no longer needed in new organisms or they are replaced by new pathways.
-Therefore, it is expected that most of the essential genes are conserved in different organisms from the same family.
-We have tested this idea by comparing the essentiality and conservation of genes in Enterobacteriaceae family.
-%
-%
-%\subsubsection{Gene classes}
-%%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
-Plenty of methods are proposed for this purpose.
-Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
-We have used Hieranoid to cluster the sets of orthologous genes.
-In addition, we intended to study the essentiality of genes in paralogous genes.
-For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
-%In order to study the relationship between essentiality and conservation, we needed to evaluate the essentiality and conservation of genes.
-For this, we divided the genes into different levels of essentiality (essential genes, ambiguous, non-essential genes, beneficial losses).
-We first normalised the biases that exist in the data using the method explained in Section~\ref{sec:biascorrection} and then defined the essentiality level of genes using the method explained in Section~\ref{sec:essentiality}.
-Moreover, we grouped the genes into different classes of conservation (genus specific, single copy, multi-copy) using the method explaind in Section~\ref{sec:conservationclasses}.
-%
-%The results for comparing four levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
-The high number of single copy genes in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality.
-However, the relatively high number of essential genes in genus specific class implies that each genus has a set of essential genes that makes it distinct.
-The figure also shows that beneficial losses are over represented in genus specific class.
-Therefore, beneficial losses are mostly recent genes that the organism tends to lose in the long run.
-Besides, most of the multi-copy clusters are non-essential and there are only a few multi-copy clusters that are essential.
-This can be explained by the redundancy that genes can keep even after $\sim100$ million years \cite{dean_pervasive_2008}.
-In the presence of two redundant variations of one gene, if we knock out one copy by transposon mutagenesis, the other copy compensates and the organism can still survive.
-%
-%\begin{figure}
-%\centering
-%\includegraphics[scale=0.5]{cluster-essentiality.pdf}
-%\caption{The genes have been clustered into orthologous groups using Hieranoid and paralogous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
-Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
-The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
-%\label{fig:iidist}
-%\end{figure}
-%
-%To study which functions are enriched in each class of essentiality, we used the word enrichment method explained in Section~\ref{sec:enrichment}.
-Fig.\@ \ref{fig:essentiality-pval} shows the top 20 enriched words for each essentiality class.
-The results show an enrichment of the genes related to replication, transcription, translation, division, and rod shape determining proteins in essential class.
-The non-essential genes are mostly membrane associated proteins, flagellar proteins, ATPase, and DNA repair proteins.
-Beneficial losses are enriched in transposase enzymes, putative and hypothetical proteins, and mobile elements.
-Beneficial losses also contain many fimbrial proteins which probably has occurred because these proteins are not needed in a rich lab medium \textcolor{red}{\{TRUE?\}}.
-%%We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}.
-It shows that most of the beneficial losses are mobile elements.
-Most of these mobile elements are not included in KEGG database.
-%
-%We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}.
-The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes.
-However, as mobile genetic elements are not stored in KEGG database \cite{kanehisa_kegg:_2000}, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{essential-pval.pdf}&
-%\includegraphics[scale=0.4]{non-essential-pval.pdf}\\
-%\includegraphics[scale=0.4]{beneficialloss-pval.pdf}&
-%\end{tabular}
-%\caption{Word enrichment analysis for essential genes, non-essential genes, and beneficial losses compared to other genes.
-The red line shows P-value = 0.05.
-The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-%\label{fig:essentiality-pval}
-%\end{figure*}
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{essential-pathways.pdf}&
-%\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
-%\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
-%\end{tabular}
-%\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
-The red line shows P-value = 0.05.
-The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-%\label{fig:essentiality-pathway}
-%\end{figure*}
-%
-%\subsubsection{The evolution of essentiality}
-%%In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
-If a gene is essential in a genome, its corresponding cell in the matrix has value 1 and 0 otherwise.
-Then we have used the Bray-Curtis distance to generate a distance matrix for the species under study.
-Finally, a tree has been generated using PHYLIP Neighbor.
-The resulting tree is not similar to the species tree.
-%%
-%%\begin{figure}
-%%\centering
-%%\includegraphics[scale=0.2]{neighbor-joining-essentiality-tree.pdf}
-%%\caption{The tree is generated from all the genes that are copied only once per genome.
-We have made a binary matrix from the essentiality of these genes.
-Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree using PHYLIP Neighbor which uses a neighbour joining method.}
-%%\label{fig:neighborjoining}
-%%\end{figure}
-%
-%In this section, we compared the number of genes that were essential and conserved at each level of the phylogenetic tree.
-We were interested to see if the trend of the essentiality variation between different organisms is in agreement with the phylogenetic tree.
-In other words, we have tested if organisms close together have more essential genes in common than organisms that have separated earlier in the phylogenetic tree.
-%
-%We needed to cluster sets of orthologous genes in the bacteria that we were studying.
-Homologous clusters introduced in ~\ref{sec:conservationclasses} were not useful for this purpose as sets of paralogous genes with different essentiality levels can make essentiality inference ambiguous.
-Plenty of methods are proposed for this purpose.
-Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
-We used Hieranoid for clustering orthologous genes.
-Hieranoid need a species tree for clustering genes.
-We used the method explained in Section~\ref{sec:phylogenetic} to generate this tree.
-%
-%In order to test if the essentiality of genes follows a tree-like trend, we compared the number of genes that were conserved in different bacteria in our study and the number of genes that were essential in these bacteria.
-The genes that are conserved in all bacteria being studied are called core genes and the genes that are conserved and essential are called core essential genes.
-We counted the number of genes that were core between every combination of bacteria, and the number of core essential genes between those combinations.
-We used UpSetR package \cite{conway_upsetr:_2016} in R to visualise the results in Fig.\@ \ref{fig:upsetr}.
-As shown in the figures, among 1908 genes that are core between all the bacteria under study, only 184 are core essential.
-We looked at subsets of the genes that were core essential (core) in every combination of our bacterial strains to see whether they were phylogenetically informative or not.
-A phylogenetically informative subset is a subset that is core essential (core) in two or more bacteria but not in all bacteria.
-We have marked the phylogenetically informative sets of genes with ticks and the uninformative ones with crosses.
-The results propose that although conservation of genes follows a tree-like trend with many phylogenetically informative sets of genes with high cardinality, the essentiality does not show a tree-like signal and most of the large sets of core essential genes belong to only one bacteria.
-We believe this is due to the small number of essential genes.
-Each bacterium has about 300 to 500 essential genes among which 184 is core essential between all bacteria.
-A portion of the remaining essential genes are specific to each bacterium and many are shared between all bacteria except one.
-The number of remaining essential genes is so low that causes the essentiality trend not to be tree-like.
-Furthermore, some of the predicted essential genes might be artefacts of transposon mutagenesis method.
-%
-%\begin{figure*}
-%\centering
-%\includegraphics[scale=0.6, page=2]{upsetr-maximiseMCC.pdf}\\
-%\includegraphics[scale=0.6, page=3]{upsetr-maximiseMCC.pdf}
-%\caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.
-The bars show the number of genes that are core between the strains marked with black circles.
-The tick marks show phylogenetically informative columns and the cross marks show non-informative columns.}
-%\label{fig:upsetr}
-%\end{figure*}
-%
-%To further study the trend of essentiality changes, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
-We used three different methods for inferring core genes and core essential genes.
-These methods are explained in Section~\ref{sec:core}.
-The numbers predicted using these three methods are shown in Fig.\@ \ref{fig:annotated-speciestree}.
-As this figure shows, the ratio between core essential and core genes is almost constant using the intersection and dollo method; however, this ratio increases as we go higher in the tree using the ancestral insertion index method.
-%
-%%\begin{figure}
-%%\centering
-%%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap-annotated.pdf}
-%%\caption{The tree shows the species tree in Fig.\@\ref{fig:species-tree} annotated by the number of core essential genes and core genes at each node.
-We used three methods to define core essential genes and core genes.
-The numbers at the leaves are the same using all these three methods.
-At the internal nodes, red shows the numbers using the intersection method, green shows the ancestral insertion index method, and turquoise shows fdollop method.}
-%%\label{fig:annotated-speciestree}
-%%\end{figure}
-%
-%As these three methods lead to conflicting results, we compared the differences between the genes found in these three methods.
-For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods.
-Then, we performed word enrichment analysis explained in Section~\ref{sec:enrichment} on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method.
-Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure.
-The results are depicted in Fig.\@\ref{fig:method-difference}.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{ancestral-intersection-pval.pdf}&
-%\includegraphics[scale=0.4]{dollo-ancestral-pval.pdf}\\
-%\includegraphics[scale=0.4]{dollo-intersection-pval.pdf}&
-%\end{tabular}
-%\caption{The figure shows the difference between core essential genes in intersection and ancestral insertion index methods, ancestral insertion index and fdollop methods, and intersection and fdollop methods.
-The red line shows P-value = 0.05.
-The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure.
-The top left figure shows words that are enriched in core essential genes found using ancestral insertion index method but not enriched in core essential genes found using intersection method.
-The top right figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using ancestral insertion index method.
-The bottom left figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using intersection method.}
-%\label{fig:method-difference}
-%\end{figure*}
-%
-%\section{Materials and Methods}
-%\subsection{Transposon mutagenesis}\label{sec:transposoninsertion}
-%We studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains (Fig.\@ \ref{fig:species-tree}) and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
-These strains are all selected from Enterobacteriaceae family and a transposon mutagenesis study have been performed on them.
-We generated single inserted mutants using Tn5 transposon and placed the mutants in a selective media for Tn5.
-Then, we picked the mutants and pooled them and performed PCR enrichment using the method described in \cite{barquist_tradis_2016}.
-We sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
-The number of insertions in a gene implies the degree of essentiality for that gene.
-%
-%\subsection{Generating the species tree}\label{sec:phylogenetic}
-%We first ran search and align commands from PhyloSift package \cite{darling_phylosift:_2014} to select gene markers for generating a phylogenetic tree and aligning them.
-Then, we concatenated the protein alignments for all 14 genomes and ran RaxML \cite{stamatakis_raxml_2014} with "PROTGAMMALG4M" amino acid substitution model, "a" algorithm and 100 alternative runs on distinct starting trees.
-%
-%\subsection{Essentiality levels}\label{sec:essentiality}
-%The more transposon insertions we observe in a gene after sequencing the genomes, the less essential the gene is.
-In order to quantify the essentiality of genes, we used a measure named insertion index which is proportional to the number of insertions in a gene.
-%
-%To calculate the insertion index for each gene, we summed up the number of transposon insertion sites observed in that gene.
-Since the lengths of the genes are different, the insertion indices were then normalised by dividing them by gene length.
-Our experiment is performed on different strains and the library density is different in each experiment.
-Therefore, in order to make the insertion indices comparable in all the strains, we normalised the insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
-%
-%Based on insertion indices, the genes were divided into four groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
-We utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
-The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
-A loess curve was fitted to this distribution to find where the two peaks separate from each other.
-The first peak shows the genes with no or just a few insertions which are considered as essential genes.
-We fitted an exponential distribution to the first peak and a gamma distribution to the second one.
-Then, we calculated the log odds ratio for belonging to each of these distributions for each gene.
-The region that has log odds value between -2 and 2 is called the ambiguous region, the genes belonging to the first peak are essential and the rest of the genes are not essential.
-Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=2, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=3, scale=0.25]{essentiality.pdf}\\
-%\includegraphics[page=4, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=5, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=6, scale=0.25]{essentiality.pdf}\\
-%\includegraphics[page=7, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=8, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=9, scale=0.25]{essentiality.pdf}\\
-%\includegraphics[page=10, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=11, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=12, scale=0.25]{essentiality.pdf}\\
-%&\includegraphics[page=13, scale=0.25]{essentiality.pdf}&\\
-%\end{tabular}
-%\caption{Plots show the insertion index distribution for each genome.
-The plots are divided into 4 regions using red lines.
-These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-%\label{fig:iidist-species}
-%\end{figure*}
-%
-%\subsection{Bias correction}\label{sec:biascorrection}
-%We observed a distance from origin of replication bias and also a positional bias within genes.
-Hence, these biases needed to be corrected before inferring the essentiality of genes from their insertion indices.
-We did not include genes shorter than 100 base-pairs in our study as they might not be targeted by any transposon due to their shortness.
-%
-%To overcome the distance from origin of replication bias, we divided the value of insertion index for a gene in a specific position by the predicted value by loess for that position.
-This value was then multiplied by the average insertion index.
-To overcome the positional bias within genes, we calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes.
-The insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=2, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=3, scale=0.25]{essentiality-normalised.pdf}\\
-%\includegraphics[page=4, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=5, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=6, scale=0.25]{essentiality-normalised.pdf}\\
-%\includegraphics[page=7, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=8, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=9, scale=0.25]{essentiality-normalised.pdf}\\
-%\includegraphics[page=10, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=11, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=12, scale=0.25]{essentiality-normalised.pdf}\\
-%&\includegraphics[page=13, scale=0.25]{essentiality-normalised.pdf}&\\
-%\end{tabular}
-%\caption{Plots show the insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes.
-The plots are divided into 4 regions using red lines.
-These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-%\label{fig:iidist-species-normalised}
-%\end{figure*}
-%
-%\subsection{Conservation classes}\label{sec:conservationclasses}
-%To study whether each gene in the 14 organisms is conserved we proposed a program that clusters homologous proteins.
-This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
-It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
-Then, it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
-In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by running Jackhmmer with more permissive parameters.
-Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
-The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 14 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
-%
-%\begin{figure}
-%\includegraphics[scale=0.2]{homclust.pdf}
-%\caption{The steps of our proposed algorithm for clustering homologous genes.}
-%\label{fig:homclust}
-%\end{figure}
-%
-%We divided the clusters of homologous genes into three groups based on their conservation.
-Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
-These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
-%
-%\begin{figure}
-%\includegraphics[scale=0.3]{cluster-size-dist.pdf}
-%\caption{Size distribution for all clusters of homologous genes.
-Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
-%\label{fig:homclust-results}
-%\end{figure}
-%
-%\subsection{Word and pathway enrichment analysis}\label{sec:enrichment}
-%To study the enrichment of words in a group of genes compared to a second group, We gathered the descriptions of genes from their embl files.
-Then, we counted the repeat number of each word for the genes in each group and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
-The P-values were then corrected using Benjamini-Hochberg-Yekutieli procedure.
-%
-%For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database \cite{kanehisa_kegg:_2000}.
-This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.
-Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class.
-Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.
-%
-%\subsection{Defining core and core essential genes}\label{sec:core}
-%We used three different methods for defining core and core essential genes: intersection, ancestral insertion index, and Dollo law. These three methods are explained in what follows.
-%
-%The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
-%
-%The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
-In this method, we averaged over the insertion indices of the pair of closest children of the ancestral node.
-We repeated this and averaged the averages until we reached the ancestral node.
-Then, we plotted the insertion indices and fitted an exponential and a gamma distribution to the plot as described in Section~\ref{sec:essentiality} and found the essential genes at that level.
-%
-%The third method is using Dollo law to define core genes and core essential genes.
-This method, assumes that the gain of genes (gain of essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (gain of essentiality) and minimise the number of times that a gene (the essentiality of a gene) has been lost.
-Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
-%
-%\section{Conclusion}
-%In this paper, we studied the relationship between the essentiality and conservation of genes in 14 bacteria from Enterobacteriaceae family.
-We first studied the biases that can affect our results and showed that transposon insertions are more abundant near the origin of replication.
-In addition, there is a slight preferred insertion motif bias and a G-C bias in A-T rich genes.
-Moreover we showed that transposon insertions are more abundant near the ends of essential genes compared to the internal region, while it is the opposite in beneficial losses.
-%
-%After correcting biases, we studied the essentiality versus conservation by dividing the genes into three classes of conservation: genus specific, single copy, and multi-copy, dividing them into four levels of essentiality: essential, ambiguous, non-essential, and benefical losses.
-We found that essential genes are mostly single copy, however there is a considerable number of genus specific genes in essential level which help to distinguish between genera.
-Furthermore, beneficial losses are mostly genus specific which means they are mostly new genes.
-The other finding was that multi-copy genes are mostly non-essential which can happen due to redundancy.
-%
-%We also found that the pattern of essentiality changes is not in agreement with the phylogenetic tree due to the small number of essential genes.
-Also the ratio between core essential genes and core genes can be either constant or increasing depending on the method that we use for defining core and core essential genes.
-\textcolor{red}{\{WHAT?\}}
-%
-%Overall, we have found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.
-%
-%\section*{Acknowledgments}
-%%
-%%
-%\nolinenumbers
-
-%This is where your bibliography is generated. Make sure that your .bib file is actually called library.bib
 \bibliography{references}
@@
 \bibliographystyle{apalike}
-
 \end{document}

### 2017-06-20 bin/clust2plot.R

@@ -80,6 +80,11 @@ for (cpitem in clusters_path)
   belthr <- max(insertion_index[res$cluster==nes])
   essthr <- max(insertion_index[res$cluster==ess])
   nesthr <- min(insertion_index[res$cluster==nes])
+  
+  bel <- insertion_index[insertion_index>belthr]
+  res2 <- dbscan(as.matrix(bel), minPts = 100, eps = 0.1)
+  ambig <- res2$cluster[which.min(bel)]
+  belam <- max(bel[res2$cluster==ambig])
   #insertion_index=(insertion_index-mean(insertion_index))/sd(insertion_index-mean(insertion_index))
   
   # if (cpitem == clusters_path[1])
@@ -95,9 +100,9 @@ for (cpitem in clusters_path)
   par(mar = mar.default + c(0, 1, 0, 0))
   
   h <- hist(insertion_index, breaks =seq(min(insertion_index),max(insertion_index)+1,0.02), plot=FALSE)
-  cuts <- cut(h$breaks, c(-Inf,essthr, nesthr, belthr, Inf))
+  cuts <- cut(h$breaks, c(-Inf,essthr, nesthr, belthr, belam, Inf))
   screen(1)
-  plot(h, col=c("darkgoldenrod4", "black", "turquoise4", "darkmagenta")[cuts], xlab = "Insertion index", main ="All gene classes", cex.lab = 2,
+  plot(h, col=c("darkgoldenrod4", "black", "turquoise4", "black", "darkmagenta")[cuts], xlab = "Insertion index", main ="All gene classes", cex.lab = 2,
        cex.axis = 1.5, cex.main = 2, xlim=c(0,4), ylim=c(0,300), lty= "blank", axes=FALSE)
   axis(1, at=seq(0,4,1), cex.axis=1.5)
   axis(2, at=seq(0,300,100), labels=c(0,NA,NA,300), cex.axis=1.5)
@@ -154,38 +159,38 @@ for (cpitem in clusters_path)
   }
   
   h <- hist(orfans, breaks =seq(min(insertion_index),(max(insertion_index)+1),0.02), plot = FALSE)
-  cuts <- cut(h$breaks, c(-Inf,essthr, nesthr, belthr, Inf))
[…]

### 2017-06-20 bin/pathanalysis-coregenes.R

@@ -1,0 +1,99 @@
+library(ggplot2)
+library(stringr)
+dbs <- "~/EnTrI/results/KEGG/"
+essentiality <- "~/EnTrI/results/define-core-accessory-hieranoid/all/core-essential-genomes/"
+# seqdb <- "~/EnTrI/data/fasta-protein/chromosome/seqdb.fasta"
+list_of_files <- list.files(path=dbs, full.names=T, recursive=FALSE)
+db=matrix(,nrow=0,ncol=3)
+for (filename in list_of_files)
+{
+  dbfile = as.matrix(read.table(filename, as.is=TRUE, header=TRUE, sep="\t"))
+  for (i in seq(1,nrow(dbfile)))
+  {
+    dbfile[i,3] = str_match(dbfile[i,3], '([[:print:]]+)-[[ Citrobacter rodentium]]|[[ Enterobacter cloacae subsp. cloacae NCTC 9394]]|
+                            [[ Escherichia coli O78:H11:K80 H10407 (ETEC)]]|[[ Escherichia coli K\\-12 MG1655]]|
+                            [[ Salmonella enterica subsp. enterica serovar Enteritidis P125109]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhimurium D23580]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhimurium SL1344]]|
+                            [[ Salmonella enterica subsp. enterica serovar Typhi Ty2]]|
+                            [[ Escherichia coli O25b:K100:H4-ST131 EC958 (UPEC)]]')[2]
+    
+  }
+  db = rbind(db,dbfile)
+}
+list_of_files <- list.files(path=essentiality, full.names=T, recursive=FALSE)
+
+genes = c()
+for (filename in list_of_files)
+{
+  gene_names = str_match(readLines(filename)[startsWith(readLines(filename),'>')],'>([[:graph:]]+)[[:blank:]]')[,2]
+  genes <- c(genes, gene_names)
+}
+
+# allgenes <- str_match(readLines(seqdb)[startsWith(readLines(seqdb),'>')],'>([[:graph:]]+)[[:blank:]]')[,2]
+
+enrichment = data.frame(matrix(nrow=length(unique(db[,3])),ncol=4),row.names =unique(db[,3]) )
[…]

### 2017-06-20 bin/study-beneficial-losses.R

@@ -1,6 +1,11 @@
 library(ggplot2)
-codons <- "../results/word-enrichment/beneficialloss-info.txt"
-codon_table <- read.table(codons, header = TRUE)
+library(stringr)
+codons <- "../results/word-enrichment2/beneficialloss-info.txt"
+# db <- "../results/eggnog-mapper/gproNOG.annotations.tsv"
+codon_table <- read.table(codons, header = TRUE, sep = '\t', quote="")
+# eggnogdf <- read.table(db, header = FALSE, sep='\t', quote = "")
+# eggnogs <- eggnogdf[,6]
+# names(eggnogs) <- eggnogdf[,2]
 fishertable =data.frame()
 for (i in seq(1,(nrow(codon_table)-1)))
 {
@@ -11,22 +16,28 @@ for (i in seq(1,(nrow(codon_table)-1)))
 }
 fishertable[,3] = p.adjust(fishertable[,2], method = "BH")
 fishertable = fishertable[order(fishertable[,3], fishertable[,2]),]
-write.table(fishertable, file = "../results/word-enrichment/beneficialloss-pval.txt", quote = FALSE, col.names = c("word", "p-value", "corrected-p-value"), row.names = FALSE, sep="\t")
+write.table(fishertable, file = "../results/word-enrichment2/beneficialloss-pval.txt", quote = FALSE, col.names = c("word", "p-value", "corrected-p-value"), row.names = FALSE, sep="\t")
 
-df = fishertable[-seq(25,nrow(fishertable),1),]
+# df = fishertable[-seq(25,nrow(fishertable),1),]
+df = fishertable#[!is.na(fishertable$V1),]
 df[,3] = -log10(df[,3])
-df = df[,-2]
+df$V1 <- str_match(df$V1, '([[:graph:]]+([[:blank:]]+[[:graph:]]+){0,6})')[,1]
+df = df[seq(1,30),-2]
 # df = df[c(-2,-5,-9,-13),]
 names(df) = c("word", "pvalue")
+df$word[1] <- 'Not classified'
+df$word[2] <- 'pct identical to transposase Yersinia pestis'
+df$word[6] <- 'Inhibits RpoS proteolysis'
+df$word[20] <- 'Influences nickel and cobalt homeostasis'
+# df$word[30] <- 'Catalyzes the dephosphorylation of heptose(II)'
 pdf("../figures/beneficialloss-pval.pdf")
[…]

### 2017-12-24 bin/entero-deg-endo-heatmap-3.R
Removed DEG studies that did not contain essential genes in the sense that we are defining them

@@ -11,21 +11,19 @@ Heatmap(toplot, cluster_rows = FALSE, cluster_columns = FALSE, rect_gp = gpar(co
 dev.off()
 
 dict <- c('Mycobacterium tuberculosis H37Rv', 'Synechococcus elongatus PCC 7942', 'Porphyromonas gingivalis ATCC 33277',
-          'Bacteroides thetaiotaomicron VPI-5482', 'Bacteroides fragilis 638R', 'Haemophilus influenzae Rd KW20',
-          'Escherichia coli ST131 strain EC958', 'Escherichia coli MG1655', 'Salmonella enterica serovar Typhi Ty2',
-          'Salmonella enterica serovar Typhi', 'Salmonella enterica serovar Typhimurium SL1344', 'Salmonella typhimurium LT2',
-          'Salmonella enterica subsp. enterica serovar Typhimurium str. 14028S', 'Vibrio cholerae N16961', 'Shewanella oneidensis MR-1',
-          'Pseudomonas aeruginosa PAO1', 'Pseudomonas aeruginosa UCBPP-PA14', 'Acinetobacter baumannii ATCC 17978', 'Acinetobacter baylyi ADP1',
-          'Francisella novicida U112', 'Burkholderia thailandensis E264', 'Burkholderia pseudomallei K96243', 'Sphingomonas wittichii RW1',
-          'Rhodopseudomonas palustris CGA009', 'Agrobacterium fabrum str. C58', 'Caulobacter crescentus',
-          'Brevundimonas subvibrioides ATCC 15264', 'Campylobacter jejuni subsp. jejuni NCTC 11168 = ATCC 700819', 'Helicobacter pylori 26695',
-          'Mycoplasma genitalium G37', 'Mycoplasma pulmonis UAB CTIP', 'Staphylococcus aureus NCTC 8325', 'Staphylococcus aureus N315',
-          'Bacillus subtilis 168', 'Bacillus thuringiensis BMB171', 'Streptococcus agalactiae A909', 'Streptococcus pyogenes NZ131',
-          'Streptococcus pyogenes MGAS5448', 'Streptococcus sanguinis', 'Streptococcus pneumoniae')
[…]

### 2018-07-06 documents/manuscript.tex
Added an ambiguous region between non-essential and beneficial loss, used eggNOG to study beneficial losses, generated a table containing all KEGG pathways with core essential and ancestrally essential genes
Added not-analysed to DEG analysis

--- 2017-02-27 documents/manuscript.tex
@@
 \usepackage[right]{lineno}
+\linenumbers
 
@@
 \usepackage{pdflscape}
-\graphicspath{{../figures/}}
+\graphicspath{{../figures/panels/}}
 
@@
 \usepackage{amsmath}
+\usepackage{pdfpages}
 
@@
 {\Large
-\textbf\newline{Title}
+\textbf{Investigating essential genes in Enterobacteriaceae using
+transposon insertion sequencing}
 }
@@
 
-\end{flushleft}
-
-\section*{Abstract}
-%Many genes have been identified with advances in sequencing technology and genome annotation methods.
-However, not all of these genes are of the same importance.
-We have used transposon mutagenesis to investigate gene essentiality in 14 strains of Enterobacteriaceae.
-We investigated the potential biases of this approach and found an origin of replication bis, no preferred insertion motif bias, a G-C bias in low G-C genes,  and positional bias within genes.
-After correcting for these biases, we investigated the changes in the cohorts of essential genes and compared them to their conservation level.
-Surprisingly, we found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved.
-However, on average, essential genes are more likely to be conserved.
-
-% now start line numbers
-\linenumbers
-
-% the * after section prevents numbering
+\end{flushleft}% Correct contributions, italicise bacteria names, italicise gene names, abbreviate species names
+% convert MG1655(keio) to BW25113    http://mbio.asm.org/content/6/3/e00306-15.full#sec-11
+% EC958 -> CFT073    http://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1003788#ppat.1003788.s003
+\section{Abstract}
+The study of genes that are essential for the survival of a prokaryotic organism can help with the design of new antibiotics and designing minimal viable genomes that can be ``programmed'' with new genes.
+Transposon insertion is a method for defining essential genes.
+We have used this method for different bacteria in the Enterobacteriaceae family.
+Here, we show that on average a density of one insertion site every 25 nucleotides can result in accurate prediction of essential genes and introduce a new computational method for defining essential genes using transposon insertion data.
+In addition, we studied possible sources of bias in this experiment.
+These analyses show that there is a decline in the number of transposon insertions as we get further from the origin of replication.
+Moreover, insertions in the $5^\prime$ and $3^\prime$ ends of genes do not follow the same pattern as intermediate region.
+We compared the essential genes in Enterobacteriaceae with other bacteria and found that genes that are essential in all bacteria are involved in replication, transcription, translation, and cell division.
+However, many genes involved in these processes are not necessarily essential in all bacteria due to the existence of alternatives that compensate for them.
+In addition, we show that the essentiality of genes and their conservation are related: essential genes are more likely to be conserved and conserved genes are more likely to be essential.
+%\section{Contributions}
+%Lars Barquist and Paul Gardner have supervised the project and provided feedback on the manuscript.
+I have performed the analyses and written the manuscript.
+The transposon insertion sequencing data has been provided by Amy Cain, Christine Boinett, Gemma Langridge, Julian Parkhill, and Nicholas Thomson at Wellcome Sanger Institute, Moataz Abd El Ghany at King Abdullah University of Science and Technology, and Lars Barquist at Helmholtz Institute for RNA-based Infection Research.
 \section{Introduction}
-%The study of cells can give us new insights for understanding the foundation of life.
-However, the naturally existing cells are very complex which hampers the study of their constituents.
-Restricting these studies to the genes that are required for the growth of cells can reduce this complexity and enable us to study life in its simplest form \cite{juhas_essence_2011}.
-
-Generating genomic variants that carry genes responsible for essential cellular processes can open up new research directions.
-Adding genes or metabolic pathways to cells provides new variants with specialised phenotypes.
-Modified cells have different potential applications in biotechnology \cite{juhas_bacillus_2014}, fuel production \cite{seo_synthetic_2013}, healthcare, and food production \cite{juhas_meeting_2013}.
-Another important application for studying essential genes is in drug discovery \cite{juhas_essential_2012}.
-Infectious diseases are among the top major causes of mortality worldwide.
-Even though antibiotic resistance is growing among bacteria, the antibiotic discovery and development rate is diminishing \cite{fischbach_antibiotics_2009, nathan_antibiotics_2004}.
-Therefore, it is urgent to find new drugs for infectious diseases.
-New antibiotics target genes that are essential for the survival of pathogenic bacteria in order to control disease.
-Some well-known antibiotics that target essential functions are tetracyclines that bind to small ribosomal subunit and interfere with protein translation \cite{brodersen_structural_2000}, penicillins that target peptidoglycan and inhibit the cell wall synthesis \cite{chung_rapid_2009}, and quinolones that target DNA Gyrase \cite{marcusson_interplay_2009}.
-%Some candidate antibiotic targets found by investigating essential genes are trans-translation genes \textit{ssrA} and \textit{smpB} from \textit{helicobacter pylori} \cite{thibonnier_trans-translation_2008}, \textit{CcrM} and \textit{MenH} which are bacterial methyltransferases, and \textit{lptD} (previously known as \textit{ostA}) in \textit{Pseudomonas aeruginosa} \cite{srinivas_peptidomimetic_2010} whose essential role is yet under investigation \cite{werneburg_inhibition_2012}.[SOME ANTIBIOTICS THAT TARGET ESSENTIAL GENES ARE...]
+Sequencing a new genome and annotating the genomic regions is now quick and easy thanks to advances in sequencing and genome annotation methods, however the functional annotation of genomes remains challenging.
+One aspect of functional annotation is the study of genes that are needed for the viability of an organism i.e.
+essential genes.
+Detecting essential genes  is important for furthering our understanding of fundamental biology.
+One of the applications of identifying essential genes is drug discovery \cite{juhas_essential_2012}.
+New antibiotics, target genes that are essential for the survival of pathogenic bacteria in order to control disease.
+Moreover, essential genes can serve as the basis for generating minimal genomes \cite{glass_essential_2006, hutchison_global_1999}.
+Minimal genomes are appealing for two reasons.
+First, to understand how small can minimal life be and second, for improving biotechnological processes \cite{martinez-garcia_quest_2016}.
 
 In the earliest attempt for the identification of essential genes, Mushegian and Koonin compared the genomes of \textit{Haemophilus influenzae} and \textit{Mycoplasma genitalium}, assuming that the genes that are shared in these two phylogenetically distant bacteria are indispensable and reported 256 genes fulfilling this requirement \cite{mushegian_minimal_1996}.
-With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes declined to less than 50 which is not enough for performing all essential functions in a cell \cite{charlebois_computing_2004}.
-Therefore, the use of experimental methods for the identification of essential genes is vital.
-Researchers have now studied the essential genes in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods.
-Baba et al.\@ \cite{baba_construction_2006} made a library of single gene deletions for \textit{Escherichia coli} K-12.
-The 303 genes where viable \textit{Escherichia coli} colonies failed to grow are the candidate essential genes.
-Another group used an antisense RNA knockdown approach to study gene essentiality in \textit{Staphylococcus} aureus \cite{forsyth_genome-wide_2002}.
-In this method, if the expression of an antisense RNA hinders the growth of the cell, its cognate gene is known as essential.
-Both of these methods are labour intensive and are dependent on the accuracy of genome annotations.
-Another widely used procedure is transposon mutagenesis combined with high-throughput sequencing \cite{chao_design_2016, van_opijnen_transposon_2013, barquist_approaches_2013} which includes different approaches namely, Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}.
-These procedures differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}.
-Nonetheless, all share the same workflow: pools of single insertion mutants are constructed using transposon mutagenesis.
-After a growth phase, mutants that are fitter outnumber the less fit ones.
-Using high-throughput sequencing and tallying transposon junctions gives an indication of whether a genomic region is essential or not.
-A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium and conversely, a low number of transposon insertions indicates that a gene is essential in the medium.
+With the advent of sequencing technologies and availability of more genome sequences, the number of core genes in different prokaryotic genomes has declined to less than 100 \cite{brown_universal_2001, koonin_comparative_2003, harris_genetic_2003}.
+This number does not seem to be enough for performing all essential functions in a cell since many of the essential functions can be performed by non-orthologous genes.
+Therefore, computational methods are not capable of defining essential genes and the use of experimental methods for the identification of essential genes is important.
+By combining computational and experimental studies, Gil et al.
+introduced a set of 206 genes that can theoretically be used to design a genetically minimal bacterium \cite{gil_determination_2004}.
+They have divided the genes responsible for essential functions into 5 groups: 1) information storage and processing which can be related to DNA or RNA metabolism, 2) protein processing, folding, and secretion, 3) cell structure and cellular processes, 4) energetic and intermediary metabolism, and 5) poorly characterised genes.
+However, in practice the number of genes that appear to be required for the smallest viable synthetic bacterium is 437 \cite{hutchison_design_2016}.
 
-There are two groups of essential genes: core essential genes are indispensable for all cells, and accessory essential genes that are required for some organisms.
-Core essential genes can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003} and have been used for the synthesis of minimal cells \cite{hutchison_design_2016}.
-On the other hand, accessory essential genes are helpful in the study of specific lineages.
-Accessory essential genes may be useful in species-specific antibiotic discovery.
-Antibiotics that target core essential genes in pathogens may not be ideal, as they may attack homologous genes in their hosts or commensal bacteria.
+Essential genes have been studied in organisms from all three domains of life \cite{luo_deg_2014} using a number of different methods.
+Baba et al.
+\cite{baba_construction_2006} made a library of single gene deletion mutants for \textit{E.
+coli} K-12 BW25113, called Keio collection, and identified 303 genes that hindered the growth of viable \textit{E.
+coli} colonies as candidate essential genes.
+Antisense RNA knockdowns have also been used to study gene essentiality in \textit{Staphylococcus aureus} \cite{forsyth_genome-wide_2002}.
+With this approach, if the expression of an antisense RNA hampers cell growth, its cognate gene is thought to be essential.
+Both methods are labour intensive and depend on the accuracy of genome annotation.
+A less targeted procedure for identifying essential genes is transposon mutagenesis combined with high-throughput sequencing \cite{barquist_approaches_2013, van_opijnen_transposon_2013, chao_design_2016} which is used in several different approaches, namely Tn-Seq \cite{van_opijnen_tn-seq:_2009}, INSeq \cite{goodman_identifying_2009}, HITS \cite{gawronski_tracking_2009} and TraDIS \cite{langridge_simultaneous_2009}.
+These methods differ in the type of transposon, sample preparation methods, and data analysis \cite{van_opijnen_transposon_2013}.
+Nonetheless, all share a similar workflow: pools of single insertion mutants are constructed using transposon mutagenesis and screened using a selective media for the inserted transposons.
+After a recovery phase and library construction, high-throughput sequencing is used to identify transposon insertion junctions, which are then tallied to indicate whether a genomic region is essential or not.
+A high number of transposon insertions in a gene indicates that the gene is not essential in its growth medium, and conversely a low number of transposon insertions indicates that a gene is essential in the medium.
 
-%Curtis and Brun \cite{curtis_identification_2014} have studied the essentiality changes in cell cycle genes of three alpha-proteobacteria strains: \textit{Caulobacter crescentus}, \textit{Brevundimonas subvibrioides}, and \textit{Agrobacterium tumefaciens} and concluded that although essential genes responsible for cell functions are conserved, there are many essential genes that are specific to each organism.
+The study of core essential genes i.e.
+genes that are essential in all bacteria can shed light on the genome structure of the last universal common ancestor and the evolution of living cells \cite{koonin_comparative_2003}, and provide insight into the requirements for the synthesis of a universal minimal cell.
+Knowledge of accessory essential genes, which are only essential in specific lineages, may help identify species-specific antibiotic targets which would not adversely affect the host or commensal bacteria.
+A number of studies have used transposon insertion experiments to compare cohorts of essential genes and showed that closely related organisms can have specific essential genes.
 Freed et al.
-\cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} 2a 2457T and \textit{Escherichia coli} K12 BW25113 and shown that there are no genes that are essential in \textit{Escherichi coli} and not essential in \textit{Shigella flexneri}, while some genes are only essential in \textit{Shigella flexneri}.
-These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and soxR-reducing complex, many of which are essential in \textit{Shigella flexneri} due to the absence of paralogs or other alternative systems that exist in \textit{Escherichia coli}.
-%The difference in essentiality stems from four reasons: the disruption of many of the genes in \textit{Shigella flexneri} 2a 2457T impede the growth, but these genes are not actually essential; the conditions were different in the two experiments done on \textit{Escherichia coli} K12 BW25113 and \textit{Shigella flexneri} 2a 2457T; many of the genes were not actually essential for growth, but essential for transposon insertion and antibiotic resistance; and
+\cite{freed_combining_2016} have investigated the difference between essential genes in \textit{Shigella flexneri} and \textit{E.
+coli} K-12 BW25113 and shown that there are no genes that are essential in \textit{E.
+coli} and not essential in \textit{S.
+flexneri}, while some genes are only essential in \textit{S.
+flexneri}.
+These include a group of genes involved in cysteine, proline and sugar nucleotide biosynthesis, acetate utilisation, translation elongation, aminoacyl tRNA synthetase, murein DD-endopeptidase, and the soxR-reducing complex.
+Many of these are essential in \textit{S.
+flexneri} due to the absence of paralogs or other alternative pathways that exist in \textit{E.
+coli}.
 
-%PPG: Reduce, summarise main findings of Canals & Barquist. e.g. Most essential genes are also essential in related species, yet there are some serovar-specific essential genes [GIVE SOME NUMBERS]... 
-Canals et al.\@ \cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} typhimurium ATCC 14028, two isolates of \textit{Salmonella typhi} Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{Escherichia coli} K12 BW25113 and found that these cells share 268 essential genes.
-Nine genes are essential in \textit{Escherichia coli} and not essential in \textit{Salmonella}s, three of which can tolerate insertions only in some parts and one has distant paralogs in \textit{Salmonella}s that is missing in \textit{Escherichia coli}.
-For other genes, there are a few insertions in \textit{Salmonella}s, but the number is not small enough to call those genes essential.
-Moreover, 159 genes were almost essential in all \textit{Salmonella}s but not in \textit{Escherichia coli}.
+In another study, Canals et al.
+\cite{canals_high-throughput_2012} have compared the essentiality of genes in \textit{Salmonella} Typhimurium, two isolates of \textit{S.} Typhi Ty2 (varying in \textit{htrA}, \textit{aroC} and \textit{aroD} genes), and \textit{E.
+coli} K-12 BW25113.
+Overall, 268 essential genes were found to be shared between these strains.
+Nine genes are essential in \textit{E.
+coli} and not essential in \textit{Salmonella}.
+These genes are mostly near essential in \textit{Salmonella} as well.
+%, three of which can tolerate insertions only in some parts and one has distant paralogs in Salmonellas that is missing in \textit{E.
+coli}.
+For other genes, there are a few insertions in Salmonellas, but the number is not small enough to call those genes essential.
+Moreover, 159 genes were essential in \textit{Salmonella} but not in \textit{E.
+coli}.
 These include genes involved in replication and genes related to ribosome and its accessory proteins.
-The authors also found 26 genes that are under greater selection in \textit{Salmonella} Typhimurium compared to \textit{S.
-typhi} and 10 genes vice versa.
+The authors also found 26 genes that are under greater selection in \textit{S.} Typhimurium compared to \textit{S.} Typhi and 10 genes vice versa.
 Barquist et al.
-\cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{Salmonella} Typhi Ty2, \textit{Salmonella} Typhimurium SL1344, and \textit{Escherichia coli} K12 BW25113.
-%THIS IS IMPORTANT:
+\cite{barquist_comparison_2013} have used transposon-directed insertion-site sequencing to compare the essentiality of genes in \textit{S.} Typhi, \textit{S.} Typhimurium, and \textit{E.
+coli} K-12 BW25113.
 These genomes share 228 essential genes which are mostly involved in cell division, transcription, translation, and fatty acid and peptidoglycan biosynthesis.
-Additionally, many of the serovar-specific essential genes in \textit{Salmonella}s are phage repressors.
-Another key difference between these two serovars is that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{Salmonella} Typhimurium and not essential in \textit{salmonella} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.
+Additionally, many of the serovar-specific essential genes in \textit{Salmonella} are phage repressors.
+These proteins keep the phages in lysogenic phase and stops them from entering the lytic cycle that destructs the cell \cite{echols_establishment_1971}.
+Another key difference in the two \textit{Salmonella} serovars is the differences in the essentiality of genes involved in iron utilisation and cell surface structure that gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.
+%that a set of genes that are putatively involved in cell wall biosynthesis are essential in \textit{S.} Typhimurium and not essential in \textit{S.} Typhi which gives an indication of the adaptation of these two \textit{Salmonella} serovars to their niches.
 
-Although there are some studies on differentiation of essentiality in different organisms, these studies usually include a few genomes.
-We are going to extend these studies to a larger scale in Enterobacteriaceae family.
-Our aim is to study the essentiality of genes in in 16 different organisms.
-These include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{Klebsiella pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{Salmonella} Typhimurium SL1344, \textit{Salmonella} Typhimurium SL3261, \textit{Salmonella} Typhimurium D23580, \textit{Salmonella} Typhimurium A130, \textit{Salmonella} Enteritidis P125109, \textit{Salmonella} Typhi Ty2, \textit{Escherichia coli} ST131 EC958, \textit{Escherichia coli} UPEC ST131, \textit{Escherichia coli} ETEC CS17, \textit{Escherichia coli} ETEC H10407, \textit{Escherichia coli} BW25113, and \textit{Escherichia coli} K-12 MG1655.
+Although there are some studies on differentiation of essentiality in different organisms, their scope is limited to a handful of species/genomes.
+We aimed to extend these studies to a larger scale by investigating the essentiality of genes in 14 different organisms in the Enterobacteriaceae family of bacteria using TraDIS.
+These bacteria include \textit{Enterobacter cloacae} NCTC 9394, \textit{Klebsiella pneumoniae} Ecl8, \textit{K.
+pneumoniae} RH201207, \textit{Citrobacter rodentium} ICC168, \textit{S.} Typhimurium SL1344, \textit{S.} Typhimurium SL3261, \textit{S.} Typhimurium D23580, \textit{S.} Typhimurium A130, \textit{S.} Enteritidis P125109, \textit{S.} Typhi Ty2, \textit{E.
+coli} UPEC CFT073, \textit{E.
+coli} UPEC ST131, and \textit{E.
+coli} K-12 BW25113 which is the parent of \textit{E.
+coli} strain used in Keio collection \cite{baba_construction_2006}.
+We have also used the essentiality data of \textit{E.
+coli} K-12 BW25113 from EcoGene database \cite{zhou_ecogene_2013} which improves the essentiality data in Keio collection \cite{baba_construction_2006}.
+We first introduce a computational method for calling essential genes, compare it with the existing methods and investigate the biases that may affect our results.
+In addition, we find the best insertion density for transposon insertion experiments in bacteria.
+We then compare the list of core and accessory essential in different lineages and investigate the implications.
+%genes to core essential genes in all bacteria and core genes in Gammaproteobacteria endosymbionts to find which genes are essential in which lineages.
+\section{Results and discussion}
+In this section we will compare different methods for evaluating gene essentiality and describe our approach.
+Then, we will discuss the ideal insertion rate for transposon insertion experiments.
+Furthermore, we will review different sources of bias in transposon insertion experiments and study the essentiality of genes in different bacterial lineages.
+\subsection{Benchmarking different essentiality evaluators}
+The number of transposon insertions detected in a gene is negatively correlated with the essentiality level of that gene.
+The more insertions exist in a gene, the less essential that gene is.
+A number of methods have been used for evaluating the essentiality of genes using transposon insertion data.
+Barquist et al.
+\cite{barquist_tradis_2016} use insertion indices which are calculated by dividing the number of insertion sites in a gene by gene length.
+Plotting the insertion indices for all genes in a genome gives a bimodal plot (Figure~\ref{fig:fig1}a), each mode representing a group of genes (essential or non-essential).
+They have fitted two gamma distributions to these modes and calculated the log odds value to select a cut-off on the insertion index that distinguishes between essential and non-essential genes.
 
-%WHAT FATEMEH DID....
+Freed et al.
+\cite{freed_combining_2016} compared eleven features that can quantify the essentiality of genes, among which the average distance between insertion sites and the largest uninterrupted fraction were the best predictors.
+Turner et al.
+\cite{turner_essential_2015} used Monte Carlo method to randomly simulate insertions in a genome and using the DESeq package \cite{anders_differential_2010} calculated log fold changes for the actual number of insertions compared to the expected number of insertions obtained from the randomisation method, and clustered the log fold changes using Mclust package in R \cite{scrucca_mclust_2016}.
 
-%Enterobacteriaceae is a well characterised family of Gram-negative bacteria with a variety of host ranges and pathogenecity \cite{brenner_bergeys_2006}.
-In addition, we added the essential genes of \textit{Escherichia coli} K-12 MG1655 from EcoGene database \cite{zhou_ecogene_2013} to our study.
-In EcoGene the essentiality of genes suggested as essential by Baba et al.\@ \cite{baba_construction_2006} has been further studied and only 299 out of 303 genes are marked as essential.
-We first performed a detailed study of biases than can influence the inference of essentiality.
-Then, we normalised our data for the biases and investigated the essentiality of genes in three classes of genes: genus specific (genes that are present only in one genus), single copy genes (genes with about one instance per genome in all of the genomes that we are studying), and multi-copy genes(genes that are copied multiple times per genome).
-We have also investigated how essentiality changes in the phylogenetic tree for these organisms.
+We compared the predictive power of the average distance between insertion sites in a gene, the largest uninterrupted fraction, insertion index, and Monte Carlo method combined with DESeq.
+To evaluate the accuracy of these methods, we compared essential genes predicted with each method to the essential genes in \textit{E.
+coli} K-12 BW25113 in the EcoGene database \cite{zhou_ecogene_2013}.
+In this database, the essential genes from Keio collection \cite{baba_construction_2006} are further studied and some are omitted from the list (e.g.
+a non-essential signal peptide needed for the translation of an essential gene, or pseudogenes) while some other genes are added (mostly due to the existence of other copies that compensate for them when deleted \cite{yamamoto_update_2009}).
+The results of our comparison are depicted in Figure~\ref{fig:fig1}b and Supplementary Figure 1.
+Most of these methods have very close accuracies, but on average, insertion index is the most accurate predictor as it has the highest average Area Under the Curve (AUC) over all 13 datasets (Supplementary Figure 2).
+We also combined all these methods using Principal Component Analysis (PCA).
+We ran the prcomp function in R and selected the first principal component for our analysis.
+The average AUC is 0.9608 for PCA.
+As the random sampling method is too slow, our PCA method runs very slowly with only a small improvement to the results.
+Therefore, we tried PCA excluding random sampling which resulted in an average AUC higher than insertion index (Supplementary Figure 2).
+The difference between average AUCs of PCA excluding random sampling and insertion index is small and insertion index is a straightforward and more intuitive measure.
+Therefore, we used insertion index for identifying essential genes.
+As our dataset is composed of the results of different transposon insertion experiments on different genomes, and the number of transposon insertion sites and genome length varies in these experiments, we have divided the insertion indices in each genome by $n_o/l_o$ where $n_o$ is the number of insertion sites in genome and $l_o$ is the length of the genome.
+The distribution of the insertion indices for \textit{E.
+coli} BW25113 is plotted in Figure~\ref{fig:fig1}a.
 
-%\begin{figure}
-%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap.pdf}
-%\caption{The species tree containing the 16 genomes used in this study.
-We have generated the tree by running RAxML \cite{stamatakis_raxml_2014} on Phylosift \cite{darling_phylosift:_2014} amino acid markers.}
-%\label{fig:species-tree}
-%\end{figure}
-
-\section{Results and Discussion}
-\subsection{Incorporating multiple measures of gene essentiality improves prediction}
-In this section we will compare a number of methods of quantifying the essentiality of genes based on transposon insertion data and introduce our own method.
-\subsubsection{Comparison of old methods}
-A number of methods have been used for evaluating the essentiality of genes using transposon insertion data.
-Freed et al.
-\cite{freed_combining_2016} have compared eleven features that can quantify the essentiality of genes.
-These include: the number of insertion sites within genes; the mean distance between insertion sites, the median distance between insertion sites, the number of base pairs before the first insertion in the $5^\prime$ end, the ratio between the number of base pairs before the first insertion in the $5^\prime$ end and gene length, the number of base pairs before the second insertion in the $5^\prime$ end, the ratio between the number of base pairs before the second insertion in the $5^\prime$ end and gene length, the length of the longest uninterrupted region in the gene, the ratio between the length of the longest uninterrupted region in the gene and gene length, the length of the longest region in the gene containing at most one insertion divided by gene length, and the number of insertion sites divided by gene length.
-Among these, the average distance between insertion sites and the length of the longest uninterrupted region were the best predictors.
-
-Barquist et al.
-\cite{barquist_tradis_2016} have used insertion indices which are calculated by dividing the number of insertion sites by gene length for defining the essentiality of genes.
-Plotting all insertion indices for all genes in a genome gives a bimodal plot, each mode representing a group of genes (essential or non-essential).
-They have fitted two gamma distributions to these modes and calculated the log odds value for each gene to test which mode it belongs to.
-
-Turner et al.
-\cite{turner_essential_2015} have randomly sampled insertions in a genome and using DESeq package \cite{anders_differential_2010} calculated log fold changes and P-values for the actual number of insertions compared to the expected number of insertions obtained from the sampling method.
-
-We have compared the predictive power of the average distance between insertion sites in a gene, the length of the longest uninterrupted region, insertion index, log odds value obtained from insertion index value, and P-value and log fold change calculated using DESeq package after sampling.
-To evaluate the accuracy of these methods, we have compared the predicted essential genes to the essential genes in \textit{Escherichia coli} K-12 MG1655 in EcoGene database \cite{zhou_ecogene_2013}.
-The results are depicted in Figure~\ref{fig:benchmark}.
-Among all methods, those that try to fit a distribution to data and predict essentiality based on that, namely log odds value from insertion indices and P-value from DESeq after sampling, are the least accurate methods.
-Sampling and then calculating log fold changes using DESeq package is the most accurate predictor and the other three methods are also good predictors despite their simplicity.
-\subsubsection{PCA based method}\label{sec:essentiality-call}
-We selected four of the predictors in previous section: insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region.
-These predictors were selected as they were more accurate and less dependent on each other.
-Afterwards, we ran principal component analysis on these predictors using \emph{prcomp} function in R and selected the first principal component.
-As Figure~\ref{fig:benchmark} shows, PCA results gave us the most powerful tool for quantifying the essentiality of genes compared to other methods.
-We have plotted the results for one sample genome and the results for other genomes are shown in {\color{red}SUPPLEMENTARY}.
-Figure~\ref{fig:pca} shows the distribution of PCA results after zero mean unit variance normalisation.
-We call this value NPEQ (Normalised Pca based Essentiality Quantifier) in this paper.
-The genes whose NPEQ values are less than -1.644854 after normalisation are beneficial losses and the genes with NPEQ greater than 1.644854 are essential.
-The rest are non-essential genes.
-The cutoff 1.644854 has been selected because if we assume that NPEQ distribution is normal, this cutoff shows the P-value 0.95.
-Moreover, by trying to define a cutoff that maximises Matthews correlation coefficient for each genome using NPEQ, we get 1.650449 as the average cutoff which is very close to this value.
+Another challenge is to define a threshold for separating essential genes from non-essential ones.
+One method, is using gamma fits to find this threshold \cite{barquist_tradis_2016}.
+However, with this method, when gamma fits are not consistent with the sample distributions, the log odds values obtained from them are not good predictors of essentiality.
+In order to make our predictions independent of the goodness of fits, we have used DBSCAN clustering \cite{ester_density-based_1996}.
+The details of this method are provided in Methods~\ref{meth:essentiality}.
+This method clusters genes whose insertion indices are closely packed together and results in four groups of genes: essential genes, non-essential genes, beneficial losses, and ambiguous genes which are between different groups.
+Supplementary Figure 3 compares the results of essential gene prediction using gamma fits and DBSCAN.
+In this figure, we have assumed that essential genes in \textit{E.coli} K-12 BW25113 which are proposed in EcoGene are also essential in other genomes and calculated Matthews Correlation Coefficient for different cut-offs on insertion index.
+The cut-off found by DBSCAN outperforms gamma fits in predicting essential genes.
 
 \begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[page=1, scale=0.4]{essential-call-comparison-BW25113.pdf}
-  \caption{}
-  \label{fig:benchmark}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[page=10, scale=0.4]{essential-call-comparison-BW25113.pdf}
-  \caption{}
-  \label{fig:pca}
-\end{subfigure}
-\caption{(a) The accuracy of 7 different prediction methods for quantifying the essentiality of genes.
-The higher the area under the curve, the more accurate the method is.
-(b) The distribution for our proposed method (NPEQ)}
-\label{fig:essentiality-call}
+\includegraphics[scale=0.45]{fig1.pdf}
+\caption{(a) The bimodal distribution of insertion indices shows different classes of essentiality obtained using DBSCAN.
+These include essential genes which have the lowest insertion indices, non-essential genes with intermediate insertion indices, beneficial losses with high insertion indices, and ambiguous genes which are located between these groups (b) ROC curves showing the accuracy of 6 methods for predicting essential genes.
+True positives are genes that are predicted as essential and whose orthologs are essential in EcoGene database of essential genes for \textit{E.
+coli} K-12 BW25113.
+False positives are genes that are predicted as essential, but are classed as non-essential in EcoGene study.}
+\label{fig:fig1}
 \end{figure}
 
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BN373.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-BW25113.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-CS17.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-EC958.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ENC.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ERS227112.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ETEC.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-NCTC13441.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-ROD.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SEN.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL1344.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-SL3261.pdf}\\
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STM.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-STMMW.pdf}&
-%\includegraphics[page=1, scale=0.23]{essential-call-comparison-t.pdf}\\
-%\end{tabular}
-%\caption{The plots show the accuracy of 7 different prediction methods for quantifying the essentiality of genes. The higher the area under the curve, the more accurate the method is.}
-%\label{fig:benchmark}
-%\end{figure*}
+\subsection{What is an ideal insertion density?}
+If the number of transposon insertions is too low compared to genome length, the insertion index distribution is not bimodal and therefore the prediction of essential genes is not accurate.
+However, higher numbers of insertions are labour intensive and cost more due to larger mutant pools which need to be sequenced.
+Insertion density is defined as the ratio between the number of insertion sites and genome length.
+In this section, we explore the minimum insertion density which does not affect the accuracy of the experiment.
 
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BN373.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-BW25113.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-CS17.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-EC958.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ENC.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ERS227112.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ETEC.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-NCTC13441.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-ROD.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SEN.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL1344.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-SL3261.pdf}\\
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STM.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-STMMW.pdf}&
-%\includegraphics[page=10, scale=0.22]{essential-call-comparison-t.pdf}\\
-%\end{tabular}
-%\caption{The plots show the distribution of our new prediction method which is calculated by running PCA on insertion index, log fold change using sampling and DESeq, the average distance between insertion sites in a gene, and the length of the longest uninterrupted region and selecting first principal component.
-This method is called NPEQ.}
-%\label{fig:pca}
-%\end{figure*}
-\subsection{TraDIS data is biased}
-In Section~\ref{sec:essentiality-call} we proposed NPEQ method that quantifies the level of essentiality of a gene.
-However, if the transposon insertion is biased to specific regions in the genome, it can increase/decrease the values predicted by our method for genes and put them in a different level of essentiality.
-Different articles have reported biases in transposon insertion using Tn5 \cite{barquist_comparison_2013, green_insertion_2012, rubin_essential_2015, canals_high-throughput_2012, langridge_simultaneous_2009}.
+We provided 4 independent \textit{S.} Typhimurium SL1344 libraries with 50K, 100K, 250K, 500K, and 1000K Colony-Forming Unit (CFU).
+On average, these libraries have 4075282, 4160203, 4526237, 5711423, and 13813802 reads respectively which results in one insertion site in every 57.07, 52.94, 16.49, 14.17 and 8.86 nucleotides, assuming a uniform distribution of insertions.
+In order to compare different ranges of insertion densities, we simulated different libraries with one insertion site in 8.86 to 243.90 nucleotides.
+For each simulated library, we used the closest real library with more insertion sites than our simulated library and randomly omitted insertion sites until we reached the desired density.
+
+The false positive rate declines with the increase of insertion density as illustrated in Figure~\ref{fig:fig2}a and at 0.04 density it becomes almost constant.
+Figure ~\ref{fig:fig2}b shows that the true positive rate goes down as the insertion density rises to 0.01.
+This happens due to the large number of predicted essential genes which also includes many actual essential genes.
+But, as the density increases the number of predicted essential genes decreases which results in a decline in true positive rate.
+After 0.01, the true positive rate rises with increasing insertion density and it gets almost constant at about 0.04 insertion density.
+As a result of the fluctuations of the true positive rate and the false positive rate, Matthew's Correlation Coefficient goes up with higher insertion densities until it reaches 0.04.
+Then, it changes unnoticeably with more insertions (Supplementary Figure 4a).
+Finally, the percentage of genes with at least 1 insertion rises with higher insertion densities, but the curve becomes almost flat after 0.04 (Supplementary Figure 4b).
+Overall, these figures show that an insertion density about 0.04 can give us a good trade-off between cost and accuracy.
+
+\begin{figure}[ht]
+\includegraphics[scale=0.45]{fig2.pdf}
+\caption{Simulation of insertion density effects: The red and blue dots are obtained from real and simulated data, respectively.
+The black line shows the loess curve with 0.2 span and the gray regions show $95\%$ confidence intervals.
+Insertion resolution is calculated by dividing genome length by the number of insertion sites.
+(a) False positive rate decreases with the increase of the insertion density (the ratio between the number of insertion sites and genome length) and remains constant after it reaches 0.04.
+(b) True positive rate goes down as the insertion density rises to 0.01.
+Afterwards, it rises with increasing insertion density and gets almost constant at about 0.04 insertion density.}
+\label{fig:fig2}
+\end{figure}
+
+\subsection{Exploring possible sources of bias in TraDIS data}
+If the transposon insertion process is biased to specific regions in the genome, it can affect the number of insertions in a gene which changes insertion indices and may put some genes in incorrect essentiality categories.
+Different articles have reported biases in transposon insertion using Tn5 \cite{goryshin_tn5/is50_1998, langridge_simultaneous_2009, gallagher_genome-scale_2011, canals_high-throughput_2012, green_insertion_2012, zomer_essentials:_2012, barquist_comparison_2013, rubin_essential_2015}.
 We performed a detailed study of these biases.
 The biases that we studied include: origin of replication bias, preferred insertion motif bias, and positional bias within genes.
+
 \subsubsection{Distance from the origin of replication bias}
-While a study has reported Tn5 insertion bias near the origin of replication \cite{barquist_comparison_2013}, another study has reported no bias \cite{rubin_essential_2015}.
-To study the bias towards the position of gene within genome, we plotted NPEQ for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Fig.\@ \ref{fig:distance-bias}.
-The figure indicates that NPEQ increases when the genes are located further from the origin of replication.
-When the bacteria are under replication during the transposn insertion process, there are more copies of the genes close to the origin of replication than the genes further away due to the initiation of different replication forks.
-This results in more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
-%One might argue that the bias that we have observed may have happened due to the abundance of essential genes near the origin of replication.
-Rocha and Eduardo \cite{rocha_replication-related_2004} have shown that unlike highly expressed genes, essential genes are not enriched near the origin of replication.
+Different studies with conflicting results have been performed on Tn5 insertion bias near the origin of replication \cite{jacobs_comprehensive_2003, gallagher_genome-scale_2011, zomer_essentials:_2012, barquist_comparison_2013, rubin_essential_2015}.
+When the bacteria are under replication during the transposon insertion process and in the presence of multiple replication forks, there are more copies of the genes close to the origin of replication than the genes further away \cite{barquist_comparison_2013}.
+To study the insertion bias towards the position of genes within genome, we plotted insertion indices for each gene versus the distance of the gene from the origin of replication normalised by the length of the genome in Figure~\ref{fig:fig3}a.
+The figure indicates that insertion index decreases when the genes are located further from the origin of replication.
+This means that there are more insertions in the genes near the origin of replication which can influence the accuracy of our predictions.
+To correct this bias, we normalised insertion indices by the distance of genes from the origin of replication in genomes (See Methods~\ref{meth:essentiality}).
 
-\begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[page=6, scale=0.25]{biases.pdf}
-  \caption{}
-  \label{fig:distance-bias}
-\end{subfigure}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[page=5, scale=0.25]{biases.pdf}
-  \caption{}
-  \label{fig:GC-bias}
-\end{subfigure}
-\begin{subfigure}{\textwidth}
-  \centering
-  \includegraphics[scale=0.6]{100logo-prob-edited.pdf}
-  \caption{}
-  \label{fig:logos}
-\end{subfigure}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[scale=0.25, page=2]{insertion-position-bias.pdf}
-  \caption{}
-  \label{fig:insertion-position-bias-ess}
-\end{subfigure}
-\begin{subfigure}{.49\textwidth}
-  \centering
-  \includegraphics[scale=0.25, page=4]{insertion-position-bias.pdf}
-  \caption{}
-  \label{fig:insertion-position-bias-ben}
-\end{subfigure}
-\caption{(a)The distance of the genes from DnaA gene normalised by the lengths of the genomes versus NPEQ.
-The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
-The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias.
-(b) G-C content of genes against their NPEQ values.
-The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows no bias.
-(c) Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
-The height of each character shows the relative frequency of that character.
-(d) and (e) The plots show the average insertion index in percentiles of all essential genes (d) and beneficial losses (e).
-The genes are divided into 3 segments: 5\% of the genes on the $5^\prime$ end, 20\% of the genes on the $3^\prime$ end, and the rest in the middle.
-These are shown by khaki, slate gray, and violet red respectively.}
-\label{fig:biases}
-\end{figure}
-
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=2, scale=0.22]{biases.pdf}&
-%\includegraphics[page=6, scale=0.22]{biases.pdf}&
-%\includegraphics[page=10, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=14, scale=0.22]{biases.pdf}&
-%\includegraphics[page=18, scale=0.22]{biases.pdf}&
-%\includegraphics[page=22, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=26, scale=0.22]{biases.pdf}&
-%\includegraphics[page=30, scale=0.22]{biases.pdf}&
-%\includegraphics[page=34, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=38, scale=0.22]{biases.pdf}&
-%\includegraphics[page=42, scale=0.22]{biases.pdf}&
-%\includegraphics[page=46, scale=0.22]{biases.pdf}\\
-%\includegraphics[page=50, scale=0.22]{biases.pdf}&
-%\includegraphics[page=54, scale=0.22]{biases.pdf}&
-%\includegraphics[page=58, scale=0.22]{biases.pdf}\\
-%\end{tabular}
-%\caption{The plots show the distance of the genes from DnaA gene normalised by the lengths of the genomes versus the insertion indices of the genes.
-The distance from DnaA gene has been calculated in both directions and then the minimum value has been used for distance.
-The red curves show the loess curve when the smoothness parameter is $0.2$.}
-%\label{fig:distance-bias}
-%\end{figure*}
 \subsubsection{Preferred nucleotide composition bias}
-Another concern while inferring essentiality from transposon insertion data is that transposon insertion is biased to certain compositions of nucleotides and high number of insertions in genes reflects the enrichment of the motifs that transposon insertion is biased towards, instead of their essentiality level.
-During Tn5 transposition, a sequence of 9 nucleotides is duplicated.
-Lodge et al.
-\cite{lodge_transposon_1988} showed that these duplicated regions have G-C pairs at two ends and replacing these G-C pairs with A-T pairs reduces the number of transposon insertions by more than fivefold.
-Goryshin et al.
-\cite{goryshin_tn5/is50_1998} reported the palindromic sequence A-GNTYWRANC-T as the consensus target site for Tn5 transposition where the 9 letters in middle show the consensus sequence in the duplicated region.
-%They have also showed that insertion sites are usually grouped together and have 5 base pairs distance from each other.
-In some other research a similar consensus motif has been found for Tn5 \cite{canals_high-throughput_2012} which is CGCGCA-GTTYWRAAC-TGCGCG.
-Others \cite{green_insertion_2012,rubin_essential_2015} have not found such a sequence but shown that the duplicated regions are G-C rich.
-We used Weblogo~\cite{crooks_weblogo:_2004} to generate a logo from duplicated regions and 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
-The results in Figure \ref{fig:logos} show the consensus motif that we have found is very similar to \cite{green_insertion_2012} and the only difference is in positions 3 and 7 within the duplicated region.
-
-%\begin{figure*}
-%\includegraphics[scale=0.9]{100logo-prob-edited.pdf}
-%\caption{Sequence logo plots generated using sequences from 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
-The height of each character shows the relative frequency of that character.}
-%\label{fig:logos}
-%\end{figure*}
+Another concern while inferring essentiality from transposon insertion data is that transposon insertion is biased to certain compositions of nucleotides and high number of insertions in genes reflects their enrichment in these motifs, instead of their essentiality level.
+When Tn5 transposon is inserted into DNA, a region of 9 nucleotides is duplicated on both sides of the transposon \cite{goryshin_tn5/is50_1998}.
+Some preferred sequence motifs have been proposed previously \cite{goryshin_tn5/is50_1998, canals_high-throughput_2012} which are illustrated in Figure~\ref{fig:fig3}c.
+Other studies have found no compelling evidence of a preferred insertion motif, but showed the existence of G-C pairs at the two ends of duplicated regions \cite{green_insertion_2012, rubin_essential_2015}.
+We used Weblogo \cite{crooks_weblogo:_2004} to generate a logo from duplicated regions and 10 nucleotides flanking the 100 top most frequent insertion sites in each genome.
+The results in Figure~\ref{fig:fig3}c show that the nucleotides in the consensus motifs are just slightly overrepresented compared to the background distribution, however when corrected for background nucleotide composition, there is very little signal (see ``bits'' in Figure~\ref{fig:fig3}c).
 
@@
 \cite{rubin_essential_2015} have reported that the number of Tn5 insertions rises with the increase of G-C content and Green et al.
-\cite{green_insertion_2012} have shown that the highest number of insertions occur in high G-C content regions.
+\cite{green_insertion_2012} have shown that the highest number of insertions occurs in the regions in genome with highest G-C content.
 On the other hand, Langridge et al.
-\cite{langridge_simultaneous_2009} have seen an increase in the number of Tn5 insertions in 40\% G-C content.
-In Figure \ref{fig:GC-bias}, we plotted the G-C content of genes versus their NPEQ values.
-The red curves are loess curves with smoothness parameter 0.2 and the green line shows no bias.
-%In most cases (BW25113, EC958, ENC, ETEC, NCTC13441, ROD, SEN, STM, and t)
-NPEQ decreases gradually (which means an increase in the number of insertions) as G-C content decreases and then somewhere between 40\% and 50\% G-C content, NPEQ starts to rise again in almost all genomes.
-%In three cases (SL1344, SL3261, and STMMW) NPEQ loess curve starts to fall off again at about 60\% G-C content which is probably due to a small number of genes with low NPEQ values in that region.
-The loess curve for BN373 is almost flat.
-In CS17, the loess curve goes down until it reaches the minimum value (at about 48\%) and then remains unchanged and in ERS227112, the NPEQ value is low at the beginning and then it increases at around 50\% G-C content.
-In the region where most of the genes are packed, the loess curve is almost flat.
+\cite{langridge_simultaneous_2009} have seen a high number of Tn5 insertions in between $40\%$ and $50\%$ G-C content and lower number of insertions for G-C content higher than $50\%$.
+In Figure~\ref{fig:fig3}b, we plotted the G-C content of genes versus their insertion indices for non-essential genes and beneficial losses in a representative genome.
+The red lines are generalised additive model (GAM) curves with formula $y \sim s(x)$.
+Insertion indices increase gradually as G-C content rises up to $40\%$-$50\%$ G-C content, after which insertion indices begin to decrease in almost all genomes.
+In the region where the genes are most densely packed, the GAM curve is almost flat.
+In \textit{E.
+coli} BW25113, only 1365 genes have a GC content of less than $50\%$ or greater than $60\%$ which means they are in the region where the density is not flat.
 On the left side of this flat region, there are genes with different G-C content which are enriched in mobile genetic elements.
-So, we expect to see more insertions (smaller NPEQ values) in this region.
-We have low NPEQ values between 40\% and 50\% G-C content which is expected.
-However, in most cases when we have less than 40\% G-C content, NPEQ value is high.
-A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which reduces the insertions in A-T rich regions.
-This has been shown for Tn10 transposon \cite{kimura_nucleoid_2016}, but not for Tn5 transposon, yet.
-Overall, the results are consistent with Langridge et al.
-\cite{langridge_simultaneous_2009}.
-The G-C content of the most of the genes with large NPEQ values is between 50\% and 60\% which is inconsistent with Green et al.
-\cite{green_insertion_2012} as this region contains most of the genes and is not considered as a high G-C content region but rather an average G-C content region.
-%Even though most of the regression lines are consistent with Rubin et al.
-results \cite{rubin_essential_2015}, the figures show that a linear regression cannot predict G-C bias as its slope changes dramatically in different datasets.
-
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.23]{biases.pdf}&
-%\includegraphics[page=5, scale=0.23]{biases.pdf}&
-%\includegraphics[page=9, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=13, scale=0.23]{biases.pdf}&
-%\includegraphics[page=17, scale=0.23]{biases.pdf}&
-%\includegraphics[page=21, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=25, scale=0.23]{biases.pdf}&
-%\includegraphics[page=29, scale=0.23]{biases.pdf}&
-%\includegraphics[page=33, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=37, scale=0.23]{biases.pdf}&
-%\includegraphics[page=41, scale=0.23]{biases.pdf}&
-%\includegraphics[page=45, scale=0.23]{biases.pdf}\\
-%\includegraphics[page=49, scale=0.23]{biases.pdf}&
-%\includegraphics[page=53, scale=0.23]{biases.pdf}&
-%\includegraphics[page=57, scale=0.23]{biases.pdf}\\
-%\end{tabular}
-%\caption{The plots show G-C content of genes against their NPEQ values.
-The red curves show the loess curve when the smoothness parameter is $0.2$ and the green line shows the linear regression line.}
-%\label{fig:GC-bias}
-%\end{figure*}
+So, we expect to see more insertions in this region.
+We have high insertion indices between $40\%$ and $50\%$ G-C content which is expected.
+However, in most cases when we have less than $40\%$ G-C content, insertion index is low.
+A possible reason for this phenomena is the association of A-T rich sequences and histone-like nucleotide structuring (H-NS) proteins, which may inhibit the insertions in A-T rich regions.
+This has been proposed by Barquist et al.
+for Tn5 \cite{barquist_comparison_2013} and shown for Tn10 transposon \cite{kimura_nucleoid_2016}.
 
 \subsubsection{Positional bias within genes}
-Some research has indicated that himar1 transposons are more probable to get inserted into the two ends of a gene compared to the middle \cite{griffin_high-resolution_2011}.
-We have tested this hypothesis using our TraDIS data.
+Different studies have suggested that transposons are more likely to insert into the ends of a gene compared to the middle \cite{jacobs_comprehensive_2003, hutchison_global_1999, griffin_high-resolution_2011, gallagher_genome-scale_2011, zomer_essentials:_2012}.
+We have tested this hypothesis using TraDIS data.
 We divided every gene into 100 fragments with equal lengths (percentiles) and calculated the mean insertion index for each percentile.
-Insertion index is calculated using $\frac{\frac{n_p}{l_p}}{\frac{n_g}{l_g}}$, where $n_p$ is the number of insertion sites in a specific percentile, $l_p$ is the length of that percentile, $n_g$ is the number of insertion sites in the whole genome and $l_g$ is genome length.
+Insertion indices were calculated using $\frac{n_p/l_p}{n_g/lg}$, where $n_p$ is the number of insertion sites in a specific percentile, $l_p$ is the length of that percentile, $n_g$ is the number of insertion sites in the whole genome and $l_g$ is genome length.
 Mean insertion index for each percentile is calculated by averaging over all insertion indices for that specific percentile of genes.
-We saw almost no bias towards any location when considering all genes together (Fig.\@ \ref{fig:insertion-position-bias} {\color{red}SUPPLEMENTARY}).
+We observed almost no bias towards any location when considering all genes together (Supplementary Figure 7a).
 We studied the bias in three different groups of genes: essential genes which have no or just a few insertions, non-essential genes which have an intermediate number of insertions, and beneficial losses which have a high number of insertions.
-The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the $5^\prime$ and $3^\prime$ ends (Fig.\@ \ref{fig:insertion-position-bias-ess}) while it is the opposite in beneficial losses (Fig.\@ \ref{fig:insertion-position-bias-ben}).
+The results imply that the number of insertions in the internal region of the essential genes is outnumbered by the number of insertions in the $5^\prime$ and $3^\prime$ ends (Figure~\ref{fig:fig3}d) while it is the opposite in beneficial losses (Figure~\ref{fig:fig3}e).
 High number of insertions at the $3^\prime$ end of essential genes implies that the functional domains are located before the insertions and the insertions are not interfering with them.
-On the other hand, high number of insertions at the $5^\prime$ end of the essential genes indicates there might be alternative start codons in the $5^\prime$ end or it might be because of annotation errors that have predicted the start codon in an incorrect place before the actual start codon.
-
-%\begin{figure*}
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4, page=1]{insertion-position-bias.pdf}&
-%\includegraphics[scale=0.4, page=2]{insertion-position-bias.pdf}\\
-%\includegraphics[scale=0.4, page=3]{insertion-position-bias.pdf}&
-%\includegraphics[scale=0.4, page=4]{insertion-position-bias.pdf}
-%\end{tabular}
-%\caption{The plots show the average insertion index in the percentiles of all genes (top left), essential genes (top right), non-essential genes (bottom left), and beneficial losses (bottom right).
-The genes are divided into 3 segments: 5\% of
-%the genes on the 5' end, 20\% of the genes on the 3' end, and the rest in the middle. These are shown by khaki, slate gray, and violet red respectively.}
-%\label{fig:insertion-position-bias}
-%\end{figure*}
-\subsection{Most essential genes are ubiquitously essential in Enterobacteriaceae}
-Previous studies of gene essentiality in Enterobacteriaceae family have compared essential genes in different genomes in this family and studied the sets of core essential genes and accessory essential genes \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}.
-Core essential genes are responsible for essential processes such as cell division, DNA replication, transcription and translation and some important pathways like peptidoglycan and fatty acid biosynthesis \cite{barquist_comparison_2013}.
-Accessory essential genes differ in genomes due to different reasons such as niche adaptation, gene duplication and the existence of alternative pathways \cite{freed_combining_2016, canals_high-throughput_2012, barquist_comparison_2013}.
-Another group of accessory essential genes are phage repressors \cite{barquist_comparison_2013}.
-Even though these genes are not essential for the growth of a cell, once phages are introduced to a cell, they become essential as long as the phage remains in the cell.
-
-In this study, we have compared the genes in 16 bacteria from Enterobacteriaceae family.
-For this we needed to study core and accessory sets of genes.
-Moreover, in the presence of two redundant variations of one gene, if we knock out one copy using TraDIS, the other copy compensates and the organism can still survive \cite{dean_pervasive_2008}.
-This leads to a different essentiality inference using TraDIS.
-Therefore, in addition to core and accessory genes, we have studied duplicate genes.
-To study whether each gene in the 16 organisms is core, we used Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to iteratively search for homologous proteins in our dataset and cluster them.
-We divided the clusters of homologous genes into three groups based on their conservation.
-Genus specific class contains genes that are present only in one genus, the genes in single copy class are present in more than one genus and more than 70\% of them are not duplicated (core genes), and  the genes in multi-copy class are present in more than one genus and more than 30\% of them are duplicated (duplicate genes).
-
-We also evaluated the essentiality of genes and divided the genes into different levels of essentiality using NPEQ values.
-If NPEQ is less than -1.644854, it means that the gene has tolerated many insertions, so it is beneficial for the organism to lose this gene in the rich medium that we used.
-However, if NPEQ is greater than 1.644854, the gene can tolerate very few or no insertions indicating that the gene is essential for cell viability in our test medium.
-Any other NPEQ value shows an intermediate number of insertions in genes meaning that it is not beneficial for the organism to lose these genes, but they are not essential, too.
-We have called this group of genes non-essential.
-
-The results for comparing three levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
-The high number of single copy clusters in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality.
-However, there are also many essential genus specific genes.
-The figure also shows that beneficial losses are over represented in genus specific class.
-Therefore, beneficial losses are mostly recent genes that the organism tends to lose in the long run.
-Besides, most of the multi-copy clusters are non-essential and there are only a few multi-copy clusters that are essential.
-This can be explained by the redundancy that duplicate genes can keep even after $\sim100$ million years \cite{dean_pervasive_2008}.
+On the other hand, high number of insertions at the $5^\prime$ end of the essential genes indicates there might be alternative start codons in the $5^\prime$ end, or because of annotation errors that have predicted the start codon in an incorrect place before the actual start codon.
+In our dataset, the number of start codons or alternative start codons (ttg, ctg, or gtg) after insertion sites (in the $5^\prime$-most $10\%$ of the coding sequence) is significantly higher in genes with insertions in their $5^\prime$ end compared to genes without insertions in their $5^\prime$ end (Fisher's exact test P-value = $6.591 \times 10^{-4}$), meaning that alternative start codons and annotation errors are likely to be responsible for high numbers of insertions in the $5^\prime$ end of genes.
+In order to correct this bias, we trimmed the ends of genes and calculated insertion indices for the remaining parts (see Methods~\ref{meth:essentiality}).
 
 \begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{cluster-essentiality.pdf}
-  \caption{}
-  \label{fig:iidist}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{essential-pathways.pdf}
-  \caption{}
-  \label{fig:essentiality-pathway-ess}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{non-essential-pathways.pdf}
-  \caption{}
-  \label{fig:essentiality-pathway-nes}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{beneficialloss-pval.pdf}
-  \caption{}
-  \label{fig:essentiality-pathway-ben}
-\end{subfigure}
-\caption{(a) The genes have been clustered into homologous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
-Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
-The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.
-(b) and (c) KEGG Pathways enriched in essential genes (b) and non-essential genes (c).
-(d) The words enriched in the description of beneficial losses in their embl files compared to other genes.
-The red line shows P-value = 0.05.
-The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-\label{fig:gene-classes}
+\includegraphics[scale=0.34]{fig3.pdf}
+\caption{(a) The distance of \textit{E.
+coli} genes from the \textit{dnaA} gene (usually found near the origin of replication) versus insertion index.
+The red line shows the GAM curve when the formula is $y \sim s(x)$ and the pink shading shows $95\%$ confidence interval.
+(b) G-C content of genes against their insertion indices.
+The red line shows the GAM curve when the formula is $y \sim s(x)$.
+(c) Sequence logo plots generated using sequences from the 10 nucleotides flanking the 100 top most frequent insertion sites from each genome.
+The relative height of each character shows their frequency.
+(d) and (e) show the average insertion index (mean ii) in percentiles of all essential genes (d) and beneficial losses (e).
+The genes are divided into 3 segments: $5\%$ of the genes on the $5^\prime$ end, $20\%$ of the genes on the $3^\prime$ end, and the rest in the middle.
+These are shown by brown, blue, and red respectively.}
+\label{fig:fig3}
 \end{figure}
 
-%\begin{figure}
-%\centering
-%\includegraphics[scale=0.5]{cluster-essentiality.pdf}
-%\caption{The genes have been clustered into homologous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
-Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
-The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
-%\label{fig:iidist}
-%\end{figure}
+\subsection{Essentiality of genes in Enterobacteriaceae}
+Previous studies of gene essentiality in representative Enterobacteriales have compared essential genes in different genomes in this family and studied the sets of core essential genes (genes that are conserved in all genomes under study) and accessory essential genes (genes that are conserved in some genomes) \cite{canals_high-throughput_2012, barquist_comparison_2013, freed_combining_2016}.
+These core essential genes are responsible for essential processes such as cell division, DNA replication, transcription and translation and some important metabolic pathways such as peptidoglycan and fatty acid biosynthesis \cite{barquist_comparison_2013}.
+Accessory essential genes differ in genomes due to niche adaptation, functional redundancy and the existence of alternative pathways \cite{canals_high-throughput_2012, bergmiller_patterns_2012, barquist_comparison_2013, freed_combining_2016}.
+Another group of accessory essential genes are phage repressors \cite{barquist_comparison_2013}.
+Although these genes are not essential for cell growth, once phages are introduced to a cell, they become essential as long as the phage remains in the cell.
+In this study, we have used transposon insertion coupled with high-throughput sequencing in 13 bacteria from the Enterobacteriaceae family in addition to the list of essential genes in \textit{E.
+coli} K-12 BW25113 in the EcoGene database \cite{zhou_ecogene_2013} to identify and compare the essentiality of genes in these 14 bacteria, and to identify genes that are essential in all Enterobacteriaceae.
+Supplementary Figure 8 shows all genes that are essential in at least one bacteria in Enterobacteriaceae and the pathways in which they are involved.
+\subsubsection{Core essential genes}
+All living organisms need to perform functions that allow them to survive, reproduce, and evolve.
+Therefore, there is a set of functions that are conserved universally and enable organisms to maintain these essential properties.
+These functions can be performed by various genes.
+In this study core essential genes in a clade are defined as genes that are essential in more than $80\%$ of the genomes that we have data for in that clade (Methods~\ref{meth:corecompare}).
+We have studied core essential genes in different clades in the bacterial evolutionary tree using this dataset, a set of Gammaproteobacteria endosymbionts (Supplementary Figure 11), and the DEG database of essential genes (Supplementary Figure 10).
+These genes can be helpful to generate a minimal genome.
+Here, we discuss genes that are essential in different clades which are Enterobacteriaceae, Gammaproteobacteria, Proteobacteria, and all bacteria to see which genes differentiate these clades.
 
-To study which functions are enriched in each class of essentiality, we used KEGG pathway enrichment analysis \cite{kanehisa_kegg:_2000} and compared the genes in each class with the databases that were available for the genomes in this study.
-The results show that essential genes are enriched in pathways related to genetic information processing such as replication (DNA replication, homologous recombination and mismatch repair), transcription (RNA polymerase), translation (ribosome), and protein export.
-These are the essential functions that every cell needs for its viability.
-Other enriched pathways are mostly metabolism related.
-These include fatty acid biosynthesis that produces cell membrane, peptidoglycan and lipopolysaccharide biosynthesis that are essential components of cell wall, terpenoid backbone biosynthesis which feeds peptidoglycan biosynthesis, nucleotide and amino acid metabolism, and the metabolism of important cofactors and vitamins like riboflavin, biotin,porphyrin and chlorophyll (Figure~\ref{fig:essentiality-pathway-ess}).
-{\color{red} CAN WE HAVE SOMETHING LIKE THIS? http://www.nature.com/articles/srep00125}
+There are 14 core essential genes in all bacteria.
+These genes are enriched in complexes and pathways related to the ribosome (\textit{rplF}, \textit{rpsC}, \textit{rpsD}, and \textit{rpsE}), aminoacyl-tRNA biosynthesis (\textit{gltx}, \textit{ileS}, \textit{pheS}, and \textit{valS}), and RNA polymerase (\textit{rpoA} and \textit{rpoB}).
+Other genes in this group are involved in DNA replication (\textit{dnaE}, \textit{gyrB} and \textit{ligA}) and cell division (\textit{ftsZ}).
+Overall, genes that are essential in all bacteria are involved in replication, transcription, translation, and cell division.
 
-Non-essential genes are mostly involved in carbohydrate, energy, and lipid metabolism, membrane transport, cell motility and cellular community.
-Even though these functions are important, they are not essential in a rich medium in lab.
+Genes that are core essential in Proteobacteria, but not core essential in all bacteria are enriched in pathways and complexes related to ribosome (\textit{rplE}, \textit{rplK}, \textit{rplM}, \textit{rplO}, \textit{rplP}, \textit{rplQ}, \textit{rplX}, \textit{rpsG}, \textit{rpsK}, and \textit{rpsM}).
+Although the ribosome is an essential macromolecule, previous studies have shown that not all of the genes in this macromolecule are essential \cite{shoji_systematic_2011, akanuma_inactivation_2012}.
+Our study shows that a group of genes in ribosome are essential in all bacteria while some others are only essential in some clades, probably due to redundancy or differences in bacteria life-styles.
+Another pathway enriched in core essential genes in Proteobacteria is terpenoid backbone biosynthesis (\textit{dxr}, \textit{dxs}, and \textit{uppS}).
+There are two pathways for terpenoid backbone biosynthesis, the mevalonate pathway and the 2C-methyl-D-erythritol 4-phosphate (MEP) pathway.
+Most of the time only one of these pathways is used \cite{heuston_isoprenoid_2012}, and as a result these genes are not classed as core essential in all groups of bacteria.
+Similarly, genes that are core essential in Gammaproteobacteria and not core essential in Proteobacteria are enriched in ribosomal genes (\textit{rplB}, \textit{rplL}, \textit{rplN}, \textit{rplR}, \textit{rplT}, \textit{rplV}, \textit{rpmD}, \textit{rpsH}, \textit{rpsL}, \textit{rpsN}, \textit{rpsP}, and \textit{rpsQ}) and terpenoid backbone biosynthesis (\textit{ispE}, \textit{ispF}, \textit{ispG}, and \textit{ispH}) pathways.
 
-As most of beneficial losses do not have homologs in other genomes, most of them are not studied and therefore there is no KEGG pathway for them.
-Because of this reason, we studied the description of these genes in their embl files and found the words that were enriched.
-These are shown in Figure~\ref{fig:essentiality-pathway-ben}.
-The results show that most of beneficial losses are mobile genetic elements like transposases and insertion elements which are not essential in their host genomes, genes that are not essential in a rich medium like stress protein and acid-resistance protein, and hypothetical proteins.
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{essential-pathways.pdf}&
-%\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
-%\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
-%\includegraphics[scale=0.4]{beneficialloss-pval.pdf}
-%\end{tabular}
-%\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
-The red line shows P-value = 0.05.
-The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-%\label{fig:essentiality-pathway}
-%\end{figure*}
+Core essential genes in Enterobacteriaceae that are also core in Gammaproteobacteria endosymbionts are enriched in pathways related to riboflavin metabolism (\textit{ribA}, \textit{ribB}, \textit{ribD}, \textit{ribE}, \textit{ribF}, \textit{ribH}), ribosomal genes (\textit{rplC}, \textit{rplD}, \textit{rplJ}, \textit{rplS}, \textit{rplU}, \textit{rplW}, \textit{rpmA}, \textit{rpmB}, \textit{rpsA}, \textit{rpsB}, \textit{rpsJ}, \textit{rpsR}, \textit{rpsS}, and \textit{rpsU}), DNA replication (\textit{dnaB}, \textit{dnaG}, \textit{dnaN}, \textit{dnaX}, \textit{holB}, and \textit{ssb}), peptidoglycan biosynthesis (\textit{ftsI}, \textit{mraY}, \textit{murA}, \textit{murD}, \textit{murE}, \textit{murI}, and \textit{mviN}), and lysine biosynthesis (\textit{asd}, \textit{dapA}, \textit{dapD}, \textit{dapE}, \textit{murE}).
+Riboflavin metabolism is an essential pathway in bacteria that can be a useful drug target \cite{long_riboflavin_2010}, however the loss of this pathway can be compensated when riboflavin transporters are present and riboflavin is available in the environment \cite{gutierrez-preciado_extensive_2015} which is mostly the case in rich media.
+Lysine biosynthesis is another essential pathway and drug target \cite{gillner_lysine_2013, hutton_inhibition_2007}.
+There are three different pathways for this process, the succinylase pathway which is the most common pathway, and the dehydrogenase and acetylase pathways.
+Some small groups of species (e.g.
+\textit{Bacillus}) use only the dehydrogenase and acetylase pathways.
+The existence of lysine in the media and alternative pathways might make genes in this pathway non-essential in some genomes.
+Lysine biosynthesis feeds into the peptidoglycan biosynthesis pathway, another core essential pathway in this group,  which forms the cell wall.
+Overall, this group is enriched in genes involved in replication, translation, and cell division.
 
-\subsection{Phylogenetic analysis of gene essentiality identifies that essential genes are more likely to be conserved}
-When a mutation is introduced to a gene, the fixation rate depends on how tolerable this mutation is \cite{wilson_biochemical_1977}.
-As mutations in essential genes are less likely to be tolerable, less mutations should be fixed in these genes, and therefore essential genes should evolve more slowly than non-essential genes.
-This is known as knockout rate hypothesis \cite{krylov_gene_2003}.
-Different studies have verified this hypothesis in prokaryotes \cite{jordan_essential_2002, gerdes_experimental_2003, rocha_analysis_2004, silander_constancy_2009}.
+Finally, genes that are only core essential in Enterobacteriaceae are enriched in pathways related to fatty acid biosynthesis and fatty acid metabolism (\textit{accB}, \textit{accC}, \textit{fabA}, \textit{fabD}, and \textit{fabH}).
+Although this pathway is essential in gram-negative bacteria, it can be bypassed in gram-positive bacteria as they can incorporate exogenous fatty acids \cite{parsons_identification_2014, parsons_incorporation_2014}.
+In addition, many of fatty acid related genes might be lost during the reduction of endosymbionts.
+Supplementary Figure 9 shows the list of genes that are core essential in every clade discussed here.
 
-Here, we have tested this hypothesis by comparing the number of core essential genes and core genes in the phylogenetic tree as we go up to the root.
-We have used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes) to define if a gene is core essential or core at each level in the phylogenetic tree.
-If we get 1 at the top of the tree while studying essentiality of a gene using Fitch algorithm, it means that the gene is core essential and otherwise, the gene is non-essential.
-The same applies to the study of core genes.
-The phylogenetic tree has been annotated with the number of core essential genes (red) and the number of core genes (blue) at each level in Figure~\ref{fig:tree}.
-We then plotted the ratios at each level in the phylogenetic tree in Figure~\ref{fig:fitch} and connected the medians in each level.
-The connecting line shows that the ratio between core essential genes and core genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.
+\subsubsection{Ancestrally essential genes}
+As essential genes are responsible for key functions in the cell, it is expected that they are more conserved than  other genes during the evolution of organisms.
+We have tested this by comparing the number of essential and conserved genes in the common ancestor of our representative \textit{Enterobacter} genomes and in each level in the phylogenetic tree.
+For this, we used Fitch's algorithm \cite{fitch_toward_1971} to define ancestrally essential and ancestrally conserved genes.
+The process is explained in Methods~\ref{meth:fitch}.
+%Table 1 shows which of the genes involved in important biological processes are ancestrally essential.
+
+The phylogenetic tree in Figure~\ref{fig:fig4}a has been annotated with the number of ancestrally essential genes (red) and the number of ancestral genes (blue) at each level.
+We then plotted the ratios at each level in the phylogenetic tree in Figure~\ref{fig:fig4}b and connected the medians in each level.
+The connecting line shows that the ratio between ancestrally essential genes and ancestrally conserved genes rises as we go higher in the phylogenetic tree which means essential genes are more likely to be conserved in genomes compared to non-essential genes.
+
+To answer the opposite question, are conserved genes more likely to be essential, we gathered the genes that are conserved in Enterobacteriaceae and defined their conservation level in all bacteria in EMBL database \cite{stoesser_embl_2002} (see: Methods~\ref{meth:conservation}).
+Figure~\ref{fig:fig4}c shows that conserved genes are more likely to be essential.
 
 \begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.6\textwidth}
-  \centering
-  \includegraphics[scale=0.12]{phylosift-aa-raxmlbootstrap-annotated.pdf}
-  \caption{}
-  \label{fig:tree}
-\end{subfigure}
-\begin{subfigure}{.4\textwidth}
-  \centering
-  \includegraphics[scale=0.4]{fitch.pdf}
-  \caption{}
-  \label{fig:fitch}
-\end{subfigure}
+\includegraphics[scale=0.4]{fig4.pdf}
 \caption{(a) The species tree for all genomes in this study.
-Numbers in red show the number of core essential genes at each level and numbers in blue show the number of core genes at each level.
-(b) The ratio between core essential genes and core genes at each level in the species tree.
-The dots in the strain level show the ratios for all 16 bacteria; the dots in the species level show the ratios for Enterobacter, Klebsiella, Citrobacter, Salmonella, and Escherichia; the subfam.
-level shows the ratios for the common ancestor of Enterobacter and klebsiella, the common ancestor of Citrobacter and Salmonella, and the common ancestor of Citrobacter, Salmonella, and Escherichia; and finally the dot in the family level shows the ratio for the root.
-The line connects the medians in each level.}
-\label{fig:essentiality-phylogeny}
+Numbers in red show the number of ancestrally essential genes at each level and numbers in blue show the number of ancestral genes at each level.
+(b) The ratio between ancestrally essential genes and ancestral genes at each level in the species tree.
+The dots at the strain level show the ratios for all 14 bacteria; the dots at the species level show the ratios for \textit{Enterobacter}, \textit{Klebsiella}, \textit{Citrobacter}, \textit{Salmonella}, and \textit{Escherichia}; the subfamily level shows the ratios for the common ancestor of \textit{Enterobacter} and \textit{Klebsiella}, the common ancestor of \textit{Citrobacter} and \textit{Salmonella}, and the common ancestor of \textit{Citrobacter}, \textit{Salmonella}, and \textit{Escherichia}; and finally the dot in the family level shows the ratio for the root.
+The line connects the medians in each level.
+(c) A comparison of essentiality and conservation.
+Genes conserved in the Enterobacteriaceae family are sorted based on their conservation level in all bacteria from left to right, then a walking hypergeometric test is done from left to right regarding the number of essential genes observed at each step.
+The khaki line shows the maximum P-value obtained and the blue line shows the gene with median conservation level.
+Most of the essential genes and the maximum P-value are at the left side of the median showing that conserved genes are more likely to be essential.}
+\label{fig:fig4}
 \end{figure}
 
-We have studied the essentiality status of genes involved in important biological processes in Table~\ref{tab:biological-processes}.
-These processes include cell division, DNA replication, transcription, translation, and important metabolic pathways such as peptidoglycan and fatty acid biosynthesis.
-If a gene was not core essential using Fitch algorithm, we looked at its essentiality in every genome: if it was essential in some of the genomes we classified it as accessory essential and if it was not essential in any genome we classified it as non-essential.
-FtsA and zipA genes are essential for cell division.
-FtsA is essential in all the genomes in our study, however zipA is essential everywhere except for Salmonella typhimurium A130 in which this gene is near essential.
-We investigated if this gene has a homolog in Salmonella typhimurium A130 and found no homologs.
-It has been shown that 21 mutations in ftsA gene can make it independent of zipA gene so that the zipA gene is no longer essential \cite{geissler_gain--function_2003, pichoff_ftsa_2012}.
-However, ftsA gene does not have any of these mutations in Salmonella typhimurium A130, so there might be another factor that is causing zipA not to be essential.
-\textcolor{red}{ADD MORE EXAMPLES HERE}
+\subsubsection{Accessory essential genes}
+Accessory essential genes are important for the design of specialised antibiotics, and can give us insights into the evolution of different genera.
+Here, we have termed the genes that are ancestrally essential in one genus and not essential in other genera ``accessory essential''.
+We have used Fitch's algorithm (Materials and Methods) to define ancestrally essential genes in each genus of Enterobacteriaceae for which we had data (\textit{Citrobacter}, \textit{Salmonella}, \textit{Escherichia}, \textit{Klebsiella}, and \textit{Enterobacter}) and studied the genes that are only ancestrally essential in one genus (Supplementary Figure 12).
+There are five genes that are exclusively essential in \textit{Citrobacter} (\textit{gutQ}, \textit{ihfA}, \textit{ihfB}, \textit{priB}, and \textit{secB}).
+Among these \textit{ihfA}, \textit{ihfB} and \textit{priB} are related to homologous recombination, \textit{secB} is related to quorum sensing and \textit{gutQ} is arabinose 5-phosphate isomerase.
+The reason for the essentiality of genes related to homologous recombination might be that \textit{C.
+rodentium} is a newly evolved and unstable pathogen \cite{petty_citrobacter_2011} and homologous recombination helps with keeping its stability as it can repair DNA damage \cite{darmon_bacterial_2014}.
 
-So far we have shown that essential genes are less likely to be lost during evolution.
-The other question is if essential genes tend to keep their essentiality in different genomes.
-Bergmiller et al.
-\cite{bergmiller_patterns_2012} have shown that only essential genes whose functions cannot be taken over by other genes keep their essentiality and others are possible to be lost or non-essential in other genomes.
-The genes that can replace other genes are not necessarily homologous to them, but they have similar functions.
+Ten genes are only ancestrally essential in \textit{Salmonella} which are shown in Supplementary Figure 12.
+Among these are \textit{fepC}, \textit{fepD}, and \textit{fepG}.
+\textit{FepB} is also essnetial in 3 out of 6 \textit{Salmonella} genomes in this study, but is not defined as ancestrally essential using Fitch's algorithm.
+These four genes have been previously shown to be essential in \textit{S.} Typhi and non-essential in \textit{S.} Typhimurium SL1344 \cite{barquist_comparison_2013}.
+Our analysis shows that they are also essential in \textit{S.} Enteritidis, \textit{S.} Typhimurium D23580, and \textit{S.} Typhimurium A130.
+This shows the importance of the import of Ferric iron in these genomes.
+\textit{FtsE} and \textit{ftsX} are other accessory essential genes in \textit{Salmonella}.
+These genes are ABC transporters that facilitate cell division.%These genes are mostly ABC transporters.
+As ABC transporters are a large family of genes, some genes in this family might be able to perform the same functions in other genera.
 
-We have compared the number of genes that are shared between our bacterial genomes (\ref{fig:upsetr-conservation}) and the number of genes essential in them (\ref{fig:upsetr-essentiality}) and used UpSetR package \cite{conway_upsetr:_2016} to visualise the results in Fig.\@ \ref{fig:upsetr}.
-As shown in the figures, among 2162 genes that are shared between all the bacteria under study, only 135 are essential everywhere.
-We looked at subsets of the genes that were shared in every combination of our bacterial strains to see whether they show a trend similar to the phylogenetic tree.
-The results propose that although conservation of genes follows a tree-like trend, the essentiality does not show such a trend.
-The reason is that many of the essential genes that are not shared in all genomes are probably compensable by other genes and so the essentiality does not look tree-like.
+Seven genes are exclusively essential in \textit{Escherichia}.
+One of these genes is \textit{hipB} which is an antitoxin and is essential in \textit{Escherichia} whenever the toxin (hipA) is present.
+Another gene in this group is \textit{rpsI} which is 30S ribosomal protein S9.
+Although this gene has appeared as essential in our study and Keio collection \cite{baba_construction_2006}, it is shown that \textit{E.
+coli} is viable without this gene despite its slow growth \cite{bubunenko_essentiality_2007, shoji_systematic_2011}.
+Two other genes in this group are \textit{PhoU} which is a phosphate specific transport system accessory protein whose mutation causes the accumulation of inorganic phosphate in \textit{E.
+coli} \cite{morohoshi_accumulation_2002} and \textit{crr} which is a component of glucose-specific phosphotransferase enzyme.
+%which are involved in different processes and complexes, such as ribosome, quorum sensing, and glycolysis.
 
-\begin{landscape}
-\begin{table}
-\small
-\caption{The essentiality status of genes involved in important biological processes}
-\label{tab:biological-processes}
-\begin{tabular}[t]{l l l l l}
+The seven genes that are only essential in \textit{Klebsiella} are involved in lipopolysaccharide (LPS) biosynthesis.
+Bacterial lipopolysaccharides are composed of three components namely, lipid A, core, and O-antigen and core is further divided into two regions, inner core, and outer core \cite{caroff_structure_2003}.
+These seven essential genes belong to the inner core.
+LPS is important for stabilising membrane in gram negative bacteria \cite{salton_structure_1996}.
+%It has previously been shown that \textit{Klebsiella} produces an extracellular toxic complex that contains LPS \cite{straus_importance_1985, straus_production_1987}.
+This complex is responsible for the damage to lung tissues which causes pneumonia and its toxicity is associated to LPS \cite{straus_importance_1985, straus_production_1987}.
+We hypothesise that the importance of this complex and its correlation with the virulence of \textit{Klebsiella} \cite{domenico_extracellular_1985} is the reason for the essentiality of 7 LPS related genes specifically in \textit{Klebsiella}.
+
+%and finally, the 22 genes that are only essential in Enterobacter are involved in different functions e.g. streptomycin biosynthesis, sulfur relay system, and aminoacyl-tRNA biosynthesis.
+Finally, the 22 genes that are only essential in \textit{Enterobacter} are involved in different functions e.g.
+sulfur relay system and aminoacyl-tRNA biosynthesis.
+Further studies are needed to find out the reason for the essentiality of these genes in \textit{Enterobacter}.
+Moreover, as the \textit{Enterobacter} genus contains only one genome in our dataset, the essentiality of these genes should be studied on different species in this genus.
+
+\subsubsection{Genus-specific, conserved single-copy and conserved multi-copy genes}
+To study the relationship between copy number and essentiality of genes, we divided the genes in this study into 3 groups: genus-specific genes that are present only in one genus, conserved single-copy genes that are present once almost in all genomes in our study, and conserved multi-copy genes that are conserved and have multiple copies in each genome.
+The procedure for defining these three groups is described in Methods~\ref{meth:homclust}.
+The distribution of each class is depicted in Figure~\ref{fig:fig5}.
+
+The high number of conserved single-copy essential clusters (380 single-copy essential clusters out of a total of 612 essential clusters), indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and remained essential.
+However, there are also many essential genus-specific genes (214 genus-specific essential clusters).
+Most of the conserved multi-copy clusters are non-essential according to TraDIS.
+This can be explained by the redundancy of function that duplicate genes can retain even after long divergence times \cite{dean_pervasive_2008} which allows duplicates of knocked-out genes to compensate.
+Therefore, if a gene is essential and has multiple copies in the genome, transposon insertion sequencing is generally unable to detect its essentiality.
+
+Figure~\ref{fig:fig5} also shows that beneficial losses are over represented in the genus-specific class.
+Therefore, beneficial losses are mostly newer genes that are deleted or mutate over time.
+Moreover, the figure shows that conserved multi-copy genes have few clusters tagged as beneficial loss, meaning that beneficial losses are less likely to have multiple copies in an organism.
+We tested whether the reason for this phenomenon is that beneficial losses are copied on plasmid which causes these genes have fewer copies due to high cellular copy number.
+For this, we compared the number of clusters tagged as beneficial loss that had at least one homolog on plasmids of the genomes in this study, the number of clusters tagged as beneficial loss with no homologs on plasmids, the number of clusters that are not grouped as beneficial loss with at least one homolog on plasmids, and the number of clusters that are not grouped as beneficial loss with no homologs on plasmids using Fisher's exact test.
+The resulting P-value is 0.31 which shows no statistically significant difference between copies of genes on plasmids in beneficial loss group and others.
+
+As beneficial losses are mostly genus specific, there is not much information about them and most of them do not exist in databases like KEGG and GO.
+Therefore, in order to investigate beneficial losses we studied the words that were enriched in the description of gene product in EMBL files.
+The list of enriched words is provided in Supplementary Table 1.
+The table shows that beneficial losses are mostly mobile genetic elements (e.g.
+transposable elements, pathogenicity islands, and bacteriophages), putative or hypothetical proteins, and bacterial adhesins (e.g.
+fimbria, curli, and pilin).
+Mobile elements are newly acquired genes, so the organism does not generally require them.
+Bacterial adhesins help the bacteria to adhere together and colonise.
+Even though these genes are required when the bacteria are living in their host, they are not essential in rich media.
+
+\begin{figure}
+\includegraphics[scale=0.8]{fig5.pdf}
+\caption{Genes clustered into homologous groups using Jackhmmer and divided into 3 groups: genus-specific, conserved single-copy, and conserved multi-copy genes.
+The essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.}
+\label{fig:fig5}
+\end{figure}
+
+\section{Conclusion}
+Our analysis shows what is needed in order to accurately identify essential genes in a transposon insertion experiment.
+Transposon insertion density is a major factor for the accuracy of this experiment.
+Our study shows that on average having one insertion site in every 25 nucleotides can accurately differentiate between essential and non-essential genes.
+When transposon insertion sequencing is performed, a measure is needed for distinguishing between essential and non-essential genes.
+Despite its simplicity, insertion index has high accuracy and can quickly evaluate the essentiality of a gene.
+However, there are biases that can affect the accuracy of this method.
+We studied different biases and showed that the distance from the origin of replication can affect the number of insertions observed in a gene.
+To overcome this bias, we normalised insertion indices by the distance of genes from the origin of replication in genomes.
+In addition, the two ends of genes do not have insertion patterns similar to the intermediate region.
+Therefore, we calculated insertion indices after trimming the ends.
+
+This study shows the sets of essential genes in 14 bacteria from Enterobacteriaceae and shows that these genes are involved in pathways related to replication, transcription, translation, and protein export.
+Other enriched pathways are predominantly involved in metabolic processes.
+These include fatty acid biosynthesis that produces cell membrane, peptidoglycan and lipopolysaccharide biosynthesis that are essential components of cell wall, terpenoid backbone biosynthesis which feeds peptidoglycan biosynthesis, nucleotide and amino acid metabolism, and the metabolism of important cofactors and vitamins like riboflavin, biotin, and porphyrin.
+
+Moreover, we studied the accessory essential genes in each genus of Enterobacteriaceae and showed that the genes that are exclusively essential in each genus help with the specific lifestyle of the genomes in that group.
+Genes related to homologous recombination are essential in \textit{Citrobacter}.
+The reason might be that they reduce genome instability by repairing DNA damage \cite{darmon_bacterial_2014}.
+In addition, genes related to LPS, which help with membrane stability \cite{salton_structure_1996}, are essential in \textit{Klebsiella}.
+The essential genes in \textit{Salmonella} are mostly related to the import of Ferric iron.
+
+The comparison of the essential genes in Enterobacteriaceae with the essential genes in Gammaproteobacteria, Proteobacteria, and all bacteria shows that many genes are not essential in all groups because of the existence of other genes and pathways that compensate for them.
+The genes that are essential in all bacteria are involved in ribosome complex, aminoacyl-tRNA biosynthesis, RNA polymerase, DNA replication, and replication.
+Some genes in terpenoid backbone biosynthesis pathway are not essential in all bacteria but core essential in Proteobacteria or Gammaproteobacteria.
+The reason might be that, even though terpenoid backbone biosynthesis is an important pathway for the formation of cell wall, there are two pathways for this process, most of the time only one of which is used in a genome \cite{heuston_isoprenoid_2012}.
+There are genes related to lysine biosynthesis that are core essential in Enterobacteriaceae and core in Gammaproteobacteria endosymbionts, but not core essential in all bacteria.
+These genes are related to the production of diaminopimelate (DAP) which is important in cell wall biosynthesis.
+The reason these genes are not core essential in all genomes might be the presence of lysine in the media and alternative pathways.
+Finally, many fatty acid biosynthesis genes are only core essential in Enterobacteriaceae.
+This pathway is only essential in gram negative bacteria \cite{parsons_identification_2014, parsons_incorporation_2014}.
+%Moreover, we studied the essential genes in each genus of Enterobacteriaceae family and showed that the genes that are exclusively essential in each genus help with the specific lifestyle of the genomes in that group e.g.
+by reducing genome instability, and increasing virulence of organisms.
+%Moreover, genes related to fatty acid metabolism and biosynthesis pathways are only core essential in Enterobacteriaceae.
+Bacterial type II fatty acid synthesis (FASII) is one of the targets for antibacterial agents \cite{yao_how_2015, freiberg_novel_2006, schiebel_rational_2014}.
+Although this pathway is essential in gram-negative bacteria, it can be bypassed in gram-positive bacteria as they can incorporate exogenous fatty acids \cite{parsons_identification_2014, parsons_incorporation_2014}.
+The FASII cycle involves Acetyl-CoA carboxylase (accABCD), and fatty acid biosynthesis genes (fabABDFGHIKLVZ) of which accABCD, fabH, fabI, and fabBF have been targeted by antibacterials \cite{yao_how_2015}.
+AccB, accC, and fabH are only core essential in Enterobacteriaceae, while accA and accD are core essential in Enterobacteriaceae and Gammaproteobacteria.
+However, these genes are not core in endosymbionts which might be because they can provide their needed fatty acid from their host.
+fabI and fabB are core essential in Enterobacteriaceae and endosymbionts and fabF is not core essential anywhere.
+
+The comparison of conservation and essentiality of genes shows that essential genes are less likely to be lost over time compared to non-essential genes.
+The reverse applies as well; genes that are conserved are more likely to be essential.
+Our analyses also show that essential genes identified using transposon insertion sequencing include genes that are conserved in all Enterobacteriaceae, or exclusive to single strains.
+But very few conserved multi-copy genes are essential since the other copy can compensate for the knocked out gene.
+Finally, most of the genes that an organism tends to lose are genus-specific.
+We also showed that genes that are beneficial to lose in a rich medium are mostly related to mobile genetic elements or bacterial adhesion.
+
+\section{Materials and methods}
+In this section we will firstly discuss how the essentiality of genes in 13 bacteria from Enterobacteriaceae were evaluated.
+Secondly, we will explain the methods used to cluster orthologous and homologous genes.
+Afterwards, the ancestrally essential and ancestral genes are defined.
+Subsequently, we will explain how core genes are defined in our dataset of 14 bacteria from the Enterobacteriaceae family, DEG database, and our dataset of endosymbionts.
+In addition, the process for calculating conservation levels of genes is described and finally we will describe the method used for pathway enrichment analysis.
+
+\subsection{Identifying essential genes}\label{meth:essentiality}
+In order to find essential genes for this study, we require a metric to evaluate if a gene is essential or not.
+For this, we used insertion index as it is an accurate and simple predictor of gene essentiality (Figure~\ref{fig:fig1}a).
+The insertion index for each gene in each genome is calculated using $\frac{n_g/l_g}{n_o/l_o}$ where $n_g$ is the number of insertion sites in the gene after trimming $5\%$ from the $5^\prime$ end and $20\%$ from the $3^\prime$ end (because the regions close to the $5^\prime$ and $3^\prime$ ends are tolerant of insertions, even in essential genes as shown in Figure~\ref{fig:fig3}d), $l_g$ is the length of the gene after cutting $5\%$ from the $5^\prime$ end and $20\%$ from the $3^\prime$ end, $n_o$ is the number of insertion sites in the genome, and $l_o$ is the length of the genome.
+Since the insertion of transposons is biased towards the distance from origin of replication, we should normalise insertion indices.
+In each genome, we fitted a GAM curve when the formula is $y \sim s(x)$ to the distance vs.
+insertion index plot using R.
+Then, we calculated the distance-normalised insertion indices by dividing insertion indices by predicted values using the GAM curve and then multiplying them to average insertion index.
+
+We used DBSCAN R package \cite{ester_density-based_1996} and clustered the normalised insertion indices using DBSCAN function with parameters minPts = 200 and eps = 0.05.
+DBSCAN groups all the genes whose distance from each other is less than or equal to eps.
+If the number of genes in each group is greater than or equal to minPts, these genes are in one cluster and otherwise, they are identified as noise.
+This algorithm finds two clusters which are essential genes and non-essential genes shown by brown and blue in Figure~\ref{fig:fig1}a, respectively.
+The noise data is scattered in two regions; between essential and non-essential clusters and on the right side of non-essential genes.
+These are shown by both gray and red in Figure~\ref{fig:fig1}a.
+The noises between essential and non-essential genes were termed ambiguous.
+Since the noises in the right side of the plot involved some parts of the second peak in insertion index distribution (the rightmost gray region in Figure~\ref{fig:fig1}a) and the long tail on the right (the red region in Figure~\ref{fig:fig1}a), we separated these two regions and called the genes in the right side of the plot beneficial losses.
+For this, we reran DBSCAN on the group of genes with high insertion index with minPts = 100 and eps = 1.
+This gives us a cluster of genes that are between the non-essential genes and the long tail of the insertion index distribution.
+The genes in this cluster are called ambiguous and the rest of the genes on the right side of this ambiguous region are beneficial losses.
+
+\subsection{Genus-specific, conserved single-copy and conserved multi-copy genes}\label{meth:homclust}
+In order to test if copy number can affect the essentiality of genes identified using transposon insertion sequencing, we studied the essentiality of genes in three groups, genus-specific, conserved single-copy, and conserved multi-copy genes.
+For this we needed to cluster homologs (both orthologs and paralogs).
+We developed a homologous protein clustering methods (HomClust) using Jackhmmer from HMMER package \cite{eddy_accelerated_2011} and clustered genes based on Jackhmmer results.
+This program iteratively searches for homologous proteins in a dataset.
+We first used Jackhmmer with 5 iterations, E-value threshold set to $10^{-10}$, and MSV threshold (F1) equal to $10^{-3}$ to find all genes in all genomes that were homologous to \textit{C.
+rodentium} genes.
+Other parameters used in Jackhmmer were the default parameters.
+Then, we selected genes that were not clustered in the first step and reran Jackhmmer with the same parameters to compare these genes with all genes and cluster them.
+If the length of the homologous domain was less than $80\%$ of the length of the gene itself, we omitted that gene from the cluster.
+
+The above clustering, results in some very large clusters (e.g.
+ABC transporters, transcriptional regulators, and dehydrogenases), which can be divided into closer groupings.
+We collected the clusters that had more than 48 members and reran Jackhmmer on them with 1 iteration and more stringent thresholds, $10^{-20}$ E-value threshold and $10^{-6}$ MSV threshold.
+Then, we collected all the clusters with less than 5 members and generated a dataset that contains the members of these clusters.
+We reran Jackhmmer on this dataset with more permissive parameters which is 5 iterations, $10^{-5}$ E-value threshold, $10^{-3/2}$ MSV threshold, and $10^{-3/2}$ inclusion E-value threshold (the default value for inclusion E-value threshold which was used in previous steps is $10^{-3}$).
+Jackhmmer produces multiple local alignments on the same genes resulting in clusters containing whole genes as well as shorter segments within genes.
+In order to solve this problem, merged the members of clusters in case they had more than $80\%$ overlap.
+Finally, clusters with more than $80\%$ sequences in common got merged.
+
+In order to study the essentiality of different classes of genes, we divided the clusters of homologous genes into three classes based on their conservation and copy number; genus-specific class, conserved single-copy class, and conserved multi-copy class.
+The genus-specific class contains clusters that include genes that are present only in one genus.
+In order to separate conserved single-copy from conserved multi-copy genes, we define a threshold of $70\%$.
+%If a single gene is duplicated in 2 of 12 genomes, should that cluster be called conserved single-copy or conserved multi-copy?
+In order to solve this problem, we called clusters conserved single-copy if a minority of genes in them were duplicated.
+For this, we used a $70\%$ threshold.
+The genes in the conserved single-copy class are present in more than one genus and more than $70\%$ of them are not duplicated, while the genes in conserved multi-copy class are present in more than one genus and more than $30\%$ of them are duplicated (duplicate genes).
+
+\subsection{Clustering orthologous sequences}\label{meth:ortholog}
+To compare the essentiality of genes in different organisms, we needed to cluster orthologous genes.
+For this purpose, we used Hieranoid \cite{schreiber_hieranoid:_2013} with default parameters.
+Clustering with Hieranoid can be performed either using BLAST \cite{altschul_basic_1990, altschul_gapped_1997} or USEARCH \cite{edgar_search_2010}.
+We used BLAST as Hieranoid's similarity search tool.
+Hieranoid needs a species tree for clustering.
+To generate the species tree, we first ran PhyloSift-search on all the genome sequences to find the orthologs of marker genes in PhyloSift database \cite{darling_phylosift:_2014}.
+Subsequently, we ran PhyloSift-align to align the genes found in previous step.
+Finally, we concatenated the alignments and ran RaXMLHPC \cite{stamatakis_raxml_2014} to generate a tree.
+The parameters used for these tools are listed in Table \ref{tab:param}.
+
+\begin{table}[]
+\caption{Tools used for generating the species tree for genomes used in this study and their parameters}
+\label{tab:param}
+\begin{tabular}{ll}
+\textbf{Tool} & \textbf{Parameters}\\
 \hline
-\textbf{Biological process} & \textbf{Subprocess} & \textbf{Core essential} & \textbf{Accessory essential} & \textbf{Non-essential}\\
 \hline
-Cell Division & & \begin{tabular}[t]{@{}l@{}}ftsAHLQWYZ, minE,\\ mukB, zipA \end{tabular}& ftsKNX, minD & \begin{tabular}[t]{@{}l@{}}CedA, ftsJ, minC,\\ sdiA, sulA\end{tabular}\\
-DNA replication & Polymerases I, II, III & dnaENQX, holABD, polA & holC & holE, polB\\	
-& Supercoiling & gyrAB, parCE & & \\
-& Primosome-associated & dnaBCGT, priA, ssb & priB, rep & priC\\
-Transcription & RNA polymerase & rpoABC &  & \\
-& \begin{tabular}[t]{@{}l@{}}Sigma, elongation, anti- \\ and termination factors\end{tabular} & nusABG, rho, rpoDH & rpoEN & rpoS\\	
-Translation & tRNA-synthetases & \begin{tabular}[t]{@{}l@{}}alaS, argS, asnS, aspS, cysS, \\ glnS, gltX, glyQS, hisS, ileS, \\ leuS, lysS, metG, pheST, \\ proS, serS, thrS, tyrS, valS \end{tabular} & trpS	 & \\
-& Ribosome components & \begin{tabular}[t]{@{}l@{}}rplBCDEFJKLMNOPQRSTUV,\\ rplWXY, rpmABCD,\\ rpsABCDEFGHJKLMNPQRSTU\end{tabular} & rplA, rpmEGHI, rpsIO & rplI, rpmF\\
-& \begin{tabular}[t]{@{}l@{}}Initiation, elongation \\ and peptide chain \\ release factors\end{tabular} & fusA, infABC, prfAB, tsf & efp & prfC, selB, tufAB \\
-Biosynthetic pathways & & & & \\
-Peptidoglycan & & MraY, murABCDEFGI & & ddlAB\\	
-Fatty acids & & accABCD, fabABDGHIZ & & \\
+PhyloSift-search & isolate \\
+ & besthit \\
 \hline
+PhyloSift-align & isolate \\
+ & besthit \\
+\hline
+RaXMLHPC & m PROTGAMMALG4M \\
+ & f a \\
+ & $\#$ 100\\
 \end{tabular}
 \end{table}
-\end{landscape}
 
-\begin{figure}
-\captionsetup[subfigure]{justification=centering}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.45,page=2]{upsetr.pdf}
-  \caption{}
-  \label{fig:upsetr-conservation}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-  \includegraphics[scale=0.45,page=3]{upsetr.pdf}
-  \caption{}
-  \label{fig:upsetr-essentiality}
-\end{subfigure}
-\caption{(a) The number of genes and (b) essential genes shared between different groups of bacteria.
-The bars show the number of genes and the filled circles show which bacteria are sharing those genes.}
-\label{fig:upsetr}
-\end{figure}
+\subsection{Defining ancestrally essential and ancestrally conserved genes}\label{meth:fitch}
+In order to compare the number of genes that are essential at each level in the phylogenetic tree with the number of genes that are conserved, we used Fitch's algorithm \cite{fitch_toward_1971} with a binary alphabet on both essentiality (0 for non-essential and 1 for essential) and conservation (1 for the presence and 0 for the absence of genes).
+The Fitch's algorithm defines the status of each node in the tree by finding the order that minimises the number of mutations.
+Here, mutations are changes in essentiality or the conservation of genes.
+Therefore, by using Fitch's algorithm on gene essentiality, we minimise the number of times a gene has lost or gained essentiality during evolution assuming that this process is rare.
+Likewise, by using this algorithm on gene conservation, we minimise the number of times a gene has been lost or gained throughout time.
+Here, we describe how we used Fitch's algorithm for defining ancestrally essential genes.
+Defining ancestrally conserved genes is the same.
 
-%\section{Discussion}
-\section{Materials and Methods}
-%\subsection{Transposon insertion}
-%\section{Results and discussion}
-%Throughout time, species can gain or lose genes.
-We investigated if these gene gain and losses are related to the essentiality of the genes.
-In this section, we have first described the biases that can affect our study, and then evaluated the essentiality of genes and their conservation and the relationship between these two.
-%
-%
-%%\subsection{Length bias}
-%%Short genes are less probable to be hit by transposons and so the genes that are very short might have no insertions.
-As a result, normalising the number of insertions by the length of the genes does not affect the essentiality level of these genes.
-%%In Fig.\@ \ref{fig:length-bias} we have studied if the length of genes can influence their essentiality and we have found no correlation between length and insertion index in short genes.
-It seems that the insertion density in our experiment is high enough to make it possible for every gene to be hit by transposons.
-%%\begin{figure*}
-%%\centering
-%%\includegraphics[page=42, scale=0.3]{biases.pdf}
-%%\caption{The plot shows the lengths of the genes against their insertion indices. The red line shows the loess curve when the smoothness parameter is $0.2$.}
-%%\label{fig:length-bias}
-%%\end{figure*}
-%
-%
-%\subsubsection{Preferred insertion motif bias}
+For every gene, considering the phylogenetic tree of the genomes in this study, we assigned a set that contains only one member to each leaf of the tree.
+If the desired gene was essential in the genome at each leaf, the assigned set contained zero, and if the gene was not essential, the set contained one.
+The parent of every two nodes was labelled with the intersection of its children's assigned sets, if the intersection was not empty.
+Otherwise, it was labelled with the union of its children's assigned sets.
+We continued this process until we reached the root of the tree.
+At the root, if the assigned set only contained 1, the gene was considered ancestrally essential.
+Otherwise, it was considered ancestrally non-essential.
 
-%
-%\begin{itemize}
-%\item model H-NS binding sites? CGWTWHWww Lang et al (2007)
-%\item seems unlikely -- show bulk of genes are around 50\% G+C (add box-whisker plots to scatter diagrams?)
-%\item check Freed, Silander paper -- the missing piece of genome, was this low G+C? It is not mentioned in the paper.
-%\end{itemize}
-%
-%\begin{figure*}
-%\includegraphics[scale=0.55]{lowgc-pval.pdf}
-%\caption{Word enrichment analysis for low G-C genes compared to genes with interquartile G-C level.
-The red line shows P-value = 0.05.
-The P-values have been calculated using Fisher's exact test and corrected using Benjamini-Hochberg-Yekutieli.}
-%\label{fig:gc-pval}
-%\end{figure*}
-%
+\subsection{Comparison of core essential genes in Enterobacteriaceae, endosymbionts, and DEG database}\label{meth:corecompare}
+In order to see how essentiality of genes has changed during evolution, we compared core essential genes at different levels of bacterial phylogeny which includes Enterobacteriaceae, endosymbionts, Gammaproteobacteria, Proteobacteria, and all bacteria.
+The data for essentiality of genes in Enterobacteriaceae comes from the 14 genomes in our study and the data for other groups excluding endosymbionts is obtained from DEG database \cite{luo_deg_2014}.
+We only included data from studies that were performed in a rich medium.
+For some genomes there were multiple studies of essential genes.
+There were two studies on \textit{E.
+coli} BW25113 in DEG \cite{gerdes_experimental_2003, baba_construction_2006}, from which we only included the Keio collection \cite{baba_construction_2006} as it uses single gene knockout and is considered to be high accuracy.
+%Among the three studies on \textit{Mycobacterium tuberculosis} H37Rv \cite{sassetti_genes_2003, griffin_high-resolution_2011, zhang_global_2012} which were either using transposon mutagenesis coupled with high-throughput sequencing or transposon site hybridization, we used the most recent one which uses transposon mutagenesis along with high-throughput sequencing \cite{zhang_global_2012}.
+Likewise, among the two studies on \textit{Porphyromonas gingivalis} ATCC 33277 that used similar methods \cite{klein_identification_2012, hutcherson_comparison_2016}, we selected the most recent one \cite{hutcherson_comparison_2016}.
+Finally, there were two studies on \textit{Pseudomonas aeruginosa} PAO1 using Tn-Seq \cite{gallagher_genome-scale_2011, turner_essential_2015}, among which we only used the most recent one \cite{turner_essential_2015}.
+When multiple studies have investigated the essentiality of genes of the same organism, the most recent one was favoured (i.e.
+\cite{zhang_global_2012, hutcherson_comparison_2016, turner_essential_2015}).
+Supplementary Figure 10 shows all the genomes and studies from which the essential genes were identified.
+The orthologous genes are found using Hieranoid as explained in Section~\ref{meth:ortholog}.
 
-%\subsection{Essentiality and conservation}
-%Essential genes are needed for the growth of organisms.
-Because of that, one might think that essential genes should not be lost in a short period of time throughout evolution, unless they are no longer needed in new organisms or they are replaced by new pathways.
-Therefore, it is expected that most of the essential genes are conserved in different organisms from the same family.
-We have tested this idea by comparing the essentiality and conservation of genes in Enterobacteriaceae family.
-%
-%
-%\subsubsection{Gene classes}
-%%We needed to cluster sets of orthologous genes in our strains to study the conservation of essentiality.
-Plenty of methods are proposed for this purpose.
-Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
-We have used Hieranoid to cluster the sets of orthologous genes.
-In addition, we intended to study the essentiality of genes in paralogous genes.
-For this, we have developed a program that clusters all homologous genes using Jackhmmer from HMMER3 package \cite{mistry_challenges_2013}.
-%In order to study the relationship between essentiality and conservation, we needed to evaluate the essentiality and conservation of genes.
-For this, we divided the genes into different levels of essentiality (essential genes, ambiguous, non-essential genes, beneficial losses).
-We first normalised the biases that exist in the data using the method explained in Section~\ref{sec:biascorrection} and then defined the essentiality level of genes using the method explained in Section~\ref{sec:essentiality}.
-Moreover, we grouped the genes into different classes of conservation (genus specific, single copy, multi-copy) using the method explaind in Section~\ref{sec:conservationclasses}.
-%
-%The results for comparing four levels of essentiality and three classes of conservation are depicted in Fig.\@ \ref{fig:iidist}.
-The high number of single copy genes in essential level, indicates that there is a set of essential genes in Enterobacteriaceae that are conserved and inclined to keep their essentiality.
-However, the relatively high number of essential genes in genus specific class implies that each genus has a set of essential genes that makes it distinct.
-The figure also shows that beneficial losses are over represented in genus specific class.
-Therefore, beneficial losses are mostly recent genes that the organism tends to lose in the long run.
-Besides, most of the multi-copy clusters are non-essential and there are only a few multi-copy clusters that are essential.
-This can be explained by the redundancy that genes can keep even after $\sim100$ million years \cite{dean_pervasive_2008}.
-In the presence of two redundant variations of one gene, if we knock out one copy by transposon mutagenesis, the other copy compensates and the organism can still survive.
-%
-%\begin{figure}
-%\centering
-%\includegraphics[scale=0.5]{cluster-essentiality.pdf}
-%\caption{The genes have been clustered into orthologous groups using Hieranoid and paralogous groups using Jackhmmer and divided into 3 groups: genus specific, single copy, and multi-copy genes.
-Then, the essentiality of the clusters has been defined using the insertion indices of the genes in the clusters.
-The figure shows that most of the essential genes are in single copy group, while most of the beneficial losses are genus-specific.}
-%\label{fig:iidist}
-%\end{figure}
-%
-%To study which functions are enriched in each class of essentiality, we used the word enrichment method explained in Section~\ref{sec:enrichment}.
-Fig.\@ \ref{fig:essentiality-pval} shows the top 20 enriched words for each essentiality class.
-The results show an enrichment of the genes related to replication, transcription, translation, division, and rod shape determining proteins in essential class.
-The non-essential genes are mostly membrane associated proteins, flagellar proteins, ATPase, and DNA repair proteins.
-Beneficial losses are enriched in transposase enzymes, putative and hypothetical proteins, and mobile elements.
-Beneficial losses also contain many fimbrial proteins which probably has occurred because these proteins are not needed in a rich lab medium \textcolor{red}{\{TRUE?\}}.
-%%We have done a word enrichment analysis on the description of the beneficial losses in Fig.\@ \ref{fig:essentiality-pval}.
-It shows that most of the beneficial losses are mobile elements.
-Most of these mobile elements are not included in KEGG database.
-%
-%We also conducted a pathway enrichment analysis for these three groups which is explained in Section~\ref{sec:enrichment}.
-The results (Fig.\@ \ref{fig:essentiality-pathway}) suggest similar results to the enrichment analysis that we had done on the description of the genes.
-However, as mobile genetic elements are not stored in KEGG database \cite{kanehisa_kegg:_2000}, the pathway enrichment analysis does not show the enrichment of mobile genetic elements in beneficial-losses.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{essential-pval.pdf}&
-%\includegraphics[scale=0.4]{non-essential-pval.pdf}\\
-%\includegraphics[scale=0.4]{beneficialloss-pval.pdf}&
-%\end{tabular}
-%\caption{Word enrichment analysis for essential genes, non-essential genes, and beneficial losses compared to other genes.
-The red line shows P-value = 0.05.
-The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-%\label{fig:essentiality-pval}
-%\end{figure*}
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{essential-pathways.pdf}&
-%\includegraphics[scale=0.4]{non-essential-pathways.pdf}\\
-%\includegraphics[scale=0.4]{beneficialloss-pathways.pdf}&
-%\end{tabular}
-%\caption{Pathway enrichment analysis for beneficial losses, essential genes, and non-essential genes compared to other genes.
-The red line shows P-value = 0.05.
-The P-values are calculated using hypergeometric test and then corrected using Benjamini-Hochberg-Yekutieli procedure.}
-%\label{fig:essentiality-pathway}
-%\end{figure*}
-%
-%\subsubsection{The evolution of essentiality}
-%%In order to test if the essentiality of genes follows a tree-like trend we have gathered all the genes that are copied once per genome and made a matrix with rows showing genomes and columns showing genes.
-If a gene is essential in a genome, its corresponding cell in the matrix has value 1 and 0 otherwise.
-Then we have used the Bray-Curtis distance to generate a distance matrix for the species under study.
-Finally, a tree has been generated using PHYLIP Neighbor.
-The resulting tree is not similar to the species tree.
-%%
-%%\begin{figure}
-%%\centering
-%%\includegraphics[scale=0.2]{neighbor-joining-essentiality-tree.pdf}
-%%\caption{The tree is generated from all the genes that are copied only once per genome.
-We have made a binary matrix from the essentiality of these genes.
-Then, we have generated a distance matrix from these values using Bray-Curtis distance and plotted a phylogenetic tree using PHYLIP Neighbor which uses a neighbour joining method.}
-%%\label{fig:neighborjoining}
-%%\end{figure}
-%
-%In this section, we compared the number of genes that were essential and conserved at each level of the phylogenetic tree.
-We were interested to see if the trend of the essentiality variation between different organisms is in agreement with the phylogenetic tree.
-In other words, we have tested if organisms close together have more essential genes in common than organisms that have separated earlier in the phylogenetic tree.
-%
-%We needed to cluster sets of orthologous genes in the bacteria that we were studying.
-Homologous clusters introduced in ~\ref{sec:conservationclasses} were not useful for this purpose as sets of paralogous genes with different essentiality levels can make essentiality inference ambiguous.
-Plenty of methods are proposed for this purpose.
-Altenhoff et al.\@ have compared 15 of these methods \cite{altenhoff_standardized_2016} and shown that Hieranoid \cite{schreiber_hieranoid:_2013} is among three methods that keep a balance between precision and recall.
-We used Hieranoid for clustering orthologous genes.
-Hieranoid need a species tree for clustering genes.
-We used the method explained in Section~\ref{sec:phylogenetic} to generate this tree.
-%
-%In order to test if the essentiality of genes follows a tree-like trend, we compared the number of genes that were conserved in different bacteria in our study and the number of genes that were essential in these bacteria.
-The genes that are conserved in all bacteria being studied are called core genes and the genes that are conserved and essential are called core essential genes.
-We counted the number of genes that were core between every combination of bacteria, and the number of core essential genes between those combinations.
-We used UpSetR package \cite{conway_upsetr:_2016} in R to visualise the results in Fig.\@ \ref{fig:upsetr}.
-As shown in the figures, among 1908 genes that are core between all the bacteria under study, only 184 are core essential.
-We looked at subsets of the genes that were core essential (core) in every combination of our bacterial strains to see whether they were phylogenetically informative or not.
-A phylogenetically informative subset is a subset that is core essential (core) in two or more bacteria but not in all bacteria.
-We have marked the phylogenetically informative sets of genes with ticks and the uninformative ones with crosses.
-The results propose that although conservation of genes follows a tree-like trend with many phylogenetically informative sets of genes with high cardinality, the essentiality does not show a tree-like signal and most of the large sets of core essential genes belong to only one bacteria.
-We believe this is due to the small number of essential genes.
-Each bacterium has about 300 to 500 essential genes among which 184 is core essential between all bacteria.
-A portion of the remaining essential genes are specific to each bacterium and many are shared between all bacteria except one.
-The number of remaining essential genes is so low that causes the essentiality trend not to be tree-like.
-Furthermore, some of the predicted essential genes might be artefacts of transposon mutagenesis method.
-%
-%\begin{figure*}
-%\centering
-%\includegraphics[scale=0.6, page=2]{upsetr-maximiseMCC.pdf}\\
-%\includegraphics[scale=0.6, page=3]{upsetr-maximiseMCC.pdf}
-%\caption{The first figure shows the number of core genes between each group of species and the second figure shows the number of core essential genes.
-The bars show the number of genes that are core between the strains marked with black circles.
-The tick marks show phylogenetically informative columns and the cross marks show non-informative columns.}
-%\label{fig:upsetr}
-%\end{figure*}
-%
-%To further study the trend of essentiality changes, we looked at different levels in the species tree and calculated the ratio of the number of core essential genes to the number of core genes in each level.
-We used three different methods for inferring core genes and core essential genes.
-These methods are explained in Section~\ref{sec:core}.
-The numbers predicted using these three methods are shown in Fig.\@ \ref{fig:annotated-speciestree}.
-As this figure shows, the ratio between core essential and core genes is almost constant using the intersection and dollo method; however, this ratio increases as we go higher in the tree using the ancestral insertion index method.
-%
-%%\begin{figure}
-%%\centering
-%%\includegraphics[scale=0.2]{phylosift-aa-raxmlbootstrap-annotated.pdf}
-%%\caption{The tree shows the species tree in Fig.\@\ref{fig:species-tree} annotated by the number of core essential genes and core genes at each node.
-We used three methods to define core essential genes and core genes.
-The numbers at the leaves are the same using all these three methods.
-At the internal nodes, red shows the numbers using the intersection method, green shows the ancestral insertion index method, and turquoise shows fdollop method.}
-%%\label{fig:annotated-speciestree}
-%%\end{figure}
-%
-%As these three methods lead to conflicting results, we compared the differences between the genes found in these three methods.
-For this, we compared the set of core essential genes resulted from intersection and ancestral insertion index methods.
-Then, we performed word enrichment analysis explained in Section~\ref{sec:enrichment} on the 184 genes in intersection method and 89 genes that are core essential using ancestral insertion index and not core essential using the intersection method.
-Moreover, the intersection and fdollop methods and also ancestral insertion index and fdollop method were compared using the same procedure.
-The results are depicted in Fig.\@\ref{fig:method-difference}.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c}
-%\includegraphics[scale=0.4]{ancestral-intersection-pval.pdf}&
-%\includegraphics[scale=0.4]{dollo-ancestral-pval.pdf}\\
-%\includegraphics[scale=0.4]{dollo-intersection-pval.pdf}&
-%\end{tabular}
-%\caption{The figure shows the difference between core essential genes in intersection and ancestral insertion index methods, ancestral insertion index and fdollop methods, and intersection and fdollop methods.
-The red line shows P-value = 0.05.
-The P-values have been calculated using Fisher's exact test and then corrected using Benjamini-Hochberg-Yekutieli procedure.
-The top left figure shows words that are enriched in core essential genes found using ancestral insertion index method but not enriched in core essential genes found using intersection method.
-The top right figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using ancestral insertion index method.
-The bottom left figure shows words that are enriched in core essential genes found using fdollop method but not enriched in core essential genes found using intersection method.}
-%\label{fig:method-difference}
-%\end{figure*}
-%
-%\section{Materials and Methods}
-%\subsection{Transposon mutagenesis}\label{sec:transposoninsertion}
-%We studied 2 \textit{Klebsiella} strains, an \textit{Enterobacter} strain, a \textit{Citrobacter} strain, 6 \textit{Salmonella} strains, and 3 \textit{Escherichia} strains (Fig.\@ \ref{fig:species-tree}) and compared the essentiality of genes in these strains and \textit{Escherichia coli} K-12 MG1655 from another study \cite{baba_construction_2006}.
-These strains are all selected from Enterobacteriaceae family and a transposon mutagenesis study have been performed on them.
-We generated single inserted mutants using Tn5 transposon and placed the mutants in a selective media for Tn5.
-Then, we picked the mutants and pooled them and performed PCR enrichment using the method described in \cite{barquist_tradis_2016}.
-We sequenced the fragments and mapped them back to the genome to figure out the number of insertions that have been tolerated in each position of the genome.
-The number of insertions in a gene implies the degree of essentiality for that gene.
-%
-%\subsection{Generating the species tree}\label{sec:phylogenetic}
-%We first ran search and align commands from PhyloSift package \cite{darling_phylosift:_2014} to select gene markers for generating a phylogenetic tree and aligning them.
-Then, we concatenated the protein alignments for all 14 genomes and ran RaxML \cite{stamatakis_raxml_2014} with "PROTGAMMALG4M" amino acid substitution model, "a" algorithm and 100 alternative runs on distinct starting trees.
-%
-%\subsection{Essentiality levels}\label{sec:essentiality}
-%The more transposon insertions we observe in a gene after sequencing the genomes, the less essential the gene is.
-In order to quantify the essentiality of genes, we used a measure named insertion index which is proportional to the number of insertions in a gene.
-%
-%To calculate the insertion index for each gene, we summed up the number of transposon insertion sites observed in that gene.
-Since the lengths of the genes are different, the insertion indices were then normalised by dividing them by gene length.
-Our experiment is performed on different strains and the library density is different in each experiment.
-Therefore, in order to make the insertion indices comparable in all the strains, we normalised the insertion indices by the ratio between the number of insertions in the whole genome and the length of the genome.
-%
-%Based on insertion indices, the genes were divided into four groups: essential genes, ambiguous, non-essential genes, and beneficial losses.
-We utilised the pipeline introduced by Barquist et al.\@ \cite{barquist_tradis_2016} to evaluate the essentiality of genes.
-The insertion index distribution plot has two peaks and a heavy tail as shown in Fig.\@ \ref{fig:iidist-species}.
-A loess curve was fitted to this distribution to find where the two peaks separate from each other.
-The first peak shows the genes with no or just a few insertions which are considered as essential genes.
-We fitted an exponential distribution to the first peak and a gamma distribution to the second one.
-Then, we calculated the log odds ratio for belonging to each of these distributions for each gene.
-The region that has log odds value between -2 and 2 is called the ambiguous region, the genes belonging to the first peak are essential and the rest of the genes are not essential.
-Among genes that are not essential, any gene for which the value of the cumulative distribution function for the gamma distribution is greater than or equal to 0.99 is considered as a beneficial loss and the other genes are non-essential genes.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=2, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=3, scale=0.25]{essentiality.pdf}\\
-%\includegraphics[page=4, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=5, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=6, scale=0.25]{essentiality.pdf}\\
-%\includegraphics[page=7, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=8, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=9, scale=0.25]{essentiality.pdf}\\
-%\includegraphics[page=10, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=11, scale=0.25]{essentiality.pdf}&
-%\includegraphics[page=12, scale=0.25]{essentiality.pdf}\\
-%&\includegraphics[page=13, scale=0.25]{essentiality.pdf}&\\
-%\end{tabular}
-%\caption{Plots show the insertion index distribution for each genome.
-The plots are divided into 4 regions using red lines.
-These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-%\label{fig:iidist-species}
-%\end{figure*}
-%
-%\subsection{Bias correction}\label{sec:biascorrection}
-%We observed a distance from origin of replication bias and also a positional bias within genes.
-Hence, these biases needed to be corrected before inferring the essentiality of genes from their insertion indices.
-We did not include genes shorter than 100 base-pairs in our study as they might not be targeted by any transposon due to their shortness.
-%
-%To overcome the distance from origin of replication bias, we divided the value of insertion index for a gene in a specific position by the predicted value by loess for that position.
-This value was then multiplied by the average insertion index.
-To overcome the positional bias within genes, we calculated the insertion index for genes by ignoring $5\%$ from the 5' end and $20\%$ from the 3' end of the genes.
-The insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes is depicted in Fig.\@ \ref{fig:iidist-species-normalised}.
-%
-%\begin{figure*}
-%\centering
-%\begin{tabular}{c c c}
-%\includegraphics[page=1, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=2, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=3, scale=0.25]{essentiality-normalised.pdf}\\
-%\includegraphics[page=4, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=5, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=6, scale=0.25]{essentiality-normalised.pdf}\\
-%\includegraphics[page=7, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=8, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=9, scale=0.25]{essentiality-normalised.pdf}\\
-%\includegraphics[page=10, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=11, scale=0.25]{essentiality-normalised.pdf}&
-%\includegraphics[page=12, scale=0.25]{essentiality-normalised.pdf}\\
-%&\includegraphics[page=13, scale=0.25]{essentiality-normalised.pdf}&\\
-%\end{tabular}
-%\caption{Plots show the insertion index distribution for each genome after correcting for distance from the origin of replication bias and bias towards the position of insertion within genes.
-The plots are divided into 4 regions using red lines.
-These regions from left to right are: essential, ambiguous, non-essential, and beneficial loss.}
-%\label{fig:iidist-species-normalised}
-%\end{figure*}
-%
-%\subsection{Conservation classes}\label{sec:conservationclasses}
-%To study whether each gene in the 14 organisms is conserved we proposed a program that clusters homologous proteins.
-This program uses Jackhmmer from HMMER package~\cite{eddy_accelerated_2011} to compare protein sequences.
-It first compares a set of query proteins against all given proteins and clusters homologous proteins using Jackhmmer.
-Then, it selects all sequences that were not selected in the first step and compares them together and clusters those protein sequences.
-In the next step, it breaks down large clusters by using Jackhmmer with more stringent parameters within the clusters and also merges clusters which have a single member by running Jackhmmer with more permissive parameters.
-Finally, the program merges overlapping sequences in each cluster and combines similar clusters.
-The program is summarised in Fig.\@ \ref{fig:homclust} and the distribution of cluster lengths after clustering the genes of 14 strains under study is plotted in Fig.\@ \ref{fig:homclust-results}.
-%
-%\begin{figure}
-%\includegraphics[scale=0.2]{homclust.pdf}
-%\caption{The steps of our proposed algorithm for clustering homologous genes.}
-%\label{fig:homclust}
-%\end{figure}
-%
-%We divided the clusters of homologous genes into three groups based on their conservation.
-Genus specific clusters contain genes that are present only in one genus, the genes in single copy clusters are present in more than one genus and more than 70\% of them are not duplicated, and  the genes in multi-copy clusters are present in more than one genus and less than 70\% of them are not duplicated.
-These three groups are depicted in Fig.\@ \ref{fig:homclust-results}.
-%
-%\begin{figure}
-%\includegraphics[scale=0.3]{cluster-size-dist.pdf}
-%\caption{Size distribution for all clusters of homologous genes.
-Genus specific genes are genes that are present only in one genus, single copy genes are present in more than one genus and more than 70\% of them are not duplicated, and multi-copy genes are present in more than one genus and less than 70\% of them are not duplicated.}
-%\label{fig:homclust-results}
-%\end{figure}
-%
-%\subsection{Word and pathway enrichment analysis}\label{sec:enrichment}
-%To study the enrichment of words in a group of genes compared to a second group, We gathered the descriptions of genes from their embl files.
-Then, we counted the repeat number of each word for the genes in each group and the number of all other words in these two groups and used a Fisher's exact test to calculate P-values.
-The P-values were then corrected using Benjamini-Hochberg-Yekutieli procedure.
-%
-%For pathway enrichment analysis, we downloaded pathway datasets for strains that were available in KEGG database \cite{kanehisa_kegg:_2000}.
-This includes pathways for Cirobacter rodentium ICC168, Salmonella Enteritidis P125109, Enterobacter cloacae NCTC 9394, Salmonella Typhimurium D23580, Escherichia coli ETEC H10407, Salmonella Typhimurium SL1344, Escherichia coli K-12 MG1655, and Salmonella Typhi Ty2.
-Then we merged these databases and used the hypergeometric test to find which pathways were enriched in each essentiality class.
-Finally, we corrected the P-values using Benjamini-Hochberg-Yekutieli.
-%
-%\subsection{Defining core and core essential genes}\label{sec:core}
-%We used three different methods for defining core and core essential genes: intersection, ancestral insertion index, and Dollo law. These three methods are explained in what follows.
-%
-%The first method was intersecting over core genes and core essential genes, so, genes are core in a node if and only if they are core in all the descendants of that node and are core essential if and only if they are core essential in all the descendants of that node.
-%
-%The second method which is called ancestral insertion index uses intersection for core genes but a different definition for core essential genes.
-In this method, we averaged over the insertion indices of the pair of closest children of the ancestral node.
-We repeated this and averaged the averages until we reached the ancestral node.
-Then, we plotted the insertion indices and fitted an exponential and a gamma distribution to the plot as described in Section~\ref{sec:essentiality} and found the essential genes at that level.
-%
-%The third method is using Dollo law to define core genes and core essential genes.
-This method, assumes that the gain of genes (gain of essentiality) is highly improbable, so it tries to have up to one occurrence of gain of genes (gain of essentiality) and minimise the number of times that a gene (the essentiality of a gene) has been lost.
-Using this method, we can predict which genes were present in the common ancestor of our strains and which genes were essential in it.
-%
-%\section{Conclusion}
-%In this paper, we studied the relationship between the essentiality and conservation of genes in 14 bacteria from Enterobacteriaceae family.
-We first studied the biases that can affect our results and showed that transposon insertions are more abundant near the origin of replication.
-In addition, there is a slight preferred insertion motif bias and a G-C bias in A-T rich genes.
-Moreover we showed that transposon insertions are more abundant near the ends of essential genes compared to the internal region, while it is the opposite in beneficial losses.
-%
-%After correcting biases, we studied the essentiality versus conservation by dividing the genes into three classes of conservation: genus specific, single copy, and multi-copy, dividing them into four levels of essentiality: essential, ambiguous, non-essential, and benefical losses.
-We found that essential genes are mostly single copy, however there is a considerable number of genus specific genes in essential level which help to distinguish between genera.
-Furthermore, beneficial losses are mostly genus specific which means they are mostly new genes.
-The other finding was that multi-copy genes are mostly non-essential which can happen due to redundancy.
-%
-%We also found that the pattern of essentiality changes is not in agreement with the phylogenetic tree due to the small number of essential genes.
-Also the ratio between core essential genes and core genes can be either constant or increasing depending on the method that we use for defining core and core essential genes.
-\textcolor{red}{\{WHAT?\}}
-%
-%Overall, we have found that conserved genes are not necessarily essential, and essential genes are not necessarily conserved. However, on average, essential genes are more likely to be conserved.
-%
-%\section*{Acknowledgments}
-%%
-%%
-%\nolinenumbers
+Endosymbionts were selected from Gammaproteobacteria.
+Supplementary Figure 11 shows all of the endosymbionts in this study.
+We have included endosymbionts since they have extremely reduced genomes, that have lost many non-essential genes.
+To define core essential genes in each sub-clade, we have calculated the ratio between the number of genomes in which the gene is essential in the sub-clade and the number of all genomes in the sub-clade.
+The gene was considered core essential if this ratio is greater than 0.8.
 
-%This is where your bibliography is generated. Make sure that your .bib file is actually called library.bib
-\bibliography{references}
+\subsection{Defining the conservation level of gene clusters}\label{meth:conservation}
+For the study of the relationship between conservation and essentiality, we defined the conservation level of each gene as below.
 
-%This defines the bibliographies style. Search online for a list of available styles.
+To define the distance between every pair of bacteria, we used cmsearch from infernal package \cite{nawrocki_infernal_2013} to find the best hit to 16S rRNA (RF00177 Rfam CM model) in each bacterial genome in EMBL database \cite{stoesser_embl_2002} downloaded at 14/4/2016 and then aligned these sequences to the 16S model using cmalign from Infernal.
+Subsequently, a distance matrix was generated using dnadist from the PHYLIP package \cite{felsenstein_phylip-phylogeny_1989} and the F84 model that defines the distances.
+Afterwards, we gathered all 3164 gene clusters that are ancestrally conserved in Enterobacteriaceae and generated HMM profiles for each of these clusters using HMMER \cite{eddy_accelerated_2011}.
+The resulting HMM profiles were compared to the genes in all bacterial genomes in EMBL and the maximum distance between all hits was used as the conservation level for each gene.
+
+\subsection{Pathway enrichment analysis}
+We used KEGG \cite{kanehisa_kegg:_2000} pathway enrichment analysis to study different groups of genes and their functions.
+Only pathways for \textit{C.
+rodentium} ICC168, \textit{S.} Typhi Ty2, \textit{S.} Enteritidis P125109, \textit{S.} Typhimurium SL1344, \textit{S.} Typhimurium D23580, \textit{E.
+coli} UPEC CFT073, \textit{E.
+coli} K-12 BW25113 from Keio collection, and \textit{E.
+cloacae} NCTC 9394 genomes were available in KEGG database.
+The available pathways were downloaded using the KEGGREST package in R and then the pathways related to the desired groups of genes were compared to all pathways using Fisher's exact test.
+We adjusted the obtained P-values using the Benjamini, Hochberg, and Yekutieli method.
+Pathways with corrected P-values over 0.05 were considered to be enriched.
+%----------------------------------------------------------------------------------------
+%	BIBLIOGRAPHY
+%----------------------------------------------------------------------------------------
+\newpage
 \bibliographystyle{apalike}
+%\bibliographystyle{natbib}
+\bibliography{refs2.bib}
+
+\includepdf[pages=1,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=2,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=3,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=4,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=5,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=6,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=7,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=8,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=9,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=10,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=11,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=12,fitpaper]{supplementary/supplementary.pdf}
+\includepdf[pages=13,fitpaper]{supplementary/supplementary.pdf}
 

### 2018-07-06 bin/define-core-accessory-hieranoid-fitch-deg.py

@@ -47,15 +47,15 @@ outdir = '/home/fatemeh/EnTrI/results/deg/define-core-accessory-hieranoid-fitch'
 makedir(outdir)
 speciestreedir = '/home/fatemeh/EnTrI/results/deg/trees/'
 sequences = read_fasta_sequences(seqdb)
-species_names = {'all': ['DEG1001', 'DEG1008', 'DEG1020', 'DEG1029', 'DEG1038', 'DEG1045', 'DEG1002', 'DEG1012',
-                         'DEG1021', 'DEG1031', 'DEG1039', 'DEG1046', 'DEG1003', 'DEG1013', 'DEG1023', 'DEG1034',
-                         'DEG1040', 'DEG1047', 'DEG1005', 'DEG1014', 'DEG1024', 'DEG1035', 'DEG1041', 'DEG1006',
-                         'DEG1015', 'DEG1027', 'DEG1036', 'DEG1042', 'DEG1007', 'DEG1017', 'DEG1028', 'DEG1037',
-                         'DEG1043'], 'proteobacteria': ['DEG1035', 'DEG1024','DEG1012', 'DEG1005', 'DEG1003', 'DEG1029',
-                                                        'DEG1015', 'DEG1036', 'DEG1013', 'DEG1043', 'DEG1020', 'DEG1046',
-                                                        'DEG1041', 'DEG1045', 'DEG1028', 'DEG1031', 'DEG1008'],
-                 'gammaproteobacteria': ['DEG1012', 'DEG1005', 'DEG1003', 'DEG1029', 'DEG1015', 'DEG1036', 'DEG1013',
-                                         'DEG1043']}
+species_names = {'all': ['DEG1027', 'DEG1040', 'DEG1022', 'DEG1023', 'DEG1034', 'DEG1020', 'DEG1046', 'DEG1041',
+                         'DEG1045', 'DEG1024', 'DEG1035', 'DEG1012', 'DEG1029', 'DEG1003', 'DEG1005', 'DEG1048',
+                         'DEG1019', 'DEG1032', 'DEG1033', 'DEG1016', 'DEG1043', 'DEG1036', 'DEG1008', 'DEG1006',
+                         'DEG1014', 'DEG1042', 'DEG1037', 'DEG1038', 'DEG1017', 'DEG1002', 'DEG1001'],
[…]

### 2018-07-06 results/deg/deg-literature/34.py

@@ -1,0 +1,36 @@
+import pandas as pd
+from Bio import SeqIO
+
+def read_fasta_sequences(filepath):
+    with open(filepath, 'rU') as fasta_file:
+        fasta_dict = SeqIO.to_dict(SeqIO.parse(fasta_file, 'fasta'))
+    return fasta_dict
+
+excluded = ['M5005_Spy_0009', 'M5005_Spy_0015c', 'M5005_Spy_0067', 'M5005_Spy_0072c', 'M5005_Spy_0074',
+            'M5005_Spy_0075', 'M5005_Spy_0076', 'M5005_Spy_0098', 'M5005_Spy_0144c', 'M5005_Spy_0154', 'M5005_Spy_0174',
+            'M5005_Spy_0181', 'M5005_Spy_0183', 'M5005_Spy_0184', 'M5005_Spy_0189', 'M5005_Spy_0190', 'M5005_Spy_0210',
+            'M5005_Spy_0211', 'M5005_Spy_0234c', 'M5005_Spy_0241', 'M5005_Spy_0254c', 'M5005_Spy_0255c',
+            'M5005_Spy_0259', 'M5005_Spy_0268', 'M5005_Spy_0294c', 'M5005_Spy_0333', 'M5005_Spy_0346', 'M5005_Spy_0350',
+            'M5005_Spy_0396c', 'M5005_Spy_0401c', 'M5005_Spy_0405', 'M5005_Spy_0412', 'M5005_Spy_0431',
+            'M5005_Spy_0448', 'M5005_Spy_0481c', 'M5005_Spy_0492c', 'M5005_Spy_0494c', 'M5005_Spy_0507c',
+            'M5005_Spy_0535', 'M5005_Spy_0650', 'M5005_Spy_0666c', 'M5005_Spy_0667c', 'M5005_Spy_0669',
+            'M5005_Spy_0675', 'M5005_Spy_0742', 'M5005_Spy_0744', 'M5005_Spy_0775', 'M5005_Spy_0797', 'M5005_Spy_0799',
+            'M5005_Spy_0801', 'M5005_Spy_0806', 'M5005_Spy_0819', 'M5005_Spy_0853', 'M5005_Spy_0877', 'M5005_Spy_0887c',
+            'M5005_Spy_0888c', 'M5005_Spy_0904', 'M5005_Spy_0912', 'M5005_Spy_0969c', 'M5005_Spy_0979c',
[…]

### 2018-08-03 documents/manuscript.tex
Compared essentiality prediction of genes using different bias corrections

@@
 % line numbers
-\usepackage[right]{lineno}
-\linenumbers
+%\usepackage[right]{lineno}
+%\linenumbers
 
@@
 {\Large
-\textbf{Investigating essential genes in Enterobacteriaceae using
-transposon insertion sequencing}
+\textbf{Investigating essential genes in Enterobacteriaceae using transposon insertion sequencing}
 }
@@
 Among these are \textit{fepC}, \textit{fepD}, and \textit{fepG}.
-\textit{FepB} is also essnetial in 3 out of 6 \textit{Salmonella} genomes in this study, but is not defined as ancestrally essential using Fitch's algorithm.
+\textit{FepB} is also essential in 3 out of 6 \textit{Salmonella} genomes in this study, but is not defined as ancestrally essential using Fitch's algorithm.
 These four genes have been previously shown to be essential in \textit{S.} Typhi and non-essential in \textit{S.} Typhimurium SL1344 \cite{barquist_comparison_2013}.
@@
 Seven genes are exclusively essential in \textit{Escherichia}.
-One of these genes is \textit{hipB} which is an antitoxin and is essential in \textit{Escherichia} whenever the toxin (hipA) is present.
+One of these genes is \textit{hipB} which is an antitoxin and is essential in \textit{Escherichia} whenever the toxin (\textit{hipA}) is present.
 Another gene in this group is \textit{rpsI} which is 30S ribosomal protein S9.

### 2018-08-03 bin/compare-bias-predictions.R

@@ -1,0 +1,134 @@
+library("dbscan")
+library("ROCR")
+library("ggplot2")
+getmode <- function(v) {
+  uniqv <- unique(v)
+  uniqv[which.max(tabulate(match(v, uniqv)))]
+}
+notrimpath = '../results/biases/check-biases_not-trimmed/BW25113.txt'
+trimpath = '../results/biases/check-biases/BW25113.txt'
+infotab <- read.table(notrimpath, header = FALSE, stringsAsFactors = FALSE)
+trimtab <- read.table(trimpath, header = FALSE, stringsAsFactors = FALSE)
+colnames(infotab) <- c("name", "ii", "dist", "gc", "length", "pos")
+colnames(trimtab) <- c("name", "ii", "essentiality", "dist", "gc", "length", "pos")
+labpath <- '../results/ecogenecounterparts/BW25113.txt'
+essentiality <- read.table(labpath, header = FALSE, stringsAsFactors = FALSE)
+sp = 0.2
+
+biasestable <- as.data.frame(matrix(0, ncol = 6, nrow = length(infotab$ii)))
+colnames(biasestable) <- c("expected", "none", "dist", "gc", "pos", "all")
+biasestable$expected <- essentiality$V2
+labels <- essentiality$V2
+row.names(biasestable) <- essentiality$V1
+
+ii <- infotab$ii
+res <- dbscan(as.matrix(infotab$ii), minPts = 200, eps = 0.05)
+ess <- res$cluster[which.min(infotab$ii)]
+for (i in (1:length(infotab$ii)))
+{
+  if (res$cluster[i] == ess)
+  {
+    biasestable$none[i] = 1
+  } else
+  {
+    biasestable$none[i] = 0
+  }
+}
+
+fit <- loess(infotab$ii~infotab$dist, span = sp)
+loessprediction <- predict(fit, infotab$dist)
+ii_dnormalised = infotab$ii/(loessprediction/mean(infotab$ii))
+res <- dbscan(as.matrix(ii_dnormalised), minPts = 200, eps = 0.05)
+ess <- res$cluster[which.min(ii_dnormalised)]
+for (i in (1:length(ii_dnormalised)))
+{
+  if (res$cluster[i] == ess)
+  {
+    biasestable$dist[i] = 1
+  } else
+  {
+    biasestable$dist[i] = 0
+  }
+}
+
[…]
